### 2015-06-01 abstract.tex
Copied files from svn

%!TEX root =  matrix-est.tex
We propose an efficient procedure for estimating the mixing time of a Markov chain.
The novelty of our approach lies in several key aspects.
Crucially,
we do not assume any a priori bounds
on the mixing time,
which
sets our approach apart from
most others that do
require some advance knowledge of this quantity.
Another
advantage of our result over
common practice is the sufficiency of a single long sampling sequence,
without access to a restart mechanism.
Perhaps most importantly, we overcome an obstacle inherent in the chicken-and-egg nature of
estimating mixing coefficients: to give confidence bounds one typically needs to know the mixing time.
We provide confidence guarantees that are fully data-dependent in the sense that they yield non-trivial
finite-sample deviation bounds without any knowledge of the mixing coefficients.
Upper and lower sample complexity bounds are provided and challenging open problems are posed.

### 2015-06-01 conclusions.tex
Copied files from svn

%!TEX root =  matrix-est.tex

The algorithm from Section~\ref{sec:empirical} demonstrates the
feasibility of fully empirical finite-sample confidence bounds on
Markov chain mixing time.
This makes it possible to apply learning methods that require
empirical bounds to many non-iid settings that are commonplace in many
applications.
%One limitation of our techniques is that they are limited to Markov
%chains over finite state spaces.
We expect that future results will generalize and tighten our
analysis, possibly extend it to large (or infinite) state spaces by
adding extra assumptions.

A salient limitation of our approach is the reversibility assumption.
Without the latter, one could still obtain non-trivial finite-sample
estimates.
One possible starting point is to work with a pointwise estimate $\wh
P_{ij}$ and invoke the Ostrowski-Elsner theorem
\citep{stewart1990matrix} to bound the deviation between the empirical
$\hatgap$ and the true $\gap$.
Unfortunately, this bound has an exponential dependence on $d$.
Another problem we leave open for now is to close the gap between the
upper bound in Theorem~\ref{thm:err} and the lower bounds in
Theorems~\ref{thm:lb-pimin} and \ref{thm:lb-gap}. 
\todoc{We need to move the references needed in the appendix to after the appendix. 
This should be possible with some latex package for splitting references..}
%and investigate
%whether the width of the fully empirical confidence interval can be
%shortened to match the width of the theoretical confidence interval.

%The upshot is that now the various Markov PAC learning results stated in
%terms of unknown mixing coefficients
%\citep{gamarnik03,MoRo09}
%can be applied to give nontrivial data-dependent bounds,
%albeit in a limited setting.

### 2015-06-01 empirical.tex
Copied files from svn

%!TEX root =  matrix-est.tex
In this section, we address the shortcoming of \cref{thm:err} and give
fully empirical confidence intervals for the stationary probabilities
and the spectral gap $\gap$.
The main idea is to use the Markov property to eliminate the
dependence of the confidence intervals on the unknown quantities
(including $\pimin$ and $\gap$).
Specifically, we estimate the transition probabilities from the sample
path using simple frequency estimates: as a consequence of the Markov
property, for each state, the frequency estimates converge at a rate
that depends only on the number of visits to the state, and in
particular the rate (given the visit count of the state) is
independent of the mixing time of the chain.

As discussed in \cref{sec:rates}, it is possible to form a confidence
interval for $\gap$ based on the eigenvalues of an estimated
transition probability matrix by appealing to the
Ostrowski-Elsner theorem.
However, as explained earlier, this would lead to a slow
$O(n^{-1/(2d)})$ rate.
We avoid this slow rate by using an estimate of the symmetric matrix
$\vL$, and employing a stronger perturbation result (namely Weyl's
inequality) available for symmetric matrices.

To form an estimate of $\vL$ based on an estimate of the transition
probabilities, one possibility is to estimate $\vpi$ using a
frequency-based estimate for $\vpi$ as was done in \cref{sec:rates},
and appeal to the relation $\vL = \Diag(\vpi)^{1/2} \vP
\Diag(\vpi)^{-1/2}$ to form a plug-in estimate.
However, as noted in \cref{sec:rates-upper}, confidence intervals for
the entries of $\vpi$ formed this way may depend on the mixing time.
Indeed, such an estimate of $\vpi$ does not exploit the Markov
property.

We adopt a different strategy for estimating $\vpi$, which leads to
our construction of empirical confidence intervals, detailed in
\cref{alg:empest}.
We form the matrix $\wh\vP$ using smoothed frequency estimates of
$\vP$ (Step~\ref{step:P}), then 
compute the so-called group inverse $\giAh$ of $\wh\vA = I - \wh\vP$,
followed by finding the unique stationary
distribution $\hat\vpi$ of $\wh\vP$ (Step~\ref{step:pi}).
The group inverse $\giAh$ of $\wh\vA$ is uniquely defined and if $\wh\vP$ defines an ergodic chain
(which is the case here due to the use of the smoothed estimates),
$\giAh$ can be computed at the cost of inverting an $(d-1)\times (d-1)$ matrix
(cf. Theorem 5.2 in \cite{meyer1975role}).
Further, once given $\giAh$, the unique stationary distribution of $\wh\vP$ can be 
read out from the last row of $\giAh$ (Theorem 5.3 of  \cite{meyer1975role}).
The group inverse will also be used later to compute the sensitivity of $\vpi$
based on a result from \cite{cho2001comparison}.
Based on $\hat\vpi$ and $\wh\vP$, we construct the plug-in estimate
$\wh\vL$ of $\vL$, and use the eigenvalues of its symmetrization to
form the estimate $\hatgap$ of the spectral gap (Steps~\ref{step:eig}
and~\ref{step:gap}).
In the remaining steps, 
we use perturbation analyses to relate $\hat\vpi$ and $\vpi$, viewing
$\vP$ as the perturbation of $\wh\vP$; and also to relate $\hatgap$
and $\gap$, viewing $\vL$ as a perturbation of $\Sym(\wh\vL)$.
Both analyses give error bounds entirely in terms of observable
quantities, tracing back to empirical error bounds for the smoothed
frequency estimates of $\vP$.
Note that the most expensive step in the algorithm is the computation of the group inverse,
which, as noted reduces to matrix inversion.
Thus, with a standard implementation of matrix inversion, the algorithm's complexity becomes 
$O( n + d^3)$, while the space complexity is $O(d^2)$.
\todoc{Implement and test?}

\begin{algorithm}[t]
\caption{Empirical confidence intervals}
\label{alg:empest}
\begin{algorithmic}[1]
  \renewcommand\algorithmicrequire{\textbf{Input}:}
  \REQUIRE
    Sample path $(X_1,X_2,\dots,X_n)$,
    confidence parameter $\delta \in (0,1)$.

  \STATE Compute state visit counts and smoothed transition
  probability estimates:
  \begin{equation}
    \begin{aligned}
      N_i & :=
      \Abs{
        \Braces{
          t \in [n-1] : X_t = i
        }
      }
      , \quad i \in [d] ; \\
      N_{i,j} & :=
      \Abs{
        \Braces{
          t \in [n-1] : (X_t,X_{t+1}) = (i,j)
        }
      }
      , \quad (i,j) \in [d]^2 ; \\
      \wh{P}_{i,j}
      & :=
      \frac{N_{i,j} + 1/d}{N_i + 1}
      , \quad (i,j) \in [d]^2 .
    \end{aligned}
    %\label{eq:frequencies}
    \notag
  \end{equation}
  \label{step:P}

  \STATE Let $\giAh$ be the group inverse of $\wh\vA := \vI -
  \wh\vP$.
  \label{step:gi}

  \STATE Let $\hat\vpi \in \Delta^{d-1}$ be the unique stationary
  distribution for $\wh\vP$.
  \label{step:pi}

  \STATE Compute eigenvalues $\hat\lambda_1 {\geq} \hat\lambda_2
  {\geq} \dotsb {\geq} \hat\lambda_d$ of $\Sym(\wh\vL)$, where $\wh\vL
  := \Diag(\hat\vpi)^{1/2} \wh\vP \Diag(\hat\vpi)^{-1/2}$.
  \label{step:eig}

  \STATE Spectral gap estimate:
  \[ \hatgap := 1 - \max\braces{ \hat\lambda_2,\, |\hat\lambda_d| } . \]
  \label{step:gap}

  \STATE Empirical bounds for $|\wh{P}_{i,j}-P_{i,j}|$ for each $(i,j)
  \in [d]^2$:
  \begin{equation}
    \wh{B}_{i,j}
    :=
    \sqrt{\frac{9\wh{P}_{i,j}(1-\wh{P}_{i,j})\ln(2d/\delta)}{2N_i}}
    + \frac{4\ln(2d/\delta)+1.5\abs{(1/d)-\wh{P}_{i,j}}}{N_i}
    .
%    \label{eq:P-obs-bound}
    \notag
  \end{equation}
  \label{step:P-bound}

%$\qquad$ $\wh\Delta_{i,j}(\delta) \gets \sqrt{\frac{9\wh{P}_{i,j}(1-\wh{P}_{i,j})\ln(2d/\delta)}{2N_i}}
%  + \frac{4\ln(2d/\delta)+1.5\abs{\alpha-d\alpha\wh{P}_{i,j}}}{N_i}$%,  $(i,j) \in [d]^2$.
%\STATE   $\qquad$ $\wh\Delta_{i,j}(\delta) \gets \min( \wh\Delta_{i,j}(\delta) , \max(\wh{P}_{i,j}, 1-\wh{P}_{i,j}))$,   $(i,j) \in [d]^2$.

%  \STATE Empirical bound for $\max_{i,j}|\wh{P}_{i,j}-P_{i,j}|$:
%  \quad $B(\delta) := \max_{i,j} B_{i,j}(\delta)$.
  %$\wh\Delta_{\vP}(\delta) \gets \sqrt{\sum_{i,j} \wh\Delta_{i,j}^2(\delta)}$,
%  $\wh\Delta^*_{\vP}(\delta) \gets \max_{i,j} \wh\Delta_{i,j}(\delta)$.

  \STATE Relative sensitivity of $\vpi$:
  \begin{equation}
    \hat\kappa :=
    \frac12
    \max
    \Braces{
      \wh{A}_{j,j}^\# - \min\Braces{ \wh{A}_{i,j}^\# : i \in [d] }
      : j \in [d]
    } 
    .
%    \label{eq:hatkappa}
    \notag
  \end{equation}
  \label{step:sens}

  \STATE Empirical bounds for $\max_{i \in [d]} |\hat{\pi}_i -
  \pi_i|$ and
  $\max\bigcup_{i\in[d]}
  \braces{
    \abs{\sqrt{\pi_i/\hat\pi_i}-1},\,
    \abs{\sqrt{\hat\pi_i/\pi_i}-1}
  }$:
  \begin{equation}
    \hat{b} := \hat\kappa \max\Braces{
      \wh{B}_{i,j}
      : (i,j) \in [d]^2
    }
    , \qquad
    \hat\rho := \frac12 \max \bigcup_{i\in[d]}
    \Braces{
      \frac{\hat{b}}{\hat\pi_i},\,
      \frac{\hat{b}}{\brackets{\hat\pi_i-\hat{b}}_+}
    }
    .
%    \label{eq:pi-obs-bound}
    \notag
  \end{equation}
  \label{step:pi-bound}

  \STATE Empirical bounds for $\abs{\hatgap-\gap}$:
  \begin{equation}
    \hat{w} := 2\hat\rho + \hat\rho^2
    + (1+2\hat\rho+\hat\rho^2)
    \Biggl(
      \sum_{(i,j)\in[d]^2} \frac{\hat\pi_i}{\hat\pi_j} \hat{B}_{i,j}^2
    \Biggr)^{1/2} .
%    \label{eq:gap-obs-bound}
    \notag
  \end{equation}
  \label{step:gap-bound}

%  \STATE Empirical bounds for $\hat{\pi}_i$:
%  \begin{align*}
%    \rho_i(\delta)
%    & := \frac{\wh\kappa B(\delta)}{\hat\pi_i}
%    , \quad i \in [d] ;
%    &
%    \rho_i'(\delta)
%    & := \frac{\wh\kappa B(\delta)}{[\hat{\pi}_i-\wh\kappa B(\delta)]_+}
%    , \quad i \in [d] .
%  \end{align*}

%\STATE $\rho(\delta) \gets \frac12 \max_{i\in \iset{d}} \rho_i(\delta)$; $\rho'(\delta) \gets \frac12 \max_{i\in \iset{d}} \rho_i'(\delta)$.
%\STATE $\hat{\rho}(\delta) \gets 2\max(\rho(\delta),\rho'(\delta)) + \max(\rho(\delta),\rho'(\delta))^2$.
%%$\hat{\rho}(\delta) \gets \rho(\delta) + \rho'(\delta) + \rho'(\delta)\rho(\delta)$. % THIS WAS BUGGY
%\STATE Half-width of interval: $w \gets 
%\hat{\rho}(\delta)+ \parens{ 1+ \hat{\rho}(\delta) } \, \sqrt{ \sum_{i,j} \frac{\hat{\pi}_i}{\hat{\pi}_j} \wh{\Delta}^2_{i,j}(\delta)}
%$.
%\STATE Confidence interval for $\gap$: $U_n(\delta) \gets [\hatgap-w,\hatgap+w]\cap [0,1]$.
\end{algorithmic}
\end{algorithm}

The main result of this section is the following result:
\todoc{Conditions?}
\todod{I think they are fine now. Ergodicity is defined in section 3
now, and we discuss why it already implies the $\pimin>0$ and
$\gap>0$.}
\begin{theorem}
  \label{thm:empirical}
  Suppose \cref{alg:empest} is given as input a sample path of length
  $n$ from an ergodic and reversible Markov chain and confidence
  parameter $\delta \in (0,1)$.
  Let $\gap>0$ denote the spectral gap, $\vpi$ the unique stationary
  distribution, and $\pimin>0$ the minimum stationary probability.
  Then, on an event of probability at least $1-\delta$,
  \[
    \pi_i \in [\hat\pi_i-\hat{b},\hat\pi_i+\hat{b}]
    \quad \text{for all $i \in [d]$} ,
    \qquad\text{and}\qquad
    \gap \in [\hatgap-\hat{w},\hatgap+\hat{w}]
    .
  \]
  Moreover, $\hat{b}$ and $\hat{w}$ almost surely satisfy (as $n \to
  \infty$)
  \[
    \hat{b}
    =
    O\Parens{
      \max_{(i,j) \in [d]^2}
      \kappa
      \sqrt{\frac{P_{i,j}\log(d/\delta)}{\pi_i n}}
    }
    ,
    \quad
    \hat{w}
    =
     O\Parens{ 
     \frac{\kappa}{\pimin} \sqrt{\frac{\ln(d/\delta)}{\pimin n}} + 
      \sqrt{ \frac{d\ln(d/\delta)}{ \pimin n}}
    }
    ,
  \]
  where $\kappa$ is analogous to the $\hat{\kappa}$ computed
  by~\cref{alg:empest} with $\giAh$ replaced by the group inverse
  $\giA$ of $\vA := \vI - \vP$.
\end{theorem}
The proof of \cref{thm:empirical} is given in \cref{app:empirical}.
As mentioned above, the obstacle encountered in \cref{thm:err} is
avoided by exploiting the Markov property.
We establish fully observable upper and lower bounds on the entries of
$\vP$ that converge at a $\sqrt{n}$ rate using standard martingale
tail inequalities; this justifies the validity of the bounds from
Step~\ref{step:P-bound}.
Properties of the group inverse~\citep{meyer1975role,cho2001comparison} and eigenvalue
perturbation theory~\citep{stewart1990matrix} are used to validate the
empirical bounds on $\pi_i$ and $\gap$ developed in the remaining
steps of the algorithm.

The first part of \cref{thm:empirical} provides valid empirical
confidence intervals for each $\pi_i$ and for $\gap$, which are
simultaneously valid at confidence level $\delta$.
The second part of \cref{thm:empirical} shows that the width of the
intervals decrease as the sequence length increases.
It is known that $\kappa \le d/\gap$ (see \cref{sec:asymptotic}),
which gives
\[
  \hat{b}
  =
  O\Parens{
    \max_{(i,j) \in [d]^2}
    \frac{d}{\gap}
    \sqrt{\frac{P_{i,j}\log(d/\delta)}{\pi_i n}}
  }
  , \quad
  \hat{w}
  =
  O\Parens{ 
    \frac{d}{\pimin\gap} \sqrt{\frac{\log(d/\delta)}{\pimin n}}
  }
  .
\] 
The widths of these intervals are somewhat larger than the corresponding
(non-empirical) intervals from \cref{thm:err}.
\todoc{This needs to be updated once we finalize the results. By how much are these larger?}

An easy way to improve the asymptotic width is to combine
\cref{thm:err,thm:empirical}.
Specifically, determine lower bounds on $\pimin$ and $\gap$ using
\cref{alg:empest},
\[
  \pimin \geq \min_{i \in [d]} \brackets{ \hat\pi_i - \hat{b} }_+
  , \quad
  \gap \geq \brackets{ \hatgap - \hat{w} }_+
  ;
\]
then plug-in these lower bounds for $\pimin$ and $\gap$ in the
deviation bounds in \cref{eq:gapbound} from \cref{thm:err}.
This yields a new interval centered around the estimate of $\gap$ from
\cref{thm:err}, and it no longer depends on unknown quantities.
The interval is a valid $1-2\delta$ probability confidence interval
for $\gap$, and for sufficiently large $n$, the width shrinks at the
rate given in \cref{eq:gapbound}.
We can similarly construct an empirical confidence interval for
$\pimin$ using \cref{eq:piminbound}, which is valid on the same
$1-2\delta$ probability event.\footnote{%
  For the $\pimin$ interval, we only plug-in lower bounds on $\pimin$
  and $\gap$ only where these quantities appear as $1/\pimin$ and
  $1/\gap$ in \cref{eq:piminbound}.
  It is then possible to ``solve'' for observable bounds on $\pimin$.
  See \cref{app:combined} for details.%
}
This is summarized in the following \lcnamecref{thm:combined}, which
we prove in \cref{app:combined}.
\begin{theorem}
  \label{thm:combined}
  The following holds under the same conditions as
  \cref{thm:empirical}.
  For any $\delta \in (0,1)$, the confidence intervals $\wh{U}$ and
  $\wh{V}$ described above for $\pimin$ and $\gap$, respectively,
  satisfy $\pimin \in \wh{U}$ and $\gap \in \wh{V}$ with probability
  at least $1-2\delta$.
  Furthermore, the widths of these intervals almost surely satisfy
  (as $n \to \infty$) \todoc{Do we also care about $\pimin\to 0$, $\gap\to 0$? Do we need then other terms for both? Replicating the form of bounds from \cref{thm:err}?}
  \todod{I don't understand the first question. For the second, we
  could do this, but it would take some non-trivial work\ldots}
  \[
    |\wh{U}|
    =
    O\Parens{
      \sqrt{\frac{\pimin\log\frac{d}{\pimin\delta}}{\gap n}}
    }
    ,
    \quad
    |\wh{V}|
    =
    O\Parens{
      \sqrt{\frac{\log\frac{d}{\delta}\cdot\log(n)}{\pimin\gap n}}
    }
    .
  \]
\end{theorem}

%The rest of the section is devoted to the proof of \cref{thm:empirical}.

%In what follows, we will (re-)define  $\wh{\Delta}_{i,j}(\delta)$ to be the RHS of \eqref{eq:P-obs-bound} when $N_i>0$,
% \todoc{Abbrev RHS should be introduced.}
%while we let $\wh{\Delta}_{i,j}(\delta)=\max(\wh{P}_{i,j},1-\wh{P}_{i,j})$ otherwise.
%%Clearly, $\wt{\Delta}_{i,j}(\delta) \ge \wh{\Delta}_{i,j}(\delta)$.
%%In what follows, we calculate with the tighter $\wh{\Delta}_{i,j}(\delta)$. 
%%In fact, the algorithm was presented for $\wt{\Delta}_{i,j}(\delta)$ for easier comprehension only.
%

### 2015-06-01 intro-nips.tex
Copied files from svn

%!TEX root =  matrix-est.tex

%Performance guarantees play a key role in the analysis and design of machine learning algorithms.

In this paper we study the construction of fully empirical bounds on the mixing time of Markov chains based on a single sample path.
Let $(X_t)_{t=1,2,\dots}$ be an irreducible, aperiodic time-homogeneous Markov chain on a finite state space $[d]$ of $d$ states numbered by positive integers with transition matrix $\vP$. Under this assumption, the chain converges to its unique stationary \todoc{Maybe steady state is better?}
distribution $\vpi = (\pi_i)_{i=1}^d$ regardless of the distribution $\vq$ of $X_1$: 
$\lim_{t\to\infty} \Pr_{\vq}\Parens{X_t = i} = \lim_{t\to\infty} (\vq \vP^t)_i = \pi_i$, $1\le i \le d$.
The mixing time $\tmix$ of the Markov chain is the time that it takes for the chain to reach its stationary distribution, say,
\[
  \tmix
  :=
  \min\Braces{
    t \in \bbN :
    \sup_{\vq}
    \max_{A\subset [d]} |\textstyle\Pr_{\vq}\Parens{ X_t \in A } - \vpi(A)|
%    \Norm{
%      \vq \vP^t - \vpi
%    }_{\tv} 
    \leq 1/4
  }\,.
\]
Here $\vpi(A) = \sum_{i\in A} \pi_i$ is the total probability assigned to set $A$ by $\vpi$ and the supremum is over
all possible initial distributions $\vq$.
The problem studied in the paper is the construction of a confidence set $C_n = C_n(X_1,\dots,X_n,\delta) \subset [0,\infty]$ based on the knowledge of $X_1,\dots,X_n$ only that captures the mixing time $\tmix$ with a pre-specified probability $0\le \delta \le 1$. 
In the remainder of the introduction, we describe the background of why this problem might be interesting, followed by a 
brief description of our main results.

In numerous scientific applications and some machine learning tasks
one is interested in estimating the mean $\vpi(f) = \sum_i \pi_i f(i)$ of some function $f$ of the states of a Markov chain. 
For example, this is the realms of the celebrated Markov Chain Monte Carlo (MCMC) paradigm, but 
the problem also arises in performance prediction when working with time correlated data, as is common in,
for example, in reinforcement learning.
%as 
% (cf. \cref{sec:mot}) 
Under our assumptions,
the law of large numbers guarantees that the sample mean $\vpi_n(f) \defeq \frac1n \sum_{t=1}^n f(X_t)$ 
converges almost surely to $\vpi(f)$, while the central limit theorem tells us that as $n\to \infty$, 
the distribution of the deviation $\sqrt{n}( \vpi_n(f)-\vpi(f))$ will be normal with zero mean and asymptotic
variance $\sigma^2_{\mathrm{as}}(f) \defeq \lim_{n\to\infty} n \Var\Parens{ \vpi_n(f) }$. \todoc{shall we add a citation here?}
While these results help us to understand how the sample mean over a Markov chain behaves in the limit,
they tell us little about the finite time, non-asymptotic behavior of the sample mean,
which are needed for prudent evaluation of one's method or even for algorithm design 
\cite{
MCMCDiscussion93%
,DBLP:conf/valuetools/KontoyiannisLM06%
,BBL06%
,MniSzeAu08%
,MauPo09%
,LiLiWaSt11:KWIK%
,flegal2011implementing%
,Gyori-paulin15%
,SwaJoa15:LoggedBandit%
}. %BaHaVa10:activelearning,ShaVo08:conformalpred
To address this need, numerous works have been devoted to deriving
Chernoff-type, exponential bounds for upper bounding the tail probabilities $p(r) = \Pr\Parens{ |\vpi_n(f)-\vpi(f)|> r }$,
also known as ``concentration bounds'', e.g., 
\cite{gillman1998chernoff,leon2004optimal,DBLP:conf/valuetools/KontoyiannisLM06,paulin15}.
Compared with the case when $(X_t)_t$ is an independent, identically distributed (iid) sequence of
random variables, these tail probabilities must be larger due to the temporal dependence of the sequence $(X_t)_t$.
Intuitively, for the Markov chain to forget the state where it is at time $t$, or in other words to get a 
``fresh'' sample $X_{t'}$ that is approximately ``independent'' from $X_t$, one must wait $\Theta(\tmix)$ time steps.
Thus, in the bounds on $p(r)$, one expects to see that the sample size $n$ 
gets appropriately reduced as $\tmix$ grows and this is indeed the case in bounds mentioned above.
is replaced by $n/\tmix$, which one
might think of as an approximate sample size. 
Indeed, all the above mentioned works get bounds on $p(r)$ that depend on various functions of $\vP$.
Namely, Kontoyiannis et al. obtain bounds that depend on certain information-theoretic quantities defined implicitly
as a function of $\vP$ and $f$, which are after simplification is bounded based on quantities related to 
a regeneration-type quantity derived from $\vP$
\cite{DBLP:conf/valuetools/KontoyiannisLM06}, while the other works mentioned above, in the case of reversible chains,
derive bounds that depend on the spectral gap of $\vP$.
Note that the spectral gap for such chains is essentially inversely related to the mixing time of the chain (see \cref{sec:setting}).
%\cite{gillman1998chernoff,leon2004optimal,,paulin15}.
Much research have been devoted to finding appropriate \emph{a priori} bounds on the various quantities that appear 
in these bounds, mostly in the theory community (e.g., \cite{saloff2004random}). 
To address the lack or looseness of such a priori bounds, a significant effort also went into developing methods
that estimate these quantities based on data, mostly in the MCMC community in an effort to develop
effective methods that are able to diagnose whether a chain has reached its stationary state 
(e.g., \cite{MCMCDiscussion93,flegal2011implementing,Gyori-paulin15}).
While the typical results state the asymptotic consistency of these estimators, or a central limit theorem (which 
are of little help to achieve a fully empirical, non-asymptotic bound on $p(r)$),
Gyori and Paulin give a finite time error bound for these estimates. 
Unsurprisingly, these error bounds themselves
depend on other unknown quantities of the Markov chain (which can be bounded in terms of the mixing time, again)
\cite{Gyori-paulin15}.

This situation is similar to what appears when one attempts to use a Bernstein-type deviation bound to 
control the deviation between the sample mean of a sequence of iid random variables from their common mean
with a known range, but unknown variance: For a tight bound, one wishes to use the unknown variance, but in the deviation
bound for the error of a variance estimate, the variance appears again.
While in the case of Bernstein-bounds this circular reference leads to a nontrivial bound
\cite{audibert2009}, it appears that the situation is different in the Markov setting.
Focusing on the spectral gap, we will see that if one attempts to use a similar technique,
the set of feasible values based on the two error bounds always will never rule out that the spectral gap is zero,
or that the mixing time is infinite. Can one even hope to obtain 
after finitely many samples a nontrivial upper bound on the mixing time?
Obviously, for any finite sample size, if the mixing time of the chain is larger than the sample size than a sound method
must return infinity as the upper bound on the mixing time. Hence, the question is whether one can design a method
that never fails and if the sample size is appropriately large, returns a nontrivial bound on the mixing time, adapting the time it needs to trap the mixing time in a fixed size confidence interval to the unknown mixing time.

\hide{
\subsection{Motivation}
\label{sec:mot}
One key competence in statistical learning is the ability to predict one's performance on future data.
To make performance prediction possible, one starts by posing some assumptions on how the data arises.
Simpler assumptions lead to simpler algorithms and analysis, while they may also limit applicability of the
obtained results and algorithms.
The most common assumption employed postulates that all data points, including past and future data points,
are sampled independently of each other from a common underlying distribution (e.g., \cite{SSBD14:UnderstandingML}).
This is helpful on two accounts: 
The lack of dependence between the data points means that sample averages of bounded functions of the data 
will converge at a uniform rate (irrespective of the distribution), while the assumption that all data points share the same
distribution means that one can learn about the distribution of future data points based on past data.
}

\hide{
It is well-known that contrary to the idealized assumptions of the standard
learning model, real data is almost never iid.
Fortunately, many classical results of statistics and statistical
learning theory extend to the dependent data in a more or less
seamless fashion, at least when the data is sufficiently ``mixing''.
Such results include the central limit theorem, 
exponential tail bounds for empirical processes, 
performance guarantees for complexity regularization
using sieves or roughness penalties, as well as 
results based on stability arguments or Rademacher complexity 
\citep{Woo86,Dou94,Yu94,ZheChu96,dlPenaGine99,Mei00,
  MR1921877,vid2003,gamarnik03,LoKuScha05,MoRo08,MoRo09,
  DBLP:conf/nips/SteinwartC09,Steinwart2009175,
  FaSze11,
  london:nips12asalsn,london:icml13,
  DBLP:conf/nips/ShaliziK13}.

Unfortunately, a distribution-free theory of non-iid learning is simply
impossible, when arbitrary dependence among the sample points is allowed.
Hence, one must quantify this dependence via the various mixing coefficients,
which appear in the non-iid generalization bounds mentioned above.
A frequently raised concern is that now the bounds depend on {\em
unknown} properties of the distribution, which precludes their
actual use in algorithmic applications (e.g., active
learning~\citep{BBL06,DHM07}).
%, and any algorithm aiming at a fully empirical error bound must
%also estimate the mixing coefficients.
In addition to learning-theoretic generalization bounds, the need for
fully empirical confidence bounds arises in Markov Chain Monte Carlo
(MCMC) simulations---so as to halt the run when sufficiently close to
the stationary distribution \citep{galin06}, and in Markov Reward
Processes \citep{sutton98}, as this will affect the learner's policy.
As \citet{Mei00} notes, in the absence of accurate estimates of the
mixing time, such fully empirical bounds remain ``a theoretician's
dream''.

Our goal here is to develop an algorithm that, given a single sample
path of a Markov chain, computes non-trivial bounds on the mixing time
of the Markov chain.
We remark on the inherent chicken-and-egg nature of the challenge:
confidence intervals derived from a Markov chain---say, on the the
chain's mixing time---are typically given in terms of the chain's
mixing time.
Our way out of this conundrum is to give {\em empirical} (i.e.,
\emph{data-dependent}) as opposed to {\em a priori} guarantees.
To illustrate the intuition, consider a Markov chain on two states,
$\braces{0,1}$.
A learner who is stuck in state $0$ for a long time learns very little
about the chain and hence cannot be expected to give meaningful upper
bounds on its mixing time.
Consider, however, a learner who is lucky enough
to be in a chain that switches frequently between the two states.
Surely at some finite time, the learner will possess meaningful and nontrivial
information about the chain's mixing time --- crucially, without any
{\em prior} knowledge regarding this quantity. This simple example captures
the nature of our results: in some cases, we will not be able to say
anything meaningful (but the learner will recognize that this is the case),
while in ``luckier'' situations, we will be able to provide nontrivial
upper estimates on the mixing time.
}

\paragraph{Main results.}
We propose what is apparently the first efficient and fully empirical
estimator of Markov mixing time.
Consider a reversible ergodic Markov chain on $d$ states with spectral
gap $\gap$ (which is well-known to govern mixing time)
and stationary distribution minorized by $\pimin$.
Our main results are summarized as follows.
\begin{enumerate}
  \item
    In \cref{thm:err}, we give an estimator for $\gap$ that
    achieves multiplicative accuracy from \emph{a single sample path}
    of length $\tilde{O}\parens{1/(\pimin\gap^3)}$.\footnote{The
    $\tilde{O}(\cdot)$ notation suppresses logarithmic factors.}
    We also provide an estimator for $\pimin$ that requires a sample
    path of length $\tilde{O}\parens{1/(\pimin\gap)}$.
    However, the valid confidence intervals suggested by
    \cref{thm:err} depend on the unknown quantities $\pimin$ and
    $\gap$.

  \item
    In \cref{thm:lb-pimin,thm:lb-gap}, we show that $\Omega\parens{
    (d\ln d)/\gap + 1/\pimin }$ observations are necessary for
    constant multiplicative accuracy in estimating $\gap$.

  \item
    In \cref{thm:empirical}, we develop a procedure for computing
    \emph{valid empirical confidence intervals} for $\pimin$ and
    $\gap$.
    The procedure uses just a single sample path and does not require
    any prior knowledge about $\pimin$ or $\gap$.
    We also quantify the rate at which the width of the confidence
    intervals shrink as a function of the length of the sample path.

    We show in \cref{thm:combined} how to combine these empirical
    confidence intervals with the non-empirical bounds from
    \cref{thm:err} to produce valid empirical confidence intervals;
    the width of these intervals converge to zero asymptotically as
    fast as those from \cref{thm:err}.

%    which is
%    $     O\Parens{
%      \frac1{\pimin}\sqrt{\frac{d\ln(d/\delta)}{ n}}
%      +
%      \frac{d}{\pimin\gap} \sqrt{\frac{\ln(d/\delta)}{\pimin n}}
%    }.$

%    \todoc{please add rate}

\end{enumerate}

\paragraph{Related work.} \todoc{Move this later? It also needs to be updated, to 
adjust it to the current intro.}
Owing to the wide applicability of MCMC methods to various computational
problems, much effort has been devoted to developing effective convergence
diagnostic methods the aim of deriving stopping rules or stationarity
diagnoses. While a large number of heuristic techniques have been developed
\citep[e.g.,][]{MCMCDiscussion93,CoCa96}, it is widely recognized that all
these methods may fail to meet the guarantees that they are designed for.
Some recent advance in this area \citep{1209.0703,Gyori-paulin15} have
yielded practical algorithms for estimating the rates of convergence
of various MCMC algorithms, but they fall short of providing intervals
that are valid with any finite length sequence.

In the theory community, \citet{BaFoRuSmiWhi00,BaFoRuSmiWhi13}
developed algorithms for testing whether a given Markov chain over a
finite state space of size $d$ is rapidly mixing.
In particular, assuming access to a sampling oracle that can generate
independent transitions from any given state, their algorithm can
distinguish between chains that in $t$ steps for a fixed $t$ reach a
distribution close to the stationary distribution using $\tilde{O}(t
d^{5/3})$ oracle calls.
It is remarkable that when $t$ is small enough, the procedure requires
sublinearly many samples in the total number of transitions in the
Markov chain.
For Markov chains with exponentially-large state spaces (with a
circuit-based description of the transition probability structure),
\citet{BhaBoMo11} establishes complexity-theoretic barriers for
similar MCMC diagnostic problems.
%A similar question was studied by \citet{BhaBoMo11} in the context of
%Markov chains with very large state spaces.
%They assume an circuit based description of the
%transition probability structure of the matrix and derive several hardness
%results for Markov chains whose state space consists of all binary strings of
%a given length. For example,
%they prove that given a Markov chain that mixes
%rapidly, it is $\mathsf{SZK}$-hard to distinguish whether the chain is close to
%stationarity by time $t$ or far from stationarity at time $ct$ for a
%constant $c$ when started from a given initial state ($\mathsf{SZK}
%\subset \mathsf{AM} \cap \mathsf{coAM}$, which is conjectured to
%extend beyond $\mathsf{BPP}$ \citep{GoVa11}).\footnote{%
%  $\mathsf{SZK}$ stands for statistical zero-knowledge, $\mathsf{BPP}$
%  stands for probabilistic polynomial-time, $\mathsf{AM}$ stands for
%  ``Arthur-Merlin protocol''.%
%}

Other works include that of \citet{McDoShaSche11} who propose to estimate
the $\beta$-mixing coefficients of a stationary process possessing a density
based on histogram estimates of the underlying joint density. Their main
result is a consistency result for a fixed ``lag''. For fast mixing processes
(e.g., for $k$-step Markov processes) their result yields a convergence rate,
which unfortunately is given in terms of the unknown mixing coefficients.
%Catch-22 problem remains: The rate of convergence of their estimates depends on the unknown mixing coefficients
%(Theorem 2.4)
% $\exp(W(\ln(n)))$ is the window length, where $W$ is Lambert's $W$-function (i.e., this is between $\ln \ln(n)$ and $\ln n$).
The earlier study of \citet{Nobel06} considers the generalization of
existing discernibility results for iid processes to weakly uniformly
ergodic families, and shows how this can be used to study---and in
some cases estimate---polynomial mixing rates for dependent processes.

### 2015-06-01 matrix-est.tex
Copied files from svn

%\documentclass[11pt]{article}
%\documentclass[anon,12pt]{colt2015} 
\documentclass{article}
\usepackage{nips15submit_e,times}
\input{__includes}
\renewcommand\citet\citep % since NIPS is using numbered refs

%\title{Single sample path based estimation of the mixing time in finite, reversible Markov chains}
\title{%
  Mixing time estimation in reversible Markov chains from a single
  sample path%
}

\author{
David S.~Hippocampus\thanks{ Use footnote for providing further information
about author (webpage, alternative address)---\emph{not} for acknowledging
funding agencies.} \\
Department of Computer Science\\
Cranberry-Lemon University\\
Pittsburgh, PA 15213 \\
\texttt{hippo@cs.cranberry-lemon.edu} \\
\And
Coauthor \\
Affiliation \\
Address \\
\texttt{email} \\
\AND
Coauthor \\
Affiliation \\
Address \\
\texttt{email} \\
\And
Coauthor \\
Affiliation \\
Address \\
\texttt{email} \\
\And
Coauthor \\
Affiliation \\
Address \\
\texttt{email} \\
(if needed)\\
}

\begin{document}
\maketitle

\begin{abstract} 
\input{abstract}
\end{abstract}

\if0
\begin{keywords}
  Markov chain; mixing time; spectral gap;
  fully data-dependent confidence interval
\end{keywords}
\fi

\section{Introduction}\label{sec:intro}
\input{intro-nips}

\section{Notation}\label{sec:notation}
\input{notation}

\section{Setting}\label{sec:setting}
\input{setting}

%\section{Results}
%\input{results}

\section{Convergence rates}\label{sec:rates}
\input{rates}

\section{Fully empirical confidence intervals}\label{sec:empirical}
\input{empirical}

\section{Discussion}\label{sec:discussion}
\input{conclusions}

\subsubsection*{References}
{\def\section*#1{}\small \bibliography{all} \bibliographystyle{myunsrt}}

\newpage

\appendix

\section{Proof of \cref{thm:err}}\label{app:upper}
\input{upperbound-new}

\section{Proofs of lower bounds}\label{app:lower}
\input{lowerbound}

\section{Proof of \cref{thm:empirical}}\label{app:empirical}
\input{appendix-empirical}

\section{Proof of \cref{thm:combined}}\label{app:combined}
\input{appendix-combined}

\end{document}

### 2015-06-01 notation.tex
Copied files from svn

%!TEX root =  matrix-est.tex
We denote the set of positive integers by $\bbN$, and
the set of the first $d$ positive integers $\{1,2,\dotsc,d\}$ by $\iset{d}$.
The non-negative part of a real number $x$ is $[x]_+ := \max\{0,x\}$.
Boldface symbols are used for vectors and matrices (e.g., $\vv$,
$\vM$), and their entries are referenced by subindexing (e.g., $v_i$,
$M_{i,j}$).
%The $i$-th coordinate basis vector is denoted by $\ve_i$.
For a vector $\vv$, $\norm{\vv}$ denotes its Euclidean norm; for a
matrix $\vM$, $\norm{\vM}$ denotes its spectral norm.
We use $\Diag(\vv)$ to denote the diagonal matrix whose $(i,i)$-th
entry is $v_i$.
The probability simplex is denoted by $\Delta^{d-1} = \{ \vp
\in [0,1]^d : \sum_{i=1}^d p_i = 1 \}$, and we regard vectors in
$\Delta^{d-1}$ as row vectors.
Finally, the total variation between two probability distributions
$\vp, \vq \in \Delta^{d-1}$ is denoted by $\norm{\vp-\vq}_{\tv}$.

### 2015-06-01 rates.tex
Copied files from svn

%!TEX root =  matrix-est.tex

In this section, we present upper and lower bounds on achievable rates
for estimating the spectral gap as a funciton of the length of the
sample path $n$.

\subsection{Upper bounds}
\label{sec:rates-upper}

Our upper bound is demonstrated using estimators based on natural
plug-in estimates.
(We also provide an estimator for the minimum stationary probability
in the process.)

Define the random matrix $\wh\vM \in [0,1]^{d \times d}$ and random
vector $\hat\vpi \in \Delta^{d-1}$ by
\begin{align*}
  \wh{M}_{i,j}
  & := \frac{|\{ t \in [n-1] : (X_t,X_{t+1}) = (i,j) \}|}{n-1}
  , \quad i,j \in [d]\,,
  \\
  \hat{\pi}_i
  & := \frac{|\{ t \in [n] : X_t = i \}|}{n}
  , \quad i \in [d]
  .
\end{align*}
Furthermore, define
\[
  \Sym(\wh\vL) := \frac12 \parens{ \wh\vL + \wh\vL^\t }
\]
to be the symmetrized version of the (possibly non-symmetric) matrix
\[
  \wh\vL := \Diag(\hat\vpi)^{-1/2} \wh\vM \Diag(\hat\vpi)^{-1/2}
  .
\]
Let $\hat\lambda_1 \geq \hat\lambda_2 \geq \dotsb \geq \hat\lambda_d$
be the eigenvalues of $\Sym(\wh\vL)$.
Our estimator of the minimum stationary probability $\pimin$ is
\[
  \hatpimin := \min_{i \in [d]} \hat\pi_i ,
\]
and our estimator of the spectral gap $\gap$ is
\[
  \hatgap := 1 - \max\{ \hat\lambda_2, |\hat\lambda_d| \} .
\]

These estimators have the following accuracy guarantees. 
\begin{theorem}
  \label{thm:err}
  There exists an absolute constant $C>0$ such that the following
  holds.
  Assume the estimators $\hatpimin$ and $\hatgap$ described above are
  formed from a sample path of length $n$ from an ergodic and
  reversible Markov chain.
  Let $\gap>0$ denote the spectral gap and $\pimin>0$ the minimum
  stationary probability.
  For any $\delta \in (0,1)$, with probability at least $1-\delta$,
  \begin{equation}\label{eq:piminbound}
    \Abs{\hatpimin-\pimin}
    \le
    C \,
    \Parens{
      \sqrt{\frac{\pimin\log\frac{d}{\pimin\delta}}{\gap n}}
      +
      \frac{\log\frac{d}{\pimin\delta}}{\gap n}
    }
  \end{equation}
  and
  \begin{equation}\label{eq:gapbound}
    \Abs{\hatgap-\gap}
    \leq
    C \,
    \Parens{
      \sqrt{\frac{\log\frac{d}{\delta}\cdot\log\frac{n}{\pimin\delta}}{\pimin\gap n}}
      + \frac{\log\frac{d}{\delta}\cdot\log\frac{n}{\pimin\delta}}{\pimin\gap n}  
      + \frac{\log\frac{1}{\gap}}{\gap n}  
    }
    \,.
  \end{equation}
\end{theorem}
\todoc{In \eqref{eq:gapbound}, we should be able to merge the second term into the first, right?}
\todod{No, not for all values of $n$.}

\Cref{thm:err} implies that the sequence lengths required to estimate
$\pimin$ and
$\gap$ to within constant multiplicative factors are, respectively,
\[
  \tilde{O}\Parens{\frac1{\pimin\gap}}
  \quad\text{and}\quad
  \tilde{O}\Parens{\frac1{\pimin\gap^3}}
  .
\]

The proof of \cref{thm:err} is based on analyzing the
convergence of the sample averages $\wh{\vM}$ and
$\hat\vpi$ to their expectation, and then using perturbation bounds
for eigenvalues to derive a bound on the error of $\hatgap$.
However, since these averages are formed using a \emph{single sample
path} from a (possibly) non-stationary Markov chain, we cannot use
standard large deviation bounds; moreover applying Chernoff-Hoeffding
bounds for Markov chains to each entry of $\wh{\vM}$ would result in a
significantly worse sequence length requirement, roughly a factor of
$d$ larger.
%\todoc{In terms of the dependence on the number of states, right? Can we be more specific? Factor of $d$ more, right?}
Instead, we adapt probability tail bounds for sums of independent
random matrices~\citep{tropp2015intro} to our non-iid setting by
directly applying a blocking technique of~\citet{Bernstein27} as
described in the article of~\citet{Yu94}.
The proof is given in \cref{app:upper}.

Note that because the eigenvalues of $\vL$ are the same as that of the
transition probability matrix $\vP$, 
we could have instead opted to
estimate $\vP$, say, using simple frequency estimates obtained from
the sample path, and then computing the second largest eigenvalue of
this empirical estimate $\wh\vP$.
The difficulty with this approach is that $\vP$ lacks the structure
required by certain strong eigenvalue perturbation results.
One could instead invoke the Ostrowski-Elsner theorem
\citep[cf.~Theorem 1.4 on Page 170 of][]{stewart1990matrix}, which
bounds the \emph{matching distance} between the eigenvalues of a
matrix $\vA$ and its perturbation $\vA+\vE$ by $O(\norm{\vE}^{1/d})$.
%\footnote{Actually, this result is not even readily applicable: 
%the Haussdorff distance is actually an optimistic estimate of the deviation between
%$\gap$ and its estimate.}
Since $\norm{\wh\vP-\vP}$ is expected to be of size $O(n^{-1/2})$,
this approach will give a confidence interval for $\gap$ whose width
shrinks at a rate of $O(n^{-1/(2d)})$---an exponential slow-down
compared to the rate from \cref{thm:err}.
As demonstrated through an example from \citet{stewart1990matrix}, the
dependence on the $d$-th root of the norm of the perturbation cannot
be avoided in general.
Our approach based on estimating a symmetric matrix affords us the use
of perturbation results that exploit more structure.
%(In \cref{sec:empirical}, we develop another estimator of
%$\vL$---in order to exploit symmetry---that is based on estimating
%$\vP$ for reasons discussed below.)

Unfortunately, \cref{thm:err} falls short of being directly
usable for obtaining a fully empirical confidence interval for $\gap$
and $\pimin$, at least without further assumptions.
This is because the deviation terms themselves depend inversely both
on $\gap$ and $\pimin$, and hence can never rule out $0$ (or an
arbitrarily small positive value) as a possibility for $\gap$ or
$\pimin$.\footnote{%
  Using \cref{thm:err}, it is possible to trap $\gap$ in the
  union of \emph{two} empirical confidence intervals---one around
  $\hatgap$ and the other around zero, both of which shrink in width
  as the sequence length increases.%
}
In effect, the fact that the Markov chain could be slow mixing and the
long-term frequency of some states could be small makes it difficult
to be confident in the estimates provided by $\hatgap$ and
$\hatpimin$.
This suggests that in order to obtain fully empirical confidence
intervals, we need an estimator that is not subject to such
effects---we pursue this in \cref{sec:empirical}.
\Cref{thm:err} thus primarily serves as a point of comparison
for what is achievable in terms of estimation accuracy when one does
not need to provide empirical confidence bounds.

\subsection{Lower bounds}

We complement \cref{thm:err} with lower bounds showing the
necessity of some of the quantities appearing in the upper bound.
We begin with a result showing that any algorithm for estimating
$\gap$ (even for just constant additive accuracy) requires an
observation sequence of length $\Omega(1/\pimin)$
even when the number of states is only two.

\begin{theorem}
  \label{thm:lb-pimin}
  There is an absolute constant $c>0$ such that the following holds.
  Pick any $\bar\pi \in (0,1/4)$.
  Consider any estimator $\hatgap$ that takes as input a random sample
  path of length $n < c/\bar\pi$ from a Markov chain starting from any
  desired initial state distribution.
  There exists a two-state ergodic and reversible Markov chain
  distribution with spectral gap $\gap \geq 1/2$ and minimum
  stationary probability $\pimin \geq \bar\pi$ such that
  \[
    \Pr\Brackets{ |\hatgap - \gap| \geq 1/6 } \geq 1/8 .
  \]
\end{theorem}

Next, we show that in order to estimate
$\gap$ to within a constant multiplicative accuracy in a $d$ state chain,
a sequence of length $\Omega(d\log(d)/\gap)$ is required.
Essentially, the sequence may have to visit all $d$ states $\log(d)/\gap$ times each on average.
This hold \emph{even} if $\pimin$ is with a factor of $1/2$ of the \emph{largest} possible value of $1/d$ that it can take,
i.e., when $\vpi$ is nearly uniform.
\begin{theorem}
  \label{thm:lb-gap}
  There is an absolute constant $c>0$ such that the following holds.
  Pick any positive integer $d \geq 3$ and any $\bar\gamma \in
  (0,1/2)$.
  Consider any estimator $\hatgap$ that takes as input a random sample
  path of length $n < c d\ln(d) / \bar\gamma$ from a $d$-state
  reversible Markov chain starting from any desired initial state
  distribution.
  There is an ergodic and reversible Markov chain distribution
  with spectral gap $\gap \in [\bar\gamma,2\bar\gamma]$ and minimum
  stationary probability $\pimin \geq 1/(2d)$ such that
  \[
    \Pr\Brackets{ |\hatgap - \gap| \geq \bar\gamma} \geq 1/4 .
  \]
\end{theorem}

The proofs of \cref{thm:lb-pimin,thm:lb-gap} are given in
\cref{app:lower}.

### 2015-06-01 scratch/aryeh-scratch-notes.tex
Copied files from svn

\documentclass[11pt]{article}
\usepackage{todonotes}
\newcommand{\todoc}[2][]{\todo[size=\scriptsize,color=green!10!white,#1]{Cs: #2}} % Csaba's comments
\newcommand{\todot}[2][]{\todo[size=\scriptsize,inline,color=blue!20!white,#1]{D: #2}} % Daniel's comments
\newcommand{\todor}[2][]{\todo[size=\scriptsize,color=orange!20!white,#1]{A: #2}} % Aryeh's comments

\usepackage{amsmath,amsbsy,amsfonts,amssymb,amsthm,color,dsfont}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% PACKAGES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{latexsym}
\usepackage{fullpage}
\usepackage{amsmath}
\usepackage{amssymb}
%\usepackage{mathtools}
\usepackage{enumerate}
\usepackage{accents}
\usepackage{tikz}
%\usepackage{pgfplots}
\usepackage[ruled]{algorithm}
\usepackage{algpseudocode}
\usepackage{dsfont}
\usepackage[bf]{caption}
\usepackage{hyperref}
\hypersetup{
    bookmarks=true,         % show bookmarks bar?
    unicode=false,          % non-Latin characters in Acrobats bookmarks
    pdftoolbar=true,        % show Acrobats toolbar?
    pdfmenubar=true,        % show Acrobats menu?
    pdffitwindow=false,     % window fit to page when opened
    pdfstartview={FitH},    % fits the width of the page to the window
    pdftitle={My title},    % title
    pdfauthor={Author},     % author
    pdfsubject={Subject},   % subject of the document
    pdfcreator={Creator},   % creator of the document
    pdfproducer={Producer}, % producer of the document
    pdfkeywords={keyword1} {key2} {key3}, % list of keywords
    pdfnewwindow=true,      % links in new window
    colorlinks=true,       % false: boxed links; true: colored links
    linkcolor=red,          % color of internal links (change box color with linkbordercolor)
    citecolor=blue,        % color of links to bibliography
    filecolor=magenta,      % color of file links
    urlcolor=cyan           % color of external links
}
\usepackage{amsthm}
\usepackage{times}
\usepackage{natbib}
\usepackage{nicefrac}
\usepackage{wrapfig}
\usepackage[capitalize]{cleveref}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% MACROS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\defined}{\vcentcolon =}
\newcommand{\rdefined}{=\vcentcolon}
%\newcommand{\E}{\mathbb E}
\newcommand{\Var}{\operatorname{Var}}
\newcommand{\calF}{\mathcal F}
\newcommand{\calR}{\mathcal R}
\newcommand{\sr}[1]{\stackrel{#1}}
\newcommand{\set}[1]{\left\{#1\right\}}
\newcommand{\ind}[1]{\mathds{1}\!\!\set{#1}}
\newcommand{\argmax}{\operatornamewithlimits{arg\,max}}
\newcommand{\argmin}{\operatornamewithlimits{arg\,min}}
\newcommand{\floor}[1]{\left \lfloor {#1} \right\rfloor}
\newcommand{\ceil}[1]{\left \lceil {#1} \right\rceil}

\def\subsubsect#1{\vspace{1ex plus 0.5ex minus 0.5ex}\noindent{\bf\boldmath{#1.}}}

\renewcommand{\P}[1]{\mathbb{P}\left\{#1\right\}}
\newcommand{\Prob}[1]{\mathbb{P}\left\{#1\right\}}
\newcommand{\EE}[1]{\mathbb{E}\left[#1\right]}

\let\temp\epsilon
\let\epsilon\varepsilon
\newcommand{\eps}{\varepsilon}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% THEOREMS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{example}[theorem]{Example}
\theoremstyle{remark}

\newcommand{\R}{\mathds{R}}

\newcommand\wh{\ensuremath{\widehat}}
\newcommand\wt{\ensuremath{\widetilde}}
\newcommand\norm[1]{\left\| #1 \right\|}
\newcommand\tvnorm[1]{\left\| #1 \right\|_{\mathrm{TV}}}
\newcommand\parens[1]{\left( #1 \right)}
\DeclareMathOperator{\Diag}{Diag}
\DeclareMathOperator{\Sym}{Sym}
\DeclareMathOperator*{\E}{\text{\bf E}}
\renewcommand\t{{\ensuremath{\scriptscriptstyle{\top}}}}
\newcommand\be{\ensuremath{\mathbf{e}}}
\newcommand\tmix{\ensuremath{t_{\mathrm{mix}}}}
\newcommand\htmix{\ensuremath{\hat{t}_{\mathrm{mix}}}}
\newcommand{\od}{\bar{d}}
\newcommand{\tcouple}{\tau_{\mathrm{couple}}}
\newcommand{\ZZ}{\mathcal{Z}}
\newcommand\trel{\ensuremath{t_{\mathrm{rel}}}}
\newcommand{\ip}[1]{\langle#1\rangle}
\newcommand{\DF}{\mathcal{E}}
\newcommand\ttmix{\ensuremath{t_{\mathrm{mix},2}}}

\newcommand{\paren}[1]{\left( #1 \right)}
\newcommand{\sqprn}[1]{\left[ #1 \right]}
\newcommand{\tlprn}[1]{\left\{ #1 \right\}}
%\newcommand{\set}[1]{\left\{ #1 \right\}}
\newcommand{\oo}[1]{\frac{1}{#1}}
\newcommand{\nrm}[1]{\left\Vert #1 \right\Vert}
\newcommand{\tnrm}[1]{\Vert #1 \Vert}
\newcommand{\abs}[1]{\left| #1 \right|}
\newcommand{\trn}{^\intercal} %operator transpose

\newcommand{\beq}{\begin{eqnarray*}}
\newcommand{\eeq}{\end{eqnarray*}}
\newcommand{\beqn}{\begin{eqnarray}}
\newcommand{\eeqn}{\end{eqnarray}}

\def\ddefloop#1{\ifx\ddefloop#1\else\ddef{#1}\expandafter\ddefloop\fi}
% \bbA, \bbB, ...
\def\ddef#1{\expandafter\def\csname bb#1\endcsname{\ensuremath{\mathbb{#1}}}}
\ddefloop ABCDEFGHIJKLMNOPQRSTUVWXYZ\ddefloop

\title{Aryeh's scratch notes}
%\author{Daniel, Aryeh, Csaba}

\begin{document}
\maketitle

%\begin{abstract}
%An abstract
%\end{abstract}
\newcommand{\XX}{\mathbb{X}}
\newcommand{\one}[1]{\mathbb{I}\left\{#1\right\}}
\newcommand{\N}{\mathbb{N}}
For any $\mu,a$ such that $n = 2\mu a$ and $\delta>2(\mu-1)\beta(a)$,
with probability at least $1-\delta$,
\begin{align}
\label{eq:xbound}
|\bar{X}_n - \EE{f(X_1)}|
\le c_1\sqrt{\frac{1}{\mu}} + c_2\sqrt{\frac{\ln(2/\delta')}{2\mu}}\,,
\end{align}
where $\delta'=\delta-2(\mu-1)\beta(a)$
If $\beta(k)$ denotes the $k$th $\beta$-mixing coefficient of $(X_i)$ then 
\begin{align}
\label{eq:betabound}
 \beta(k) \le \oo{\pi_*}e^{-k \gamma_*}
\end{align}
where  $\gamma_* = 1-\lambda_*$ is the spectral gap ($\lambda_*$ is the second largest eigenvalue of $T$)
and  $\pi_*  = \min_i \pi_i>0$ is the critical probability underlying the chain
(see the $\beta$-mixing note).
For a given value of $\pi_*,\gamma_*$, $\delta$, let $B_n(\pi_*,\gamma_*,\delta)$ be the optimal value of the optimization problem
\begin{align*}
\lefteqn{ \min_{\mu,a} \min\left(1, 
c_1\sqrt{\frac{1}{\mu}} + c_2\sqrt{\frac{\ln(2/\delta')}{2\mu}}\right) 
} \\
& \qquad \text{s.t.} \quad n = 2\mu a, \quad \delta>2(\mu-1)\oo{\pi_*}e^{-a \gamma_*},
\quad \delta' = \delta-2(\mu-1)\oo{\pi_*}e^{-a \gamma_*}, \quad \mu,a \in \N\,.
\end{align*}

\paragraph{We will upper-bound} $B_n$ as follows.
Choose
$$ a=(2\ln n)/\gamma_*.$$
Then
$\mu=n/2a=\gamma_*n/4\ln n$
and
\beq
\delta' &=& \delta-2(\mu-1)\oo{\pi_*}e^{-a \gamma_*} \\
&=& \delta - 2\paren{\frac{\gamma_*n}{4\ln n}-1}\oo{n^2\pi_*} \\
%&=& \delta - {\frac{\gamma_*}{2\pi_*n\ln n}} + O(n^{-2}) \\
&=&
\delta - \sqprn{
\frac{\gamma_*}{2\pi_*n\ln n}-\frac{2}{\pi_*n^2}
}
=:\delta-x.
%\frac{2}{\delta'}
%&=&
%\frac{2}{\delta} + 
\eeq
Now
\beq
\ln\frac{2}{\delta-x} &=& \ln\frac{2}{\delta}+\frac{x}{\delta^2}+O(x^2), \\
\sqrt{\ln\frac{2}{\delta-x}} &=& 
\sqrt{\ln\frac{2}{\delta}}
+
\frac{x}{2\delta\sqrt{\ln(2/\delta)}}+O(x^2), \\
\eeq
and hence,
\beq
B_n(\pi_*,\gamma_*,\delta)
&\le&
2\sqrt{\frac{\ln n}{\gamma_*n}}
\sqprn{
c_1
+ c_2
\sqrt{\ln(2/\delta')/2}
}\\
&=&
2\sqrt{\frac{\ln n}{\gamma_*n}}
\sqprn{
c_1
+ c_2
\paren{
\sqrt{
\oo2
\ln\frac{2}{\delta}}
+
\frac{x}{2\delta\sqrt{2\ln(2/\delta)}}+O(x^2)
}
}\\
&=&
2\sqrt{\frac{\ln n}{\gamma_*n}}
\sqprn{
c_1
+ c_2
\paren{
\sqrt{\oo2\ln\frac{2}{\delta}}
+
\frac{
\gamma_*
}{4{\pi_*n\ln n}\delta\sqrt{2\ln(2/\delta)}}+O(n^{-2})
}
}\\
&=&
2c_1\sqrt{\frac{\ln n}{\gamma_*n}}
+
2c_2\sqrt{\frac{\ln n}{2\gamma_*n}
\ln\frac{2}{\delta}}
+
c_2\frac{\sqrt{\gamma_*}}
{2\pi_* n^{3/2}\delta\sqrt{2\ln(n) \ln(2/\delta)}}
+
\tilde O(n^{-2.5}).
\eeq

\paragraph{Estimating the bias in $\hat T_{ij}$.}
This analysis assumes stationarity. 
For $1\le i,j\le d$, define the random variable $N_{ij}=\#(i,j)$ to be the number of consecutive $(i,j)$
pairs in the observed sequence; define $N_i=\#(i)$ analogously.
We have
\beq
\hat T_{ij} = \frac{ N_{ij}}{N_i}.
\eeq
We will estimate the bias $\E[\hat T_{ij}]-T_{ij}$ as follows.
First, compute the expectations of the numerator and the denominator:
\beq
\E[N_{ij}]
&=& 
(n-1)\pi_i T_{ij} 
\\
\E[N_{i}]
&=& n\pi_i
,\\
\frac{\E[N_{ij}]}{\E[N_{i}]}
&=&
\frac{(n-1)\pi_i T_{ij}}{n\pi_i} = T_{ij}+O\paren{\oo n}.
\eeq
To estimate $\E[\hat T_{ij}]$, we are going to claim that both the numerator and the denominator
are concentrated about their means.
In particular, both $N_{ij}$ and $N_i$ are Lipschitz (w.r.t. Hamming) functions of the state sequence
(with constants $2$ and $1$, resp.),
and hence, invoking \citet[Theorem 1.1]{meroi2012-jap} with $G=1/\pi_*$ and $\theta=e^{-\gamma_*}$,
\beq
\P{ \abs { 
%\oo n\E[N_{ij}] - (1-\oo n)\pi_i T_{ij} 
N_{ij}-\E[N_{ij}]
} > \eps n
} &\le& 2\exp\paren{ -\frac{n(1-e^{-\gamma_*})^2\pi_*^2\eps^2}{8} } \\
\P{ \abs { 
%\oo n\E[N_i] - \pi_i  
N_{i}-\E[N_{i}]
} > 
\eps n} &\le& 2\exp\paren{ -\frac{n(1-e^{-\gamma_*})^2\pi_*^2\eps^2}{2} } \\
\eeq
Thus, with probability at least
\beq
1-\delta &:=& 1-4 \exp\paren{ -\frac{n(1-e^{-\gamma_*})^2\pi_*^2\eps^2}{8} },
\eeq
%all of 
the empirical counts $N_{ij}$ and $N_i$ are $\eps n$-close to their expectations.
Finally, since for all $a,b,x>0$, we have
\beq
\abs{ \frac{a\pm x}{b\pm x} - \frac a b} \le \frac{a+b}{b^2}x+O(x^2), 
\eeq
it follows that for all $i,j$,
\beq
\abs{
\E\sqprn{\frac{ N_{ij}}{N_i}}
-
\frac{\E[N_{ij}]}{\E[N_{i}]}
}
&\le&
\delta + \frac{\E[N_{ij}]+\E[N_i]}{\E[N_i]^2}\eps + O(\eps^2) \\
&=& \delta + \frac{ (n-1)\pi_i T_{ij} + n\pi_i}{n^2\pi_i^2}\eps + O(\eps^2) \\
&\le& \delta + \frac{ (1-1/n)T_{ij} + 1}{n\pi_*}\eps + O(\eps^2).
\eeq

Thus,
\beq
\abs{\E[\hat T_{ij}]-T_{ij}} &\le&
%\delta + \frac{ (1-1/n)T_{ij} + 1}{n\pi_*}\eps + O\paren{\eps^2+\oo n}.
O\paren{
\exp\paren{ -\frac{n(1-e^{-\gamma_*})^2\pi_*^2\eps^2}{8} }
+\frac{\eps}{n\pi_*}}.
\eeq
Choosing $\eps=O\paren{\frac{1}{\pi_*(1-e^{-\gamma^*})\sqrt{n\ln n}}}$ yields
\beq
\abs{\E[\hat T_{ij}]-T_{ij}} &\le&
O\paren{
\frac{1}{\pi_*^2(1-e^{-\gamma^*})n\sqrt{n\ln n}}}.
\eeq

\paragraph{Stationariy assumption:}
One approach is to let the chain run for some time to ``burn in'',
until it reaches the distribution $\pi'$. If $\pi$ is the true stationary distribution,
then all of our ``w.h.p.'' results will hold with an additive correction of 
$\tvnorm{\pi-\pi'}$. The problem is that we won't know for how long to burn in
and also won't know $\tvnorm{\pi-\pi'}$ --- a chicken and egg problem.

Another idea is to use ``coupling from the past'':\\
\url{http://www.cc.gatech.edu/~vigoda/MCMC_Course/CouplingFromPast.pdf}
\\
\url{http://people.csail.mit.edu/costis/6896sp11/lec8s.pdf}
\\
This requires multiple restarts but otherwise works well with finite state spaces.
It gives us guaranteed perfect samples from the stationary distribution,
and from that point the chain can be run for $n$ steps to obtain our statistics.

\paragraph{Still need to do:}
Analytic upper bound on $\EE{ \hat{B}_n(\delta)}$.
Matching lower bound.

\bibliographystyle{plainnat}
\bibliography{all}

\end{document}

Thus, we need estimates of $\gamma_*$ and $\pi_*$.
To estimate them we may want to estimate the stationary probabilities and the transition matrix $T$.
Let $\hat \pi$ denote the estimate of the former, while let $\hat T$ be the estimate of the latter.
In particular, $\hat \pi_i$ is taken to be the empirical frequency of the visits of state $i$ in $(X_t)$,
while we let $\hat T_{i,j}$ the frequency when 
$X_{t+1}=j$ happens given that $X_t = i$ ($t=1,\dots,n-1$). 
When $X_t = i$ never happens, we take $\hat T_{i,\cdot}$ to be the uniform distribution.
Since $\Prob{\sum_{t=1}^n \one{X_t = i}=0}=\Prob{X_1\ne i, \dots, X_n\ne i}>0$ for each $i$, $\hat T_{ij}$ is a biased estimate of $T_{ij}$, although the bias decreases exponentially fast.

For $x\in \R^d$, let $D_x$ be the diagonal matrix with $x$ in its diagonal. 
Define $L = D_\pi^{-1/2} T  D_\pi^{1/2}$, $\hat{L} = D_{\hat{\pi}}^{-1/2} \hat{T} D_{\hat{\pi}}^{1/2}$, 
By Weyl's inequality (see notes.pdf),
\begin{align}
\label{eq:gammabound}
|\hat{\gamma}_* - \gamma_* | = |\hat{\lambda}_* - \lambda_* | \le \tnrm{\hat L - L }_2
\end{align}
and so it remains to bound this latter quantity.
Define $E_T = D_{\pi}^{-1/2} (\hat T - T ) D_{\pi}^{1/2}$
and $E_\pi = D_{\hat{\pi}}^{-1/2} D_\pi^{1/2} - I$.
We have (see notes.pdf),
\begin{align}
\label{eq:lbound}
\tnrm{\hat L - L}_2 \le \norm{E_T}_2 + \norm{E_{\pi}}_2 (\norm{E_T}_2+1) (\norm{E_\pi}_2+2)\,.
\end{align}
Now, $\norm{E_T}_2 = \tnrm{\hat T - T}_2$ (because the eigenvalues of $E_T$ are the same as that of $\hat T - T$).
Further, for any $\mu,a$ such that $n=2\mu a$
and $\delta>2(\mu-1)\beta(a)$,
with probability at least $1-\delta$,
\begin{align}
\label{eq:tbound}
\tnrm{T-\hat T}_2
\le 25\sqrt{\frac{d}{\mu}} + 2\sqrt{\frac{\ln(2/\delta')}{2\mu}}\,,
\end{align}
where $\delta'=\delta-2(\mu-1)\beta(a)$ (see mohri-notes.pdf). \todoc{This does not take into account the bias.}
\todoc[inline]{Next problem: For a fixed $n$, this will limit the range of $\delta$.
For any $n$, there will be a range of $\delta$ in the neighbourhood of $0$ for which the algorithm
will need to return ``I don't know''.
}
Similarly, for any $\mu',a'$ such that $n=2\mu' a'$
and $\delta>2(\mu'-1)\beta(a')$,
with probability at least $1-\delta$,
\begin{align}
\label{eq:pibound}
|\hat\pi_i - \pi_i | \le C \left( \sqrt{\frac{d}{\mu'}} + \sqrt{\frac{\ln(2/\delta'')}{2\mu'}}\right)\,,
\end{align}
where $\delta''=\delta-2(\mu'-1)\beta(a')$.

By combining~\eqref{eq:betabound},
\eqref{eq:gammabound}, \eqref{eq:lbound}, \eqref{eq:tbound} and~\eqref{eq:pibound}, we can derive 
that for any $0\le \delta \le 1$,
 $\mu,a,\mu',a'$  such that $n = 2\mu a$ and
 $\frac{2}{\pi_*} \max( (\mu-1)e^{-a\gamma_*}, (\mu'-1)e^{-a'\gamma_*} )<\delta$, 
with probability $1-\delta$,
\begin{align*}
|\hat{\gamma}_* - \gamma_* | &\le u( \gamma_*, \pi_*, \hat{\gamma}, \hat\pi, \hat T, n, \mu, a, \mu', a', \delta )\,,\\
|\hat{\pi}_* - \pi_* | &\le v( \gamma_*, \pi_*, \hat{\gamma}, \hat\pi, \hat T, n, \mu, a, \mu', a', \delta )\,,
\end{align*}
for some functions $u$ and $v$.
Reordering (and relaxing) the  inequalities we get
\begin{align*}
\gamma_* &\ge \hat\gamma_* - u( \gamma_*, \pi_*, \hat{\gamma}, \hat\pi, \hat T, n, \mu, a, \mu', a', \delta )\,,\\
 \pi_*  &\ge \hat{\pi}_* - v( \gamma_*, \pi_*, \hat{\gamma}, \hat\pi, \hat T, n, \mu, a, \mu', a', \delta )\,.
\end{align*}
Now, let $\hat{B}_n(\delta)$ be the optimal value of the optimization problem 
\begin{align*}
\lefteqn{\max B_n(g,p,\delta)\qquad \text{s.t.} }\\
&\qquad  g \ge \hat\gamma_* - u( g, p, \hat{\gamma}, \hat\pi, \hat T, n, \mu, a, \mu', a', \delta )\,,\\
&\qquad p  \ge \hat{\pi}_* - v( g, p, \hat{\gamma}, \hat\pi, \hat T, n, \mu, a, \mu', a', \delta )\,,\\
&\qquad n = 2\mu a, \quad n = 2\mu' a'\,,\\
&\qquad \frac{2}{p} \max( (\mu-1)e^{-ag}, (\mu'-1)e^{-a'g} )<\delta\,,\\
&\qquad g,p >0, \mu,a,\mu',a'\in \N\,.
\end{align*}
Then, w.p. $1-\delta$, it holds that
\[
|\bar{X}_n - \EE{f(X_1)}| \le \hat{B}_n(\delta)\,.
\]

Next goals: Analytic upper bound on $\EE{ \hat{B}_n(\delta)}$.
Matching lower bound.

\bibliographystyle{plainnat}
\bibliography{all}

\end{document}

### 2015-06-01 scratch/beta-mixing.tex
Copied files from svn

--- 2015-06-01 scratch/aryeh-scratch-notes.tex
@@
 \usepackage{amssymb}
-%\usepackage{mathtools}
+\usepackage{mathtools}
 \usepackage{enumerate}
@@
 \usepackage{tikz}
-%\usepackage{pgfplots}
+\usepackage{pgfplots}
 \usepackage[ruled]{algorithm}
@@
     bookmarks=true,         % show bookmarks bar?
-    unicode=false,          % non-Latin characters in Acrobats bookmarks
-    pdftoolbar=true,        % show Acrobats toolbar?
-    pdfmenubar=true,        % show Acrobats menu?
+    unicode=false,          % non-Latin characters in AcrobatÕs bookmarks
+    pdftoolbar=true,        % show AcrobatÕs toolbar?
+    pdfmenubar=true,        % show AcrobatÕs menu?
     pdffitwindow=false,     % window fit to page when opened
@@
 \newcommand{\rdefined}{=\vcentcolon}
-%\newcommand{\E}{\mathbb E}
+\newcommand{\E}{\mathbb E}
 \newcommand{\Var}{\operatorname{Var}}
 \newcommand{\calF}{\mathcal F}
-\newcommand{\calR}{\mathcal R}
 \newcommand{\sr}[1]{\stackrel{#1}}
@@
 \DeclareMathOperator{\Sym}{Sym}
-\DeclareMathOperator*{\E}{\text{\bf E}}
 \renewcommand\t{{\ensuremath{\scriptscriptstyle{\top}}}}
@@
 \newcommand{\oo}[1]{\frac{1}{#1}}
-\newcommand{\nrm}[1]{\left\Vert #1 \right\Vert}
-\newcommand{\tnrm}[1]{\Vert #1 \Vert}
-\newcommand{\abs}[1]{\left| #1 \right|}
-\newcommand{\trn}{^\intercal} %operator transpose
-
-\newcommand{\beq}{\begin{eqnarray*}}
-\newcommand{\eeq}{\end{eqnarray*}}
-\newcommand{\beqn}{\begin{eqnarray}}
-\newcommand{\eeqn}{\end{eqnarray}}
 
@@
 
-\title{Aryeh's scratch notes}
-%\author{Daniel, Aryeh, Csaba}
+\title{Statistical Estimation of Mixing Times}
+\author{Daniel, Aryeh, Csaba}
 
@@
 
-%\begin{abstract}
-%An abstract
-%\end{abstract}
-\newcommand{\XX}{\mathbb{X}}
-\newcommand{\one}[1]{\mathbb{I}\left\{#1\right\}}
-\newcommand{\N}{\mathbb{N}}
-For any $\mu,a$ such that $n = 2\mu a$ and $\delta>2(\mu-1)\beta(a)$,
-with probability at least $1-\delta$,
-\begin{align}
-\label{eq:xbound}
-|\bar{X}_n - \EE{f(X_1)}|
-\le c_1\sqrt{\frac{1}{\mu}} + c_2\sqrt{\frac{\ln(2/\delta')}{2\mu}}\,,
-\end{align}
-where $\delta'=\delta-2(\mu-1)\beta(a)$
-If $\beta(k)$ denotes the $k$th $\beta$-mixing coefficient of $(X_i)$ then 
-\begin{align}
-\label{eq:betabound}
- \beta(k) \le \oo{\pi_*}e^{-k \gamma_*}
-\end{align}
-where  $\gamma_* = 1-\lambda_*$ is the spectral gap ($\lambda_*$ is the second largest eigenvalue of $T$)
-and  $\pi_*  = \min_i \pi_i>0$ is the critical probability underlying the chain
-(see the $\beta$-mixing note).
-For a given value of $\pi_*,\gamma_*$, $\delta$, let $B_n(\pi_*,\gamma_*,\delta)$ be the optimal value of the optimization problem
-\begin{align*}
-\lefteqn{ \min_{\mu,a} \min\left(1, 
-c_1\sqrt{\frac{1}{\mu}} + c_2\sqrt{\frac{\ln(2/\delta')}{2\mu}}\right) 
-} \\
-& \qquad \text{s.t.} \quad n = 2\mu a, \quad \delta>2(\mu-1)\oo{\pi_*}e^{-a \gamma_*},
-\quad \delta' = \delta-2(\mu-1)\oo{\pi_*}e^{-a \gamma_*}, \quad \mu,a \in \N\,.
-\end{align*}
+\begin{abstract}
+An abstract
+\end{abstract}
 
-\paragraph{We will upper-bound} $B_n$ as follows.
-Choose
-$$ a=(2\ln n)/\gamma_*.$$
-Then
-$\mu=n/2a=\gamma_*n/4\ln n$
-and
-\beq
-\delta' &=& \delta-2(\mu-1)\oo{\pi_*}e^{-a \gamma_*} \\
-&=& \delta - 2\paren{\frac{\gamma_*n}{4\ln n}-1}\oo{n^2\pi_*} \\
-%&=& \delta - {\frac{\gamma_*}{2\pi_*n\ln n}} + O(n^{-2}) \\
-&=&
-\delta - \sqprn{
-\frac{\gamma_*}{2\pi_*n\ln n}-\frac{2}{\pi_*n^2}
-}
-=:\delta-x.
-%\frac{2}{\delta'}
-%&=&
-%\frac{2}{\delta} + 
-\eeq
-Now
-\beq
-\ln\frac{2}{\delta-x} &=& \ln\frac{2}{\delta}+\frac{x}{\delta^2}+O(x^2), \\
-\sqrt{\ln\frac{2}{\delta-x}} &=& 
-\sqrt{\ln\frac{2}{\delta}}
-+
-\frac{x}{2\delta\sqrt{\ln(2/\delta)}}+O(x^2), \\
-\eeq
-and hence,
-\beq
-B_n(\pi_*,\gamma_*,\delta)
-&\le&
-2\sqrt{\frac{\ln n}{\gamma_*n}}
-\sqprn{
-c_1
-+ c_2
-\sqrt{\ln(2/\delta')/2}
-}\\
-&=&
-2\sqrt{\frac{\ln n}{\gamma_*n}}
-\sqprn{
-c_1
-+ c_2
-\paren{
-\sqrt{
-\oo2
-\ln\frac{2}{\delta}}
-+
-\frac{x}{2\delta\sqrt{2\ln(2/\delta)}}+O(x^2)
-}
-}\\
-&=&
-2\sqrt{\frac{\ln n}{\gamma_*n}}
-\sqprn{
-c_1
-+ c_2
-\paren{
-\sqrt{\oo2\ln\frac{2}{\delta}}
-+
-\frac{
-\gamma_*
-}{4{\pi_*n\ln n}\delta\sqrt{2\ln(2/\delta)}}+O(n^{-2})
-}
-}\\
-&=&
-2c_1\sqrt{\frac{\ln n}{\gamma_*n}}
-+
-2c_2\sqrt{\frac{\ln n}{2\gamma_*n}
-\ln\frac{2}{\delta}}
-+
-c_2\frac{\sqrt{\gamma_*}}
-{2\pi_* n^{3/2}\delta\sqrt{2\ln(n) \ln(2/\delta)}}
-+
-\tilde O(n^{-2.5}).
-\eeq
+\section{Beta coefficients of reversible, irreducible, stationary Markov chains}
 
-\paragraph{Estimating the bias in $\hat T_{ij}$.}
-This analysis assumes stationarity. 
-For $1\le i,j\le d$, define the random variable $N_{ij}=\#(i,j)$ to be the number of consecutive $(i,j)$
-pairs in the observed sequence; define $N_i=\#(i)$ analogously.
-We have
-\beq
-\hat T_{ij} = \frac{ N_{ij}}{N_i}.
-\eeq
-We will estimate the bias $\E[\hat T_{ij}]-T_{ij}$ as follows.
-First, compute the expectations of the numerator and the denominator:
-\beq
-\E[N_{ij}]
-&=& 
-(n-1)\pi_i T_{ij} 
-\\
-\E[N_{i}]
-&=& n\pi_i
-,\\
-\frac{\E[N_{ij}]}{\E[N_{i}]}
-&=&
-\frac{(n-1)\pi_i T_{ij}}{n\pi_i} = T_{ij}+O\paren{\oo n}.
-\eeq
-To estimate $\E[\hat T_{ij}]$, we are going to claim that both the numerator and the denominator
-are concentrated about their means.
-In particular, both $N_{ij}$ and $N_i$ are Lipschitz (w.r.t. Hamming) functions of the state sequence
-(with constants $2$ and $1$, resp.),
-and hence, invoking \citet[Theorem 1.1]{meroi2012-jap} with $G=1/\pi_*$ and $\theta=e^{-\gamma_*}$,
-\beq
-\P{ \abs { 
-%\oo n\E[N_{ij}] - (1-\oo n)\pi_i T_{ij} 
-N_{ij}-\E[N_{ij}]
-} > \eps n
-} &\le& 2\exp\paren{ -\frac{n(1-e^{-\gamma_*})^2\pi_*^2\eps^2}{8} } \\
-\P{ \abs { 
-%\oo n\E[N_i] - \pi_i  
-N_{i}-\E[N_{i}]
-} > 
-\eps n} &\le& 2\exp\paren{ -\frac{n(1-e^{-\gamma_*})^2\pi_*^2\eps^2}{2} } \\
-\eeq
-Thus, with probability at least
-\beq
-1-\delta &:=& 1-4 \exp\paren{ -\frac{n(1-e^{-\gamma_*})^2\pi_*^2\eps^2}{8} },
-\eeq
-%all of 
-the empirical counts $N_{ij}$ and $N_i$ are $\eps n$-close to their expectations.
-Finally, since for all $a,b,x>0$, we have
-\beq
-\abs{ \frac{a\pm x}{b\pm x} - \frac a b} \le \frac{a+b}{b^2}x+O(x^2), 
-\eeq
-it follows that for all $i,j$,
-\beq
-\abs{
-\E\sqprn{\frac{ N_{ij}}{N_i}}
--
-\frac{\E[N_{ij}]}{\E[N_{i}]}
-}
-&\le&
-\delta + \frac{\E[N_{ij}]+\E[N_i]}{\E[N_i]^2}\eps + O(\eps^2) \\
-&=& \delta + \frac{ (n-1)\pi_i T_{ij} + n\pi_i}{n^2\pi_i^2}\eps + O(\eps^2) \\
-&\le& \delta + \frac{ (1-1/n)T_{ij} + 1}{n\pi_*}\eps + O(\eps^2).
-\eeq
+Suppose our Markov chain is reversible, irreducible and stationary (RIS),
+with stationary distribution $\pi$. We would like to upper-bound its $k$th $\beta$-mixing coefficient.
+The latter may be defined,
+for any stationary sequence $\set{Z_t}_{t=-\infty}^{\infty}$,
+ as follows:
+$$ \beta(k) = \sup_n \E_B \sqprn{ \sup_A |\P{A|B}-\P{A}|},
+$$
+where $B$ is an event in the $\sigma$-algebra of $Z_{-\infty}^n$ and $A$ an event in $\sigma(Z_{n+k}^\infty)$.
+Clearly, $\beta(k)\le\phi(k)$, where the $\phi$-mixing coefficient is defined analogously to $\beta$,
+with $\sup_B$ in place of
+$\E_B$  \cite{bradley05}.
 
-Thus,
-\beq
-\abs{\E[\hat T_{ij}]-T_{ij}} &\le&
-%\delta + \frac{ (1-1/n)T_{ij} + 1}{n\pi_*}\eps + O\paren{\eps^2+\oo n}.
-O\paren{
-\exp\paren{ -\frac{n(1-e^{-\gamma_*})^2\pi_*^2\eps^2}{8} }
-+\frac{\eps}{n\pi_*}}.
-\eeq
-Choosing $\eps=O\paren{\frac{1}{\pi_*(1-e^{-\gamma^*})\sqrt{n\ln n}}}$ yields
-\beq
-\abs{\E[\hat T_{ij}]-T_{ij}} &\le&
-O\paren{
-\frac{1}{\pi_*^2(1-e^{-\gamma^*})n\sqrt{n\ln n}}}.
-\eeq
+For a RIS chain, we have
+$$ \tmix(\eps) \le  \oo{1-\lambda_2}\ln\left( \frac{1}{\eps \pi_* }\right).$$
 
-\paragraph{Stationariy assumption:}
-One approach is to let the chain run for some time to ``burn in'',
-until it reaches the distribution $\pi'$. If $\pi$ is the true stationary distribution,
-then all of our ``w.h.p.'' results will hold with an additive correction of 
-$\tvnorm{\pi-\pi'}$. The problem is that we won't know for how long to burn in
-and also won't know $\tvnorm{\pi-\pi'}$ --- a chicken and egg problem.
-
-Another idea is to use ``coupling from the past'':\\
-\url{http://www.cc.gatech.edu/~vigoda/MCMC_Course/CouplingFromPast.pdf}
-\\
-\url{http://people.csail.mit.edu/costis/6896sp11/lec8s.pdf}
-\\
-This requires multiple restarts but otherwise works well with finite state spaces.
-It gives us guaranteed perfect samples from the stationary distribution,
-and from that point the chain can be run for $n$ steps to obtain our statistics.
-
-\paragraph{Still need to do:}
-Analytic upper bound on $\EE{ \hat{B}_n(\delta)}$.
-Matching lower bound.
+Inverting for $\eps$, we have
+$$ \beta(k) \le \phi(k) \le \oo{\pi_*}e^{-k(1-\lambda_2)}.$$
 
 \bibliographystyle{plainnat}
-\bibliography{all}
+\bibliography{../all}
 
 \end{document}
-
-Thus, we need estimates of $\gamma_*$ and $\pi_*$.
-To estimate them we may want to estimate the stationary probabilities and the transition matrix $T$.
-Let $\hat \pi$ denote the estimate of the former, while let $\hat T$ be the estimate of the latter.
-In particular, $\hat \pi_i$ is taken to be the empirical frequency of the visits of state $i$ in $(X_t)$,
-while we let $\hat T_{i,j}$ the frequency when 
-$X_{t+1}=j$ happens given that $X_t = i$ ($t=1,\dots,n-1$). 
-When $X_t = i$ never happens, we take $\hat T_{i,\cdot}$ to be the uniform distribution.
-Since $\Prob{\sum_{t=1}^n \one{X_t = i}=0}=\Prob{X_1\ne i, \dots, X_n\ne i}>0$ for each $i$, $\hat T_{ij}$ is a biased estimate of $T_{ij}$, although the bias decreases exponentially fast.
-
-For $x\in \R^d$, let $D_x$ be the diagonal matrix with $x$ in its diagonal. 
-Define $L = D_\pi^{-1/2} T  D_\pi^{1/2}$, $\hat{L} = D_{\hat{\pi}}^{-1/2} \hat{T} D_{\hat{\pi}}^{1/2}$, 
-By Weyl's inequality (see notes.pdf),
-\begin{align}
-\label{eq:gammabound}
-|\hat{\gamma}_* - \gamma_* | = |\hat{\lambda}_* - \lambda_* | \le \tnrm{\hat L - L }_2
-\end{align}
-and so it remains to bound this latter quantity.
-Define $E_T = D_{\pi}^{-1/2} (\hat T - T ) D_{\pi}^{1/2}$
-and $E_\pi = D_{\hat{\pi}}^{-1/2} D_\pi^{1/2} - I$.
-We have (see notes.pdf),
-\begin{align}
-\label{eq:lbound}
-\tnrm{\hat L - L}_2 \le \norm{E_T}_2 + \norm{E_{\pi}}_2 (\norm{E_T}_2+1) (\norm{E_\pi}_2+2)\,.
-\end{align}
-Now, $\norm{E_T}_2 = \tnrm{\hat T - T}_2$ (because the eigenvalues of $E_T$ are the same as that of $\hat T - T$).
-Further, for any $\mu,a$ such that $n=2\mu a$
-and $\delta>2(\mu-1)\beta(a)$,
-with probability at least $1-\delta$,
-\begin{align}
-\label{eq:tbound}
-\tnrm{T-\hat T}_2
-\le 25\sqrt{\frac{d}{\mu}} + 2\sqrt{\frac{\ln(2/\delta')}{2\mu}}\,,
-\end{align}
-where $\delta'=\delta-2(\mu-1)\beta(a)$ (see mohri-notes.pdf). \todoc{This does not take into account the bias.}
-\todoc[inline]{Next problem: For a fixed $n$, this will limit the range of $\delta$.
-For any $n$, there will be a range of $\delta$ in the neighbourhood of $0$ for which the algorithm
-will need to return ``I don't know''.
-}
-Similarly, for any $\mu',a'$ such that $n=2\mu' a'$
-and $\delta>2(\mu'-1)\beta(a')$,
-with probability at least $1-\delta$,
-\begin{align}
-\label{eq:pibound}
-|\hat\pi_i - \pi_i | \le C \left( \sqrt{\frac{d}{\mu'}} + \sqrt{\frac{\ln(2/\delta'')}{2\mu'}}\right)\,,
-\end{align}
-where $\delta''=\delta-2(\mu'-1)\beta(a')$.
-
-By combining~\eqref{eq:betabound},
-\eqref{eq:gammabound}, \eqref{eq:lbound}, \eqref{eq:tbound} and~\eqref{eq:pibound}, we can derive 
-that for any $0\le \delta \le 1$,
- $\mu,a,\mu',a'$  such that $n = 2\mu a$ and
- $\frac{2}{\pi_*} \max( (\mu-1)e^{-a\gamma_*}, (\mu'-1)e^{-a'\gamma_*} )<\delta$, 
-with probability $1-\delta$,
-\begin{align*}
-|\hat{\gamma}_* - \gamma_* | &\le u( \gamma_*, \pi_*, \hat{\gamma}, \hat\pi, \hat T, n, \mu, a, \mu', a', \delta )\,,\\
-|\hat{\pi}_* - \pi_* | &\le v( \gamma_*, \pi_*, \hat{\gamma}, \hat\pi, \hat T, n, \mu, a, \mu', a', \delta )\,,
-\end{align*}
-for some functions $u$ and $v$.
-Reordering (and relaxing) the  inequalities we get
-\begin{align*}
-\gamma_* &\ge \hat\gamma_* - u( \gamma_*, \pi_*, \hat{\gamma}, \hat\pi, \hat T, n, \mu, a, \mu', a', \delta )\,,\\
- \pi_*  &\ge \hat{\pi}_* - v( \gamma_*, \pi_*, \hat{\gamma}, \hat\pi, \hat T, n, \mu, a, \mu', a', \delta )\,.
-\end{align*}
-Now, let $\hat{B}_n(\delta)$ be the optimal value of the optimization problem 
-\begin{align*}
-\lefteqn{\max B_n(g,p,\delta)\qquad \text{s.t.} }\\
-&\qquad  g \ge \hat\gamma_* - u( g, p, \hat{\gamma}, \hat\pi, \hat T, n, \mu, a, \mu', a', \delta )\,,\\
-&\qquad p  \ge \hat{\pi}_* - v( g, p, \hat{\gamma}, \hat\pi, \hat T, n, \mu, a, \mu', a', \delta )\,,\\
-&\qquad n = 2\mu a, \quad n = 2\mu' a'\,,\\
-&\qquad \frac{2}{p} \max( (\mu-1)e^{-ag}, (\mu'-1)e^{-a'g} )<\delta\,,\\
-&\qquad g,p >0, \mu,a,\mu',a'\in \N\,.
-\end{align*}
-Then, w.p. $1-\delta$, it holds that
-\[
-|\bar{X}_n - \EE{f(X_1)}| \le \hat{B}_n(\delta)\,.
-\]
-
-Next goals: Analytic upper bound on $\EE{ \hat{B}_n(\delta)}$.
-Matching lower bound.
-
-\bibliographystyle{plainnat}
-\bibliography{all}
-
-\end{document}

### 2015-06-01 scratch/mohri-notes.tex
Copied files from svn

--- 2015-06-01 scratch/aryeh-scratch-notes.tex
@@
 \newcommand{\nrm}[1]{\left\Vert #1 \right\Vert}
-\newcommand{\tnrm}[1]{\Vert #1 \Vert}
 \newcommand{\abs}[1]{\left| #1 \right|}
@@
 
-\title{Aryeh's scratch notes}
+\title{Notes on the Mohri-Rostamizadeh approach}
 %\author{Daniel, Aryeh, Csaba}
@@
 %\end{abstract}
-\newcommand{\XX}{\mathbb{X}}
-\newcommand{\one}[1]{\mathbb{I}\left\{#1\right\}}
-\newcommand{\N}{\mathbb{N}}
-For any $\mu,a$ such that $n = 2\mu a$ and $\delta>2(\mu-1)\beta(a)$,
-with probability at least $1-\delta$,
-\begin{align}
-\label{eq:xbound}
-|\bar{X}_n - \EE{f(X_1)}|
-\le c_1\sqrt{\frac{1}{\mu}} + c_2\sqrt{\frac{\ln(2/\delta')}{2\mu}}\,,
-\end{align}
-where $\delta'=\delta-2(\mu-1)\beta(a)$
-If $\beta(k)$ denotes the $k$th $\beta$-mixing coefficient of $(X_i)$ then 
-\begin{align}
-\label{eq:betabound}
- \beta(k) \le \oo{\pi_*}e^{-k \gamma_*}
-\end{align}
-where  $\gamma_* = 1-\lambda_*$ is the spectral gap ($\lambda_*$ is the second largest eigenvalue of $T$)
-and  $\pi_*  = \min_i \pi_i>0$ is the critical probability underlying the chain
-(see the $\beta$-mixing note).
-For a given value of $\pi_*,\gamma_*$, $\delta$, let $B_n(\pi_*,\gamma_*,\delta)$ be the optimal value of the optimization problem
-\begin{align*}
-\lefteqn{ \min_{\mu,a} \min\left(1, 
-c_1\sqrt{\frac{1}{\mu}} + c_2\sqrt{\frac{\ln(2/\delta')}{2\mu}}\right) 
-} \\
-& \qquad \text{s.t.} \quad n = 2\mu a, \quad \delta>2(\mu-1)\oo{\pi_*}e^{-a \gamma_*},
-\quad \delta' = \delta-2(\mu-1)\oo{\pi_*}e^{-a \gamma_*}, \quad \mu,a \in \N\,.
-\end{align*}
 
-\paragraph{We will upper-bound} $B_n$ as follows.
-Choose
-$$ a=(2\ln n)/\gamma_*.$$
+Let $T$ be the true transition matrix and $\hat T$ the estimated one
+(leaving the estimation method presently unspecified). 
+Recall that for any matrix $A$, its Euclidean norm $\nrm{A}_2$ is given by
+$$ \nrm{A}_2 = \sup_{\nrm{x}\le1}\nrm{Ax} 
+= \sup_{\nrm{x},\nrm{y}\le1}x\trn A y,
+$$
+where all vector norms are $\ell_2$ unless stated otherwise.
+
+We are interested
+in bounding
+$$ \nrm{T-\hat T}_2 = \sup_{\nrm{x},\nrm{y}\le1} {x\trn(T-\hat T)y} .
+$$
+
+Trying to stay notationally close to \cite{mohri-rosta08},
+let $H=B\times B$,
+where $B$ is the Euclidean unit ball, and put
+$R(x,y)=x\trn T y$, $\hat R_S(x,y)=x\trn \hat T y$.
 Then
-$\mu=n/2a=\gamma_*n/4\ln n$
-and
+$$ \nrm{T-\hat T}_2 = \sup_{(x,y)\in H} R(x,y)-\hat R_S(x,y)=:\Phi(S),$$
+where $S$ is the observed sample sequence.
+
+We will need an analogue of Prop. 1 in \cite{mohri-rosta08}:
+$$R(x,y) = \E_{S}[\hat R_S(x,y)].$$
+I believe that for $S$ sampled from a stationary process
+and $\hat T_{ij}$ taken to be the sample mean, this indeed holds.
+\todoc{I don't think so. There is a bias because maybe $i$ is never visited.}
+I don't think it holds for the median-of-means estimate.
+
+[Note: there appears to be something wrong with the indexing
+of the blocks in (6)-(7) of \cite{mohri-rosta08}. 
+The numbering in 
+\cite{MR1258867}, p.101 makes more sense, and we will follow it.]
+Partition the sequence $S=(z_1,\ldots,z_n)$,
+$n=2a\mu$, into $2a\mu$ blocks of length $a$.
+Let $S_0$ be the sequence of odd blocks
+and $\tilde S_0$ a sequence of independent blocks of length
+$a$, where each block in $\tilde S_0$ is distributed identically to its
+corresponding block $S_0$.
+
+Since $\abs{\Phi(\cdot)}\le2$, Lemma 4.1 of
+\cite{MR1258867} implies
+$$ \abs{ \E_{S_0}[\Phi]-\E_{\tilde S_0}[\Phi]}\le 2(\mu-1)\beta(a),$$
+where $\beta(\cdot)$ is the beta-mixing coefficient.
+
+By Lemma 1 of \cite{mohri-rosta08}, we have,
+for all $a,\mu,\eps>0$ with $2\mu a=n$ 
+%and 
+%$\eps>\E_{\tilde S_0}[\Phi(\tilde S_0)]$,
+that
+$$ \P{\Phi(S)>\eps}
+\le 2\P{ \Phi(\tilde S_0)
+%-\E_{\tilde S_0}[\Phi(\tilde S_0}]
+>\eps
+}
++
+2(\mu-1)\beta(a).
+$$
+
+Proposition 2 ibid. yields, for all
+$\eps>\E_{\tilde S_0}[\Phi(\tilde S_0)]$,
+
+$$ \P{\Phi(S)>\eps} \le
+2\exp(-\mu{\eps'}^2/2)+2(\mu-1)\beta(a),
+$$
+where $\eps'=\eps-\E_{\tilde S_0}[\Phi(\tilde S_0)]$.
+
+Now invoke Lemma 2 ibid. to bound the expectation over iid blocks:
+$$ \E_{\tilde S_0}[\Phi(\tilde S_0)] \le \calR_\mu(H),$$
+where $\cal_\mu(\cdot)$ is the ordinary (iid) Rademacher complexity of
+$H$.
+Now it is well-known \cite{MR1965359} that the $\ell_2$
+$t$-covering number of the $d$-dimensional Euclidean unit ball is
+bounded by $(3/t)^d$. 
+To bound the $\ell_2$ covering number of $H=B_d\times B_d$,
+fix a radius $t$ and
+%recall that
+%$$ 
+%%\frac{\sqrt{a}+\sqrt{b}}{\sqrt 2}
+%%\le 
+%\sqrt{a+b}\le \sqrt a+\sqrt b$$
+%for all $a,b>0$.
+%Hence
+observe that
 \beq
-\delta' &=& \delta-2(\mu-1)\oo{\pi_*}e^{-a \gamma_*} \\
-&=& \delta - 2\paren{\frac{\gamma_*n}{4\ln n}-1}\oo{n^2\pi_*} \\
-%&=& \delta - {\frac{\gamma_*}{2\pi_*n\ln n}} + O(n^{-2}) \\
-&=&
-\delta - \sqprn{
-\frac{\gamma_*}{2\pi_*n\ln n}-\frac{2}{\pi_*n^2}
-}
-=:\delta-x.
-%\frac{2}{\delta'}
-%&=&
-%\frac{2}{\delta} + 
+\nrm{(x,y)}_2 &=& \sqrt{ \sum_i x_i^2 + \sum_j y_j^2} \\
+&\le& \sqrt2 t.
 \eeq
-Now
+Hence,
+$$ B_d\times B_d \subset \sqrt2 B_{2d} $$
+and $H$
+can be covered by
+$(\sqrt 6/t)^{2d}$ balls of radius $t$.
+Using Dudley's chaining integral to estimate
+the Rademacher complexity, we have
+
 \beq
-\ln\frac{2}{\delta-x} &=& \ln\frac{2}{\delta}+\frac{x}{\delta^2}+O(x^2), \\
-\sqrt{\ln\frac{2}{\delta-x}} &=& 
-\sqrt{\ln\frac{2}{\delta}}
-+
-\frac{x}{2\delta\sqrt{\ln(2/\delta)}}+O(x^2), \\
+\calR_\mu(H) &\le& \frac{12}{\sqrt\mu}\int_0^{\sqrt2}\sqrt{d\ln(\sqrt 6/t)}dt \\
+&\le& 25\sqrt{d/\mu}.
 \eeq
-and hence,
+Note that $d$ is the number of states, consistent with our other notes.
+
+Adapting Theorem 1 in \cite{mohri-rosta08} to our needs, we have
+that for any sample $S$ of length $n=2\mu a$
+and $\delta>2(\mu-1)\beta(a)$,
+the following
+holds with probability at least $1-\delta$:
 \beq
-B_n(\pi_*,\gamma_*,\delta)
-&\le&
-2\sqrt{\frac{\ln n}{\gamma_*n}}
-\sqprn{
-c_1
-+ c_2
-\sqrt{\ln(2/\delta')/2}
-}\\
-&=&
-2\sqrt{\frac{\ln n}{\gamma_*n}}
-\sqprn{
-c_1
-+ c_2
-\paren{
-\sqrt{
-\oo2
-\ln\frac{2}{\delta}}
-+
-\frac{x}{2\delta\sqrt{2\ln(2/\delta)}}+O(x^2)
-}
-}\\
-&=&
-2\sqrt{\frac{\ln n}{\gamma_*n}}
-\sqprn{
-c_1
-+ c_2
-\paren{
-\sqrt{\oo2\ln\frac{2}{\delta}}
-+
-\frac{
-\gamma_*
-}{4{\pi_*n\ln n}\delta\sqrt{2\ln(2/\delta)}}+O(n^{-2})
-}
-}\\
-&=&
-2c_1\sqrt{\frac{\ln n}{\gamma_*n}}
-+
-2c_2\sqrt{\frac{\ln n}{2\gamma_*n}
-\ln\frac{2}{\delta}}
-+
-c_2\frac{\sqrt{\gamma_*}}
-{2\pi_* n^{3/2}\delta\sqrt{2\ln(n) \ln(2/\delta)}}
-+
-\tilde O(n^{-2.5}).
+%\Phi(S) 
+\nrm{T-\hat T}_2
+\le 25\sqrt{\frac{d}{\mu}} + 2\sqrt{\frac{\ln(2/\delta')}{2\mu}},
 \eeq
+where $\delta'=\delta-2(\mu-1)\beta(a)$.
 
-\paragraph{Estimating the bias in $\hat T_{ij}$.}
-This analysis assumes stationarity. 
-For $1\le i,j\le d$, define the random variable $N_{ij}=\#(i,j)$ to be the number of consecutive $(i,j)$
-pairs in the observed sequence; define $N_i=\#(i)$ analogously.
-We have
-\beq
-\hat T_{ij} = \frac{ N_{ij}}{N_i}.
-\eeq
-We will estimate the bias $\E[\hat T_{ij}]-T_{ij}$ as follows.
-First, compute the expectations of the numerator and the denominator:
-\beq
-\E[N_{ij}]
-&=& 
-(n-1)\pi_i T_{ij} 
-\\
-\E[N_{i}]
-&=& n\pi_i
-,\\
-\frac{\E[N_{ij}]}{\E[N_{i}]}
-&=&
-\frac{(n-1)\pi_i T_{ij}}{n\pi_i} = T_{ij}+O\paren{\oo n}.
-\eeq
-To estimate $\E[\hat T_{ij}]$, we are going to claim that both the numerator and the denominator
-are concentrated about their means.
-In particular, both $N_{ij}$ and $N_i$ are Lipschitz (w.r.t. Hamming) functions of the state sequence
-(with constants $2$ and $1$, resp.),
-and hence, invoking \citet[Theorem 1.1]{meroi2012-jap} with $G=1/\pi_*$ and $\theta=e^{-\gamma_*}$,
-\beq
-\P{ \abs { 
-%\oo n\E[N_{ij}] - (1-\oo n)\pi_i T_{ij} 
-N_{ij}-\E[N_{ij}]
-} > \eps n
-} &\le& 2\exp\paren{ -\frac{n(1-e^{-\gamma_*})^2\pi_*^2\eps^2}{8} } \\
-\P{ \abs { 
-%\oo n\E[N_i] - \pi_i  
-N_{i}-\E[N_{i}]
-} > 
-\eps n} &\le& 2\exp\paren{ -\frac{n(1-e^{-\gamma_*})^2\pi_*^2\eps^2}{2} } \\
-\eeq
-Thus, with probability at least
-\beq
-1-\delta &:=& 1-4 \exp\paren{ -\frac{n(1-e^{-\gamma_*})^2\pi_*^2\eps^2}{8} },
-\eeq
-%all of 
-the empirical counts $N_{ij}$ and $N_i$ are $\eps n$-close to their expectations.
-Finally, since for all $a,b,x>0$, we have
-\beq
-\abs{ \frac{a\pm x}{b\pm x} - \frac a b} \le \frac{a+b}{b^2}x+O(x^2), 
-\eeq
-it follows that for all $i,j$,
-\beq
-\abs{
-\E\sqprn{\frac{ N_{ij}}{N_i}}
--
-\frac{\E[N_{ij}]}{\E[N_{i}]}
-}
-&\le&
-\delta + \frac{\E[N_{ij}]+\E[N_i]}{\E[N_i]^2}\eps + O(\eps^2) \\
-&=& \delta + \frac{ (n-1)\pi_i T_{ij} + n\pi_i}{n^2\pi_i^2}\eps + O(\eps^2) \\
-&\le& \delta + \frac{ (1-1/n)T_{ij} + 1}{n\pi_*}\eps + O(\eps^2).
-\eeq
-
-Thus,
-\beq
-\abs{\E[\hat T_{ij}]-T_{ij}} &\le&
-%\delta + \frac{ (1-1/n)T_{ij} + 1}{n\pi_*}\eps + O\paren{\eps^2+\oo n}.
-O\paren{
-\exp\paren{ -\frac{n(1-e^{-\gamma_*})^2\pi_*^2\eps^2}{8} }
-+\frac{\eps}{n\pi_*}}.
-\eeq
-Choosing $\eps=O\paren{\frac{1}{\pi_*(1-e^{-\gamma^*})\sqrt{n\ln n}}}$ yields
-\beq
-\abs{\E[\hat T_{ij}]-T_{ij}} &\le&
-O\paren{
-\frac{1}{\pi_*^2(1-e^{-\gamma^*})n\sqrt{n\ln n}}}.
-\eeq
-
-\paragraph{Stationariy assumption:}
-One approach is to let the chain run for some time to ``burn in'',
-until it reaches the distribution $\pi'$. If $\pi$ is the true stationary distribution,
-then all of our ``w.h.p.'' results will hold with an additive correction of 
-$\tvnorm{\pi-\pi'}$. The problem is that we won't know for how long to burn in
-and also won't know $\tvnorm{\pi-\pi'}$ --- a chicken and egg problem.
-
-Another idea is to use ``coupling from the past'':\\
-\url{http://www.cc.gatech.edu/~vigoda/MCMC_Course/CouplingFromPast.pdf}
-\\
-\url{http://people.csail.mit.edu/costis/6896sp11/lec8s.pdf}
-\\
-This requires multiple restarts but otherwise works well with finite state spaces.
-It gives us guaranteed perfect samples from the stationary distribution,
-and from that point the chain can be run for $n$ steps to obtain our statistics.
-
-\paragraph{Still need to do:}
-Analytic upper bound on $\EE{ \hat{B}_n(\delta)}$.
-Matching lower bound.
+Recall that our other calculation (``beta-mixing'') gave
+$$ \beta(a) \le \oo{\pi_*}e^{-a(1-\lambda_2)}.$$
 
@@
 \end{document}
-
-Thus, we need estimates of $\gamma_*$ and $\pi_*$.
-To estimate them we may want to estimate the stationary probabilities and the transition matrix $T$.
-Let $\hat \pi$ denote the estimate of the former, while let $\hat T$ be the estimate of the latter.
-In particular, $\hat \pi_i$ is taken to be the empirical frequency of the visits of state $i$ in $(X_t)$,
-while we let $\hat T_{i,j}$ the frequency when 
-$X_{t+1}=j$ happens given that $X_t = i$ ($t=1,\dots,n-1$). 
-When $X_t = i$ never happens, we take $\hat T_{i,\cdot}$ to be the uniform distribution.
-Since $\Prob{\sum_{t=1}^n \one{X_t = i}=0}=\Prob{X_1\ne i, \dots, X_n\ne i}>0$ for each $i$, $\hat T_{ij}$ is a biased estimate of $T_{ij}$, although the bias decreases exponentially fast.
-
-For $x\in \R^d$, let $D_x$ be the diagonal matrix with $x$ in its diagonal. 
-Define $L = D_\pi^{-1/2} T  D_\pi^{1/2}$, $\hat{L} = D_{\hat{\pi}}^{-1/2} \hat{T} D_{\hat{\pi}}^{1/2}$, 
-By Weyl's inequality (see notes.pdf),
-\begin{align}
-\label{eq:gammabound}
-|\hat{\gamma}_* - \gamma_* | = |\hat{\lambda}_* - \lambda_* | \le \tnrm{\hat L - L }_2
-\end{align}
-and so it remains to bound this latter quantity.
-Define $E_T = D_{\pi}^{-1/2} (\hat T - T ) D_{\pi}^{1/2}$
-and $E_\pi = D_{\hat{\pi}}^{-1/2} D_\pi^{1/2} - I$.
-We have (see notes.pdf),
-\begin{align}
-\label{eq:lbound}
-\tnrm{\hat L - L}_2 \le \norm{E_T}_2 + \norm{E_{\pi}}_2 (\norm{E_T}_2+1) (\norm{E_\pi}_2+2)\,.
-\end{align}
-Now, $\norm{E_T}_2 = \tnrm{\hat T - T}_2$ (because the eigenvalues of $E_T$ are the same as that of $\hat T - T$).
-Further, for any $\mu,a$ such that $n=2\mu a$
-and $\delta>2(\mu-1)\beta(a)$,
-with probability at least $1-\delta$,
-\begin{align}
-\label{eq:tbound}
-\tnrm{T-\hat T}_2
-\le 25\sqrt{\frac{d}{\mu}} + 2\sqrt{\frac{\ln(2/\delta')}{2\mu}}\,,
-\end{align}
-where $\delta'=\delta-2(\mu-1)\beta(a)$ (see mohri-notes.pdf). \todoc{This does not take into account the bias.}
-\todoc[inline]{Next problem: For a fixed $n$, this will limit the range of $\delta$.
-For any $n$, there will be a range of $\delta$ in the neighbourhood of $0$ for which the algorithm
-will need to return ``I don't know''.
-}
-Similarly, for any $\mu',a'$ such that $n=2\mu' a'$
-and $\delta>2(\mu'-1)\beta(a')$,
-with probability at least $1-\delta$,
-\begin{align}
-\label{eq:pibound}
-|\hat\pi_i - \pi_i | \le C \left( \sqrt{\frac{d}{\mu'}} + \sqrt{\frac{\ln(2/\delta'')}{2\mu'}}\right)\,,
-\end{align}
-where $\delta''=\delta-2(\mu'-1)\beta(a')$.
-
-By combining~\eqref{eq:betabound},
-\eqref{eq:gammabound}, \eqref{eq:lbound}, \eqref{eq:tbound} and~\eqref{eq:pibound}, we can derive 
-that for any $0\le \delta \le 1$,
- $\mu,a,\mu',a'$  such that $n = 2\mu a$ and
- $\frac{2}{\pi_*} \max( (\mu-1)e^{-a\gamma_*}, (\mu'-1)e^{-a'\gamma_*} )<\delta$, 
-with probability $1-\delta$,
-\begin{align*}
-|\hat{\gamma}_* - \gamma_* | &\le u( \gamma_*, \pi_*, \hat{\gamma}, \hat\pi, \hat T, n, \mu, a, \mu', a', \delta )\,,\\
-|\hat{\pi}_* - \pi_* | &\le v( \gamma_*, \pi_*, \hat{\gamma}, \hat\pi, \hat T, n, \mu, a, \mu', a', \delta )\,,
-\end{align*}
-for some functions $u$ and $v$.
-Reordering (and relaxing) the  inequalities we get
-\begin{align*}
-\gamma_* &\ge \hat\gamma_* - u( \gamma_*, \pi_*, \hat{\gamma}, \hat\pi, \hat T, n, \mu, a, \mu', a', \delta )\,,\\
- \pi_*  &\ge \hat{\pi}_* - v( \gamma_*, \pi_*, \hat{\gamma}, \hat\pi, \hat T, n, \mu, a, \mu', a', \delta )\,.
-\end{align*}
-Now, let $\hat{B}_n(\delta)$ be the optimal value of the optimization problem 
-\begin{align*}
-\lefteqn{\max B_n(g,p,\delta)\qquad \text{s.t.} }\\
-&\qquad  g \ge \hat\gamma_* - u( g, p, \hat{\gamma}, \hat\pi, \hat T, n, \mu, a, \mu', a', \delta )\,,\\
-&\qquad p  \ge \hat{\pi}_* - v( g, p, \hat{\gamma}, \hat\pi, \hat T, n, \mu, a, \mu', a', \delta )\,,\\
-&\qquad n = 2\mu a, \quad n = 2\mu' a'\,,\\
-&\qquad \frac{2}{p} \max( (\mu-1)e^{-ag}, (\mu'-1)e^{-a'g} )<\delta\,,\\
-&\qquad g,p >0, \mu,a,\mu',a'\in \N\,.
-\end{align*}
-Then, w.p. $1-\delta$, it holds that
-\[
-|\bar{X}_n - \EE{f(X_1)}| \le \hat{B}_n(\delta)\,.
-\]
-
-Next goals: Analytic upper bound on $\EE{ \hat{B}_n(\delta)}$.
-Matching lower bound.
-
-\bibliographystyle{plainnat}
-\bibliography{all}
-
-\end{document}

### 2015-06-01 scratch/notes.tex
Copied files from svn

\documentclass[11pt]{article}
\usepackage{todonotes}
\newcommand{\todoc}[2][]{\todo[size=\scriptsize,color=green!10!white,#1]{Cs: #2}} % Csaba's comments
\newcommand{\todot}[2][]{\todo[size=\scriptsize,inline,color=blue!20!white,#1]{D: #2}} % Daniel's comments
\newcommand{\todor}[2][]{\todo[size=\scriptsize,color=orange!20!white,#1]{A: #2}} % Aryeh's comments

\usepackage{amsmath,amsbsy,amsfonts,amssymb,amsthm,color,dsfont}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% PACKAGES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{latexsym}
\usepackage{fullpage}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{enumerate}
\usepackage{accents}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage[ruled]{algorithm}
\usepackage{algpseudocode}
\usepackage{dsfont}
\usepackage[bf]{caption}
\usepackage{hyperref}
\hypersetup{
    bookmarks=true,         % show bookmarks bar?
    unicode=false,          % non-Latin characters in AcrobatÕs bookmarks
    pdftoolbar=true,        % show AcrobatÕs toolbar?
    pdfmenubar=true,        % show AcrobatÕs menu?
    pdffitwindow=false,     % window fit to page when opened
    pdfstartview={FitH},    % fits the width of the page to the window
    pdftitle={My title},    % title
    pdfauthor={Author},     % author
    pdfsubject={Subject},   % subject of the document
    pdfcreator={Creator},   % creator of the document
    pdfproducer={Producer}, % producer of the document
    pdfkeywords={keyword1} {key2} {key3}, % list of keywords
    pdfnewwindow=true,      % links in new window
    colorlinks=true,       % false: boxed links; true: colored links
    linkcolor=red,          % color of internal links (change box color with linkbordercolor)
    citecolor=blue,        % color of links to bibliography
    filecolor=magenta,      % color of file links
    urlcolor=cyan           % color of external links
}
\usepackage{amsthm}
\usepackage{times}
\usepackage{natbib}
\usepackage{nicefrac}
\usepackage{wrapfig}
\usepackage[capitalize]{cleveref}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% MACROS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\defined}{\vcentcolon =}
\newcommand{\rdefined}{=\vcentcolon}
\newcommand{\E}{\mathbb E}
\newcommand{\Var}{\operatorname{Var}}
\newcommand{\calF}{\mathcal F}
\newcommand{\sr}[1]{\stackrel{#1}}
\newcommand{\set}[1]{\left\{#1\right\}}
\newcommand{\ind}[1]{\mathds{1}\!\!\set{#1}}
\newcommand{\argmax}{\operatornamewithlimits{arg\,max}}
\newcommand{\argmin}{\operatornamewithlimits{arg\,min}}
\newcommand{\floor}[1]{\left \lfloor {#1} \right\rfloor}
\newcommand{\ceil}[1]{\left \lceil {#1} \right\rceil}

\def\subsubsect#1{\vspace{1ex plus 0.5ex minus 0.5ex}\noindent{\bf\boldmath{#1.}}}

\renewcommand{\P}[1]{\mathbb{P}\left\{#1\right\}}
\newcommand{\Prob}[1]{\mathbb{P}\left\{#1\right\}}
\newcommand{\EE}[1]{\mathbb{E}\left[#1\right]}

\let\temp\epsilon
\let\epsilon\varepsilon
\newcommand{\eps}{\varepsilon}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% THEOREMS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{example}[theorem]{Example}
\theoremstyle{remark}

\newcommand{\R}{\mathds{R}}

\newcommand\wh{\ensuremath{\widehat}}
\newcommand\wt{\ensuremath{\widetilde}}
\newcommand\norm[1]{\| #1 \|}
\newcommand\Norm[1]{\left\| #1 \right\|}
\newcommand\tvnorm[1]{\left\| #1 \right\|_{\mathrm{TV}}}
\newcommand\Parens[1]{\left( #1 \right)}
\newcommand\parens[1]{( #1 )}
\newcommand\Abs[1]{\left| #1 \right|}
\newcommand\abs[1]{| #1 |}
\DeclareMathOperator{\Diag}{Diag}
\DeclareMathOperator{\Sym}{Sym}
\renewcommand\t{{\ensuremath{\scriptscriptstyle{\top}}}}
\newcommand\be{\ensuremath{\mathbf{e}}}
\newcommand\tmix{\ensuremath{t_{\mathrm{mix}}}}
\newcommand\htmix{\ensuremath{\hat{t}_{\mathrm{mix}}}}
\newcommand{\od}{\bar{d}}
\newcommand{\tcouple}{\tau_{\mathrm{couple}}}
\newcommand{\ZZ}{\mathcal{Z}}
\newcommand\trel{\ensuremath{t_{\mathrm{rel}}}}
\newcommand{\ip}[1]{\langle#1\rangle}
\newcommand{\DF}{\mathcal{E}}
\newcommand\ttmix{\ensuremath{t_{\mathrm{mix},2}}}

\def\ddefloop#1{\ifx\ddefloop#1\else\ddef{#1}\expandafter\ddefloop\fi}
% \bbA, \bbB, ...
\def\ddef#1{\expandafter\def\csname bb#1\endcsname{\ensuremath{\mathbb{#1}}}}
\ddefloop ABCDEFGHIJKLMNOPQRSTUVWXYZ\ddefloop

\title{Statistical Estimation of Mixing Times}
\author{Daniel, Aryeh, Csaba}

\begin{document}
\maketitle

\begin{abstract}
An abstract
\end{abstract}

\section{Problem Definition}
All unattributed statements of this section can be found in \citet{LePeWi08}.
Let $T\in [0,1]^{d\times d}$ be a Markov transition kernel with a single essential communicating class.
Then, as it is well known (e.g., Proposition 1.26 of \citet{LePeWi08}), $T$ has a unique stationary distribution $\pi\in[0,1]^d$:
$\pi \ge 0$, $\pi \be = 1$, $\pi T = \pi$ (we view $\pi$ as a row vector, while $\be = (1,1,\ldots,1)^\t$).

Let $T^t(i,\cdot)$ \todoc{I would prefer to use $P$ to $T$ to make this ``look'' better..}
be the probability distribution over the states 
obtained by running a chain from initial state $i$ for $t$ steps. 
Let 
\[
d(t) = \max_{1\le i \le d} \tvnorm{ T^t(i,\cdot) - \pi }.
\]
The $\eps$-mixing time is defined to be 
\[
\tmix(\eps) = \min\{ t\ge 0: d(t)\le \eps \},
\]
while we define ``the'' mixing time to be 
\[
\tmix = \tmix(1/4).
\]
As it is well known, $d(	\ell \tmix ) \le 2^{-\ell}$, and $\tmix(\eps)\le \ceil{ \ln_2 \eps^{-1} } \, \tmix$.
We are interested in estimating the mixing time based on a finite sample, as the mixing time plays a critical role when developing finite sample confidence bounds for estimating the mean of a function with respect to the stationary distribution using a sample mean obtained by averaging the function values over a single sample path of a Markov chain underlying $T$.
\todoc{Add some results showing this.}

\paragraph{Problem 1:}
Given a sample path $(X_n)_{n\ge 0}$ of $T$ a confidence parameter $0<\delta<1$,
design an algorithm that 
stops at time $\tau >0$ and outputs an estimate $\htmix$ of the mixing time $\tmix$ such that 
$\Prob{  \tmix \le \htmix } \ge 1-\delta$ while keeping $\EE{\tau}$ is as small as possible.
Ideally, we want to find an algorithm whose performance is within a constant factor of that of a minimax optimal algorithm.
Of interest are the running time and the memory requirements of the algorithms.

\noindent Variations:
\begin{enumerate}[{\emph (i)}]
\item Require that $\Prob{ \frac12 \htmix \le \tmix \le \htmix } \ge 1-\delta$ so that $\htmix$ is not a ridiculous overestimate of $\tmix$.
\item Get rid of $\delta$, change the requirement to $\tmix \le \EE{\htmix}$.
\item KWIK-like: Get rid of $\delta$, allow the algorithm output ``I don't know'' (IDK). Require that if the algorithm outputs a real number, it should be an overestimate of $\tmix$. Measure the performance with the probability of outputting IDK.
\item Allow multiple sample paths: The algorithm may stop a sample path and ask for a new one, starting from a fixed known/unknown/chosen probability distribution. If the initial distribution can be chosen, it is likely that the algorithms will sample single transitions only.
\end{enumerate}

\section{Background}
Define $\od(t) = \max_{i,j} \tvnorm{ T^t(i,\cdot) - T^t(j,\cdot) }$. It is well known that $d(t)\le \od(t) \le 2d(t)$ (Eq. (4.23) of \citet{LePeWi08}). Further, the function $\od$ is submultiplicative.
A \emph{coupling of Markov chains} with the transition kernel $T$ is a stochastic process $(X_t,Y_t)_{t\ge 0}$ such that both $(X_t)_t$ and $(Y_t)_t$ are Markov chains with transition kernel $T$. A coupling is canonical if $X_s = Y_s$ holds for any $s\ge t$ such that $X_t = Y_t$, i.e., the chains stay together after the first time they meet. The corresponding coupling time is defined to be $\tcouple = \min \{ t\ge 0: X_t = Y_t \}$.
We have the following result (cf. Corollary~5.3 of \citet{LePeWi08}):
\begin{theorem}
Let $\tcouple(i,j)$ be the coupling time of a canonical coupling $(X_t,Y_t)_t$ of Markov chains with kernel $T$ where $X_0= i$ and $Y_0=j$. Then $\od(t) \le \max_{1\le i, j \le d} \Prob{\tcouple(i,j) > t}$.
\end{theorem}
\todoc[inline]{How tight is this theorem?}
Hence, to test whether $\od(t)\le c$ for some specific time $t$, it suffices to construct canonical couplings with start states $(i,j)$ and estimate the probabilities $\Prob{\tcouple(i,j)>t}$. 
For this, it is probably best to use a so-called \emph{grand coupling} $(X_t^i)_{1\le i \le d, t\ge 0}$ where for each $i$, $(X_t^i)$ is a Markov chain with transition kernel $T$ started from $i$. This can be done by using a random mapping representation of a chain: $X_{t+1}^i = f(X_t^i,Z_{t+1})$, $t\ge 0$, where $(Z_t)_{t\ge 0}$ is an i.i.d. $\ZZ$-valued sequence and $f: [d] \times \ZZ \to [d]$.
Given a grand coupling, we can test for $\max_{1\le i, j \le d} \Prob{\tcouple(i,j) > t}\le 1/4$ at times $t=1,2,2^2,2^3,\ldots$ by running $m$ independent grand couplings while distributing the error probability $\delta = \sum_{k=0}^\infty \delta_k$ to the test times $t = 2^k$ (e.g., $\delta_k = \delta/(k(k+1))$)
 and stopping when at some of the test times $t = 2^k$, $\hat{p}(t)  \le 1/4$, where $\hat{p}(t)$ is an upper estimate of $\max_{i,j} \Prob{\tcouple(i,j)>t}$ with probability $1-\delta_k$.
\todoc[inline]{How does this compare to estimating the transition kernel?}
\todoc[inline]{Can we still use this when there is a single sample path?}

An alternative idea is to use \emph{strong stationary times} to bound $d(t)$:
A strong stationary time for a Markov chain $(X_t)_t$ with $X_0 = i$ and stationary distribution $\pi$ is a randomized stopping time $\tau = \tau(i)$ that may depend on $i$ and is such that 
$\Prob{\tau = t, X_\tau = j} = \Prob{\tau = t} \pi(j)$.
If $\tau$ is a strong stationary time then $d(t) \le \max_i \Prob{\tau(i)> t}$. 
\todoc[inline]{Is this tighter than the previous inequality?}
\todoc[inline]{Is there a general way of constructing strong stationary times?}
\todoc[inline]{Section 6.6 of  \citet{LePeWi08} defines Cesaro mixing times $\tmix^*(\eps)$: These are mixing times defined for the average of the distributions $1/t \sum_{s=0}^{t-1} T^s(i,\cdot)$. Then, if $\tau(i)$ is a stationary time for a chain started from $i$ (i.e., $\Prob{X_\tau = j} = \pi(j)$ for all $j$) then $\tmix^*(1/4) \le 4 \max_i \EE{ \tau(i) }+1$. Is this useful? Can the Cesaro mixing time be used in finite sample bounds for sample means?}

\subsection{Time Reversals and Reversible Chains}
The time reversal of $T$ is defined to be $T^*\in \R^{d\times d}$ satisfying 
$T^*_{i,j} = \pi_j T_{j,i}/\pi_i$. The kernel $T$  is called reversible if $T = T^*$.
If $T$ is reversible and irreducible with stationary distribution $\pi$,
then 
\[
(\trel-1)\, \ln\left(\frac{1}{2\eps}\right)\le \tmix(\eps) \le \trel\, \ln\left( \frac{1}{\eps \pi^* }\right),
\]
where $\pi^* = \min_i \pi_i$ and $\trel = 1/\gamma_*$ is the relaxation time of the Markov chain with kernel $T$ and \emph{absolute spectral gap} $\gamma_*$, which is defined by $\gamma_* = 1-\lambda_*$, where $\lambda_* = \max\{ |\lambda| : \lambda \text{ is an eigenvalue of } T, \lambda \ne 1 \}$. ($T$ being aperiodic and irreducible implies that $\gamma_*>0$.)

Define the $\pi$-weighted inner product by $\ip{u,v}_{\pi} = \sum_i u_i v_i \pi_i$ and denote the corresponding norm by $\norm{\cdot}_{2,\pi}$. More generally, we denote the $\pi$-weighted $p$-norm by $\norm{\cdot}_{p,\pi}$.
Let $T$ be a reversible transition matrix, with eigenvalues $1 = \lambda_1 > \lambda_2 \ge \ldots \lambda_d \ge -1$ and associated eigenvectors $\{v_j\}_j$ with respect to $\ip{\cdot,\cdot}_\pi$. Then,
\[
\tvnorm{ T^t(i,\cdot) - \pi}^2 
= \frac14  \Norm{\frac{T^t(i,\cdot)}{\pi(\cdot)} - \be }_{1,\pi}^2  
\le \frac14  \Norm{\frac{T^t(i,\cdot)}{\pi(\cdot)} - \be }_{2,\pi}^2
= \frac14 \sum_{j=2}^d v_j^2 \lambda_j^{2t}\,
\]
(cf. Lemma~12.16 of \citet{LePeWi08}).
\if0
Furthermore, if the chain is transitive then 
\[
\tvnorm{ T^t(i,\cdot) - \pi}^2 \le \frac14  \Norm{\frac{T^t(i,\cdot)}{\pi_i} - \be }_2^2 \frac14 \sum_{j=2}^d \lambda_j^{2t}\,.
\]
\fi
We shall let $d_2(t) =  \max_i \Norm{\frac{T^t(i,\cdot)}{\pi(\cdot)} - \be }_{2,\pi}$ and the $\ell^2$ mixing time by $\ttmix(\eps) = \min\{ t: d_2(t)\le \eps\}$.
Denote $\norm{T^*} = \sup_{v\ne 0, v^\top \pi = 0} \frac{\norm{T^* v}_{2,\pi}}{\norm{v}_{2,\pi}}$.
Then, 
\[
\ttmix(\eps) \le \ceil{ \frac{1}{1-\norm{T^*}} \ln \frac{1}{\eps\sqrt{\pi_*}}} 
%=  \ceil{\frac1{\gamma_{T T^*}}\ln \frac1{\eps\sqrt{\pi_*}}} \,,
\]
(cf. Proposition~1.2 \citet{MoTe06}). 
\if0
Here, for a Markov chain with kernel $K$ and stationary distribution $\pi$,
 $\gamma_{K}$ is defined by
\[
\gamma_{K} = \inf_v \frac{\mathrm{Var}_\pi(v)}{\DF_K(v,v)}\,,
\]
where $\mathrm{Var}_\pi(v) = \sum_i (v_i - \ip{v,\pi})^2 \pi_i$ and
$\DF_K(u,v) = \ip{ u, (I-K)v }_\pi$ is the \emph{Dirichlet form} underlying $K$.
If $T$ is reversible, $\gamma_{T T^*} =\gamma_{T^2} = \gamma_*^2$.
\fi
If the chain is non-reversible, the above bound may be trivial (even though the chain may be contracting). For example, this happens when $d = 3$, $T_{1,2} = T_{2,3} = 1$, $T_{3,3} = T_{3,1} = 1/2$. The origin of the failure of the bound is that the methods require that the distance decreases at every step, while if the initial distribution is concentrated at state $1$, then it will take two steps before the distance decreases. This is a common problem with the bounds of this type for the nonreversible case.
%Then, $\pi_1 = \pi_2 = 1/4$, $\pi_3 = 1/2$.

\section{The matrix estimation method}

Let $T$ be the probability kernel for a $d$-state reversible Markov
chain, so
\[ \Pr(X_{n+1} = j | X_n = i) = T_{i,j} . \]
and
\[
  \pi_i T_{i,j} = \pi_j T_{j,i}
\]
for some stationary distribution given by $\pi$.

Let $P := \Diag(\pi) T$ (i.e., $P_{i,j} := \pi_i T_{i,j}$), so $L =
\Diag(\pi)^{-1/2} P \Diag(\pi)^{-1/2}$.
We are interested in approximating the eigenvalues of the matrix $L :=
\Diag(\pi)^{1/2} T \Diag(\pi)^{-1/2}$ (which is symmetric by
reversibility).

Our approach is to directly estimate $P$ and $\pi$ to form an estimate
$\wh{L}$ of $L$.
Then, we use the eigenvalues of $\wh{L}$ to form our mixing time
estimate.

\subsection{Perturbation analysis}

Let $\wh{P}$ and $\hat\pi$ be estimates of $P$ and $\pi$,
respectively, and define
\begin{align*}
  \wh{L} & := \Diag(\hat\pi)^{-1/2} \wh{P} \Diag(\hat\pi)^{-1/2} \\
  E_\pi & := \Diag(\hat\pi)^{-1/2} \Diag(\pi)^{1/2} - I \\
  E_P & := \Diag(\pi)^{-1/2} (\wh{P} - P) \Diag(\pi)^{-1/2} .
\end{align*}
Then
\[
  \wh{L} - L
  = E_P + E_\pi L + L E_\pi + E_\pi L E_\pi + E_\pi E_P + E_P E_\pi +
  E_\pi E_P E_\pi
  ,
\]
so
\[
  \norm{\wh{L} - L}_2
  \leq \norm{E_P}_2 + (\norm{E_P}_2 + \norm{L}_2)
  (2\norm{E_\pi}_2 + \norm{E_\pi}_2^2)
  .
\]

Recall that $L$ is symmetric and that its eigenvalues are conained in
$[-1,1]$.
This is because $I - L$ is a normalized Laplacian matrix, which has
eigenvalues contained in the interval $[0,2]$.
In fact, the smallest eigenvalue of $I - L$ is zero, so the largest
eigenvalue of $L$ is equal to $1$.
Therefore $\norm{L}_2 = \max\{\lambda_{\max}(L),-\lambda_{\min}(L)\} =
1$, and
\[
  \norm{\wh{L} - L}_2
  \leq \norm{E_P}_2 + (\norm{E_P}_2 + 1)
  (2\norm{E_\pi}_2 + \norm{E_\pi}_2^2)
  .
\]

Let $\Sym(A) := 0.5 (A + A^\t)$ for a square matrix $A$.
By the triangle inequality,
\[
  \norm{\Sym(\wh{L}) - L}_2
  \leq \norm{\wh{L} - L}_2 .
\]
Let $\lambda_i(A)$ denote the $i$-th eigenvalue of a symmetric matrix
$A$ (assuming the eigenvalues are arranged in non-increasing order).
By Weyl's inequality,
\[
  |\lambda_i(\Sym(\wh{L})) - \lambda_i(L)| \leq
  \norm{\wh{L} - L}_2
\]
for all $i\in[d]$.

\subsection{Estimation error in iid setting}

Suppose we have $n$ iid copies of $(X_t,X_{t+1})$, where the marginal
distribution of $X_t$ is $\pi$, and $\wh{P}$ is the maximum likelihood
estimate of $P$.
Then we can write $E_P$ as the average of $n$ independent copies of the
zero-mean random matrix $Z$, where
\[
  \Pr\Parens{ Z = \frac1{\sqrt{\pi_i\pi_j}} e_i e_j^\t - L } = P_{i,j} .
\]
We can apply the matrix Bernstein inequality to bound $\norm{E_P}_2$ with
high probability.
To do this, we need to bound the following quantities:
\[
  \norm{Z}_2 , \quad
  \norm{\bbE(ZZ^\t)}_2 , \quad
  \norm{\bbE(Z^\t Z)}_2 .
\]
By the triangle inequality, we have
\[
  \norm{Z}_2
  \leq \frac{1}{\pi_{\min}} + \norm{L}_2
  = \frac{1}{\pi_{\min}} + 1
\]
where $\pi_{\min} := \min_{i\in[d]} \pi_i$.
Moreover, by reversibility,
\[
  \bbE(ZZ^\t)
  = \bbE(Z^\t Z)
  = \sum_{i=1}^d \Parens{ \sum_{j=1}^d \frac{T_{i,j}}{\pi_j} }
  e_ie_i^\t - L^2
\]
so
\[
  \norm{\bbE(ZZ^\t)}_2
  = \norm{\bbE(Z^\t Z)}_2
  \leq \max_{i \in [d]} \sum_{j=1}^d \frac{T_{i,j}}{\pi_j} + \norm{L}_2^2
  = D_{T,\pi} + 1
\]
where $D_{T,\pi} := \max_{i\in[d]} \sum_{j=1}^d T_{i,j} / \pi_j \leq
1/\pi_{\min}$.
By the matrix Bernstein inequality, there is an absolute constant
$c>0$ (maybe $2$ or so) such that with probability at least
$1-\delta$,
\begin{align*}
  \norm{E_P}_2
  & \leq c \sqrt{\Parens{ \frac1{\pi_{\min}} + 1 }\frac{\ln(4d/\delta)}{n}}
  + c \Parens{ \frac1{\pi_{\min}} + 1 } \frac{\ln(4d/\delta)}{n}
  .
\end{align*}

We can also estimate $\pi$ using a maximum likelihood estimator
$\wh\pi$.
We want a bound on
\[
  \norm{E_\pi}_2 = \max_{i\in[d]} \Abs{\sqrt{\pi_i/\wh\pi_i} - 1}
  .
\]
If we are willing to assume that $n \geq c / \pi_{\min}$ for some
suitable constant $c>0$, then by multiplicative Chernoff bounds and a
union bound, we have with probability at least $1-\delta$,
\[
  \norm{E_\pi}_2 \leq
  c' \sqrt{\frac{\ln(d/\delta)}{\pi_{\min} n}} +
  + c' \frac{\ln(d/\delta)}{\pi_{\min} n}
  .
\]
We might be able to avoid this assumption by using something like
``inverse sampling''.

\bibliographystyle{plainnat}
\bibliography{../all}

\end{document}

### 2015-06-01 scratch/old-notes.tex
Copied files from svn

--- 2015-06-01 scratch/notes.tex
@@
 \newcommand\wt{\ensuremath{\widetilde}}
-\newcommand\norm[1]{\| #1 \|}
-\newcommand\Norm[1]{\left\| #1 \right\|}
+\newcommand\norm[1]{\left\| #1 \right\|}
 \newcommand\tvnorm[1]{\left\| #1 \right\|_{\mathrm{TV}}}
 \newcommand\Parens[1]{\left( #1 \right)}
-\newcommand\parens[1]{( #1 )}
-\newcommand\Abs[1]{\left| #1 \right|}
-\newcommand\abs[1]{| #1 |}
 \DeclareMathOperator{\Diag}{Diag}
@@
 \tvnorm{ T^t(i,\cdot) - \pi}^2 
-= \frac14  \Norm{\frac{T^t(i,\cdot)}{\pi(\cdot)} - \be }_{1,\pi}^2  
-\le \frac14  \Norm{\frac{T^t(i,\cdot)}{\pi(\cdot)} - \be }_{2,\pi}^2
+= \frac14  \norm{\frac{T^t(i,\cdot)}{\pi(\cdot)} - \be }_{1,\pi}^2  
+\le \frac14  \norm{\frac{T^t(i,\cdot)}{\pi(\cdot)} - \be }_{2,\pi}^2
 = \frac14 \sum_{j=2}^d v_j^2 \lambda_j^{2t}\,
@@
 \[
-\tvnorm{ T^t(i,\cdot) - \pi}^2 \le \frac14  \Norm{\frac{T^t(i,\cdot)}{\pi_i} - \be }_2^2 \frac14 \sum_{j=2}^d \lambda_j^{2t}\,.
+\tvnorm{ T^t(i,\cdot) - \pi}^2 \le \frac14  \norm{\frac{T^t(i,\cdot)}{\pi_i} - \be }_2^2 \frac14 \sum_{j=2}^d \lambda_j^{2t}\,.
 \]
 \fi
-We shall let $d_2(t) =  \max_i \Norm{\frac{T^t(i,\cdot)}{\pi(\cdot)} - \be }_{2,\pi}$ and the $\ell^2$ mixing time by $\ttmix(\eps) = \min\{ t: d_2(t)\le \eps\}$.
+We shall let $d_2(t) =  \max_i \norm{\frac{T^t(i,\cdot)}{\pi(\cdot)} - \be }_{2,\pi}$ and the $\ell^2$ mixing time by $\ttmix(\eps) = \min\{ t: d_2(t)\le \eps\}$.
 Denote $\norm{T^*} = \sup_{v\ne 0, v^\top \pi = 0} \frac{\norm{T^* v}_{2,\pi}}{\norm{v}_{2,\pi}}$.
@@
 
-\section{The matrix estimation method}
+\section{Analysis via Estimation of $T$}
 
@@
 
-Let $P := \Diag(\pi) T$ (i.e., $P_{i,j} := \pi_i T_{i,j}$), so $L =
-\Diag(\pi)^{-1/2} P \Diag(\pi)^{-1/2}$.
+\subsection{Estimation error}
+
+We use the median-of-means technique to estimate all $d^2$ entries of
+$T$ (this is just to get a variance-only bound with high probability;
+using the empirical average also works but will involve some
+additional lower-order terms).
+That is, for each $(i,j)\in[d]^2$, we consider the $\hat{n}_i$
+times $t$ where $X_t = i$ and split them into $k$ groups each of size
+$\hat{n}_i/k$; within each group, we compute the fraction of these
+times where $X_{t+1}=j$; we then compute the median of these fractions
+and use that as the estimate of $T_{i,j}$.
+Call these estimates $\wt{T}_{i,j}$.
+Using $k := 4.5\ln(d^2/\delta)$, we have using the median-of-means
+theorem and a union bound,
+\[
+  \Parens{ \wt{T}_{i,j} - T_{i,j} }^2
+  \leq \frac{27T_{i,j}(1-T_{i,j})\ln(d^2/\delta)}{\hat{n}_i}
+\]
+for all $(i,j) \in [d]^2$, with probability $\geq1-\delta$.
+Therefore, in this event,
+\begin{align*}
+  \norm{\Diag(\pi)^{1/2} (\wt{T} - T) \Diag(\pi)^{-1/2}}_F^2
+  & = \sum_{i=1}^d \sum_{j=1}^d \frac{\pi_i}{\pi_j}
+  \Parens{ \wt{T}_{i,j} - T_{i,j} }^2 \\
+  & \leq \sum_{i=1}^d \sum_{j=1}^d
+  \frac{27\pi_i T_{i,j} (1 - T_{i,j})\ln(d^2/\delta)}{\pi_j\hat{n}_i} \\
+  & = \sum_{i=1}^d \sum_{j=1}^d
+  \frac{27\pi_j T_{j,i} (1 - T_{i,j})\ln(d^2/\delta)}{\pi_j\hat{n}_i} \\
+  & = \sum_{i=1}^d \sum_{j=1}^d
+  \frac{27T_{j,i} (1 - T_{i,j})\ln(d^2/\delta)}{\hat{n}_i} \\
+  & = \sum_{i=1}^d \frac{27d_i\ln(d^2/\delta)}{\hat{n}_i}
+\end{align*}
+where $d_i := \sum_{j=1}^d T_{j,i} (1 - T_{i,j})$.
+
+\subsection{Perturbation analysis}
+
 We are interested in approximating the eigenvalues of the matrix $L :=
@@
 
-Our approach is to directly estimate $P$ and $\pi$ to form an estimate
-$\wh{L}$ of $L$.
-Then, we use the eigenvalues of $\wh{L}$ to form our mixing time
-estimate.
+Let $\wt{T}$ and $\tilde\pi$ be estimates of $T$ and $\pi$,
+respectively.
+%Assume the Markov chain associated with $\wt{T}$ has a stationary
+%distribution denoted by $\tilde\pi$.
+%Since $I - T + 1\pi^\t$ is invertible and
+%\[
+%  (\tilde\pi - \pi)^\t (I - T + 1\pi^\t)
+%  = \tilde\pi^\t (\wt{T} - T)
+%  ,
+%\]
+%we can bound $\|\tilde\pi - \pi\|_\infty$ by
+%\[
+%  \|\tilde\pi - \pi\|_\infty
+%  \leq \|(\wt{T} - T)(I - T + 1\pi^\t)^{-1}\|_{\infty\to1}
+%  .
+%\]
+Define
+\begin{align*}
+  \wt{L}
+  & := \Diag(\tilde\pi)^{1/2} \wt{T} \Diag(\tilde\pi)^{-1/2}
+  \\
+  E_\pi & := \Diag(\pi)^{1/2} \Diag(\tilde\pi)^{-1/2} - I \\
+  E_\pi' & := \Diag(\tilde\pi)^{1/2} \Diag(\pi)^{-1/2} - I \\
+  E_T & := \Diag(\pi)^{1/2} (\wt{T} - T) \Diag(\pi)^{-1/2} .
+\end{align*}
+Then
+\[
+  \wt{L} - L
+  = E_T + E_\pi' L + L E_\pi + E_\pi' L E_\pi
+  + E_\pi' E_T + E_T E_\pi + E_\pi' E_T E_\pi
+  ,
+\]
+so
+\begin{align*}
+  \|\wt{L} - L\|_2
+  & \leq \|E_T\|_2
+  + (\|L\|_2 + \|E_T\|_2)
+  (\|E_\pi\|_2 + \|E_\pi'\|_2 + \|E_\pi\|_2\|E_\pi'\|_2)
+  .
+\end{align*}
+
+Let $\Sym(A) := 0.5 (A + A^\t)$ for a square matrix $A$.
+By the triangle inequality,
+\[
+  \norm{\Sym(\wt{L}) - L}_2
+  \leq \|\wt{L} - L\|_2 .
+\]
+Let $\lambda_i(A)$ denote the $i$-th eigenvalue of a symmetric matrix
+$A$ (assuming the eigenvalues are arranged in non-increasing order).
+By Weyl's inequality,
+\[
+  |\lambda_i(\Sym(\wt{L})) - \lambda_i(L)| \leq
+  \|\wt{L} - L\|_2
+\]
+for all $i\in[d]$.
+
+\section{Analysis via Estimation of $\Diag(\pi)T$}
+
+Let $P := \Diag(\pi) T$ (i.e., $P_{i,j} := \pi_i T_{i,j}$), so $L =
+\Diag(\pi)^{-1/2} P \Diag(\pi)^{-1/2}$.
 
@@
 \[
-  \norm{\wh{L} - L}_2
-  \leq \norm{E_P}_2 + (\norm{E_P}_2 + \norm{L}_2)
-  (2\norm{E_\pi}_2 + \norm{E_\pi}_2^2)
+  \|\wh{L} - L\|_2
+  \leq \|E_P\|_2 + (\|E_P\|_2 + \|L\|_2) (2\|E_\pi\|_2 + \|E_\pi\|_2^2)
   .
@@
 
-Recall that $L$ is symmetric and that its eigenvalues are conained in
-$[-1,1]$.
+Recall that $L$ is symmetric and that its eigenvalues are between $-1$
+and $+1$.
 This is because $I - L$ is a normalized Laplacian matrix, which has
@@
 In fact, the smallest eigenvalue of $I - L$ is zero, so the largest
-eigenvalue of $L$ is equal to $1$.
-Therefore $\norm{L}_2 = \max\{\lambda_{\max}(L),-\lambda_{\min}(L)\} =
-1$, and
+eigenvalue of $L$ is equal to $1$; the smallest eigenvalue of $L$ is
+at least $-1$.
+Therefore $\|L\|_2 = \max\{\lambda_{\max}(L),-\lambda_{\min}(L)\} =
+1$.
+Therefore
 \[
-  \norm{\wh{L} - L}_2
-  \leq \norm{E_P}_2 + (\norm{E_P}_2 + 1)
-  (2\norm{E_\pi}_2 + \norm{E_\pi}_2^2)
+  \|\wh{L} - L\|_2
+  \leq \|E_P\|_2 + (\|E_P\|_2 + 1) (2\|E_\pi\|_2 + \|E_\pi\|_2^2)
   .
 \]
-
-Let $\Sym(A) := 0.5 (A + A^\t)$ for a square matrix $A$.
-By the triangle inequality,
-\[
-  \norm{\Sym(\wh{L}) - L}_2
-  \leq \norm{\wh{L} - L}_2 .
-\]
-Let $\lambda_i(A)$ denote the $i$-th eigenvalue of a symmetric matrix
-$A$ (assuming the eigenvalues are arranged in non-increasing order).
-By Weyl's inequality,
-\[
-  |\lambda_i(\Sym(\wh{L})) - \lambda_i(L)| \leq
-  \norm{\wh{L} - L}_2
-\]
-for all $i\in[d]$.
 
@@
 \]
-We can apply the matrix Bernstein inequality to bound $\norm{E_P}_2$ with
+We can apply the matrix Bernstein inequality to bound $\|E_P\|_2$ with
 high probability.
@@
 \[
-  \norm{Z}_2 , \quad
-  \norm{\bbE(ZZ^\t)}_2 , \quad
-  \norm{\bbE(Z^\t Z)}_2 .
+  \|Z\|_2 , \quad
+  \|\bbE(ZZ^\t)\|_2 , \quad
+  \|\bbE(Z^\t Z)\|_2 .
 \]
@@
 \[
-  \norm{Z}_2
-  \leq \frac{1}{\pi_{\min}} + \norm{L}_2
+  \|Z\|_2
+  \leq \frac{1}{\pi_{\min}} + \|L\|_2
   = \frac{1}{\pi_{\min}} + 1
@@
 \[
-  \norm{\bbE(ZZ^\t)}_2
-  = \norm{\bbE(Z^\t Z)}_2
-  \leq \max_{i \in [d]} \sum_{j=1}^d \frac{T_{i,j}}{\pi_j} + \norm{L}_2^2
+  \|\bbE(ZZ^\t)\|_2
+  = \|\bbE(Z^\t Z)\|_2
+  \leq \max_{i \in [d]} \sum_{j=1}^d \frac{T_{i,j}}{\pi_j} + \|L\|_2^2
   = D_{T,\pi} + 1
@@
 By the matrix Bernstein inequality, there is an absolute constant
-$c>0$ (maybe $2$ or so) such that with probability at least
-$1-\delta$,
+$c>0$ such that with probability at least $1-\delta$,
 \begin{align*}
-  \norm{E_P}_2
-  & \leq c \sqrt{\Parens{ \frac1{\pi_{\min}} + 1 }\frac{\ln(4d/\delta)}{n}}
-  + c \Parens{ \frac1{\pi_{\min}} + 1 } \frac{\ln(4d/\delta)}{n}
+  \|E_P\|_2
+  & \leq c \sqrt{\frac{(D_{T,\pi} + 1) \ln(d/\delta)}{n}}
+  + c \Parens{ \frac1{\pi_{\min}} + 1 } \frac{\ln(d/\delta)}{n}
+  \\
+  & \leq c \sqrt{\Parens{ \frac1{\pi_{\min}} + 1 }\frac{\ln(d/\delta)}{n}}
+  + c \Parens{ \frac1{\pi_{\min}} + 1 } \frac{\ln(d/\delta)}{n}
   .
@@
 
-We can also estimate $\pi$ using a maximum likelihood estimator
-$\wh\pi$.
-We want a bound on
-\[
-  \norm{E_\pi}_2 = \max_{i\in[d]} \Abs{\sqrt{\pi_i/\wh\pi_i} - 1}
-  .
-\]
-If we are willing to assume that $n \geq c / \pi_{\min}$ for some
-suitable constant $c>0$, then by multiplicative Chernoff bounds and a
-union bound, we have with probability at least $1-\delta$,
-\[
-  \norm{E_\pi}_2 \leq
-  c' \sqrt{\frac{\ln(d/\delta)}{\pi_{\min} n}} +
-  + c' \frac{\ln(d/\delta)}{\pi_{\min} n}
-  .
-\]
-We might be able to avoid this assumption by using something like
-``inverse sampling''.
-
 \bibliographystyle{plainnat}
-\bibliography{../all}
+\bibliography{all}
 

### 2015-06-01 scratch/synthesis.tex
Copied files from svn

--- 2015-06-01 scratch/aryeh-scratch-notes.tex
@@
 
-\title{Aryeh's scratch notes}
+\title{Synthesis..}
 %\author{Daniel, Aryeh, Csaba}
@@
 \newcommand{\N}{\mathbb{N}}
-For any $\mu,a$ such that $n = 2\mu a$ and $\delta>2(\mu-1)\beta(a)$,
+Let  $X_1,\ldots,X_n$ be a  Markov chain with transition matrix $T$
+over the state space $\XX = \{1,\ldots,d \}$.
+We assume that the chain is reversible, irreducible, and stationary (RIS).
+\todoc[inline]{Fix this: We don't want stationarity.}
+Our goal is to derive a hight probability bound on the deviation 
+$\bar{X}_n - \EE{f(X_1)} = \frac1n \sum_{t=1}^n f(X_t) - \EE{f(X_1)}$
+(here, $f$ is some $[0,1]$-valued function).
+We know that $\bar{X}_n - \EE{f(X_1)}$ can be bounded in terms of the $\beta$-mixing coefficients of $(X_i)$,
+following the ``blocking'' method of \cite{MR1258867}.
+In particular, for any $\mu,a$ such that $n = 2\mu a$ and $\delta>2(\mu-1)\beta(a)$,
 with probability at least $1-\delta$,
@@
 \end{align}
-where $\delta'=\delta-2(\mu-1)\beta(a)$
-If $\beta(k)$ denotes the $k$th $\beta$-mixing coefficient of $(X_i)$ then 
+where $\delta'=\delta-2(\mu-1)\beta(a)$ (see mohri-notes.pdf).
+Now note that if $\beta(k)$ denotes the $k$th $\beta$-mixing coefficient of $(X_i)$ then 
 \begin{align}
 \label{eq:betabound}
- \beta(k) \le \oo{\pi_*}e^{-k \gamma_*}
+ \beta(k) \le \oo{\pi_*}e^{-k \gamma^*}
 \end{align}
-where  $\gamma_* = 1-\lambda_*$ is the spectral gap ($\lambda_*$ is the second largest eigenvalue of $T$)
-and  $\pi_*  = \min_i \pi_i>0$ is the critical probability underlying the chain
+where  $\gamma^* = 1-\lambda^*$ is the spectral gap ($\lambda^*$ is the second largest eigenvalue of $T$)
+and  $\pi^*  = \min_i \pi_i>0$ is the critical probability underlying the chain
 (see the $\beta$-mixing note).
-For a given value of $\pi_*,\gamma_*$, $\delta$, let $B_n(\pi_*,\gamma_*,\delta)$ be the optimal value of the optimization problem
+For a given value of $\pi^*,\gamma^*$, $\delta$, let $B_n(\pi^*,\gamma^*,\delta)$ be the optimal value of the optimization problem
 \begin{align*}
-\lefteqn{ \min_{\mu,a} \min\left(1, 
-c_1\sqrt{\frac{1}{\mu}} + c_2\sqrt{\frac{\ln(2/\delta')}{2\mu}}\right) 
-} \\
-& \qquad \text{s.t.} \quad n = 2\mu a, \quad \delta>2(\mu-1)\oo{\pi_*}e^{-a \gamma_*},
-\quad \delta' = \delta-2(\mu-1)\oo{\pi_*}e^{-a \gamma_*}, \quad \mu,a \in \N\,.
+\lefteqn{ \min_{\mu,a} \min\left(1, c_1\sqrt{\frac{1}{\mu}} + c_2\sqrt{\frac{\ln(2/\delta')}{2\mu}}\right) } \\
+& \qquad \text{s.t.} \quad n = 2\mu a, \quad \delta>2(\mu-1)\oo{\pi_*}e^{-a \gamma^*},
+\quad \delta' = \delta-2(\mu-1)\oo{\pi_*}e^{-a \gamma^*}, \quad \mu,a \in \N\,.
 \end{align*}
 
-\paragraph{We will upper-bound} $B_n$ as follows.
-Choose
-$$ a=(2\ln n)/\gamma_*.$$
-Then
-$\mu=n/2a=\gamma_*n/4\ln n$
-and
-\beq
-\delta' &=& \delta-2(\mu-1)\oo{\pi_*}e^{-a \gamma_*} \\
-&=& \delta - 2\paren{\frac{\gamma_*n}{4\ln n}-1}\oo{n^2\pi_*} \\
-%&=& \delta - {\frac{\gamma_*}{2\pi_*n\ln n}} + O(n^{-2}) \\
-&=&
-\delta - \sqprn{
-\frac{\gamma_*}{2\pi_*n\ln n}-\frac{2}{\pi_*n^2}
-}
-=:\delta-x.
-%\frac{2}{\delta'}
-%&=&
-%\frac{2}{\delta} + 
-\eeq
-Now
-\beq
-\ln\frac{2}{\delta-x} &=& \ln\frac{2}{\delta}+\frac{x}{\delta^2}+O(x^2), \\
-\sqrt{\ln\frac{2}{\delta-x}} &=& 
-\sqrt{\ln\frac{2}{\delta}}
-+
-\frac{x}{2\delta\sqrt{\ln(2/\delta)}}+O(x^2), \\
-\eeq
-and hence,
-\beq
-B_n(\pi_*,\gamma_*,\delta)
-&\le&
-2\sqrt{\frac{\ln n}{\gamma_*n}}
-\sqprn{
-c_1
-+ c_2
-\sqrt{\ln(2/\delta')/2}
-}\\
-&=&
-2\sqrt{\frac{\ln n}{\gamma_*n}}
-\sqprn{
-c_1
-+ c_2
-\paren{
-\sqrt{
-\oo2
-\ln\frac{2}{\delta}}
-+
-\frac{x}{2\delta\sqrt{2\ln(2/\delta)}}+O(x^2)
-}
-}\\
-&=&
-2\sqrt{\frac{\ln n}{\gamma_*n}}
-\sqprn{
-c_1
-+ c_2
-\paren{
-\sqrt{\oo2\ln\frac{2}{\delta}}
-+
-\frac{
-\gamma_*
-}{4{\pi_*n\ln n}\delta\sqrt{2\ln(2/\delta)}}+O(n^{-2})
-}
-}\\
-&=&
-2c_1\sqrt{\frac{\ln n}{\gamma_*n}}
-+
-2c_2\sqrt{\frac{\ln n}{2\gamma_*n}
-\ln\frac{2}{\delta}}
-+
-c_2\frac{\sqrt{\gamma_*}}
-{2\pi_* n^{3/2}\delta\sqrt{2\ln(n) \ln(2/\delta)}}
-+
-\tilde O(n^{-2.5}).
-\eeq
-
-\paragraph{Estimating the bias in $\hat T_{ij}$.}
-This analysis assumes stationarity. 
-For $1\le i,j\le d$, define the random variable $N_{ij}=\#(i,j)$ to be the number of consecutive $(i,j)$
-pairs in the observed sequence; define $N_i=\#(i)$ analogously.
-We have
-\beq
-\hat T_{ij} = \frac{ N_{ij}}{N_i}.
-\eeq
-We will estimate the bias $\E[\hat T_{ij}]-T_{ij}$ as follows.
-First, compute the expectations of the numerator and the denominator:
-\beq
-\E[N_{ij}]
-&=& 
-(n-1)\pi_i T_{ij} 
-\\
-\E[N_{i}]
-&=& n\pi_i
-,\\
-\frac{\E[N_{ij}]}{\E[N_{i}]}
-&=&
-\frac{(n-1)\pi_i T_{ij}}{n\pi_i} = T_{ij}+O\paren{\oo n}.
-\eeq
-To estimate $\E[\hat T_{ij}]$, we are going to claim that both the numerator and the denominator
-are concentrated about their means.
-In particular, both $N_{ij}$ and $N_i$ are Lipschitz (w.r.t. Hamming) functions of the state sequence
-(with constants $2$ and $1$, resp.),
-and hence, invoking \citet[Theorem 1.1]{meroi2012-jap} with $G=1/\pi_*$ and $\theta=e^{-\gamma_*}$,
-\beq
-\P{ \abs { 
-%\oo n\E[N_{ij}] - (1-\oo n)\pi_i T_{ij} 
-N_{ij}-\E[N_{ij}]
-} > \eps n
-} &\le& 2\exp\paren{ -\frac{n(1-e^{-\gamma_*})^2\pi_*^2\eps^2}{8} } \\
-\P{ \abs { 
-%\oo n\E[N_i] - \pi_i  
-N_{i}-\E[N_{i}]
-} > 
-\eps n} &\le& 2\exp\paren{ -\frac{n(1-e^{-\gamma_*})^2\pi_*^2\eps^2}{2} } \\
-\eeq
-Thus, with probability at least
-\beq
-1-\delta &:=& 1-4 \exp\paren{ -\frac{n(1-e^{-\gamma_*})^2\pi_*^2\eps^2}{8} },
-\eeq
-%all of 
-the empirical counts $N_{ij}$ and $N_i$ are $\eps n$-close to their expectations.
-Finally, since for all $a,b,x>0$, we have
-\beq
-\abs{ \frac{a\pm x}{b\pm x} - \frac a b} \le \frac{a+b}{b^2}x+O(x^2), 
-\eeq
-it follows that for all $i,j$,
-\beq
-\abs{
-\E\sqprn{\frac{ N_{ij}}{N_i}}
--
-\frac{\E[N_{ij}]}{\E[N_{i}]}
-}
-&\le&
-\delta + \frac{\E[N_{ij}]+\E[N_i]}{\E[N_i]^2}\eps + O(\eps^2) \\
-&=& \delta + \frac{ (n-1)\pi_i T_{ij} + n\pi_i}{n^2\pi_i^2}\eps + O(\eps^2) \\
-&\le& \delta + \frac{ (1-1/n)T_{ij} + 1}{n\pi_*}\eps + O(\eps^2).
-\eeq
-
-Thus,
-\beq
-\abs{\E[\hat T_{ij}]-T_{ij}} &\le&
-%\delta + \frac{ (1-1/n)T_{ij} + 1}{n\pi_*}\eps + O\paren{\eps^2+\oo n}.
-O\paren{
-\exp\paren{ -\frac{n(1-e^{-\gamma_*})^2\pi_*^2\eps^2}{8} }
-+\frac{\eps}{n\pi_*}}.
-\eeq
-Choosing $\eps=O\paren{\frac{1}{\pi_*(1-e^{-\gamma^*})\sqrt{n\ln n}}}$ yields
-\beq
-\abs{\E[\hat T_{ij}]-T_{ij}} &\le&
-O\paren{
-\frac{1}{\pi_*^2(1-e^{-\gamma^*})n\sqrt{n\ln n}}}.
-\eeq
-
-\paragraph{Stationariy assumption:}
-One approach is to let the chain run for some time to ``burn in'',
-until it reaches the distribution $\pi'$. If $\pi$ is the true stationary distribution,
-then all of our ``w.h.p.'' results will hold with an additive correction of 
-$\tvnorm{\pi-\pi'}$. The problem is that we won't know for how long to burn in
-and also won't know $\tvnorm{\pi-\pi'}$ --- a chicken and egg problem.
-
-Another idea is to use ``coupling from the past'':\\
-\url{http://www.cc.gatech.edu/~vigoda/MCMC_Course/CouplingFromPast.pdf}
-\\
-\url{http://people.csail.mit.edu/costis/6896sp11/lec8s.pdf}
-\\
-This requires multiple restarts but otherwise works well with finite state spaces.
-It gives us guaranteed perfect samples from the stationary distribution,
-and from that point the chain can be run for $n$ steps to obtain our statistics.
-
-\paragraph{Still need to do:}
-Analytic upper bound on $\EE{ \hat{B}_n(\delta)}$.
-Matching lower bound.
-
-\bibliographystyle{plainnat}
-\bibliography{all}
-
-\end{document}
-
-Thus, we need estimates of $\gamma_*$ and $\pi_*$.
+Thus, we need estimates of $\gamma^*$ and $\pi^*$.
 To estimate them we may want to estimate the stationary probabilities and the transition matrix $T$.
@@
 \label{eq:gammabound}
-|\hat{\gamma}_* - \gamma_* | = |\hat{\lambda}_* - \lambda_* | \le \tnrm{\hat L - L }_2
+|\hat{\gamma}^* - \gamma^* | = |\hat{\lambda}^* - \lambda^* | \le \tnrm{\hat L - L }_2
 \end{align}
@@
  $\mu,a,\mu',a'$  such that $n = 2\mu a$ and
- $\frac{2}{\pi_*} \max( (\mu-1)e^{-a\gamma_*}, (\mu'-1)e^{-a'\gamma_*} )<\delta$, 
+ $\frac{2}{\pi^*} \max( (\mu-1)e^{-a\gamma^*}, (\mu'-1)e^{-a'\gamma^*} )<\delta$, 
 with probability $1-\delta$,
 \begin{align*}
-|\hat{\gamma}_* - \gamma_* | &\le u( \gamma_*, \pi_*, \hat{\gamma}, \hat\pi, \hat T, n, \mu, a, \mu', a', \delta )\,,\\
-|\hat{\pi}_* - \pi_* | &\le v( \gamma_*, \pi_*, \hat{\gamma}, \hat\pi, \hat T, n, \mu, a, \mu', a', \delta )\,,
+|\hat{\gamma}^* - \gamma^* | &\le u( \gamma^*, \pi^*, \hat{\gamma}, \hat\pi, \hat T, n, \mu, a, \mu', a', \delta )\,,\\
+|\hat{\pi}^* - \pi^* | &\le v( \gamma^*, \pi^*, \hat{\gamma}, \hat\pi, \hat T, n, \mu, a, \mu', a', \delta )\,,
 \end{align*}
@@
 \begin{align*}
-\gamma_* &\ge \hat\gamma_* - u( \gamma_*, \pi_*, \hat{\gamma}, \hat\pi, \hat T, n, \mu, a, \mu', a', \delta )\,,\\
- \pi_*  &\ge \hat{\pi}_* - v( \gamma_*, \pi_*, \hat{\gamma}, \hat\pi, \hat T, n, \mu, a, \mu', a', \delta )\,.
+\gamma^* &\ge \hat\gamma^* - u( \gamma^*, \pi^*, \hat{\gamma}, \hat\pi, \hat T, n, \mu, a, \mu', a', \delta )\,,\\
+ \pi^*  &\ge \hat{\pi}^* - v( \gamma^*, \pi^*, \hat{\gamma}, \hat\pi, \hat T, n, \mu, a, \mu', a', \delta )\,.
 \end{align*}
@@
 \lefteqn{\max B_n(g,p,\delta)\qquad \text{s.t.} }\\
-&\qquad  g \ge \hat\gamma_* - u( g, p, \hat{\gamma}, \hat\pi, \hat T, n, \mu, a, \mu', a', \delta )\,,\\
-&\qquad p  \ge \hat{\pi}_* - v( g, p, \hat{\gamma}, \hat\pi, \hat T, n, \mu, a, \mu', a', \delta )\,,\\
+&\qquad  g \ge \hat\gamma^* - u( g, p, \hat{\gamma}, \hat\pi, \hat T, n, \mu, a, \mu', a', \delta )\,,\\
+&\qquad p  \ge \hat{\pi}^* - v( g, p, \hat{\gamma}, \hat\pi, \hat T, n, \mu, a, \mu', a', \delta )\,,\\
 &\qquad n = 2\mu a, \quad n = 2\mu' a'\,,\\

### 2015-06-01 setting.tex
Copied files from svn

%!TEX root =  matrix-est.tex
\label{sec:setting}
Let $\vP \in (\Delta^{d-1})^d \subset [0,1]^{d \times d}$ be a $d
\times d$ row-stochastic matrix for an ergodic (i.e., irreducible and
aperiodic) Markov chain.
This implies there is a unique stationary distribution $\vpi \in
\Delta^{d-1}$ with $\pi_i > 0$ for all $i \in [d]$~\citep[Corollary
1.17]{LePeWi08}.
We also assume that $\vP$ is \emph{reversible} (with respect to
$\vpi$):
\begin{align}
\label{eq:reversibility}
  \pi_i P_{i,j} = \pi_j P_{j,i} ,
  \quad i,j \in [d] .
\end{align}
The minimum stationary probability is denoted by $\pimin := \min_{i
\in [d]} \pi_i$.

Define the matrices
\begin{align*}
\vM := \Diag(\vpi) \vP \quad \text{and} \quad
\vL := \Diag(\vpi)^{-1/2} \vM \Diag(\vpi)^{-1/2}\,.
\end{align*}
The $(i,j)$th entry of the matrix $M_{i,j}$ contains the \emph{doublet
probabilities} associated with $\vP$: $M_{i,j} = \pi_i P_{i,j}$ is the
probability of seeing state $i$ followed by state $j$ when the chain
is started from its stationary distribution.
The matrix $\vM$ is symmetric on account of the reversibility of
$\vP$, and hence it follows that $\vL$ is also symmetric.
Further, $\vL = \Diag(\vpi)^{1/2} \vP \Diag(\vpi)^{-1/2}$, hence $\vL$
and $\vP$ are similar and thus their eigenvalue systems are identical.
Ergodicity and reversibility imply that the eigenvalues of $\vL$ are
contained in the interval $\opencloseint{-1,1}$, and that $1$ is an
eigenvalue of $\vL$ with multiplicity $1$~\citep[Lemmas 12.1 and
12.2]{LePeWi08}.
Denote and order the eigenvalues of $\vL$ as
\[
  1 = \lambda_1 > \lambda_2 \geq \dotsb \geq \lambda_d > -1 .
\]
Let $\slem := \max\{ \lambda_2,\, |\lambda_d| \}$, and define the
(absolute) spectral gap to be $\gap := 1-\slem$, which is strictly positive.
%Throughout the paper, both $\pimin$ and $\gap$ are assumed
%to be positive.

Let $(X_t)_{t\in\bbN}$ be a Markov chain whose transitions are governed by
$\vP$.
For each $t \in \bbN$, let $\vpi^{(t)} \in \Delta^{d-1}$ denote the
marginal distribution of $X_t$, so
\[
  \vpi^{(t+1)} = \vpi^{(t)} \vP ,
  \quad t \in \bbN .
\]
Note that the initial distribution $\vpi^{(1)}$ is arbitrary,
and need not be the stationary distribution $\vpi$.

The goal is to estimate $\pimin$ and $\gap$ from the length $n$ sample
path $(X_t)_{t \in [n]}$, and also to construct fully empirical
confidence intervals that $\pimin$ and $\gap$ with high probability;
in particular, the construction of the intervals should not depend on
any unobservable quantities, including $\pimin$ and $\gap$ themselves.
It is well-known that the \emph{mixing time} of the Markov chain $\tmix$ (defined in the introduction)
\hide{,
\[
  \tmix
  :=
  \min\Braces{
    t \in \bbN :
    \sup_{\vq \in \Delta^{d-1}}
    \Norm{
      \vq \vP^t - \vpi
    }_{\tv} \leq 1/4
  }
  ,
\]}
can be bounded in terms of $\pimin$ and $\gap$~\citep[Theorems~12.3
and~12.4]{LePeWi08}:
\[
  \Parens{\frac1{\gap} - 1} \ln2
  \ \leq \ \tmix
  \ \leq \ \frac1{\gap} \log \frac4{\pimin}
  .
\]
Moreover, convergence rates for empirical processes on Markov chain
sequences are also often given in terms of mixing coefficients that
can ultimately be bounded in terms of $\pimin$ and
$\gap$~\citep[e.g.,][]{MR1921877,MoRo08,MoRo09}. \todoc{Cite one? Btw, do these really have these results?}
Therefore, valid confidence intervals for $\pimin$ and $\gap$ can be
used to make these rates fully observable.

%Therefore, confidence intervals for $\pimin$ and $\gap$ can be used to
%make these rates data-dependent and hence usable in procedures such
%as Structural Risk Minimization.
%\todoc{citation for SRM, and other applications?}
%
%Additionally, we aim to provide
%asymptotic convergence rates
%as well as data-dependent confidence intervals;
%the latter should not depend on any unknown quantities
%(such as $\pimin$ and $\gap$).
%
%\todoc[inline]{What is the data? What is the problem?}

### 2015-06-01 upperbound.tex
Copied files from svn

%!TEX root =  matrix-est.tex

In this section, we prove Theorem~\ref{thm:err}.
We start by proving the deviation bound on $\pimin-\hatpimin$.

%\subsection{Accuracy of singleton frequency estimates and bounding $\pimin-\hatpimin$}
%\label{sec:singleton}

%{\bf Aryeh: I'm going to do this without blocking}
%First, 
Theorem~3.8 of \citet{paulin15} implies
\[
\bbP_{\vpi}
\Braces{
\abs{ \hat\pi_i-\pi_i} > \eps
} \le
2\exp\Parens{-\frac{
    n\gap\eps^2
  }{
4\pi_i(1-\pi_i)+10\eps
}},
\qquad i\in[d],
\]
where $\bbP_{\vpi}$ denotes the probability with respect to the stationary chain.
Moreover, by Proposition~3.14 of \citet{paulin15}, we have
\begin{eqnarray*}
\bbP
\Braces{
\abs{ \hat\pi_i-\pi_i} > \eps
}
&\le&
\sqrt{
%\Parens{\sum_{i\in[d]}}
\frac1\pimin
\bbP_{\vpi}
\Braces{
\abs{ \hat\pi_i-\pi_i} > \eps
}} 
\le
\sqrt{\frac2\pimin}\exp\Parens{-\frac{
    n\gap\eps^2
  }{
8\pi_i(1-\pi_i)+20\eps
}}.
\end{eqnarray*}

% \ln\Parens{\frac{d}\delta\sqrt{\frac{2}{\pimin}}}

Hence, with probability $1-\delta$, we have for all $i\in [d]$,
%\todoc{Is it possible to have $\pi_i$ here instead of $\pimin$ in the RHS? Probably not..}
\begin{align}
\label{eq:singletonboundsimple}
\Abs{\hat{\pi}_i-\pi_i}
\le
\sqrt{ 8 \pi_i(1-\pi_i) 
\,\frac{
\ln\Parens{\frac{d}\delta\sqrt{\frac{2}{\pimin}}}
  }{\gap n}}
+
\frac{20
  \ln\Parens{\frac{d}\delta\sqrt{\frac{2}{\pimin}}}
}{\gap n}
%\frac1{(n-1)\gap\sqrt\pimin}
%+
%\sqrt{\frac2{\pimin\gap^2n}\ln(2d/\delta)}
\le
23 \sqrt{  \pi_i 
	\frac{\ln\Parens{\frac{d}\delta\sqrt{\frac{2}{\pimin}}}
           }{\gap n}
       }\,,
\end{align}
where the second inequality holds as soon as 
\[
n \ge
	\frac{\ln\Parens{\frac{d}\delta\sqrt{\frac{2}{\pimin}}}
           }{\gap \pimin}\,.
\]
Consider now the following lemma:
\begin{lemma}
Let  $\pi_i,\wh\pi_i$, $i\in [2]$, $C>0$ be
such that $|\pi_i - \wh\pi_i| \le C \sqrt{\pi_i}$, $i\in [2]$.
Assume that $\pi_1\le \pi_2$.
Then, if $C\le 2\sqrt{\pi_1}$ then $\Abs{\pi_1 - \min(\wh\pi_1,\wh\pi_2) } \le C \sqrt{\pi_1}$.
\end{lemma}
\begin{proof}
Let $\hatpimin = \min(\wh\pi_1,\wh\pi_2)$.
We have
\begin{align*}
\hatpimin \le \hat{\pi}_1 \le  \pi_1 + C \sqrt{\pi_1}
\end{align*}
so it remains to show that
\begin{align*}
\pi_1 - C \sqrt{\pi_1} \le \hatpimin\,.
\end{align*}
Note that $\hat\pi_2 \ge \pi_2- C \sqrt{\pi_2 } $. Now,
our condition on $C$ implies that $C \le 2 \sqrt{\pi_1} = \sqrt{\pi_1}+\sqrt{\pi_2} = \frac{\pi_2 - \pi_1}{\sqrt{\pi_2}-\sqrt{\pi_1}}$. 
\begin{align*}
\pi_2 - C \sqrt{\pi_2  } \ge \pi_1 - C \sqrt{\pi_1 }
\end{align*}
holds if and only if
%\begin{align*}
%\pi_i - \pi_1 \ge C (\sqrt{\pi_i} - \sqrt{\pi_1}) \sqrt{\eps}\,,
%\end{align*}
%which in turn is equivalent to 
\begin{align*}
\sqrt{\pi_2} + \sqrt{\pi_1} = \frac{\pi_2 - \pi_1}{\sqrt{\pi_2} -\sqrt{\pi_1}} \ge C \,,
\end{align*}
which is implied by our condition on $\eps$. Thus, under our condition $\hat\pi_2 \ge \pi_2 - C  \sqrt{\pi_2 } \ge \pi_1 - C \sqrt{\pi_1}$, finishing the proof.
\end{proof}
By this lemma, as soon as $n\ge (11.5)^2 \frac{\ln( \frac{d}{\delta} )+\frac12\ln(\frac2\pimin) }{ \gap\pimin}$,
then
\[
\Abs{\pimin - \hatpimin} \le 23 \sqrt{ \pimin \frac{\ln( \frac{d}{\delta} )+\frac12\ln(\frac2\pimin) }{ \gap n} }\,,
\]
For $n$ smaller than the above critical value, the second term in \eqref{eq:piminbound} is lower bounded by one.
Noting that $\Abs{\pimin-\hatpimin}\le 1$ thus shows that \eqref{eq:piminbound} holds for any $n$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Let us now turn to proving \eqref{eq:gapbound}, i.e., 
the bound on the error of the spectral gap estimate $\hatgap$.
The accuracy of $\hatgap$ is based on the accuracy of $\Sym(\wh\vL)$
in approximating $\vL$ via Weyl's inequality:
\begin{align*}%\label{eq:Weyl}
  |\hat\lambda_i - \lambda_i|
  \leq \norm{\Sym(\wh\vL) - \vL}
  \quad \text{for all }i \in [d] .
\end{align*}
Moreover, the triangle inequality implies that symmetrizing $\wh\vL$
can only help:
\begin{align*}
%\label{eq:symm}
  \norm{\Sym(\wh\vL) - \vL} \leq \norm{\wh\vL - \vL} .
\end{align*}
Therefore, the bound on the spectral gap in \cref{thm:err} is readily obtained from the
following lemma which bounds $\norm{\wh\vL - \vL}$ as well as $\abs{
\hat\pi_i/\pi_i - 1 }$ for each $i \in [d]$ with high probability.
\begin{lemma}
  \label{lem:err}
  There exists an absolute constant $C>0$ such that the following
  holds.
  For any $\delta \in (0,1)$, with probability at least $1-\delta$,
  \begin{equation*}
    \norm{\wh\vL - \vL}
    \leq
    \frac{C}{\pimin}
      \sqrt{\frac{\ln(d/\delta)\ln\frac{n}{\delta\pimin}}{n\pimin\gap}} \,.
  \end{equation*}
	 \todoc{Note that the bound that was here essentially got multiplied by $1/\pimin$!}
\end{lemma}
The remainder of this section is devoted to proving this \lcnamecref{lem:err}.

The error $\wh\vL - \vL$ may be written as
\[
  \wh\vL - \vL
  = \errm + \errp \vL + \vL \errp + \errp \vL \errp
  + \errp \errm + \errm \errp + \errp \errm \errp\,,
\]
where
\begin{align*}
  \errp & := \Diag(\hat\vpi)^{-1/2} \Diag(\vpi)^{1/2} - \vI , \\
  \errm & := \Diag(\vpi)^{-1/2} \Parens{
    \wh\vM - \vM
  } \Diag(\vpi)^{-1/2} .
\end{align*}
Therefore
\begin{align*}%\label{eq:vlbound}
  \norm{\wh\vL - \vL}
  \leq \norm{\errm} +
  \Parens{
    \norm{\errm} + \norm{\vL}
  }
  \Parens{
    2\norm{\errp} + \norm{\errp}^2
  }
  .
\end{align*}
When  $\norm{\errm}\le 1$ and $\norm{\errp}\le 1$ both hold, the previous inequality implies that
\begin{align}\label{eq:vlsimple}
\norm{\wh\vL - \vL} \le C(\norm{\errm}+\norm{\errp})\,.
\end{align}
Here and in what follows we use $C>0$ to denote some universal constant, whose value does not depend on the problem parameters, but may change from line to line.
Since in addition to $\norm{\vL}\le 1$,  $\norm{\wh\vL}\le 1$ also holds (this again follows from the Perron-Frobenius theorem), in fact  \eqref{eq:vlsimple}  always holds as long as $C\ge 2$.

\subsection{A bound on $\norm{\errp}$}
We have
\begin{align*}
  \norm{\errp}
  = \max_{i\in[d]} \Abs{
    \sqrt{\frac{\pi_i}{\hat\pi_i}} - 1\,.
  }
\end{align*}
%The purpose of this section is to develop a bound on the RHS of \eqref{eq:singleton} so as to bound $\norm{\errp}$.
First note that for $x$ real, $\Abs{x-1}\le 1/2$ implies that $\max(\Abs{x-1},\Abs{1/x-1}) \le 2 \Abs{x-1}$.
Hence, it suffices to develop a bound on $ \Abs{ \frac{\hat\pi_i}{\pi_i} - 1 }$;
the final bound will follow under the condition that the developed bound is below $1/2$.
The required bound is readily obtained from~\eqref{eq:singletonboundsimple}:
As long as $n = \Omega \Parens{
	\frac{\ln\Parens{\frac{d}\delta\sqrt{\frac{2}{\pimin}}}
           }{\gap \pimin} }$, \todoc{First condition on $n$.}
$
\Abs{\frac{\hat{\pi}_i}{\pi_i}-1}
\le
23 \sqrt{  
	\frac{\ln\Parens{\frac{d}\delta\sqrt{\frac{2}{\pimin}}}
           }{\pimin\gap n}
       }\,,
$
and thus
\begin{align}
\label{eq:singletonbound}
\norm{\errp} \le C \sqrt{ \frac{\ln\Parens{\frac{d}\delta} + \ln \frac{1}{\pimin} }{\pimin\gap n}} \,.
\end{align}
We now turn to bounding  $\norm{\errm}$.
%Note that although the entries of $\wh{\vM}$ and $\hat\vpi$ are simply
%sample averages, they are formed using a \emph{single sample path}
%from a (possibly) non-stationary Markov chain.
%Applying Chernoff-Hoeffding bounds for Markov chains to each entry of
%$\wh{\vM}$ would result in a
%significantly worse sample complexity bound.
%%vastly worse sequence length requirement.
%Instead, we adapt probability tail bounds for sums of independent
%random matrices to our setting.

\subsection{Accuracy of doublet frequency estimates (bounding $\norm{\errm}$)}
\label{sec:pairwise}
In this section we prove a bound on $\norm{\errm}$.
For this, we decompose
$\errm = \Diag(\vpi)^{-1/2}(\wh\vM-\vM)\Diag(\vpi)^{-1/2} $
into $\EE{\errm}$ and 
$
\errm-\EE{\errm}% =  \Diag(\vpi)^{-1/2}(\wh\vM - \EE{\wh\vM}) \Diag(\vpi)^{-1/2} ,
$, the first measuring the effect of a non-stationary start of the chain,
while the second measuring the variation due to randomness.
\if0
into the sum of
\[
  \Diag(\vpi)^{-1/2}
  \left(\bbE( \wh\vM )- \vM\right)
  \Diag(\vpi)^{-1/2}
\]
and
\begin{multline*}
  \Diag(\vpi)^{-1/2} \left(\wh\vM- \bbE(\wh\vM)\right) \Diag(\vpi)^{-1/2}
  \\
  =
  \frac{n_H}{n-1}
  \Diag(\vpi)^{-1/2}\Parens{
    \wh{\vM}_H - \bbE(\wh{\vM}_H)
  }\Diag(\vpi)^{-1/2}
  \\
  + \frac{n_T}{n-1}
  \Diag(\vpi)^{-1/2}\Parens{
    \wh{\vM}_T - \bbE(\wh{\vM}_T)
  }\Diag(\vpi)^{-1/2}
  .
\end{multline*}
\fi
\subsubsection{Bounding $\norm{\EE{\errm}}$: The price of a non-stationary start.}
Let $\vpi^{(t)}$ be the distribution of states at time step $t$.
We will make use of the following proposition, which can be derived by following 
Proposition 1.12 of \citet{MoTe06}:
\begin{proposition}
For $t\ge 1$, let $\vUpsilon^{(t)}$ be the vector with $\Upsilon^{(t)}_i= \frac{\pi_i^{(t)}}{\pi_i}$ and let
$\norm{\cdot}_{2,\vpi}$ denote the $\vpi$-weighted $2$-norm
\begin{align}
\label{eq:vv}
  \norm{\vv}_{2,\vpi} := \Parens{ \sum_{i=1}^d \pi_i v_i^2 }^{1/2}
  .
\end{align}
%\todoc{Move up to notation?}
 Then,
\begin{equation}
  \label{eq:L2-contraction}
  \norm{ \vUpsilon^{(t)} - \v1 }_{2,\vpi} \leq
  \frac{(1-\gap)^{t-1}}{\sqrt{\pimin}} \,.
\end{equation}
\end{proposition}
%Corollary 1.15:
An immediate corollary of this result is that 
\begin{align}
\label{eq:dvpi-ratio-bound}
\Norm{  \Dvpit \Dvpi^{-1} - \vI  } \le \frac{(1-\gap)^{t-1}}{\sqrt{\pimin}}\,.
\end{align}
Now note that
\begin{align*}
\bbE (\wh\vM) = \frac{1}{n-1} \sum_{t=1}^{n-1} \Dvpit \vP\,
\end{align*}
and thus
\begin{align*}
\EE{\errm} &=
  \Dvpi^{-1/2}
  \left(\bbE( \wh\vM )- \vM\right)
  \Dvpi^{-1/2}\\
& =  
\frac{1}{n-1} \sum_{t=1}^n
  \Dvpi^{-1/2} (\Dvpit  - \Dvpi) \vP \Dvpi^{-1/2} \\
& =  
\frac{1}{n-1} \sum_{t=1}^n
  \Dvpi^{-1/2} (\Dvpit \Dvpi^{-1} - \vI) \vM \Dvpi^{-1/2} \\
& =  
\frac{1}{n-1} \sum_{t=1}^n
   (\Dvpit \Dvpi^{-1} - \vI) \vL \,.
\end{align*}
Combining this, $\norm{\vL}\le 1$ and~\eqref{eq:dvpi-ratio-bound}, we get
\begin{align}
\label{eq:biasbound}
\norm{\bbE(\errm)} \le \frac{1}{(n-1)\sqrt{\pimin}}\, \sum_{t=1}^n (1-\gap)^{t-1} \le  \frac{1}{(n-1)\gap\sqrt{\pimin}}\, .
\end{align}
\if0
There is a bias in the estimates that arises when the initial state
distribution is not the stationary distribution.
The first block $F$ is most subject to non-stationarity effects, while
the subsequent blocks $H_s$ and $T_s$ are much less affected as long
as the block lengths are long enough.
Fortunately, the impact of the first block is small when there are
enough blocks.
Choosing an appropriate block length will balance these effects.
(The block length also plays a role in the dependence between blocks,
but this is dealt with later.)

Define
\begin{align*}
  \vpi^{(H_s)} & := \frac1{|H_s|} \sum_{t \in H_s} \vpi^{(t)} ,
  & \vpi^{(T_s)} & := \frac1{|T_s|} \sum_{t \in T_s} \vpi^{(t)} , \\
  \vpi^{(H)} & := \sum_{s=1}^{\mu_H} w_{H,s} \vpi^{(H_s)} ,
  & \vpi^{(T)} & := \sum_{s=1}^{\mu_T} w_{T,s} \vpi^{(T_s)} ,
\end{align*}
and
\begin{align*}
  \vpi^{(H,T)} & :=
  \frac{n_H}{n_H + n_T} \vpi^{(H)}
  + \frac{n_T}{n_H + n_T} \vpi^{(T)}
  .
\end{align*}
Then
\begin{multline*}
  \Diag(\vpi)^{-1/2}
  \bbE\Parens{ \wh\vM \,\Big\vert\, \wh{\vM}_F }
  \Diag(\vpi)^{-1/2} - \vL
  \\
  =
  \frac{|F|}{n-1}\Parens{
    \Diag(\vpi)^{-1/2} \wh{\vM}_F \Diag(\vpi)^{-1/2} - \vL
  }
  \\
  + \Parens{ 1 - \frac{|F|}{n-1} }
  \Parens{
    \Diag(\vpi)^{-1/2} \Diag(\vpi^{(H,T)}) \Diag(\vpi)^{-1/2} - \vI
  } \vL
  .
\end{multline*}
Since $\norm{\wh{\vM}_F} \leq 1$ and $\norm{\vL} \leq 1$, it follows
that
\begin{multline*}
  \Norm{
    \Diag(\vpi)^{-1/2}
    \bbE\Parens{ \wh\vM \,\Big\vert\, \wh{\vM}_F }
    \Diag(\vpi)^{-1/2} - \vL
  }
  \\
  \leq
  \frac{|F|}{n-1}
  \Parens{ \frac1{\pimin} + 1 }
  +
  \Parens{ 1 - \frac{|F|}{n-1} }
  \max_{i \in [d]} \Abs{ \frac{\pi_i^{(H,T)}}{\pi_i} - 1 }
  .
\end{multline*}

For any block $H_s$,
\begin{align*}
  \Norm{ \vUpsilon^{(H_s)} - \v1 }_{2,\vpi}
  & =
  \Norm{
    \frac1{|H_s|} \sum_{t \in H_s} \vUpsilon^{(t)} - \v1
  }_{2,\vpi}
  \\
  & \leq
  \frac1{|H_s|} \sum_{t \in H_s}
  \Norm{
    \vUpsilon^{(t)} - \v1
  }_{2,\vpi}
  \\
  & \leq \frac1{|H_s|} \cdot \frac{(1-\gap)^{\min H_s -
  1}}{\sqrt{\pimin}\gap}
  .
\end{align*}

Therefore, for each $i \in [d]$,
\begin{align*}
  \Abs{ \frac{\pi_i^{(H_s)}}{\pi_i} - 1 }
  & \leq
  \frac1{\sqrt{\pi_i}}
  \Norm{ \vUpsilon^{(H_s)} - \v1 }_{2,\vpi}
  \\
  & \leq
  \frac1{|H_s|}
  \frac{(1-\gap)^{\min H_s - 1}}{\sqrt{\pi_i\pimin}\gap}
  .
\end{align*}
(The same, of course, applies to $T_s$ blocks.)
By the same argument, for all $i \in [d]$,
\begin{align*}
  \Abs{ \frac{\pi_i^{(H,T)}}{\pi_i} - 1 }
  & \leq \frac1{n-1} \cdot \frac{(1-\gap)^{\min H_1 -
  1}}{\sqrt{\pi_i\pimin}\gap}
  .
\end{align*}

Suppose the first block $F$ is of length
\begin{equation}
  \label{eq:block-length}
  |F| = a \geq 1 + \frac{1}{\gap} \ln \frac1{\pimin\gap}
  .
\end{equation}
Then
\[
  \frac{(1-\gap)^{\min H_s - 1}}{\pimin\gap}
  \leq 1
\]
for all blocks $H_s$ (and same for the $T_s$ blocks), so
\begin{align}
  \max \Braces{
    \Abs{ \frac{\pi_i^{(H_s)}}{\pi_i} } ,\,
    \Abs{ \frac{\pi_i^{(T_s)}}{\pi_i} } ,\,
    \Abs{ \frac{\pi_i^{(H,T)}}{\pi_i} }
    : i \in [d] , s
  }
  & \leq 2
  .
  \label{eq:near-stationary}
\end{align}
This means that as long as the blocks are sufficiently long, then all
but the first block $F$ can effectively be thought of as being started
from the stationary distribution.
We shall henceforth assume that~\eqref{eq:block-length} holds.
\fi
\subsubsection{
Bounding $\norm{\errm-\EE{\errm}}$: 
Application of a matrix tail inequality}
\label{sec:pairwise-tailbound}

In this section we analyze the deviations of $\errm-\EE{\errm}$. By the definition of $\errm$,
\begin{align}
\label{eq:errmdev1}
\norm{\errm-\EE{\errm}}
 =  \norm{\Diag(\vpi)^{-1/2}\Parens{\wh\vM - \bbE \wh\vM } \Diag(\vpi)^{-1/2}}
 \le \frac{1}{\pimin} \norm{\wh\vM - \bbE \wh\vM }
 \,.
\end{align}
The matrix $\wh\vM - \EE{\wh\vM}$  is defined 
as a sum of dependent centered random matrices.
We will use the blocking technique of  \citet{Bernstein27} 
to relate the likely deviations of this matrix to that of
a sum of independent centered random matrices.
The deviations of these will then bounded
with the help of a Bernstein-type matrix tail inequality due to \citet{tropp2015intro}.

%We divide $(X_t)_{t\in[n]}$ into $2\mu$ blocks each of length
%$a\geq2$.
%Define
%\begin{align*}
%  H_j & := \{ t \in \bbN : 2(j-1)a + 1 \leq t \leq (2j-1)a \}
%  ,
%  \\
%  T_j & := \{ t \in \bbN : (2j-1)a + 1 \leq t \leq 2ja \}
%  ,
%  \qquad j \in [\mu] .
%\end{align*}
%Now define the corresponding blocks of random variables:
%\begin{equation*}
%  X(H_j) := ( X_t : t \in H_j )
%  , \quad
%  X(T_j) := ( X_t : t \in T_j )
%  ,
%  \qquad j \in [\mu] .
%\end{equation*}
%Let $\Xi(H_j)$ for $j \in [\mu]$ be independent copies of $X(H_j)$ for
%$j \in [\mu]$, and similarly define $\Xi(T_j)$ for $j \in [\mu]$.
%Let the distribution of $X(H_j)$ be denoted by $\cP_j$, and let the
%distribution of $X(T_j)$ be denoted by $\cQ_j$.
%Let the distributions of $(X(H_j) : j \in [\mu])$ and $(\Xi(H_j) : j
%\in [\mu])$ be denoted by $\cP$ and $\tilde{\cP} = \prod_{j=1}^{\mu}
%\cP_j$, respectively.
%Similarly define $\cQ$ and $\tilde{\cQ} = \prod_{j=1}^{\mu} \cQ_j$.
%Then for any $\veps>0$,
%\begin{align*}
%\end{align*}

Divide $[n-1]$ into several blocks each of length $a \leq n/2$ (except \todoc{Needed $a\le n/2$}
possibly the final block):
\begin{align*}
  F & := [a] , \\
  H_s & := \{ t \in [n-1] : (2s-1)a + 1 \leq t \leq 2sa \} , \\
  T_s & := \{ t \in [n-1] : 2sa + 1 \leq t \leq (2s+1) a \} ,
\end{align*}
where $s=1,2,\dots$.
Let $\mu_H$ (resp., $\mu_T$) be the number of non-empty $H_s$ (resp., $T_s$)
blocks.
Let $n_H := \sum_{s=1}^{\mu_H} |H_s|$ (resp., $n_T := \sum_{s=1}^{\mu_T}|T_s|$)
be the number of time steps in $\cup_s H_s$ (resp., in $\cup_s T_s$) and let
 $w_{H,s} := |H_s|/n_H$ (resp., $w_{T,s} := |T_s|/n_T$) be the relative weight of data points in block $H_s$
 (resp., in block $T_s$).

We have
\begin{align*}
  \wh\vM
  & = \frac1{n-1} \sum_{t=1}^{n-1} \ve_{X_t} \ve_{X_{t+1}}^\t \\
  & = \frac{|F|}{n-1} 
    \underbrace{
      \frac{1}{|F|} \sum_{t\in F} \ve_{X_t} \ve_{X_{t+1}}^\t
     }_{\wh\vM_F} +
  \frac{n_H}{n-1}
  \underbrace{
    \sum_{s=1}^{\mu_H} w_{H,s}
    \Parens{
      \frac1{|H_s|} \sum_{t \in H_s} \ve_{X_t} \ve_{X_{t+1}}^\t
    }
  }_{\wh{\vM}_H}\\
  & \qquad
  + \frac{n_T}{n-1}
  \underbrace{
    \sum_{s=1}^{\mu_T} w_{T,s}
    \Parens{
      \frac1{|T_s|} \sum_{t \in T_s} \ve_{X_t} \ve_{X_{t+1}}^\t
    }
  }_{\wh{\vM}_T}
  .
\end{align*}

Note that $\norm{\wh{\vM}_F},\norm{\EE{\vM}}\le 1$, so the first term's effect on the deviation of $\wh \vM$ from $\vM$ is bounded by $2/(n-1)$.
We now present a bound on $\wh{\vM}_H - \bbE(\wh{\vM}_H)$; the bound on (and analysis of)
 $\wh{\vM}_T - \bbE(\wh{\vM}_T)$ is identical.

Let
\[
  \vY_s := \frac1{|H_s|} \sum_{t \in H_s} \ve_{X_t} \ve_{X_{t+1}}^\t ,
  \quad s \in [\mu_H] ,
\]
so
\[
  \wh{\vM}_H = \sum_{s=1}^{\mu_H} w_{H,s} \vY_s ,
\]
a weighted average of the random matrices $\vY_s$.
For each $s \in [\mu_H]$, the random matrix $\vY_s$ is a function of
\[ (X_t : (2s-1)a + 1 \leq t \leq 2sa + 1) \]
(note the $+1$ in the upper limit of $t$).
So $\vY_s$ and $\vY_{s+1}$ are separated by $a-1$ time steps.
When $a$ is sufficiently large, we will be able to effectively treat
the random matrices $\vY_s$ as if they were independent.
In the sequel, we shall always assume that the block length $a$
satisfies
\begin{equation}
  \label{eq:block-length}
  a \geq
   1 + \frac{1}{2\gap} \ln \frac1{\pimin}\,.
%   \geq
%   1 + \frac{1}{\gap} \ln \frac1{|H_s|\sqrt{\pimin}\gap}
\end{equation}
\todoc{By using the geometric series bound,
we can derive that it is enough
if $a\ge 1 + \frac{1}{\gap} \ln \frac1{|H_s|\sqrt{\pimin}\gap}$,
which is better when $|H_s|$ is bigger than $\gap$.
}
\todoc{As we need $a\le n/2$, we will need $n = \Omega(1/\gap\ln\frac1{\pimin})$.}

Define%\todoc{Do we want this abuse?}
\begin{align*}
  \vpi^{(H_s)}  := \frac1{|H_s|} \sum_{t \in H_s} \vpi^{(t)} , \qquad \qquad
  \vpi^{(H)}  := \sum_{s=1}^{\mu_H} w_{H,s} \vpi^{(H_s)} .
\end{align*}
Observe that
\[
  \bbE(\vY_s)
  = \Diag(\vpi^{(H_s)}) \vP
\]
so
\[
  \bbE\Parens{
    \sum_{s=1}^{\mu_H} w_{H,s} \vY_s
  } = \Diag(\vpi^{(H)}) \vP
  .
\]

Define
\[
  \vZ_s
  := \Diag(\vpi)^{-1/2} \left( \vY_s 
  - \bbE(\vY_s) \right)\Diag(\vpi)^{-1/2}
  .
\]
We apply a matrix tail inequality to the
$w_{H,s}$-weighted sum of \emph{independent} copies of the $\vZ_s$'s.
More precisely, we will apply the tail inequality to independent copies $\wt{\vZ}_s$, $s\in \iset{\mu_H}$ 
of the random variables $\vZ_s$ and then relate the weighted sum of $\wt{\vZ}_s$ to that of $\vZ_s$.
The matrix tail inequality we use is as follows:
\begin{theorem}[Matrix Bernstein inequality, Theorem~6.1.1 of \citet{tropp2015intro}]
\label{thm:mxbernstein}
Let $\vX_1,\dots,\vX_n$ be a sequence of independent, random matrices with common dimension $d_1\times d_2$.
Assume that $\EE{\vX_i} = 0$ and $\Norm{\vX_i}\le R$ for each $1\le i \le n$. Let $\vS = \sum_{i=1}^n \vX_i$ and let 
\begin{align*}
v = \max\left\{ \norm{\bbE \textstyle\sum_i \vX_i \vX_i^\top  }, 
						      \norm{\bbE \textstyle\sum_i \vX_i^\top \vX_i  }
			\right\}\,.
\end{align*}
Then, for all $t\ge 0$, 
\[
\bbP\left( \Norm{\vS} \ge t \right) \le (d_1 + d_2) \exp\left( -\frac{t^2/2}{v+R t/3}\right)\,.
\]
\end{theorem}
By inverting the bound, we get that for any $0\le \delta \le 1$, with probability $1-\delta$, 
\[
\norm{\vS} \le \sqrt{2 v \ln \frac{d_1+d_2}{\delta}} + \frac{2R}{3} \ln\frac{d_1+d_2}{\delta}\,.
\]

We see, that in order to apply this result we need to bound the range of $\vZ_s$ and 
the spectral norms of the matrices $\sum_{s=1}^{\mu_H}
w_{H,s}^2 \bbE(\vZ_s \vZ_s^\t)$ and $\sum_{s=1}^{\mu_H} w_{H,s}^2
\bbE(\vZ_s^\t \vZ_s)$ (which we call variance bounds).

\paragraph{Range bound.}
%We first determine a bound on the spectral norm of $\vZ_s$.
By the triangle inequality,
\[
  \norm{\vZ_s}
  \leq \norm{\Diag(\vpi)^{-1/2} \vY_s \Diag(\vpi)^{-1/2}}
  + \norm{\Diag(\vpi)^{-1/2} \bbE(\vY_s) \Diag(\vpi)^{-1/2}}
  \,.
\]
Clearly,
\begin{align}
\label{eq:vysbound}
  \norm{\Diag(\vpi)^{-1/2} \vY_s \Diag(\vpi)^{-1/2}} \leq \frac1{\pimin} .
\end{align}
Similarly to \eqref{eq:biasbound}, using $\norm{\vL}\le 1$,
we get 
\begin{align*}
\MoveEqLeft  \norm{\Diag(\vpi)^{-1/2} (\bbE(\vY_s) -\vM)\Diag(\vpi)^{-1/2}}
= 
\norm{ \bigl(\Diag(\vpi^{(H_s)}) \Diag(\vpi)^{-1} - \vI  \bigr) \vL } \nonumber \\
&\qquad\qquad\qquad \le
\norm{ \Diag(\vpi^{(H_s)}) \Diag(\vpi)^{-1} - \vI }\,.
\end{align*}
Then, using~\eqref{eq:dvpi-ratio-bound},
\begin{align}
\norm{ \Diag(\vpi^{(H_s)}) \Diag(\vpi)^{-1} - \vI } \le
 \frac{|H_s|(1-\gap)^{(2s-1)a}}{|H_s|\sqrt{\pimin}}
\le
 \frac{(1-\gap)^{a}}{\sqrt{\pimin}} \le 1\,,
 \label{eq:ysbias}
\end{align}
where the last inequality follows from the assumption that the
block length $a$ satisfies~\eqref{eq:block-length}.

Combining this with
$\norm{\Diag(\vpi)^{-1/2} \vM\Diag(\vpi)^{-1/2}} = \norm{\vL}\le 1$,
%\if0
%
%Since
%\[
%  \Diag(\vpi)^{-1/2} \bbE(\vY_s) \Diag(\vpi)^{-1/2}
%  =
%  \Diag(\vpi)^{-1/2}
%  \Diag(\vpi^{(H_s)})
%  \Diag(\vpi)^{-1/2}
%  \vL
%\]
%and
%\[
%  \norm{\vL} \leq 1
%  ,
%\]
%it follows that
%\[
%  \norm{\Diag(\vpi)^{-1/2} \bbE(\vY_s) \Diag(\vpi)^{-1/2}}
%  \leq \max_{i \in [d]} \frac{\pi_i^{(H_s)}}{\pi_i}\,.
%%  \leq 2
%\]
%
%Suppose the first block $F$ is of length
%\begin{equation}
%  \label{eq:block-length}
%  |F| \geq 1 + \frac{1}{\gap} \ln \frac1{\pimin\gap}
%  .
%\end{equation}
%Then
%\[
%  \frac{(1-\gap)^{\min H_s - 1}}{\pimin\gap}
%  \leq 1
%\]
%for all blocks $H_s$ (and same for the $T_s$ blocks), so
%\begin{align}
%  \max \Braces{
%    \Abs{ \frac{\pi_i^{(H_s)}}{\pi_i} } ,\,
%    \Abs{ \frac{\pi_i^{(T_s)}}{\pi_i} } ,\,
%    \Abs{ \frac{\pi_i^{(H,T)}}{\pi_i} }
%    : i \in [d] , s
%  }
%  & \leq 2
%  .
%  \label{eq:near-stationary}
%\end{align}
%
%
%
%(using the assumption that~\eqref{eq:block-length} holds).
%Therefore
%\fi
it follows that
\begin{align}
\label{eq:evysbound}
\norm{\Diag(\vpi)^{-1/2} \bbE(\vY_s) \Diag(\vpi)^{-1/2}} \le 2
\end{align}
which, together with~\eqref{eq:vysbound}, gives
\[
  \norm{\vZ_s} \leq \frac1{\pimin} + 2
%  \max_{i\in[d]} \frac{\pi_i^{(H_s)}}{\pi_i}
  .
\]

\paragraph{Variance bound.}
We now determine bounds on the spectral norm of $\sum_{s=1}^{\mu_H}
w_{H,s}^2 \bbE(\vZ_s \vZ_s^\t)$ and of $\sum_{s=1}^{\mu_H} w_{H,s}^2
\bbE(\vZ_s^\t \vZ_s)$.
Observe that
\begin{align}
  \lefteqn{
    \bbE(\vZ_s\vZ_s^\t)
  } \notag \\
  & =
  \frac1{|H_s|^2} \sum_{t \in H_s}
  \bbE\Parens{
    \Diag(\vpi)^{-1/2}
    \ve_{X_t} \ve_{X_{t+1}}^\t 
    \Diag(\vpi)^{-1}
    \ve_{X_{t+1}} \ve_{X_t}^\t
    \Diag(\vpi)^{-1/2}
  }
  \label{eq:var1}
  \\
  & \quad
  + \frac1{|H_s|^2} \sum_{ \substack{ t \neq t' \\ t,t'\in H_s} } 
  \bbE\Parens{
    \Diag(\vpi)^{-1/2}
    \ve_{X_t} \ve_{X_{t+1}}^\t 
    \Diag(\vpi)^{-1}
    \ve_{X_{t'+1}} \ve_{X_{t'}}^\t
    \Diag(\vpi)^{-1/2}
  }
  \label{eq:var2}
  \\
  & \quad
  - \Diag(\vpi)^{-1/2}
  \bbE(\vY_s)
  \Diag(\vpi)^{-1} 
  \bbE(\vY_s^\t)
  \Diag(\vpi)^{-1/2}
  .
  \label{eq:var3}
\end{align}
The first sum~\eqref{eq:var1} easily simplifies to the diagonal matrix
\begin{align*}
  \lefteqn{
    \frac1{|H_s|^2}
    \sum_{t \in H_s}
    \sum_{i=1}^d \sum_{j=1}^d \Pr(X_t = i, X_{t+1} = j) \cdot
    \frac{1}{\pi_i\pi_j} \ve_i\ve_j^\t\ve_j\ve_i^\t
  } \\
  & =
  \frac1{|H_s|^2}
  \sum_{t \in H_s}
  \sum_{i=1}^d \sum_{j=1}^d \pi_i^{(t)} P_{i,j} \cdot
  \frac{1}{\pi_i\pi_j} \ve_i\ve_i^\t
  =
  \frac1{|H_s|}
  \sum_{i=1}^d \frac{\pi_i^{(H_s)}}{\pi_i}
  \Parens{ \sum_{j=1}^d \frac{P_{i,j}}{\pi_j} }
   \ve_i \ve_i^\t
  .
\end{align*}
For the second sum~\eqref{eq:var2},  a symmetric matrix, consider
\[
  \vu^\t
  \Parens{
    \frac1{|H_s|^2} \sum_{ \substack{ t \neq t' \\ t,t'\in H_s} } 
    \bbE\Parens{
      \Diag(\vpi)^{-1/2}
      \ve_{X_t} \ve_{X_{t+1}}^\t 
      \Diag(\vpi)^{-1}
      \ve_{X_{t'+1}} \ve_{X_{t'}}^\t
      \Diag(\vpi)^{-1/2}
    }
  } \vu
\]
for an arbitrary unit vector $\vu$.
By Cauchy-Schwarz and AM/GM, this is bounded from above by
\begin{multline*}
  \frac1{2|H_s|^2}
  \sum_{ \substack{ t \neq t' \\ t,t'\in H_s} }
  \biggl[
  \bbE\Parens{
    \vu^\t
    \Diag(\vpi)^{-1/2} \ve_{X_t} \ve_{X_{t+1}}^\t \Diag(\vpi)^{-1}
    \ve_{X_{t+1}} \ve_{X_t}^\t \Diag(\vpi)^{-1/2}
    \vu
  }
  \\
  +
  \bbE\Parens{
    \vu^\t
    \Diag(\vpi)^{-1/2} \ve_{X_{t'}} \ve_{X_{t'+1}}^\t \Diag(\vpi)^{-1}
    \ve_{X_{t'+1}} \ve_{X_{t'}}^\t \Diag(\vpi)^{-1/2}
    \vu
  }
  \biggr]
  ,
\end{multline*}
which simplifies to %the sum of
\[
  \frac{|H_s|-1}{|H_s|^2}
  \vu^\t \bbE\Parens{
    \sum_{t \in H_s}
    \Diag(\vpi)^{-1/2} \ve_{X_t} \ve_{X_{t+1}}^\t \Diag(\vpi)^{-1}
    \ve_{X_{t+1}} \ve_{X_t}^\t \Diag(\vpi)^{-1/2}
  } \vu\,.
\]
%and the same expression with $\vu$ replaced by $\vv$.
The expectation is the same as that from the first
term~\eqref{eq:var1}.

Finally, the spectral norm of the third term~\eqref{eq:var3} is bounded by
\[
  \norm{\Diag(\vpi)^{-1/2} \bbE(\vY_s) \Diag(\vpi)^{-1/2}}^2
  \leq 4
  %\Parens{ \max_{i \in [d]} \frac{\pi_i^{(H_s)}}{\pi_i} }^2
\]
where we used~\eqref{eq:evysbound}, which holds when~\eqref{eq:block-length} is satisfied.
%\comment{possibly improve this to depend only on $\pi^{(H)}$?}

Therefore, by the triangle inequality and some simplifications, 
\begin{align*}
  \Norm{
    \sum_{s=1}^{\mu_H} w_{H,s}^2 \bbE(\vZ_s \vZ_s^\t)
  }
  &\leq
  \Parens{ \max_s w_{H,s} } \left[
  \max_{i\in[d]}
  \Parens{ \sum_{j=1}^d \frac{P_{i,j}}{\pi_j} }
  \frac{\pi_i^{(H)}}{\pi_i}
  + 4 \right]
  \\
  &\leq
  \Parens{ \max_s w_{H,s} } \left[
  2\max_{i\in[d]}
  \Parens{ \sum_{j=1}^d \frac{P_{i,j}}{\pi_j} }
  + 4 \right]
\end{align*}
using \eqref{eq:ysbias}, which holds 
under the assumption that~\eqref{eq:block-length} is satisfied.

We can also bound $\sum_{s=1}^{\mu_H} w_{H,s}^2 \bbE(\vZ_s^\t \vZ_s)$
in a similar way; the only difference is that the reversibility needs
to be used at one place to simplify an expectation:
\begin{align*}
\MoveEqLeft
    \frac1{|H_s|^2} \sum_{t \in H_s}
    \bbE\Parens{
      \Diag(\vpi)^{-1/2}
      \ve_{X_{t+1}} \ve_{X_t}^\t
      \Diag(\vpi)^{-1}
      \ve_{X_t}\ve_{X_{t+1}}^\t
      \Diag(\vpi)^{-1/2}
    }
	\\
  & =
  \frac1{|H_s|^2} \sum_{t \in H_s}
  \sum_{i=1}^d \sum_{j=1}^d \Pr(X_t = i, X_{t+1} = j) \cdot
  \frac{1}{\pi_i\pi_j} \ve_j\ve_j^\t
  \\
  & =
  \frac1{|H_s|^2} \sum_{t \in H_s}
  \sum_{i=1}^d \sum_{j=1}^d \pi_i^{(t)} P_{i,j} \cdot
  \frac{1}{\pi_i\pi_j} \ve_j\ve_j^\t
  \\
  & =
  \frac1{|H_s|^2} \sum_{t \in H_s}
  \sum_{j=1}^d \Parens{
    \sum_{i=1}^d \frac{\pi_i^{(t)}}{\pi_i} \cdot \frac{P_{j,i}}{\pi_i}
  } \ve_j\ve_j^\t & (\text{by }\eqref{eq:reversibility})
  .
\end{align*}
As before, we get
\begin{align*}
  \Norm{
    \sum_{s=1}^{\mu_H} w_{H,s}^2 \bbE(\vZ_s^\t\vZ_s)
  }
  &\leq
  \Parens{ \max_s w_{H,s} } \left[
  \max_{i\in[d]}
  \Parens{
    \sum_{j=1}^d \frac{P_{i,j}}{\pi_j}
    \cdot \frac{\pi_j^{(H)}}{\pi_j}
  }
  + 4 \right]
  \\
&  \leq
  \Parens{ \max_s w_{H,s} }
  \left[ 2 \max_{i\in[d]}
  \Parens{
    \sum_{j=1}^d \frac{P_{i,j}}{\pi_j}
  }
  + 4\right]
\end{align*}
using again \eqref{eq:ysbias}.
%(using the assumption that~\eqref{eq:block-length} holds).

\paragraph{Independent copies bound.}
Let $\wt\vZ_s$ for $s \in [\mu_H]$ be independent copies of
$\vZ_s$ for $s \in [\mu_H]$.
Applying \cref{thm:mxbernstein} to the weighted sum of these random variables, we  get that 
with probability at least $1-\delta$, the spectral norm of
$\sum_{s=1}^{\mu_H} w_{H,s} \wt\vZ_s$ is at most
\begin{align*}
\MoveEqLeft  \Norm{\sum_{s=1}^{\mu_H} w_{H,s} \wt\vZ_s}
  \leq
  \sqrt{
    4\Parens{ \max_s w_{H,s} } \Parens{
      d_{\vP}
      + 2
    }
    \ln(4d/\delta)
  }
  \\
& \qquad\qquad\qquad  +
  \frac23 \max_s w_{H,s}
  \Parens{
    \frac1{\pimin} + 2
  }
  \ln(4d/\delta)
\end{align*}
where
\begin{align*}
  d_{\vP}
  & := \max_{i \in [d]} \sum_{j=1}^d \frac{P_{i,j}}{\pi_j}
  \leq
  \frac{1}{\pimin}
  \,.
\end{align*}

For simplicity, we assume that all of the blocks have length exactly $a$.
Under this assumption, 
the above bound (also with looser constants) gives that
with probability at least $1-\delta$:
\begin{equation}
\label{eq:indbound}
  \Norm{\frac1{\mu_H} \sum_{s=1}^{\mu_H} \wt\vZ_s}
  \leq
  \sqrt{
    \frac{12\ln(4d/\delta)}{\mu_H\pimin}
  }
  +
  \frac{2\ln(4d/\delta)}{\mu_H\pimin}
  \,.
\end{equation}

\paragraph{The actual bound.}
To bound the probability that $\norm{\sum_{s=1}^{\mu_H} w_{H,s}
\vZ_s}$ is large, we appeal to the following result; a consequence of
Corollary 2.7 of~\citet{Yu94}.
Let $\bbP$ denote the joint distribution of $(\vZ_s : s \in [\mu_H])$;
let $\bbP_s$ be its marginal over $\vZ_s$, and let $\bbP_{1:s+1}$ be
its marginal over $(\vZ_1,\vZ_2,\dotsc,\vZ_{s+1})$.
Let $\wt\bbP$ be the product distribution formed from the marginals
$\bbP_1,\bbP_2,\dotsc,\bbP_{\mu_H}$, so $\wt\bbP$ is the joint
distribution of $(\wt\vZ_s : s \in [\mu_H])$.
For any event $E$,
\[
  |\bbP(E) - \wt\bbP(E)| \leq (\mu_H-1) \beta(\bbP)
\]
where
\[
  \beta(\bbP)
  := \max_{1 \leq s \leq \mu_H-1}
  \bbE\Parens{
    \Norm{
      \bbP_{1:s+1}(\cdot\,|\vZ_1,\vZ_2,\dotsc,\vZ_s) - \bbP_{s+1}
    }_{\tv}
  }
  \,.
\]
Here, $\norm{\mu}_{\tv}$ denotes the total variation norm.
Recalling the definition of beta mixing,
the number $\beta(\bbP)$ can be recognized to be the first beta mixing coefficient of the stochastic process $\{\vZ_s\}$.
This result implies that the bound stated above for
$\norm{\sum_{s=1}^{\mu_H} w_{H,s} \wt\vZ_s}$ also holds for
$\norm{\sum_{s=1}^{\mu_H} w_{H,s} \vZ_s}$, except the failure
probability increases from $\delta$ to $\delta +
(\mu_H-1)\beta(\bbP)$.

It is not hard to see that $\beta(\bbP)$ in turn is upper bounded by 
the $(a-1)$-th beta mixing coefficient of the Markov chain $\{X_t\}$ (because the first
index in $H_{s+1}$ and the last index in $H_s$ are at a distance of $(a-1)$).
Hence, 
%of a signed measure $\mu$ over a measurable space 
%$(\Omega,\mathcal{A})$, defined by 
%$\Norm{\mu}_{\tv} = \frac12(\mu^+(\Omega)+\mu^-(\Omega))$, 
%where $\mu^+$ and $\mu^-$ are the positive and negative parts of measure $\mu$.
by a well-known result 
\[
  \beta(\bbP)
  \leq
  \frac{1}{\pimin} \exp(-(a-1)\gap)
\]
\citep[see, for instance,][]{bradley05}. \todoc{Which theorem? I did not find this result there. Can anyone please refine this citation?}
It follows that if the block length $a$ satisfies
\begin{equation}
  \label{eq:block-length2}
  a \geq 1 + \frac{1}{\gap}\ln\frac{n}{\delta\pimin}
  ,
\end{equation}
then the failure probability increases from $\delta$ to at most
$2\delta$.
%Note that this condition implies \eqref{eq:block-length}.

Combining \eqref{eq:biasbound},\eqref{eq:indbound}, we get that if
\[
  a := \left\lceil 1 + \frac{1}{\gap}\ln\frac{2n}{\pimin\delta} \right\rceil
\]
then for any $0\le \delta \le 1$, with probability at least $1-\delta$, with $\mu := \mu_H = \mu_T=n/(2a)$,
\begin{align*}
%\errm-\EE{\errm}
%& =  \Diag(\vpi)^{-1/2}\Parens{\wh\vM - \bbE[\wh\vM]} \Diag(\vpi)^{-1/2} 
\norm{\wh\vM - \bbE \wh\vM }
& \le 
  2\sqrt{
    \frac{12\ln(16d/\delta)}{\mu\pimin}
  }
  +
  \frac{4\ln(16d/\delta)}{\mu\pimin}
  + 
  \frac{2}{n-1}\,,
%  +
%  \frac1{(n-1)\gap \sqrt{\pimin}} 
\end{align*}
which, after the definition of $a$ is plugged in,
together with \eqref{eq:errmdev1},
 gives that as long as $n$ is big enough so that $2a\le n$ holds, \todoc{Second condition on $n$.}
\begin{align*}
\norm{\errm-\EE{\errm}}
\le \frac{C}{\pimin}
\sqrt{
    \frac{\ln(d/\delta)\ln(\frac{n}{\delta\pimin})}{\pimin\gap n}
  }
  \,.
\end{align*}
\subsubsection{The bound on $\norm{\errm}$}
As the RHS of the previous bound dominates the RHS of \eqref{eq:biasbound},
we get 
\begin{align}
\label{eq:doubletbound}
\norm{\errm}
\le \frac{C}{\pimin}
\sqrt{
    \frac{\ln(d/\delta)\ln(\frac{n}{\delta\pimin})}{\pimin\gap n}
  }\,.
\end{align}

\subsection{Overall error bound}
Consider an event when both~\eqref{eq:singletonbound} and \eqref{eq:doubletbound} hold.
Then, using that $\ln(d/\delta) + \ln(1/\pimin) \le \ln(d/\delta)\ln(\frac{n}{\delta\pimin})$, we get
\begin{align*}
\norm{\wh\vL - \vL} \le 
\frac{C}{\pimin}
\sqrt{
    \frac{\ln(d/\delta)\ln(\frac{n}{\delta\pimin})}{\pimin\gap n}
  }\,,
\end{align*}
which together with \eqref{eq:singletonboundsimple},
finishes the proof of \cref{lem:err}.
\hfill $\blacksquare$

\if0

Plugging the previous bound and~\eqref{eq:singletonbound} into~\eqref{eq:vlbound}, we see that
\[
  a := \left\lceil 1 + \frac{1}{\gap}\ln\frac{n}{\pimin\delta} \right\rceil
\]
so that~\eqref{eq:block-length2} holds (which implies~\eqref{eq:block-length}).
Then, we have the following error bound: for some absolute constant
$C>0$, with probability at least $1-\delta$, \todoc{From
$\norm{\errp}$, we get $(\pimin\gap^2n)^{-1/2}$! Where is the cubic term coming from? Why is it not dominated by the other term? All the subsequent error bounds will need to be updated.}
\begin{equation}
  \norm{\wh\vL - \vL}
  \leq
  C
  \sqrt{\frac{\ln(d/\delta)\ln\frac{n}{\delta\pimin}}{n\pimin\gap}}
  +
  C
  \Parens{
    \frac{\ln(d/\delta)\ln\frac{n}{\delta\pimin}}{n\pimin\gap}
  }^3
\end{equation}
and for all $i \in [d]$,
\begin{equation}
  \Abs{ \frac{\hat\pi_i}{\pi_i} - 1 }
  \leq 
  C
  \sqrt{\frac{\ln(d/\delta)\ln\frac{n}{\delta\pimin}}{n\pimin\gap}}
  +
  C
  \frac{\ln(d/\delta)\ln\frac{n}{\delta\pimin}}{n\pimin\gap}
  .
\end{equation}
In this event, we have
\begin{equation}
  |\hatgap - \gap|
  \leq
  C'
  \max\Braces{
    \sqrt{\frac{\ln(d/\delta)\ln\frac{n}{\delta\pimin}}{n\pimin\gap}}
    ,\,
    \Parens{
      \frac{\ln(d/\delta)\ln\frac{n}{\delta\pimin}}{n\pimin\gap}
    }^3
  }
\end{equation}
for some other absolute constant $C'>0$.
This completes the proof of Theorem~\ref{thm:err}.
\hfill $\blacksquare$
\fi

%In terms of matching our lower bounds for estimating $\gap$, we are
%pretty close: ignoring log factors, our sample complexity is
%\[
%  \tilde{O}\Parens{ \frac1{\pimin\gap} }
%\]
%whereas the lower bound is
%\[
%  \tilde{\Omega}\Parens{ \frac{d}{\gap} + \frac1{\pimin} }
%  .
%\]
%So when $\vpi$ is not too far from uniform---meaning that $\pimin =
%\Omega(1/d)$---we are matching the lower bound up to logarithmic
%factors.

\if0
\subsection{Empirical upper confidence bound}

Now, we actually want to be able to provide an empirical upper
confidence bound on $\gap$, which is where we had to do that
calculation from NIPS that introduced extra factors of $1/\gap$ and
$1/\pimin$.
I haven't had time to write out these details, but needless to say, it
will be messy.

We know that
\[
  \pimin \geq \hatpimin - \sqrt{\frac{C\hatpimin}{n\gap}} -
  \frac{C}{n\gap}
  \geq \frac{\hatpimin}{2} - \frac{3C}{n\gap}
  .
\]
for some suitable ``constant'' (log term) $C>0$.
Therefore, ignoring the higher-order forms, we have
\[
  |\hatgap-\gap|
  \leq \sqrt{\frac{C'}{n\hatpimin\gap/2-C''}}
  .
\]
Unfortunately, I don't think we can get an upper bound on $\gap$ this
way, since we can't really know when the denominator on the RHS is
positive. \todoc{Why not? We have to add the constraint that this is positive
and we could solve the inequalities numerically.}

Our non-rigorous approach is to use a first-order approximation.
Let $x := \gap - \hatgap$ and $y := \pimin - \hatpimin$, and use a
linear apporximation to
\[
  b(x,y) = \sqrt{\frac{C}{n(\hatgap-x)(\hatpimin-y)}}
\]
as a bound on $|x|$ and $|y|$.
The first-order approximation is
\[
  b(x,y)
  \approx \tilde{b}(x,y)
  :=
  \sqrt{\frac{C}{n\hatgap\hatpimin}}
  + \frac12\sqrt{\frac{C'}{n\hatgap^3\hatpimin}}x
  + \frac12\sqrt{\frac{C'}{n\hatgap\hatpimin^3}}y
  .
\]
The approximation is a valid upper-bound when $|x| \leq
\frac{\hatgap}2$ and $|y| \leq \frac{\hatpimin}2$; such will be true
when
\[
  \sqrt{\frac{C}{n\gap\pimin}} \leq \min\Braces{ \frac{\hatgap}{3} ,\,
  \frac{\hatpimin}{3} }
\]
i.e., when
\[
  n = \Omega\Parens{ \frac1{\gap^3\pimin} + \frac1{\gap\pimin^3} } .
\]
Assuming this, we find $\bar{x},\bar{y} \geq 0$ such that
$\tilde{b}(\bar{x},\bar{y}) = \bar{x}$ and $\tilde{b}(\bar{x},\bar{y})
= \bar{y}$:
\begin{align*}
  \Parens{
    1 - \frac12\sqrt{\frac{C}{n\hatgap^3\hatpimin}}
  } \bar{x}
  - \frac12\sqrt{\frac{C}{n\hatgap\hatpimin^3}}
  \bar{y}
  & =
  \sqrt{\frac{C}{n\hatgap\hatpimin}}
  \\
  - \frac12\sqrt{\frac{C}{n\hatgap^3\hatpimin}}
  \bar{x}
  + \Parens{
    1 - \frac12\sqrt{\frac{C}{n\hatgap\hatpimin^3}}
  }
  \bar{y}
  & =
  \sqrt{\frac{C}{n\hatgap\hatpimin}}
  .
\end{align*}
The solutions are
\begin{align*}
  \bar{x} 
  = \bar{y}
  & = \frac{1}{
    1 - \frac12\sqrt{\frac{C}{n\hatgap^3\hatpimin}}
    - \frac12\sqrt{\frac{C}{n\hatgap\hatpimin^3}}
  }
  \sqrt{\frac{C}{n\hatgap\hatpimin}}
  .
\end{align*}
\fi

### 2015-06-01 empirical.tex
minor tweak to discussion of group inverse

@@
 We form the matrix $\wh\vP$ using smoothed frequency estimates of
-$\vP$ (Step~\ref{step:P}), then 
-compute the so-called group inverse $\giAh$ of $\wh\vA = I - \wh\vP$,
-followed by finding the unique stationary
-distribution $\hat\vpi$ of $\wh\vP$ (Step~\ref{step:pi}).
-The group inverse $\giAh$ of $\wh\vA$ is uniquely defined and if $\wh\vP$ defines an ergodic chain
-(which is the case here due to the use of the smoothed estimates),
-$\giAh$ can be computed at the cost of inverting an $(d-1)\times (d-1)$ matrix
-(cf. Theorem 5.2 in \cite{meyer1975role}).
-Further, once given $\giAh$, the unique stationary distribution of $\wh\vP$ can be 
-read out from the last row of $\giAh$ (Theorem 5.3 of  \cite{meyer1975role}).
-The group inverse will also be used later to compute the sensitivity of $\vpi$
-based on a result from \cite{cho2001comparison}.
+$\vP$ (Step~\ref{step:P}), then compute the so-called group inverse
+$\giAh$ of $\wh\vA = I - \wh\vP$ (Step~\ref{step:gi}), followed by
+finding the unique stationary distribution $\hat\vpi$ of $\wh\vP$
+(Step~\ref{step:pi}).
+The group inverse $\giAh$ of $\wh\vA$ is uniquely defined; and if
+$\wh\vP$ defines an ergodic chain (which is the case here due to the
+use of the smoothed estimates), $\giAh$ can be computed at the cost of
+inverting an $(d-1)\times (d-1)$ matrix~\citep[Theorem
+5.2]{meyer1975role}.
+Further, once given $\giAh$, the unique stationary distribution
+$\hat\vpi$ of $\wh\vP$ can be read out from the last row of
+$\giAh$~\citep[Theorem 5.3]{meyer1975role}.
+The group inverse will also be used later to compute the sensitivity
+of $\vpi$ based on a result from \cite{cho2001comparison}.
 Based on $\hat\vpi$ and $\wh\vP$, we construct the plug-in estimate

### 2015-06-01 rates.tex
simplify upper bound rate
tweak to lower bound spart

@@
       \sqrt{\frac{\log\frac{d}{\delta}\cdot\log\frac{n}{\pimin\delta}}{\pimin\gap n}}
-      + \frac{\log\frac{d}{\delta}\cdot\log\frac{n}{\pimin\delta}}{\pimin\gap n}  
+%      + \frac{\log\frac{d}{\delta}\cdot\log\frac{n}{\pimin\delta}}{\pimin\gap n}  
       + \frac{\log\frac{1}{\gap}}{\gap n}  
@@
 \todoc{In \eqref{eq:gapbound}, we should be able to merge the second term into the first, right?}
-\todod{No, not for all values of $n$.}
+\todod{Oh, I see what you mean---yes, that's right.}
 
@@
 Essentially, the sequence may have to visit all $d$ states $\log(d)/\gap$ times each on average.
-This hold \emph{even} if $\pimin$ is with a factor of $1/2$ of the \emph{largest} possible value of $1/d$ that it can take,
-i.e., when $\vpi$ is nearly uniform.
+This hold \emph{even} if $\pimin$ is within a factor of two of the
+\emph{largest} possible value of $1/d$ that it can take, i.e., when
+$\vpi$ is nearly uniform.
 \begin{theorem}

### 2015-06-01 intro-nips.tex
tweaks and comments in intro/abstract

@@
 In this paper we study the construction of fully empirical bounds on the mixing time of Markov chains based on a single sample path.
-Let $(X_t)_{t=1,2,\dots}$ be an irreducible, aperiodic time-homogeneous Markov chain on a finite state space $[d]$ of $d$ states numbered by positive integers with transition matrix $\vP$.
+Let $(X_t)_{t=1,2,\dotsc}$ be an irreducible, aperiodic time-homogeneous Markov chain on a finite state space $[d]$ of $d$ states numbered by positive integers with transition matrix $\vP$.
 Under this assumption, the chain converges to its unique stationary \todoc{Maybe steady state is better?}
@@
 \]
-Here $\vpi(A) = \sum_{i\in A} \pi_i$ is the total probability assigned to set $A$ by $\vpi$ and the supremum is over
+Here $\vpi(A) = \sum_{i\in A} \pi_i$ is the total probability assigned
+to set $A$ by $\vpi$, and the supremum is over
 all possible initial distributions $\vq$.
-The problem studied in the paper is the construction of a confidence set $C_n = C_n(X_1,\dots,X_n,\delta) \subset [0,\infty]$ based on the knowledge of $X_1,\dots,X_n$ only that captures the mixing time $\tmix$ with a pre-specified probability $0\le \delta \le 1$.
+The problem studied in the paper is the construction of a confidence
+set $C_n = C_n(X_1,X_2,\dotsc,X_n,\delta) \subset [0,\infty]$ based on
+the knowledge of $X_1,X_2,\dotsc,X_n$ only that captures the mixing
+time $\tmix$ with a pre-specified probability $1-\delta \in (0,1)$. 
 In the remainder of the introduction, we describe the background of why this problem might be interesting, followed by a 
 brief description of our main results.
+\todod{``might be interesting'' sounds weak.}
 

### 2015-06-01 intro-nips.tex
.

@@
 estimator of Markov mixing time.
+\todod{``apparently'' sounds weak.}
 Consider a reversible ergodic Markov chain on $d$ states with spectral

### 2015-06-01 empirical.tex
adding and resolving todos, as per our discussion.
bib file improvements
generalized inverse clean up, both in appendix and in main text
ideas for future work in conclusion
empirical results: comment on O(.) notation
minor tweak to intro

--- 2015-06-01 empirical.tex
@@
 We avoid this slow rate by using an estimate of the symmetric matrix
-$\vL$, and employing a stronger perturbation result (namely Weyl's
-inequality) available for symmetric matrices.
+$\vL$, so that we can use a stronger perturbation result (namely Weyl's
+inequality, as in the proof of \cref{thm:err}) available for symmetric matrices.
 
@@
 We form the matrix $\wh\vP$ using smoothed frequency estimates of
-$\vP$ (Step~\ref{step:P}), then 
-compute the so-called group inverse $\giAh$ of $\wh\vA = I - \wh\vP$,
-followed by finding the unique stationary
-distribution $\hat\vpi$ of $\wh\vP$ (Step~\ref{step:pi}).
-The group inverse $\giAh$ of $\wh\vA$ is uniquely defined and if $\wh\vP$ defines an ergodic chain
-(which is the case here due to the use of the smoothed estimates),
-$\giAh$ can be computed at the cost of inverting an $(d-1)\times (d-1)$ matrix
-(cf. Theorem 5.2 in \cite{meyer1975role}).
-Further, once given $\giAh$, the unique stationary distribution of $\wh\vP$ can be 
-read out from the last row of $\giAh$ (Theorem 5.3 of  \cite{meyer1975role}).
-The group inverse will also be used later to compute the sensitivity of $\vpi$
-based on a result from \cite{cho2001comparison}.
+$\vP$ (Step~\ref{step:P}), then compute the so-called group inverse
+$\giAh$ of $\wh\vA = I - \wh\vP$ (Step~\ref{step:gi}), followed by
+finding the unique stationary distribution $\hat\vpi$ of $\wh\vP$
+(Step~\ref{step:pi}), this way decoupling the bound on the accuracy of $\hat\vpi$
+from the mixing time.
+The group inverse $\giAh$ of $\wh\vA$ is uniquely defined;
+ and if
+$\wh\vP$ defines an ergodic chain (which is the case here due to the
+use of the smoothed estimates), $\giAh$ can be computed at the cost of
+inverting an $(d-1)\times (d-1)$ matrix~\citep[Theorem
+5.2]{meyer1975role}.%
+\footnote{
+\label{ftnote:group-inverse}
+The group inverse of a square matrix $\vA$, a special case of the {\em Drazin inverse},
+is the unique matrix $\vA^\#$ satisfying
+$ \vA\vA^\#\vA = \vA$,
+$\vA^\#\vA\vA^\# = \vA^\#$ and
+$\vA^\#\vA = \vA\vA^\#$.
+%It is efficiently computable,
+%and for $\vA=\vI-\vP$, where $\vP$ is ergodic with stationary distribution $\vpi$,
+%has the
+%following simple
+%form: $\vA^\#=(\vI-\vP+\v1\vpi^\t)^{-1}-\v1\vpi^\t$
+%\citep{seneta1993sensitivity,meyer1975role}.
+}
+Further, once given $\giAh$, the unique stationary distribution
+$\hat\vpi$ of $\wh\vP$ can be read out from the last row of
+$\giAh$~\citep[Theorem 5.3]{meyer1975role}.
+The group inverse will also be used later to compute the sensitivity
+of $\vpi$ based on a result of Cho and Meyer \cite{cho2001comparison}.
 Based on $\hat\vpi$ and $\wh\vP$, we construct the plug-in estimate
@@
 Thus, with a standard implementation of matrix inversion, the algorithm's complexity becomes 
-$O( n + d^3)$, while the space complexity is $O(d^2)$.
+$O( n + d^3)$, while its space complexity is $O(d^2)$.
 \todoc{Implement and test?}
@@
 The main result of this section is the following result:
-\todoc{Conditions?}
-\todod{I think they are fine now. Ergodicity is defined in section 3
-now, and we discuss why it already implies the $\pimin>0$ and
-$\gap>0$.}
 \begin{theorem}
@@
   $\giA$ of $\vA := \vI - \vP$.
+  \todoc{
+    Precisely,
+    what we mean by the $O(\cdot)$ notation is as follows:
+  for a random sequence of nonnegative numbers $(X_n)_n$
+  and another sequence of nonrandom positive numbers $(a_n^\eps)_n$ 
+  indexed by $\eps$ we
+  say that $X_n = O(a_n^\eps)$ holds almost surely as $n\to \infty$
+  if for some universal constant $C>0$, 
+  $\limsup_{n\to \infty} X_n/a_n^\eps \le C$ holds almost surely, regardless of
+  the choice of $\eps$. I think this should be made clear.
+  }
 \end{theorem}

### 2015-06-01 intro-nips.tex
adding and resolving todos, as per our discussion.
bib file improvements
generalized inverse clean up, both in appendix and in main text
ideas for future work in conclusion
empirical results: comment on O(.) notation
minor tweak to intro

@@
 time $\tmix$ with a pre-specified probability $1-\delta \in (0,1)$. 
-In the remainder of the introduction, we describe the background of why this problem might be interesting, followed by a 
+In the remainder of the introduction, we describe our motivation of studying this problem, followed by a 
 brief description of our main results.
-\todod{``might be interesting'' sounds weak.}
+\todod{``might be interesting'' sounds weak.} \todoc{better now?}
 

### 2015-06-01 rates.tex
adding and resolving todos, as per our discussion.
bib file improvements
generalized inverse clean up, both in appendix and in main text
ideas for future work in conclusion
empirical results: comment on O(.) notation
minor tweak to intro

@@
   , \quad i \in [d]
-  .
+  \,.
 \end{align*}
@@
 
-These estimators have the following accuracy guarantees. 
+These estimators have the following accuracy guarantees:
 \begin{theorem}
@@
 \end{theorem}
-\todoc{In \eqref{eq:gapbound}, we should be able to merge the second term into the first, right?}
-\todod{Oh, I see what you mean---yes, that's right.}
 
@@
 
+\todoc{Make this paragraph into a remark?}
 Note that because the eigenvalues of $\vL$ are the same as that of the

### 2015-06-01 setting.tex
adding and resolving todos, as per our discussion.
bib file improvements
generalized inverse clean up, both in appendix and in main text
ideas for future work in conclusion
empirical results: comment on O(.) notation
minor tweak to intro

@@
 $\vP$, and hence it follows that $\vL$ is also symmetric.
+As we will see later, we will strongly exploit this later.  \todoc{Say more? Maybe not..}
+%will be advantageous due to the existence of stronger perturbation results
+%for symmetric matrices.
 Further, $\vL = \Diag(\vpi)^{1/2} \vP \Diag(\vpi)^{-1/2}$, hence $\vL$
@@
 (absolute) spectral gap to be $\gap := 1-\slem$, which is strictly positive.
+\todoc{Do our results generalize to the spectral gap? Where do we need that we talk about the absolute spectral gap?
+What is the difference between the spectral and the absolute spectral gap? Say, in their relationship to $\tmix$?}
 %Throughout the paper, both $\pimin$ and $\gap$ are assumed
@@
 can ultimately be bounded in terms of $\pimin$ and
-$\gap$~\citep[e.g.,][]{MR1921877,MoRo08,MoRo09}. \todoc{Cite one? Btw, do these really have these results?}
+$\gap$, as we will show in the proof of our first result.\todoc{I removed the citations (saving space), referring to the proof.}
+%~\citep[e.g.,][]{MR1921877,MoRo08,MoRo09}. 
 Therefore, valid confidence intervals for $\pimin$ and $\gap$ can be

### 2015-06-02 abstract.tex
tweaks and comments in intro/abstract
bla
intro and other stuff

@@
 %!TEX root =  matrix-est.tex
-We propose an efficient procedure for estimating the mixing time of a Markov chain.
-The novelty of our approach lies in several key aspects.
-Crucially,
-we do not assume any a priori bounds
-on the mixing time,
-which
-sets our approach apart from
-most others that do
-require some advance knowledge of this quantity.
-Another
-advantage of our result over
-common practice is the sufficiency of a single long sampling sequence,
-without access to a restart mechanism.
-Perhaps most importantly, we overcome an obstacle inherent in the chicken-and-egg nature of
-estimating mixing coefficients: to give confidence bounds one typically needs to know the mixing time.
-We provide confidence guarantees that are fully data-dependent in the sense that they yield non-trivial
-finite-sample deviation bounds without any knowledge of the mixing coefficients.
-Upper and lower sample complexity bounds are provided and challenging open problems are posed.
+This article provides an efficient procedure for estimating the mixing
+time of a reversible ergodic Markov chain.
+The procedure computes the estimate from a single sample path of the
+Markov chain, and does not require any \emph{a priori} bounds on the
+mixing time itself.
+This stands in contrast to previous approaches, which either require a
+reset mechanism or require additional prior knowledge.
+Most notably, the procedure produces a fully empirical confidence
+interval (i.e., it depends only on the sample path and a given
+confidence parameter $\delta$) that is always valid with probability
+$1-\delta$; and the width of the interval converges to zero at a
+$\sqrt{n}$-rate, where $n$ is the length of the sample path.
+
+%We provide confidence guarantees that are fully data-dependent in the sense that they yield non-trivial
+%finite-sample deviation bounds without any knowledge of the mixing coefficients.
+%
+%
+%The novelty of our approach lies in several key aspects.
+%Crucially,
+%we do not assume any a priori bounds
+%on the mixing time,
+%which
+%sets our approach apart from
+%most others that do
+%require some advance knowledge of this quantity.
+%Another
+%advantage of our result over
+%common practice is the sufficiency of a single long sampling sequence,
+%without access to a restart mechanism.
+%Perhaps most importantly, we overcome an obstacle inherent in the chicken-and-egg nature of
+%estimating mixing coefficients: to give confidence bounds one typically needs to know the mixing time.
+%We provide confidence guarantees that are fully data-dependent in the sense that they yield non-trivial
+%finite-sample deviation bounds without any knowledge of the mixing coefficients.
+%%Upper and lower sample complexity bounds are provided and challenging open problems are posed.
+%%\todod{I would remove the last sentence, at least the part about
+%%``challenging'' open problems.}
+%%%test

### 2015-06-02 empirical.tex
intro and other stuff

@@
   by~\cref{alg:empest} with $\giAh$ replaced by the group inverse
-  $\giA$ of $\vA := \vI - \vP$.
-  \todoc{
-    Precisely,
-    what we mean by the $O(\cdot)$ notation is as follows:
-  for a random sequence of nonnegative numbers $(X_n)_n$
-  and another sequence of nonrandom positive numbers $(a_n^\eps)_n$ 
-  indexed by $\eps$ we
-  say that $X_n = O(a_n^\eps)$ holds almost surely as $n\to \infty$
-  if for some universal constant $C>0$, 
-  $\limsup_{n\to \infty} X_n/a_n^\eps \le C$ holds almost surely, regardless of
-  the choice of $\eps$. I think this should be made clear.
-  }
+  $\giA$ of $\vA := \vI - \vP$.\footnote{%
+    In \cref{thm:empirical,thm:combined}, our use of big-$O$ notation
+    is as follows.
+    For a random sequence $(Y_n)_n$ and a (non-random) positive
+    sequence $(a_{\theta,n})_n$ parameterized by $\theta$, we say
+    ``$Y_n = O(a_{\theta,n})$ holds almost surely as $n\to\infty$'' if
+    there is some universal constant $C>0$ such that for all $\theta$,
+    $\limsup_{n\to\infty} Y_n/a_{\theta,n} \leq C$ holds almost
+    surely.%
+  }%
+%  \todoc{
+%    Precisely,
+%    what we mean by the $O(\cdot)$ notation is as follows:
+%  for a random sequence of nonnegative numbers $(X_n)_n$
+%  and another sequence of nonrandom positive numbers $(a_n^\eps)_n$ 
+%  indexed by $\eps$ we
+%  say that $X_n = O(a_n^\eps)$ holds almost surely as $n\to \infty$
+%  if for some universal constant $C>0$, 
+%  $\limsup_{n\to \infty} X_n/a_n^\eps \le C$ holds almost surely, regardless of
+%  the choice of $\eps$. I think this should be made clear.
+%  }
 \end{theorem}

### 2015-06-02 intro-nips.tex
intro and other stuff

--- 2015-06-01 intro-nips.tex
@@
 
-In this paper we study the construction of fully empirical bounds on the mixing time of Markov chains based on a single sample path.
-Let $(X_t)_{t=1,2,\dotsc}$ be an irreducible, aperiodic time-homogeneous Markov chain on a finite state space $[d]$ of $d$ states numbered by positive integers with transition matrix $\vP$.
-Under this assumption, the chain converges to its unique stationary \todoc{Maybe steady state is better?}
-distribution $\vpi = (\pi_i)_{i=1}^d$ regardless of the distribution $\vq$ of $X_1$: 
-$\lim_{t\to\infty} \Pr_{\vq}\Parens{X_t = i} = \lim_{t\to\infty} (\vq \vP^t)_i = \pi_i$, $1\le i \le d$.
-The mixing time $\tmix$ of the Markov chain is the time that it takes for the chain to reach its stationary distribution, say,
+This work considers the construction of fully empirical bounds on the
+mixing time of Markov chains based on a single sample path.
+Let $(X_t)_{t=1,2,\dotsc}$ be an irreducible, aperiodic
+time-homogeneous Markov chain on a finite state space $[d] :=
+\{1,2,\dotsc,d\}$ with transition matrix $\vP$.
+Under this assumption, the chain converges to its unique stationary
+\todoc{Maybe steady state is better?} distribution $\vpi =
+(\pi_i)_{i=1}^d$ regardless of the initial state distribution $\vq$: 
+\[
+  \lim_{t\to\infty} {\Pr}_{\vq}\Parens{X_t = i}
+  = \lim_{t\to\infty} (\vq \vP^t)_i = \pi_i
+  \quad \text{for each $i \in [d]$} .
+\]
+The mixing time $\tmix$ of the Markov chain is the time that it takes
+for the chain to (approximately) reach its stationary distribution,
 \[
@@
     \sup_{\vq}
-    \max_{A\subset [d]} |\textstyle\Pr_{\vq}\Parens{ X_t \in A } - \vpi(A)|
+    \max_{A\subset [d]}
+    \Abs{
+      \textstyle\Pr_{\vq}\Parens{ X_t \in A } - \vpi(A)
+    }
 %    \Norm{
@@
 \]
-Here $\vpi(A) = \sum_{i\in A} \pi_i$ is the total probability assigned
-to set $A$ by $\vpi$, and the supremum is over
-all possible initial distributions $\vq$.
-The problem studied in the paper is the construction of a confidence
-set $C_n = C_n(X_1,X_2,\dotsc,X_n,\delta) \subset [0,\infty]$ based on
-the knowledge of $X_1,X_2,\dotsc,X_n$ only that captures the mixing
-time $\tmix$ with a pre-specified probability $1-\delta \in (0,1)$. 
-In the remainder of the introduction, we describe the background of why this problem might be interesting, followed by a 
-brief description of our main results.
-\todod{``might be interesting'' sounds weak.}
+Here, $\vpi(A) = \sum_{i\in A} \pi_i$ is the probability assigned to
+set $A$ by $\vpi$, and the supremum is over all possible initial
+distributions $\vq$.
+The problem studied in this work is the construction of a non-trivial
+confidence interval $C_n = C_n(X_1,X_2,\dotsc,X_n,\delta) \subset
+[0,\infty]$, based only on the observed sample path
+$(X_1,X_2,\dotsc,X_n)$ and $\delta \in (0,1)$, that captures the
+mixing time $\tmix$ with the pre-specified probability $1-\delta$.
 
-In numerous scientific applications and some machine learning tasks
-one is interested in estimating the mean $\vpi(f) = \sum_i \pi_i f(i)$ of some function $f$ of the states of a Markov chain. 
-For example, this is the realms of the celebrated Markov Chain Monte Carlo (MCMC) paradigm, but 
-the problem also arises in performance prediction when working with time correlated data, as is common in,
-for example, in reinforcement learning.
-%as 
-% (cf. \cref{sec:mot}) 
-Under our assumptions,
-the law of large numbers guarantees that the sample mean $\vpi_n(f) \defeq \frac1n \sum_{t=1}^n f(X_t)$ 
-converges almost surely to $\vpi(f)$, while the central limit theorem tells us that as $n\to \infty$, 
-the distribution of the deviation $\sqrt{n}( \vpi_n(f)-\vpi(f))$ will be normal with zero mean and asymptotic
-variance $\sigma^2_{\mathrm{as}}(f) \defeq \lim_{n\to\infty} n \Var\Parens{ \vpi_n(f) }$. \todoc{shall we add a citation here?}
-While these results help us to understand how the sample mean over a Markov chain behaves in the limit,
-they tell us little about the finite time, non-asymptotic behavior of the sample mean,
-which are needed for prudent evaluation of one's method or even for algorithm design 
-\cite{
-MCMCDiscussion93%
-,DBLP:conf/valuetools/KontoyiannisLM06%
-,BBL06%
-,MniSzeAu08%
-,MauPo09%
-,LiLiWaSt11:KWIK%
-,flegal2011implementing%
-,Gyori-paulin15%
-,SwaJoa15:LoggedBandit%
-}. %BaHaVa10:activelearning,ShaVo08:conformalpred
-To address this need, numerous works have been devoted to deriving
-Chernoff-type, exponential bounds for upper bounding the tail probabilities $p(r) = \Pr\Parens{ |\vpi_n(f)-\vpi(f)|> r }$,
-also known as ``concentration bounds'', e.g., 
-\cite{gillman1998chernoff,leon2004optimal,DBLP:conf/valuetools/KontoyiannisLM06,paulin15}.
-Compared with the case when $(X_t)_t$ is an independent, identically distributed (iid) sequence of
-random variables, these tail probabilities must be larger due to the temporal dependence of the sequence $(X_t)_t$.
-Intuitively, for the Markov chain to forget the state where it is at time $t$, or in other words to get a 
-``fresh'' sample $X_{t'}$ that is approximately ``independent'' from $X_t$, one must wait $\Theta(\tmix)$ time steps.
-Thus, in the bounds on $p(r)$, one expects to see that the sample size $n$ 
-gets appropriately reduced as $\tmix$ grows and this is indeed the case in bounds mentioned above.
-is replaced by $n/\tmix$, which one
-might think of as an approximate sample size. 
-Indeed, all the above mentioned works get bounds on $p(r)$ that depend on various functions of $\vP$.
-Namely, Kontoyiannis et al. obtain bounds that depend on certain information-theoretic quantities defined implicitly
-as a function of $\vP$ and $f$, which are after simplification is bounded based on quantities related to 
-a regeneration-type quantity derived from $\vP$
-\cite{DBLP:conf/valuetools/KontoyiannisLM06}, while the other works mentioned above, in the case of reversible chains,
-derive bounds that depend on the spectral gap of $\vP$.
-Note that the spectral gap for such chains is essentially inversely related to the mixing time of the chain (see \cref{sec:setting}).
-%\cite{gillman1998chernoff,leon2004optimal,,paulin15}.
-Much research have been devoted to finding appropriate \emph{a priori} bounds on the various quantities that appear 
-in these bounds, mostly in the theory community (e.g., \cite{saloff2004random}). 
-To address the lack or looseness of such a priori bounds, a significant effort also went into developing methods
-that estimate these quantities based on data, mostly in the MCMC community in an effort to develop
-effective methods that are able to diagnose whether a chain has reached its stationary state 
-(e.g., \cite{MCMCDiscussion93,flegal2011implementing,Gyori-paulin15}).
-While the typical results state the asymptotic consistency of these estimators, or a central limit theorem (which 
-are of little help to achieve a fully empirical, non-asymptotic bound on $p(r)$),
-Gyori and Paulin give a finite time error bound for these estimates. 
-Unsurprisingly, these error bounds themselves
-depend on other unknown quantities of the Markov chain (which can be bounded in terms of the mixing time, again)
-\cite{Gyori-paulin15}.
-
-This situation is similar to what appears when one attempts to use a Bernstein-type deviation bound to 
-control the deviation between the sample mean of a sequence of iid random variables from their common mean
-with a known range, but unknown variance: For a tight bound, one wishes to use the unknown variance, but in the deviation
-bound for the error of a variance estimate, the variance appears again.
-While in the case of Bernstein-bounds this circular reference leads to a nontrivial bound
-\cite{audibert2009}, it appears that the situation is different in the Markov setting.
-Focusing on the spectral gap, we will see that if one attempts to use a similar technique,
-the set of feasible values based on the two error bounds always will never rule out that the spectral gap is zero,
-or that the mixing time is infinite. Can one even hope to obtain 
-after finitely many samples a nontrivial upper bound on the mixing time?
-Obviously, for any finite sample size, if the mixing time of the chain is larger than the sample size than a sound method
-must return infinity as the upper bound on the mixing time. Hence, the question is whether one can design a method
-that never fails and if the sample size is appropriately large, returns a nontrivial bound on the mixing time, adapting the time it needs to trap the mixing time in a fixed size confidence interval to the unknown mixing time.
-
-\hide{
-\subsection{Motivation}
-\label{sec:mot}
-One key competence in statistical learning is the ability to predict one's performance on future data.
-To make performance prediction possible, one starts by posing some assumptions on how the data arises.
-Simpler assumptions lead to simpler algorithms and analysis, while they may also limit applicability of the
-obtained results and algorithms.
-The most common assumption employed postulates that all data points, including past and future data points,
-are sampled independently of each other from a common underlying distribution (e.g., \cite{SSBD14:UnderstandingML}).
-This is helpful on two accounts: 
-The lack of dependence between the data points means that sample averages of bounded functions of the data 
-will converge at a uniform rate (irrespective of the distribution), while the assumption that all data points share the same
-distribution means that one can learn about the distribution of future data points based on past data.
-}
-
-\hide{
-It is well-known that contrary to the idealized assumptions of the standard
-learning model, real data is almost never iid.
-Fortunately, many classical results of statistics and statistical
-learning theory extend to the dependent data in a more or less
-seamless fashion, at least when the data is sufficiently ``mixing''.
-Such results include the central limit theorem, 
-exponential tail bounds for empirical processes, 
-performance guarantees for complexity regularization
-using sieves or roughness penalties, as well as 
-results based on stability arguments or Rademacher complexity 
-\citep{Woo86,Dou94,Yu94,ZheChu96,dlPenaGine99,Mei00,
-  MR1921877,vid2003,gamarnik03,LoKuScha05,MoRo08,MoRo09,
-  DBLP:conf/nips/SteinwartC09,Steinwart2009175,
-  FaSze11,
-  london:nips12asalsn,london:icml13,
-  DBLP:conf/nips/ShaliziK13}.
-
-Unfortunately, a distribution-free theory of non-iid learning is simply
-impossible, when arbitrary dependence among the sample points is allowed.
-Hence, one must quantify this dependence via the various mixing coefficients,
-which appear in the non-iid generalization bounds mentioned above.
-A frequently raised concern is that now the bounds depend on {\em
-unknown} properties of the distribution, which precludes their
-actual use in algorithmic applications (e.g., active
-learning~\citep{BBL06,DHM07}).
-%, and any algorithm aiming at a fully empirical error bound must
-%also estimate the mixing coefficients.
-In addition to learning-theoretic generalization bounds, the need for
-fully empirical confidence bounds arises in Markov Chain Monte Carlo
-(MCMC) simulations---so as to halt the run when sufficiently close to
-the stationary distribution \citep{galin06}, and in Markov Reward
-Processes \citep{sutton98}, as this will affect the learner's policy.
-As \citet{Mei00} notes, in the absence of accurate estimates of the
-mixing time, such fully empirical bounds remain ``a theoretician's
-dream''.
-
-Our goal here is to develop an algorithm that, given a single sample
-path of a Markov chain, computes non-trivial bounds on the mixing time
-of the Markov chain.
-We remark on the inherent chicken-and-egg nature of the challenge:
-confidence intervals derived from a Markov chain---say, on the the
-chain's mixing time---are typically given in terms of the chain's
-mixing time.
-Our way out of this conundrum is to give {\em empirical} (i.e.,
-\emph{data-dependent}) as opposed to {\em a priori} guarantees.
-To illustrate the intuition, consider a Markov chain on two states,
-$\braces{0,1}$.
-A learner who is stuck in state $0$ for a long time learns very little
-about the chain and hence cannot be expected to give meaningful upper
-bounds on its mixing time.
-Consider, however, a learner who is lucky enough
-to be in a chain that switches frequently between the two states.
-Surely at some finite time, the learner will possess meaningful and nontrivial
-information about the chain's mixing time --- crucially, without any
-{\em prior} knowledge regarding this quantity. This simple example captures
-the nature of our results: in some cases, we will not be able to say
-anything meaningful (but the learner will recognize that this is the case),
-while in ``luckier'' situations, we will be able to provide nontrivial
-upper estimates on the mixing time.
-}
+This problem is motivated by the numerous scientific applications and
+machine learning tasks in which one is interested in estimating the
+mean $\vpi(f) = \sum_i \pi_i f(i)$ of a function $f$ of the states of
+a Markov chain.
+This is the realms of the celebrated Markov Chain Monte Carlo (MCMC)
+paradigm~\cite{liu2001monte}, but the problem also arises in
+performance prediction involving time-correlated data, as is common in
+reinforcement learning~\cite{sutton98}.
+Observable bounds on mixing times are useful in the design and
+diagnostics of these methods; they yield effective approaches to
+assessing the estimation quality, even when \emph{a priori} knowledge
+of the mixing time or correlation structure is unavailable.
 
 \paragraph{Main results.}
-We propose what is apparently the first efficient and fully empirical
-estimator of Markov mixing time.
-Consider a reversible ergodic Markov chain on $d$ states with spectral
-gap $\gap$ (which is well-known to govern mixing time)
-and stationary distribution minorized by $\pimin$.
+We develop the first efficient procedure for constructing non-trivial
+and fully empirical confidence intervals for Markov mixing time.
+Consider a reversible ergodic Markov chain on $d$ states with absolute
+spectral gap $\gap$ and stationary distribution minorized by $\pimin$.
+(These quantities $\gap$ and $\pimin$ are well-known to govern
+mixing time.)
 Our main results are summarized as follows.
@@
     path of length $\tilde{O}\parens{1/(\pimin\gap)}$.
+    This establishes the feasibility of \emph{estimating} the mixing
+    time in this setting.
     However, the valid confidence intervals suggested by
@@
     In \cref{thm:lb-pimin,thm:lb-gap}, we show that $\Omega\parens{
-    (d\ln d)/\gap + 1/\pimin }$ observations are necessary for
-    constant multiplicative accuracy in estimating $\gap$.
+    (d\log d)/\gap + 1/\pimin}$ observations are necessary for any
+    procedure to guarantee constant multiplicative accuracy in
+    estimating $\gap$.
+    Essentially, every state may need to be visited about
+    $\log(d)/\gap$ times, on average, before an accurate estimate of
+    the mixing time can be provided.
 
@@
     \emph{valid empirical confidence intervals} for $\pimin$ and
-    $\gap$.
+    $\gap$ (\cref{alg:empest}).
     The procedure uses just a single sample path and does not require
@@
 
-    We show in \cref{thm:combined} how to combine these empirical
-    confidence intervals with the non-empirical bounds from
-    \cref{thm:err} to produce valid empirical confidence intervals;
+  \item
+    Finally, in \cref{thm:combined}, we show how to combine the
+    empirical confidence intervals from \cref{alg:empest} with the
+    non-empirical bounds from \cref{thm:err} to produce valid
+    empirical confidence intervals;
     the width of these intervals converge to zero asymptotically as
@@
 
-\paragraph{Related work.} \todoc{Move this later? It also needs to be updated, to 
-adjust it to the current intro.}
-Owing to the wide applicability of MCMC methods to various computational
-problems, much effort has been devoted to developing effective convergence
-diagnostic methods the aim of deriving stopping rules or stationarity
-diagnoses. While a large number of heuristic techniques have been developed
-\citep[e.g.,][]{MCMCDiscussion93,CoCa96}, it is widely recognized that all
-these methods may fail to meet the guarantees that they are designed for.
-Some recent advance in this area \citep{1209.0703,Gyori-paulin15} have
-yielded practical algorithms for estimating the rates of convergence
-of various MCMC algorithms, but they fall short of providing intervals
-that are valid with any finite length sequence.
+\paragraph{Related work.}
+There is a vast statistical literature on estimation in Markov chains.
+For instance, it is know that under the assumptions on the Markov
+chain $(X_t)_t$ stated above, the law of large numbers guarantees that
+the sample mean $\vpi_n(f) \defeq \frac1n \sum_{t=1}^n f(X_t)$
+converges almost surely to $\vpi(f)$~\cite{meyn1993markov}, while the
+central limit theorem tells us that as $n\to \infty$, the distribution
+of the deviation $\sqrt{n}( \vpi_n(f)-\vpi(f))$ will be normal with
+mean zero and asymptotic variance $\lim_{n\to\infty}
+n\Var\Parens{\vpi_n(f)}$~\cite{kipnis1986central}.
 
-In the theory community, \citet{BaFoRuSmiWhi00,BaFoRuSmiWhi13}
-developed algorithms for testing whether a given Markov chain over a
-finite state space of size $d$ is rapidly mixing.
-In particular, assuming access to a sampling oracle that can generate
-independent transitions from any given state, their algorithm can
-distinguish between chains that in $t$ steps for a fixed $t$ reach a
-distribution close to the stationary distribution using $\tilde{O}(t
-d^{5/3})$ oracle calls.
-It is remarkable that when $t$ is small enough, the procedure requires
-sublinearly many samples in the total number of transitions in the
-Markov chain.
-For Markov chains with exponentially-large state spaces (with a
-circuit-based description of the transition probability structure),
-\citet{BhaBoMo11} establishes complexity-theoretic barriers for
-similar MCMC diagnostic problems.
-%A similar question was studied by \citet{BhaBoMo11} in the context of
-%Markov chains with very large state spaces.
-%They assume an circuit based description of the
-%transition probability structure of the matrix and derive several hardness
-%results for Markov chains whose state space consists of all binary strings of
-%a given length. For example,
-%they prove that given a Markov chain that mixes
-%rapidly, it is $\mathsf{SZK}$-hard to distinguish whether the chain is close to
-%stationarity by time $t$ or far from stationarity at time $ct$ for a
-%constant $c$ when started from a given initial state ($\mathsf{SZK}
-%\subset \mathsf{AM} \cap \mathsf{coAM}$, which is conjectured to
-%extend beyond $\mathsf{BPP}$ \citep{GoVa11}).\footnote{%
-%  $\mathsf{SZK}$ stands for statistical zero-knowledge, $\mathsf{BPP}$
-%  stands for probabilistic polynomial-time, $\mathsf{AM}$ stands for
-%  ``Arthur-Merlin protocol''.%
+Although these asymptotic results help us understand the limiting
+behavior of the sample mean over a Markov chain, they say little about
+the finite-time non-asymptotic behavior, which is often needed for the
+prudent evaluation of a method or even its algorithmic design
+\cite{
+MCMCDiscussion93%
+,DBLP:conf/valuetools/KontoyiannisLM06%
+,BBL06%
+,MniSzeAu08%
+,MauPo09%
+,LiLiWaSt11:KWIK%
+,flegal2011implementing%
+,Gyori-paulin15%
+,SwaJoa15:LoggedBandit%
+}. %BaHaVa10:activelearning,ShaVo08:conformalpred
+To address this need, numerous works have developed Chernoff-type
+bounds on $\Pr\parens{ |\vpi_n(f)-\vpi(f)| > \eps }$, thus providing
+invaluable tools for non-asymptotic probabilistic
+analysis~\cite{gillman1998chernoff,leon2004optimal,DBLP:conf/valuetools/KontoyiannisLM06,paulin15}.
+These probability bounds are larger than corresponding bounds for
+independent and identically distributed (iid) data due to the temporal
+dependence; intuitively, for the Markov chain to yield a ``fresh''
+draw $X_{t'}$ that is approximately ``independent'' from $X_t$, one
+must wait $\Theta(\tmix)$ time steps.
+Note that the bounds generally depend on distribution-specific
+properties of the Markov chain (e.g., $\vP$, $\tmix$), which are often
+unknown \emph{a priori} in practice.
+Consequently, much effort has been put towards estimating these
+unknown quantities, especially in the context of MCMC diagnostics, in
+order to provide data-dependent assessments of estimation
+accuracy~\cite[e.g.,][]{MCMCDiscussion93,flegal2011implementing,1209.0703,Gyori-paulin15}.
+However, these approaches generally only provide asymptotic
+guarantees, and hence fall short of our goal of empirical bounds that
+are valid with any finite-length sample path.
+
+Many results from statistical learning and empirical process theory
+have been extended to sufficiently fast mixing
+data~\citep[e.g.,][]{Yu94,MR1921877,gamarnik03,MoRo08,MoRo09,DBLP:conf/nips/SteinwartC09,Steinwart2009175},
+providing learnability assurances (e.g., generalization error bounds)
+in scenarios where data are dependent.
+These results are often given in terms of mixing coefficients, which
+can be consistently estimated in some cases~\citet{McDoShaSche11}.
+However, the convergence rates of the estimates given
+in~\citet{McDoShaSche11} are given in terms of unknown mixing
+coefficients.
+When the data comes from a Markov chain, these mixing coefficients can
+often be bounded in terms of mixing times, and hence our main results
+provide a way to make them fully empirical.
+
+It is possible to eliminate many of the difficulties presented above
+with additional access to the Markov chain.
+For example, with access to a sampling oracle that generates
+independent transitions from any given state (akin to ``reset''
+mechanism), the mixing time becomes an efficiently testable property
+in the sense studied in~\citet{BaFoRuSmiWhi00,BaFoRuSmiWhi13}.
+On the other hand, when one only has a circuit-based description of
+the transition probabilities of Markov chain over an
+exponentially-large state space, there are complexity-theoretic
+barriers for many MCMC diagnostic problems~\citet{BhaBoMo11}.
+
+%%Performance guarantees play a key role in the analysis and design of machine learning algorithms.
+%
+%In this paper we study the construction of fully empirical bounds on the mixing time of Markov chains based on a single sample path.
+%Let $(X_t)_{t=1,2,\dotsc}$ be an irreducible, aperiodic time-homogeneous Markov chain on a finite state space $[d]$ of $d$ states numbered by positive integers with transition matrix $\vP$.
+Under this assumption, the chain converges to its unique stationary \todoc{Maybe steady state is better?}
+%distribution $\vpi = (\pi_i)_{i=1}^d$ regardless of the distribution $\vq$ of $X_1$: 
+%$\lim_{t\to\infty} \Pr_{\vq}\Parens{X_t = i} = \lim_{t\to\infty} (\vq \vP^t)_i = \pi_i$, $1\le i \le d$.
+%The mixing time $\tmix$ of the Markov chain is the time that it takes for the chain to reach its stationary distribution, say,
+%\[
+%  \tmix
+%  :=
+%  \min\Braces{
+%    t \in \bbN :
+%    \sup_{\vq}
+%    \max_{A\subset [d]} |\textstyle\Pr_{\vq}\Parens{ X_t \in A } - \vpi(A)|
+%%    \Norm{
+%%      \vq \vP^t - \vpi
+%%    }_{\tv} 
+%    \leq 1/4
+%  }\,.
+%\]
+%Here $\vpi(A) = \sum_{i\in A} \pi_i$ is the total probability assigned
+%to set $A$ by $\vpi$, and the supremum is over
+%all possible initial distributions $\vq$.
+%The problem studied in the paper is the construction of a confidence
+%set $C_n = C_n(X_1,X_2,\dotsc,X_n,\delta) \subset [0,\infty]$ based on
+%the knowledge of $X_1,X_2,\dotsc,X_n$ only that captures the mixing
+%time $\tmix$ with a pre-specified probability $1-\delta \in (0,1)$. 
+%In the remainder of the introduction, we describe our motivation of studying this problem, followed by a 
+%brief description of our main results.
+%\todod{``might be interesting'' sounds weak.} \todoc{better now?}
+%
+%In numerous scientific applications and some machine learning tasks
+%one is interested in estimating the mean $\vpi(f) = \sum_i \pi_i f(i)$ of some function $f$ of the states of a Markov chain. 
+%For example, this is the realms of the celebrated Markov Chain Monte Carlo (MCMC) paradigm, but 
+%the problem also arises in performance prediction when working with time correlated data, as is common in,
+%for example, in reinforcement learning.
+%%as 
+%% (cf. \cref{sec:mot}) 
+%Under our assumptions,
+%the law of large numbers guarantees that the sample mean $\vpi_n(f) \defeq \frac1n \sum_{t=1}^n f(X_t)$ 
+%converges almost surely to $\vpi(f)$, while the central limit theorem tells us that as $n\to \infty$, 
+%the distribution of the deviation $\sqrt{n}( \vpi_n(f)-\vpi(f))$ will be normal with zero mean and asymptotic
+%variance $\sigma^2_{\mathrm{as}}(f) \defeq \lim_{n\to\infty} n \Var\Parens{ \vpi_n(f) }$. \todoc{shall we add a citation here?}
+%While these results help us to understand how the sample mean over a Markov chain behaves in the limit,
+%they tell us little about the finite time, non-asymptotic behavior of the sample mean,
+%which are needed for prudent evaluation of one's method or even for algorithm design 
+%\cite{
+%MCMCDiscussion93%
+%,DBLP:conf/valuetools/KontoyiannisLM06%
+%,BBL06%
+%,MniSzeAu08%
+%,MauPo09%
+%,LiLiWaSt11:KWIK%
+%,flegal2011implementing%
+%,Gyori-paulin15%
+%,SwaJoa15:LoggedBandit%
+%}. %BaHaVa10:activelearning,ShaVo08:conformalpred
+%To address this need, numerous works have been devoted to deriving
+%Chernoff-type, exponential bounds for upper bounding the tail probabilities $p(r) = \Pr\Parens{ |\vpi_n(f)-\vpi(f)|> r }$,
+%also known as ``concentration bounds'', e.g., 
+%\cite{gillman1998chernoff,leon2004optimal,DBLP:conf/valuetools/KontoyiannisLM06,paulin15}.
+%Compared with the case when $(X_t)_t$ is an independent, identically distributed (iid) sequence of
+%random variables, these tail probabilities must be larger due to the temporal dependence of the sequence $(X_t)_t$.
+%Intuitively, for the Markov chain to forget the state where it is at time $t$, or in other words to get a 
+%``fresh'' sample $X_{t'}$ that is approximately ``independent'' from $X_t$, one must wait $\Theta(\tmix)$ time steps.
+%Thus, in the bounds on $p(r)$, one expects to see that the sample size $n$ 
+%gets appropriately reduced as $\tmix$ grows and this is indeed the case in bounds mentioned above.
+%is replaced by $n/\tmix$, which one
+%might think of as an approximate sample size. 
+%Indeed, all the above mentioned works get bounds on $p(r)$ that depend on various functions of $\vP$.
+%Namely, Kontoyiannis et al. obtain bounds that depend on certain information-theoretic quantities defined implicitly
+%as a function of $\vP$ and $f$, which are after simplification is bounded based on quantities related to 
+%a regeneration-type quantity derived from $\vP$
+%\cite{DBLP:conf/valuetools/KontoyiannisLM06}, while the other works mentioned above, in the case of reversible chains,
+%derive bounds that depend on the spectral gap of $\vP$.
+%Note that the spectral gap for such chains is essentially inversely related to the mixing time of the chain (see \cref{sec:setting}).
+%%\cite{gillman1998chernoff,leon2004optimal,,paulin15}.
+%Much research have been devoted to finding appropriate \emph{a priori} bounds on the various quantities that appear 
+%in these bounds, mostly in the theory community (e.g., \cite{saloff2004random}). 
+%To address the lack or looseness of such a priori bounds, a significant effort also went into developing methods
+%that estimate these quantities based on data, mostly in the MCMC community in an effort to develop
+%effective methods that are able to diagnose whether a chain has reached its stationary state 
+%(e.g., \cite{MCMCDiscussion93,flegal2011implementing,Gyori-paulin15}).
+%While the typical results state the asymptotic consistency of these estimators, or a central limit theorem (which 
+%are of little help to achieve a fully empirical, non-asymptotic bound on $p(r)$),
+%Gyori and Paulin give a finite time error bound for these estimates. 
+%Unsurprisingly, these error bounds themselves
+%depend on other unknown quantities of the Markov chain (which can be bounded in terms of the mixing time, again)
+%\cite{Gyori-paulin15}.
+%
+%This situation is similar to what appears when one attempts to use a Bernstein-type deviation bound to 
+%control the deviation between the sample mean of a sequence of iid random variables from their common mean
+%with a known range, but unknown variance: For a tight bound, one wishes to use the unknown variance, but in the deviation
+%bound for the error of a variance estimate, the variance appears again.
+%While in the case of Bernstein-bounds this circular reference leads to a nontrivial bound
+%\cite{audibert2009}, it appears that the situation is different in the Markov setting.
+%Focusing on the spectral gap, we will see that if one attempts to use a similar technique,
+%the set of feasible values based on the two error bounds always will never rule out that the spectral gap is zero,
+%or that the mixing time is infinite. Can one even hope to obtain 
+%after finitely many samples a nontrivial upper bound on the mixing time?
+%Obviously, for any finite sample size, if the mixing time of the chain is larger than the sample size than a sound method
+%must return infinity as the upper bound on the mixing time. Hence, the question is whether one can design a method
+%that never fails and if the sample size is appropriately large, returns a nontrivial bound on the mixing time, adapting the time it needs to trap the mixing time in a fixed size confidence interval to the unknown mixing time.
+%
+%\hide{
+%\subsection{Motivation}
+%\label{sec:mot}
+%One key competence in statistical learning is the ability to predict one's performance on future data.
+%To make performance prediction possible, one starts by posing some assumptions on how the data arises.
+%Simpler assumptions lead to simpler algorithms and analysis, while they may also limit applicability of the
+%obtained results and algorithms.
+%The most common assumption employed postulates that all data points, including past and future data points,
+%are sampled independently of each other from a common underlying distribution (e.g., \cite{SSBD14:UnderstandingML}).
+%This is helpful on two accounts: 
+%The lack of dependence between the data points means that sample averages of bounded functions of the data 
+%will converge at a uniform rate (irrespective of the distribution), while the assumption that all data points share the same
+%distribution means that one can learn about the distribution of future data points based on past data.
 %}
-
-Other works include that of \citet{McDoShaSche11} who propose to estimate
-the $\beta$-mixing coefficients of a stationary process possessing a density
-based on histogram estimates of the underlying joint density. Their main
-result is a consistency result for a fixed ``lag''. For fast mixing processes
-(e.g., for $k$-step Markov processes) their result yields a convergence rate,
-which unfortunately is given in terms of the unknown mixing coefficients.
-%Catch-22 problem remains: The rate of convergence of their estimates depends on the unknown mixing coefficients
-%(Theorem 2.4)
-% $\exp(W(\ln(n)))$ is the window length, where $W$ is Lambert's $W$-function (i.e., this is between $\ln \ln(n)$ and $\ln n$).
-The earlier study of \citet{Nobel06} considers the generalization of
-existing discernibility results for iid processes to weakly uniformly
-ergodic families, and shows how this can be used to study---and in
-some cases estimate---polynomial mixing rates for dependent processes.
+%
+%\hide{
+%It is well-known that contrary to the idealized assumptions of the standard
+%learning model, real data is almost never iid.
+%Fortunately, many classical results of statistics and statistical
+%learning theory extend to the dependent data in a more or less
+%seamless fashion, at least when the data is sufficiently ``mixing''.
+%Such results include the central limit theorem, 
+%exponential tail bounds for empirical processes, 
+%performance guarantees for complexity regularization
+%using sieves or roughness penalties, as well as 
+%results based on stability arguments or Rademacher complexity 
+%\citep{Woo86,Dou94,Yu94,ZheChu96,dlPenaGine99,Mei00,
+%  MR1921877,vid2003,gamarnik03,LoKuScha05,MoRo08,MoRo09,
+%  DBLP:conf/nips/SteinwartC09,Steinwart2009175,
+%  FaSze11,
+%  london:nips12asalsn,london:icml13,
+%  DBLP:conf/nips/ShaliziK13}.
+%
+%Unfortunately, a distribution-free theory of non-iid learning is simply
+%impossible, when arbitrary dependence among the sample points is allowed.
+%Hence, one must quantify this dependence via the various mixing coefficients,
+%which appear in the non-iid generalization bounds mentioned above.
+%A frequently raised concern is that now the bounds depend on {\em
+%unknown} properties of the distribution, which precludes their
+%actual use in algorithmic applications (e.g., active
+%learning~\citep{BBL06,DHM07}).
+%%, and any algorithm aiming at a fully empirical error bound must
+%%also estimate the mixing coefficients.
+%In addition to learning-theoretic generalization bounds, the need for
+%fully empirical confidence bounds arises in Markov Chain Monte Carlo
+%(MCMC) simulations---so as to halt the run when sufficiently close to
+%the stationary distribution \citep{galin06}, and in Markov Reward
+%Processes \citep{sutton98}, as this will affect the learner's policy.
+%As \citet{Mei00} notes, in the absence of accurate estimates of the
+%mixing time, such fully empirical bounds remain ``a theoretician's
+%dream''.
+%
+%Our goal here is to develop an algorithm that, given a single sample
+%path of a Markov chain, computes non-trivial bounds on the mixing time
+%of the Markov chain.
+%We remark on the inherent chicken-and-egg nature of the challenge:
+%confidence intervals derived from a Markov chain---say, on the the
+%chain's mixing time---are typically given in terms of the chain's
+%mixing time.
+%Our way out of this conundrum is to give {\em empirical} (i.e.,
+%\emph{data-dependent}) as opposed to {\em a priori} guarantees.
+%To illustrate the intuition, consider a Markov chain on two states,
+%$\braces{0,1}$.
+%A learner who is stuck in state $0$ for a long time learns very little
+%about the chain and hence cannot be expected to give meaningful upper
+%bounds on its mixing time.
+%Consider, however, a learner who is lucky enough
+%to be in a chain that switches frequently between the two states.
+%Surely at some finite time, the learner will possess meaningful and nontrivial
+%information about the chain's mixing time --- crucially, without any
+%{\em prior} knowledge regarding this quantity. This simple example captures
+%the nature of our results: in some cases, we will not be able to say
+%anything meaningful (but the learner will recognize that this is the case),
+%while in ``luckier'' situations, we will be able to provide nontrivial
+%upper estimates on the mixing time.
+%}
+%
+%\paragraph{Main results.}
+%We propose what is apparently the first efficient and fully empirical
+%estimator of Markov mixing time.
+%\todod{``apparently'' sounds weak.}
+%Consider a reversible ergodic Markov chain on $d$ states with spectral
+%gap $\gap$ (which is well-known to govern mixing time)
+%and stationary distribution minorized by $\pimin$.
+%Our main results are summarized as follows.
+%\begin{enumerate}
+%  \item
+%    In \cref{thm:err}, we give an estimator for $\gap$ that
+%    achieves multiplicative accuracy from \emph{a single sample path}
+%    of length $\tilde{O}\parens{1/(\pimin\gap^3)}$.\footnote{The
+%    $\tilde{O}(\cdot)$ notation suppresses logarithmic factors.}
+%    We also provide an estimator for $\pimin$ that requires a sample
+%    path of length $\tilde{O}\parens{1/(\pimin\gap)}$.
+%    However, the valid confidence intervals suggested by
+%    \cref{thm:err} depend on the unknown quantities $\pimin$ and
+%    $\gap$.
+%
+%  \item
+%    In \cref{thm:lb-pimin,thm:lb-gap}, we show that $\Omega\parens{
+%    (d\ln d)/\gap + 1/\pimin }$ observations are necessary for
+%    constant multiplicative accuracy in estimating $\gap$.
+%
+%  \item
+%    In \cref{thm:empirical}, we develop a procedure for computing
+%    \emph{valid empirical confidence intervals} for $\pimin$ and
+%    $\gap$.
+%    The procedure uses just a single sample path and does not require
+%    any prior knowledge about $\pimin$ or $\gap$.
+%    We also quantify the rate at which the width of the confidence
+%    intervals shrink as a function of the length of the sample path.
+%
+%    We show in \cref{thm:combined} how to combine these empirical
+%    confidence intervals with the non-empirical bounds from
+%    \cref{thm:err} to produce valid empirical confidence intervals;
+%    the width of these intervals converge to zero asymptotically as
+%    fast as those from \cref{thm:err}.
+%
+%%    which is
+%%    $     O\Parens{
+%%      \frac1{\pimin}\sqrt{\frac{d\ln(d/\delta)}{ n}}
+%%      +
+%%      \frac{d}{\pimin\gap} \sqrt{\frac{\ln(d/\delta)}{\pimin n}}
+%%    }.$
+%
+%
+%    
+%%    \todoc{please add rate}
+%
+%\end{enumerate}
+%
+%
+%
+%\paragraph{Related work.} \todoc{Move this later? It also needs to be updated, to 
+%adjust it to the current intro.}
+%Owing to the wide applicability of MCMC methods to various computational
+%problems, much effort has been devoted to developing effective convergence
+%diagnostic methods the aim of deriving stopping rules or stationarity
+%diagnoses. While a large number of heuristic techniques have been developed
+%\citep[e.g.,][]{MCMCDiscussion93,CoCa96}, it is widely recognized that all
+%these methods may fail to meet the guarantees that they are designed for.
+%Some recent advance in this area \citep{1209.0703,Gyori-paulin15} have
+%yielded practical algorithms for estimating the rates of convergence
+%of various MCMC algorithms, but they fall short of providing intervals
+%that are valid with any finite length sequence.
+%
+%In the theory community, \citet{BaFoRuSmiWhi00,BaFoRuSmiWhi13}
+%developed algorithms for testing whether a given Markov chain over a
+%finite state space of size $d$ is rapidly mixing.
+%In particular, assuming access to a sampling oracle that can generate
+%independent transitions from any given state, their algorithm can
+%distinguish between chains that in $t$ steps for a fixed $t$ reach a
+%distribution close to the stationary distribution using $\tilde{O}(t
+%d^{5/3})$ oracle calls.
+%It is remarkable that when $t$ is small enough, the procedure requires
+%sublinearly many samples in the total number of transitions in the
+%Markov chain.
+%For Markov chains with exponentially-large state spaces (with a
+%circuit-based description of the transition probability structure),
+%\citet{BhaBoMo11} establishes complexity-theoretic barriers for
+%similar MCMC diagnostic problems.
+%%A similar question was studied by \citet{BhaBoMo11} in the context of
+%%Markov chains with very large state spaces.
+%%They assume an circuit based description of the
+%%transition probability structure of the matrix and derive several hardness
+%%results for Markov chains whose state space consists of all binary strings of
+%%a given length. For example,
+%%they prove that given a Markov chain that mixes
+%%rapidly, it is $\mathsf{SZK}$-hard to distinguish whether the chain is close to
+%%stationarity by time $t$ or far from stationarity at time $ct$ for a
+%%constant $c$ when started from a given initial state ($\mathsf{SZK}
+%%\subset \mathsf{AM} \cap \mathsf{coAM}$, which is conjectured to
+%%extend beyond $\mathsf{BPP}$ \citep{GoVa11}).\footnote{%
+%%  $\mathsf{SZK}$ stands for statistical zero-knowledge, $\mathsf{BPP}$
+%%  stands for probabilistic polynomial-time, $\mathsf{AM}$ stands for
+%%  ``Arthur-Merlin protocol''.%
+%%}
+%
+%Other works include that of \citet{McDoShaSche11} who propose to estimate
+%the $\beta$-mixing coefficients of a stationary process possessing a density
+%based on histogram estimates of the underlying joint density. Their main
+%result is a consistency result for a fixed ``lag''. For fast mixing processes
+%(e.g., for $k$-step Markov processes) their result yields a convergence rate,
+%which unfortunately is given in terms of the unknown mixing coefficients.
+%%Catch-22 problem remains: The rate of convergence of their estimates depends on the unknown mixing coefficients
+%%(Theorem 2.4)
+%% $\exp(W(\ln(n)))$ is the window length, where $W$ is Lambert's $W$-function (i.e., this is between $\ln \ln(n)$ and $\ln n$).
+%The earlier study of \citet{Nobel06} considers the generalization of
+%existing discernibility results for iid processes to weakly uniformly
+%ergodic families, and shows how this can be used to study---and in
+%some cases estimate---polynomial mixing rates for dependent processes.
+%
+%

### 2015-06-02 setting.tex
intro and other stuff

@@
 can ultimately be bounded in terms of $\pimin$ and
-$\gap$, as we will show in the proof of our first result.\todoc{I removed the citations (saving space), referring to the proof.}
+$\gap$, as we will show in the proof of our first result.
+\todoc{I removed the citations (saving space), referring to the proof.}
 %~\citep[e.g.,][]{MR1921877,MoRo08,MoRo09}. 

### 2015-06-02 intro-nips.tex
intro and other stuff

@@
 For example, with access to a sampling oracle that generates
-independent transitions from any given state (akin to ``reset''
-mechanism), the mixing time becomes an efficiently testable property
-in the sense studied in~\citet{BaFoRuSmiWhi00,BaFoRuSmiWhi13}.
+independent transitions from any given state (akin to a ``reset''
+device), the mixing time becomes an efficiently testable property in
+the sense studied in~\citet{BaFoRuSmiWhi00,BaFoRuSmiWhi13}.
 On the other hand, when one only has a circuit-based description of

### 2015-06-02 rates.tex
intro and other stuff

@@
 path} from a (possibly) non-stationary Markov chain, we cannot use
-standard large deviation bounds; moreover applying Chernoff-Hoeffding
+standard large deviation bounds; moreover applying Chernoff-type
 bounds for Markov chains to each entry of $\wh{\vM}$ would result in a
@@
 described in the article of~\citet{Yu94}.
-The proof is given in \cref{app:upper}.
+Due to ergodicity, the convergence rate can be bounded without any
+dependence on the initial state distribution $\vpi^{(1)}$.
+The proof of \cref{thm:err} is given in \cref{app:upper}.
 

### 2015-06-02 setting.tex
intro and other stuff

@@
 $\vP$, and hence it follows that $\vL$ is also symmetric.
-As we will see later, we will strongly exploit this later.  \todoc{Say more? Maybe not..}
+(We will strongly exploit the symmetry in our results.)
+\todoc{Say more? Maybe not..}
 %will be advantageous due to the existence of stronger perturbation results
@@
 Let $\slem := \max\{ \lambda_2,\, |\lambda_d| \}$, and define the
-(absolute) spectral gap to be $\gap := 1-\slem$, which is strictly positive.
+(absolute) spectral gap to be $\gap := 1-\slem$, which is strictly
+positive on account of ergodicity.
 \todoc{Do our results generalize to the spectral gap? Where do we need that we talk about the absolute spectral gap?
@@
 
-Let $(X_t)_{t\in\bbN}$ be a Markov chain whose transitions are governed by
-$\vP$.
+Let $(X_t)_{t\in\bbN}$ be a Markov chain whose transition
+probabilities are governed by $\vP$.
 For each $t \in \bbN$, let $\vpi^{(t)} \in \Delta^{d-1}$ denote the
@@
 can ultimately be bounded in terms of $\pimin$ and
-$\gap$, as we will show in the proof of our first result.
+$\gap$ (as we will show in the proof of our first result).
 \todoc{I removed the citations (saving space), referring to the proof.}

### 2015-06-02 empirical.tex
stuff

@@
 $\giAh$~\citep[Theorem 5.3]{meyer1975role}.
-The group inverse will also be used later to compute the sensitivity
-of $\vpi$ based on a result of Cho and Meyer \cite{cho2001comparison}.
+The group inverse is also be used to compute the sensitivity of
+$\vpi$.
 Based on $\hat\vpi$ and $\wh\vP$, we construct the plug-in estimate
@@
 Both analyses give error bounds entirely in terms of observable
-quantities, tracing back to empirical error bounds for the smoothed
-frequency estimates of $\vP$.
-Note that the most expensive step in the algorithm is the computation of the group inverse,
-which, as noted reduces to matrix inversion.
-Thus, with a standard implementation of matrix inversion, the algorithm's complexity becomes 
-$O( n + d^3)$, while its space complexity is $O(d^2)$.
+quantities (e.g., $\hat\kappa$), tracing back to empirical error
+bounds for the smoothed frequency estimates of $\vP$.
+
+The most computationally expensive step in \cref{alg:empest} is the
+computation of the group inverse $\giAh$, which, as noted reduces to
+matrix inversion.
+Thus, with a standard implementation of matrix inversion, the
+algorithm's time complexity is $O(n + d^3)$, while its space
+complexity is $O(d^2)$.
 \todoc{Implement and test?}
@@
 
-The main result of this section is the following result:
+To state our main theorem concerning \cref{alg:empest}, we first define
+$\kappa$ to be analogous to $\hat{\kappa}$ from Step~\ref{step:sens},
+with $\giAh$ replaced by the group inverse $\giA$ of $\vA := \vI -
+\vP$.
+The result is as follows.
 \begin{theorem}
@@
     }
-    ,
-  \]
-  where $\kappa$ is analogous to the $\hat{\kappa}$ computed
-  by~\cref{alg:empest} with $\giAh$ replaced by the group inverse
-  $\giA$ of $\vA := \vI - \vP$.\footnote{%
+    .\footnote{%
     In \cref{thm:empirical,thm:combined}, our use of big-$O$ notation
@@
     For a random sequence $(Y_n)_n$ and a (non-random) positive
-    sequence $(a_{\theta,n})_n$ parameterized by $\theta$, we say
-    ``$Y_n = O(a_{\theta,n})$ holds almost surely as $n\to\infty$'' if
-    there is some universal constant $C>0$ such that for all $\theta$,
-    $\limsup_{n\to\infty} Y_n/a_{\theta,n} \leq C$ holds almost
-    surely.%
+    sequence $(\veps_{\theta,n})_n$ parameterized by $\theta$, we say
+    ``$Y_n = O(\veps_{\theta,n})$ holds almost surely as
+    $n\to\infty$'' if there is some universal constant $C>0$ such that
+    for all $\theta$, $\limsup_{n\to\infty} Y_n/\veps_{\theta,n} \leq
+    C$ holds almost surely.%
   }%
+  \]
 %  \todoc{

### 2015-06-02 abstract.tex
stuff

@@
 time of a reversible ergodic Markov chain.
-The procedure computes the estimate from a single sample path of the
-Markov chain, and does not require any \emph{a priori} bounds on the
-mixing time itself.
+The procedure computes the estimate from a single finite-length sample
+path from the Markov chain, and does not require any \emph{a priori}
+bounds on the mixing time itself.
 This stands in contrast to previous approaches, which either require a

### 2015-06-02 intro-nips.tex
stuff

@@
 order to provide data-dependent assessments of estimation
-accuracy~\cite[e.g.,][]{MCMCDiscussion93,flegal2011implementing,1209.0703,Gyori-paulin15}.
+accuracy~\cite[e.g.,][]{MCMCDiscussion93,jones2001,flegal2011implementing,1209.0703,Gyori-paulin15}.
 However, these approaches generally only provide asymptotic

### 2015-06-02 abstract.tex
.

@@
 $1-\delta$; and the width of the interval converges to zero at a
-$\sqrt{n}$-rate, where $n$ is the length of the sample path.
+$\sqrt{n}$ rate, where $n$ is the length of the sample path.
 

### 2015-06-02 intro-nips.tex
.

@@
 can be consistently estimated in some cases~\citet{McDoShaSche11}.
-However, the convergence rates of the estimates given
-in~\citet{McDoShaSche11} are given in terms of unknown mixing
-coefficients.
+However, the convergence rates of the estimates
+from~\citet{McDoShaSche11}, which are needed to derive confidence
+bounds, are given in terms of unknown mixing coefficients.
 When the data comes from a Markov chain, these mixing coefficients can

### 2015-06-02 intro-nips.tex
.

@@
 independent and identically distributed (iid) data due to the temporal
-dependence; intuitively, for the Markov chain to yield a ``fresh''
-draw $X_{t'}$ that is approximately ``independent'' from $X_t$, one
-must wait $\Theta(\tmix)$ time steps.
+dependence; intuitively, for the Markov chain to yield a fresh draw
+$X_{t'}$ that behaves as if it was independent of $X_t$, one must wait
+$\Theta(\tmix)$ time steps.
 Note that the bounds generally depend on distribution-specific

### 2015-06-02 abstract.tex
Mostly abstract/intro
Making abstract more precise.
New reference.
\ln -> \log replacement (should be the other way around?)
Changed some section titles

--- 2015-06-02 abstract.tex
@@
 %!TEX root =  matrix-est.tex
-This article provides an efficient procedure for estimating the mixing
-time of a reversible ergodic Markov chain.
-The procedure computes the estimate from a single sample path of the
-Markov chain, and does not require any \emph{a priori} bounds on the
-mixing time itself.
-This stands in contrast to previous approaches, which either require a
-reset mechanism or require additional prior knowledge.
-Most notably, the procedure produces a fully empirical confidence
-interval (i.e., it depends only on the sample path and a given
-confidence parameter $\delta$) that is always valid with probability
-$1-\delta$; and the width of the interval converges to zero at a
-$\sqrt{n}$-rate, where $n$ is the length of the sample path.
+This article provides the first efficient, non-heuristic procedure for computing a fully data dependent interval
+that traps the mixing time of a finite reversible ergodic Markov chain with a prescribed probability.
+The interval is computed from a single finite-length sample
+path from the Markov chain, and does not require the 
+knowledge of any parameters of the chain.
+%bounds on the mixing time itself.
+This stands in contrast to previous approaches, which either 
+only provide point estimates, or require a
+reset mechanism, or additional prior knowledge.
+%Most notably, 
+%the procedure produces a fully empirical confidence
+%interval (i.e., it depends only on the sample path and a given
+%confidence parameter $\delta$) that is always valid with probability
+%$1-\delta$; and 
+The width of the interval converges to zero at a
+$\sqrt{n}$ rate, where $n$ is the length of the sample path.
+Upper and lower bounds are given on the number of samples required
+to achieve a constant accuracy estimate in terms of the chain's parameters,
+in particular, based on the absolute spectral gap $\gap$.
+\todoc{We may need to add some explanation to connect $\gap$ estimates
+to $\tmix$ estimates.}
+Our lower bounds indicate that, 
+unless further restrictions are placed on the chain,
+no procedure can achieve a constant
+accuracy before seeing each state at least  $\Omega(1/\gap)$ times.
+Future directions of research are identified.
+\todoc{Such as: closing the gap. Extension to large state spaces, 
+based on additional assumptions. Even if $\gap$ is large, we need many samples.
+Maybe we need even stronger assumptions, like the decay rate of eigenvalues.
+}
+%If $\gap$ denotes the absolute spectral gap of the unknown Markov chain
+%and the chain's stationary distribution is minorized by $\pimin$,
+%our procedure requires $\tilde{O}( \frac{1}{\pimin\gap^3})$ observations
+% to estimate the mixing time up to a constant multiplicative accuracy, 
+%while we also establish that any algorithm to achieve the same
+%goal would need $\Omega( d\log(d)/\gap + 1/\pimin)$ samples.
+
+%and the dependence on the main  chain's
 

### 2015-06-02 intro-nips.tex
Mostly abstract/intro
Making abstract more precise.
New reference.
\ln -> \log replacement (should be the other way around?)
Changed some section titles

@@
 for the chain to (approximately) reach its stationary distribution,
-\[
+\begin{align}
+\label{eq:mixtimedef}
   \tmix
@@
   }\,.
-\]
+\end{align}
 Here, $\vpi(A) = \sum_{i\in A} \pi_i$ is the probability assigned to
@@
 \paragraph{Main results.}
-We develop the first efficient procedure for constructing non-trivial
+We develop the first efficient \todoc{Maybe leave out efficient? 
+What is efficient is a question of interpretation. We 
+should maybe rather add a note on the computational complexity.}
+procedure for constructing non-trivial
 and fully empirical confidence intervals for Markov mixing time.
 Consider a reversible ergodic Markov chain on $d$ states with absolute
-spectral gap $\gap$ and stationary distribution minorized by $\pimin$.
-(These quantities $\gap$ and $\pimin$ are well-known to govern
-mixing time.)
+spectral gap $\gap$ and stationary distribution minorized by $\pimin$. \todoc{Comment on reversibility somewhere?}
+As is well known \citep[Theorems~12.3 and~12.4]{LePeWi08},
+$(\frac{1}\gap  - 1)\log(2)  \le \tmix \le \frac{1}\gap \log \frac4\pimin$,
+hence it suffices to estimate $\gap$ and $\pimin$. 
+\todoc{I would prefer $\ln$ over $\log$ throughout the paper.}
+%these quantities $\gap$ and $\pimin$ are well-known to govern
+%mixing time.
 Our main results are summarized as follows.
@@
   \item
-    In \cref{thm:err}, we give an estimator for $\gap$ that
+    In \cref{thm:err}, we give a point-estimator for $\gap$ that
     achieves multiplicative accuracy from \emph{a single sample path}
@@
     $\tilde{O}(\cdot)$ notation suppresses logarithmic factors.}
-    We also provide an estimator for $\pimin$ that requires a sample
+    We also provide a point-estimator for $\pimin$ that requires a sample
     path of length $\tilde{O}\parens{1/(\pimin\gap)}$.
@@
     estimating $\gap$.
-    Essentially, every state may need to be visited about
+    Essentially, \emph{every} state may need to be visited about
     $\log(d)/\gap$ times, on average, before an accurate estimate of
-    the mixing time can be provided.
+    the mixing time can be provided, regardless of what
+    estimation procedure one uses.
 
@@
     The procedure uses just a single sample path and does not require
-    any prior knowledge about $\pimin$ or $\gap$.
+    any prior knowledge of the chain's parameters, such as $\pimin$ or $\gap$.
     We also quantify the rate at which the width of the confidence
@@
     fast as those from \cref{thm:err}.
+    \todoc{Is it true from the construction of \cref{thm:combined}
+    that if we had a better point estimator
+    then we would automatically get an interval estimator
+    with the same accuracy?
+    This seems to be true.
+    This would mean that to close the gap
+    one should really work on coming up with better point estimators.
+    }
 
@@
 %    $     O\Parens{
-%      \frac1{\pimin}\sqrt{\frac{d\ln(d/\delta)}{ n}}
+%      \frac1{\pimin}\sqrt{\frac{d\log(d/\delta)}{ n}}
 %      +
-%      \frac{d}{\pimin\gap} \sqrt{\frac{\ln(d/\delta)}{\pimin n}}
+%      \frac{d}{\pimin\gap} \sqrt{\frac{\log(d/\delta)}{\pimin n}}
 %    }.$
@@
 Note that the bounds generally depend on distribution-specific
-properties of the Markov chain (e.g., $\vP$, $\tmix$), which are often
+properties of the Markov chain (e.g., $\vP$, $\tmix$, $\gap$), which are often
 unknown \emph{a priori} in practice.
@@
 order to provide data-dependent assessments of estimation
-accuracy~\cite[e.g.,][]{MCMCDiscussion93,jones2001,flegal2011implementing,1209.0703,Gyori-paulin15}.
+accuracy~\cite[e.g.,][]{MCMCDiscussion93,GaSmi00:eigval,jones2001,flegal2011implementing,1209.0703,Gyori-paulin15}.
 However, these approaches generally only provide asymptotic
@@
 
+Learning with dependent data is another main motivation to our work.
 Many results from statistical learning and empirical process theory
-have been extended to sufficiently fast mixing
+have been extended to sufficiently fast mixing, dependent
 data~\citep[e.g.,][]{Yu94,MR1921877,gamarnik03,MoRo08,MoRo09,DBLP:conf/nips/SteinwartC09,Steinwart2009175},
-providing learnability assurances (e.g., generalization error bounds)
-in scenarios where data are dependent.
+providing learnability assurances (e.g., generalization error bounds).
 These results are often given in terms of mixing coefficients, which
@@
 often be bounded in terms of mixing times, and hence our main results
-provide a way to make them fully empirical.
+provide a way to make them fully empirical, at least in the limited setting we study.
 
@@
 barriers for many MCMC diagnostic problems~\citet{BhaBoMo11}.
-
+\todoc{Explain the organization of the paper?}
 %%Performance guarantees play a key role in the analysis and design of machine learning algorithms.
@@
 %    In \cref{thm:lb-pimin,thm:lb-gap}, we show that $\Omega\parens{
-%    (d\ln d)/\gap + 1/\pimin }$ observations are necessary for
+%    (d\log d)/\gap + 1/\pimin }$ observations are necessary for
 %    constant multiplicative accuracy in estimating $\gap$.
@@
 %%    $     O\Parens{
-%%      \frac1{\pimin}\sqrt{\frac{d\ln(d/\delta)}{ n}}
+%%      \frac1{\pimin}\sqrt{\frac{d\log(d/\delta)}{ n}}
 %%      +
-%%      \frac{d}{\pimin\gap} \sqrt{\frac{\ln(d/\delta)}{\pimin n}}
+%%      \frac{d}{\pimin\gap} \sqrt{\frac{\log(d/\delta)}{\pimin n}}
 %%    }.$
@@
 %%(Theorem 2.4)
-%% $\exp(W(\ln(n)))$ is the window length, where $W$ is Lambert's $W$-function (i.e., this is between $\ln \ln(n)$ and $\ln n$).
+%% $\exp(W(\log(n)))$ is the window length, where $W$ is Lambert's $W$-function (i.e., this is between $\log \log(n)$ and $\log n$).
 %The earlier study of \citet{Nobel06} considers the generalization of

### 2015-06-02 notation.tex
adding and resolving todos, as per our discussion.
bib file improvements
generalized inverse clean up, both in appendix and in main text
ideas for future work in conclusion
empirical results: comment on O(.) notation
minor tweak to intro
Mostly abstract/intro
Making abstract more precise.
New reference.
\ln -> \log replacement (should be the other way around?)
Changed some section titles

@@
 the set of the first $d$ positive integers $\{1,2,\dotsc,d\}$ by $\iset{d}$.
-The non-negative part of a real number $x$ is $[x]_+ := \max\{0,x\}$.
+The non-negative part of a real number $x$ is $[x]_+ := \max\{0,x\}$,
+$\log$ denotes the natural logarithm.
 Boldface symbols are used for vectors and matrices (e.g., $\vv$,
@@
 $\Delta^{d-1}$ as row vectors.
-Finally, the total variation between two probability distributions
-$\vp, \vq \in \Delta^{d-1}$ is denoted by $\norm{\vp-\vq}_{\tv}$.
+%Finally, the total variation between two probability distributions
+%$\vp, \vq \in \Delta^{d-1}$ is denoted by $\norm{\vp-\vq}_{\tv}$.

### 2015-06-02 rates.tex
section title

@@
 In this section, we present upper and lower bounds on achievable rates
-for estimating the spectral gap as a funciton of the length of the
+for estimating the spectral gap as a function of the length of the
 sample path $n$.
 
-\subsection{Upper bounds}
+\subsection{A plug-in based point estimator and its accuracy}
 \label{sec:rates-upper}
@@
 plug-in estimates.
-(We also provide an estimator for the minimum stationary probability
-in the process.)
+We also provide an estimator for the minimum stationary probability
+in the process, so that one can use the earlier stated bounds to trap
+the mixing time.
 
@@
 
-\todoc{Make this paragraph into a remark?}
+\todoc{Turn this paragraph into a remark?}
 Note that because the eigenvalues of $\vL$ are the same as that of the
@@
   Consider any estimator $\hatgap$ that takes as input a random sample
-  path of length $n < c d\ln(d) / \bar\gamma$ from a $d$-state
+  path of length $n < c d\log(d) / \bar\gamma$ from a $d$-state
   reversible Markov chain starting from any desired initial state

### 2015-06-02 upperbound.tex
ln -> log

@@
 
-% \ln\Parens{\frac{d}\delta\sqrt{\frac{2}{\pimin}}}
+% \log\Parens{\frac{d}\delta\sqrt{\frac{2}{\pimin}}}
 
@@
 \,\frac{
-\ln\Parens{\frac{d}\delta\sqrt{\frac{2}{\pimin}}}
+\log\Parens{\frac{d}\delta\sqrt{\frac{2}{\pimin}}}
   }{\gap n}}
@@
 \frac{20
-  \ln\Parens{\frac{d}\delta\sqrt{\frac{2}{\pimin}}}
+  \log\Parens{\frac{d}\delta\sqrt{\frac{2}{\pimin}}}
 }{\gap n}
@@
 %+
-%\sqrt{\frac2{\pimin\gap^2n}\ln(2d/\delta)}
+%\sqrt{\frac2{\pimin\gap^2n}\log(2d/\delta)}
 \le
 23 \sqrt{  \pi_i 
-	\frac{\ln\Parens{\frac{d}\delta\sqrt{\frac{2}{\pimin}}}
+	\frac{\log\Parens{\frac{d}\delta\sqrt{\frac{2}{\pimin}}}
            }{\gap n}
@@
 n \ge
-	\frac{\ln\Parens{\frac{d}\delta\sqrt{\frac{2}{\pimin}}}
+	\frac{\log\Parens{\frac{d}\delta\sqrt{\frac{2}{\pimin}}}
            }{\gap \pimin}\,.
@@
 \end{proof}
-By this lemma, as soon as $n\ge (11.5)^2 \frac{\ln( \frac{d}{\delta} )+\frac12\ln(\frac2\pimin) }{ \gap\pimin}$,
+By this lemma, as soon as $n\ge (11.5)^2 \frac{\log( \frac{d}{\delta} )+\frac12\log(\frac2\pimin) }{ \gap\pimin}$,
 then
 \[
-\Abs{\pimin - \hatpimin} \le 23 \sqrt{ \pimin \frac{\ln( \frac{d}{\delta} )+\frac12\ln(\frac2\pimin) }{ \gap n} }\,,
+\Abs{\pimin - \hatpimin} \le 23 \sqrt{ \pimin \frac{\log( \frac{d}{\delta} )+\frac12\log(\frac2\pimin) }{ \gap n} }\,,
 \]
@@
     \frac{C}{\pimin}
-      \sqrt{\frac{\ln(d/\delta)\ln\frac{n}{\delta\pimin}}{n\pimin\gap}} \,.
+      \sqrt{\frac{\log(d/\delta)\log\frac{n}{\delta\pimin}}{n\pimin\gap}} \,.
   \end{equation*}
@@
 As long as $n = \Omega \Parens{
-	\frac{\ln\Parens{\frac{d}\delta\sqrt{\frac{2}{\pimin}}}
+	\frac{\log\Parens{\frac{d}\delta\sqrt{\frac{2}{\pimin}}}
            }{\gap \pimin} }$, \todoc{First condition on $n$.}
@@
 23 \sqrt{  
-	\frac{\ln\Parens{\frac{d}\delta\sqrt{\frac{2}{\pimin}}}
+	\frac{\log\Parens{\frac{d}\delta\sqrt{\frac{2}{\pimin}}}
            }{\pimin\gap n}
@@
 \label{eq:singletonbound}
-\norm{\errp} \le C \sqrt{ \frac{\ln\Parens{\frac{d}\delta} + \ln \frac{1}{\pimin} }{\pimin\gap n}} \,.
+\norm{\errp} \le C \sqrt{ \frac{\log\Parens{\frac{d}\delta} + \log \frac{1}{\pimin} }{\pimin\gap n}} \,.
 \end{align}
@@
   \label{eq:block-length}
-  |F| = a \geq 1 + \frac{1}{\gap} \ln \frac1{\pimin\gap}
+  |F| = a \geq 1 + \frac{1}{\gap} \log \frac1{\pimin\gap}
   .
@@
   a \geq
-   1 + \frac{1}{2\gap} \ln \frac1{\pimin}\,.
+   1 + \frac{1}{2\gap} \log \frac1{\pimin}\,.
 %   \geq
-%   1 + \frac{1}{\gap} \ln \frac1{|H_s|\sqrt{\pimin}\gap}
+%   1 + \frac{1}{\gap} \log \frac1{|H_s|\sqrt{\pimin}\gap}
 \end{equation}
@@
 we can derive that it is enough
-if $a\ge 1 + \frac{1}{\gap} \ln \frac1{|H_s|\sqrt{\pimin}\gap}$,
+if $a\ge 1 + \frac{1}{\gap} \log \frac1{|H_s|\sqrt{\pimin}\gap}$,
 which is better when $|H_s|$ is bigger than $\gap$.
 }
-\todoc{As we need $a\le n/2$, we will need $n = \Omega(1/\gap\ln\frac1{\pimin})$.}
+\todoc{As we need $a\le n/2$, we will need $n = \Omega(1/\gap\log\frac1{\pimin})$.}
 
@@
 \[
-\norm{\vS} \le \sqrt{2 v \ln \frac{d_1+d_2}{\delta}} + \frac{2R}{3} \ln\frac{d_1+d_2}{\delta}\,.
+\norm{\vS} \le \sqrt{2 v \log \frac{d_1+d_2}{\delta}} + \frac{2R}{3} \log\frac{d_1+d_2}{\delta}\,.
 \]
@@
 %  \label{eq:block-length}
-%  |F| \geq 1 + \frac{1}{\gap} \ln \frac1{\pimin\gap}
+%  |F| \geq 1 + \frac{1}{\gap} \log \frac1{\pimin\gap}
 %  .
@@
     }
-    \ln(4d/\delta)
+    \log(4d/\delta)
   }
@@
   }
-  \ln(4d/\delta)
+  \log(4d/\delta)
 \end{align*}
@@
   \sqrt{
-    \frac{12\ln(4d/\delta)}{\mu_H\pimin}
+    \frac{12\log(4d/\delta)}{\mu_H\pimin}
   }
   +
-  \frac{2\ln(4d/\delta)}{\mu_H\pimin}
+  \frac{2\log(4d/\delta)}{\mu_H\pimin}
   \,.
@@
   \label{eq:block-length2}
-  a \geq 1 + \frac{1}{\gap}\ln\frac{n}{\delta\pimin}
+  a \geq 1 + \frac{1}{\gap}\log\frac{n}{\delta\pimin}
   ,
@@
 \[
-  a := \left\lceil 1 + \frac{1}{\gap}\ln\frac{2n}{\pimin\delta} \right\rceil
+  a := \left\lceil 1 + \frac{1}{\gap}\log\frac{2n}{\pimin\delta} \right\rceil
 \]
@@
   2\sqrt{
-    \frac{12\ln(16d/\delta)}{\mu\pimin}
+    \frac{12\log(16d/\delta)}{\mu\pimin}
   }
   +
-  \frac{4\ln(16d/\delta)}{\mu\pimin}
+  \frac{4\log(16d/\delta)}{\mu\pimin}
   + 
@@
 \sqrt{
-    \frac{\ln(d/\delta)\ln(\frac{n}{\delta\pimin})}{\pimin\gap n}
+    \frac{\log(d/\delta)\log(\frac{n}{\delta\pimin})}{\pimin\gap n}
   }
@@
 \sqrt{
-    \frac{\ln(d/\delta)\ln(\frac{n}{\delta\pimin})}{\pimin\gap n}
+    \frac{\log(d/\delta)\log(\frac{n}{\delta\pimin})}{\pimin\gap n}
   }\,.
@@
 Consider an event when both~\eqref{eq:singletonbound} and \eqref{eq:doubletbound} hold.
-Then, using that $\ln(d/\delta) + \ln(1/\pimin) \le \ln(d/\delta)\ln(\frac{n}{\delta\pimin})$, we get
+Then, using that $\log(d/\delta) + \log(1/\pimin) \le \log(d/\delta)\log(\frac{n}{\delta\pimin})$, we get
 \begin{align*}
@@
 \sqrt{
-    \frac{\ln(d/\delta)\ln(\frac{n}{\delta\pimin})}{\pimin\gap n}
+    \frac{\log(d/\delta)\log(\frac{n}{\delta\pimin})}{\pimin\gap n}
   }\,,
@@
 \[
-  a := \left\lceil 1 + \frac{1}{\gap}\ln\frac{n}{\pimin\delta} \right\rceil
+  a := \left\lceil 1 + \frac{1}{\gap}\log\frac{n}{\pimin\delta} \right\rceil
 \]
@@
   C
-  \sqrt{\frac{\ln(d/\delta)\ln\frac{n}{\delta\pimin}}{n\pimin\gap}}
+  \sqrt{\frac{\log(d/\delta)\log\frac{n}{\delta\pimin}}{n\pimin\gap}}
   +
@@
   \Parens{
-    \frac{\ln(d/\delta)\ln\frac{n}{\delta\pimin}}{n\pimin\gap}
+    \frac{\log(d/\delta)\log\frac{n}{\delta\pimin}}{n\pimin\gap}
   }^3
@@
   C
-  \sqrt{\frac{\ln(d/\delta)\ln\frac{n}{\delta\pimin}}{n\pimin\gap}}
+  \sqrt{\frac{\log(d/\delta)\log\frac{n}{\delta\pimin}}{n\pimin\gap}}
   +
   C
-  \frac{\ln(d/\delta)\ln\frac{n}{\delta\pimin}}{n\pimin\gap}
+  \frac{\log(d/\delta)\log\frac{n}{\delta\pimin}}{n\pimin\gap}
   .
@@
   \max\Braces{
-    \sqrt{\frac{\ln(d/\delta)\ln\frac{n}{\delta\pimin}}{n\pimin\gap}}
+    \sqrt{\frac{\log(d/\delta)\log\frac{n}{\delta\pimin}}{n\pimin\gap}}
     ,\,
     \Parens{
-      \frac{\ln(d/\delta)\ln\frac{n}{\delta\pimin}}{n\pimin\gap}
+      \frac{\log(d/\delta)\log\frac{n}{\delta\pimin}}{n\pimin\gap}
     }^3

### 2015-06-02 intro-nips.tex
minor

@@
 spectral gap $\gap$ and stationary distribution minorized by $\pimin$. \todoc{Comment on reversibility somewhere?}
-As is well known \citep[Theorems~12.3 and~12.4]{LePeWi08},
-$(\frac{1}\gap  - 1)\log(2)  \le \tmix \le \frac{1}\gap \log \frac4\pimin$,
-hence it suffices to estimate $\gap$ and $\pimin$. 
+As is well-known \citep[Theorems~12.3 and~12.4]{LePeWi08},
+\[
+  \Parens{\frac1{\gap} - 1} \ln2
+  \ \leq \ \tmix
+  \ \leq \ \frac1{\gap} \log \frac4{\pimin}
+  ,
+\]
+%$(\frac{1}\gap  - 1)\log(2)  \le \tmix \le \frac{1}\gap \log \frac4\pimin$,
+and hence it suffices to estimate $\gap$ and $\pimin$. 
 \todoc{I would prefer $\ln$ over $\log$ throughout the paper.}
@@
 There is a vast statistical literature on estimation in Markov chains.
-For instance, it is know that under the assumptions on the Markov
-chain $(X_t)_t$ stated above, the law of large numbers guarantees that
+For instance, it is known that under the assumptions on
+$(X_t)_t$ from above, the law of large numbers guarantees that
 the sample mean $\vpi_n(f) \defeq \frac1n \sum_{t=1}^n f(X_t)$
@@
 On the other hand, when one only has a circuit-based description of
-the transition probabilities of Markov chain over an
+the transition probabilities of a Markov chain over an
 exponentially-large state space, there are complexity-theoretic

### 2015-06-02 setting.tex
minor

@@
 %!TEX root =  matrix-est.tex
-\label{sec:setting}
 Let $\vP \in (\Delta^{d-1})^d \subset [0,1]^{d \times d}$ be a $d
@@
 any unobservable quantities, including $\pimin$ and $\gap$ themselves.
-It is well-known that the \emph{mixing time} of the Markov chain $\tmix$ (defined in the introduction)
+As mentioned in the introduction,
+it is well-known that the \emph{mixing time} of the Markov chain
+$\tmix$ (defined in Eq.~\ref{eq:mixtimedef})
+is bounded in terms of $\pimin$ and $\gap$, as shown in
+\cref{eq:mixing-time-bound}.
 \hide{,
@@
   ,
-\]}
+\]
 can be bounded in terms of $\pimin$ and $\gap$~\citep[Theorems~12.3
@@
   .
-\]
+\]}%
 Moreover, convergence rates for empirical processes on Markov chain

### 2015-06-02 abstract.tex
minor

@@
 to $\tmix$ estimates.}
-Our lower bounds indicate that, 
+The lower bounds indicate that, 
 unless further restrictions are placed on the chain,

### 2015-06-02 rates.tex
minor

@@
 We also provide an estimator for the minimum stationary probability
-in the process, so that one can use the earlier stated bounds to trap
+in the process, so that one can use the bounds from
+\cref{eq:mixing-time-bound} to trap
 the mixing time.

### 2015-06-02 abstract.tex
abstract

@@
 %!TEX root =  matrix-est.tex
-This article provides the first efficient, non-heuristic procedure for computing a fully data dependent interval
-that traps the mixing time of a finite reversible ergodic Markov chain with a prescribed probability.
+This article provides the first procedure for computing a fully
+data-dependent interval that traps the mixing time $\tmix$ of a finite
+reversible ergodic Markov chain at a prescribed confidence level.
 The interval is computed from a single finite-length sample
@@
 $\sqrt{n}$ rate, where $n$ is the length of the sample path.
-Upper and lower bounds are given on the number of samples required
-to achieve a constant accuracy estimate in terms of the chain's parameters,
-in particular, based on the absolute spectral gap $\gap$.
+Upper and lower bounds are given on the number of samples required to
+achieve constant-factor multiplicative accuracy.
 \todoc{We may need to add some explanation to connect $\gap$ estimates
 to $\tmix$ estimates.}
-The lower bounds indicate that, 
-unless further restrictions are placed on the chain,
-no procedure can achieve a constant
-accuracy before seeing each state at least  $\Omega(1/\gap)$ times.
+The lower bounds indicate that, unless further restrictions are placed
+on the chain, no procedure can achieve this accuracy level before
+seeing each state at least $\Omega(\tmix)$ times.
 Future directions of research are identified.

### 2015-06-02 intro-nips.tex
minor
abstract

@@
 \]
-The mixing time $\tmix$ of the Markov chain is the time that it takes
-for the chain to (approximately) reach its stationary distribution,
+The \emph{mixing time} $\tmix$ of the Markov chain is the time that it
+takes for the chain to (approximately) reach its stationary
+distribution,
 \begin{align}
@@
 \paragraph{Main results.}
-We develop the first efficient \todoc{Maybe leave out efficient? 
+We develop the first \todoc{Maybe leave out efficient? 
 What is efficient is a question of interpretation. We 
@@
 As is well-known \citep[Theorems~12.3 and~12.4]{LePeWi08},
-\[
+\begin{equation}
+  \label{eq:mixing-time-bound}
   \Parens{\frac1{\gap} - 1} \ln2
@@
   ,
-\]
+\end{equation}
 %$(\frac{1}\gap  - 1)\log(2)  \le \tmix \le \frac{1}\gap \log \frac4\pimin$,

### 2015-06-02 abstract.tex
abstract

@@
 on the chain, no procedure can achieve this accuracy level before
-seeing each state at least $\Omega(\tmix)$ times.
+seeing each state at least $\Omega(\tmix)$ times on the average.
 Future directions of research are identified.

### 2015-06-02 intro-nips.tex
main results road map

@@
   \ \leq \ \tmix
-  \ \leq \ \frac1{\gap} \log \frac4{\pimin}
+  \ \leq \ \frac1{\gap} \ln \frac4{\pimin}
   ,
@@
   \item
-    In \cref{thm:err}, we give a point-estimator for $\gap$ that
-    achieves multiplicative accuracy from \emph{a single sample path}
-    of length $\tilde{O}\parens{1/(\pimin\gap^3)}$.\footnote{The
+    In \cref{sec:rates-upper}, we give a point-estimator for $\gap$,
+    and prove in \cref{thm:err} that it achieves multiplicative
+    accuracy from \emph{a single sample path} of length
+    $\tilde{O}\parens{1/(\pimin\gap^3)}$.\footnote{The
     $\tilde{O}(\cdot)$ notation suppresses logarithmic factors.}
@@
     $\gap$.
+    We also discuss the importance of reversibility, and some possible
+    extensions to non-reversible chains.
 
   \item
-    In \cref{thm:lb-pimin,thm:lb-gap}, we show that $\Omega\parens{
-    (d\log d)/\gap + 1/\pimin}$ observations are necessary for any
-    procedure to guarantee constant multiplicative accuracy in
-    estimating $\gap$.
+    In \cref{sec:rates-lower}, we show in
+    \cref{thm:lb-pimin,thm:lb-gap} that $n = \Omega\parens{ (d\log
+    d)/\gap + 1/\pimin}$ observations are necessary for any procedure
+    to guarantee constant multiplicative accuracy in estimating $\gap$
+    (\cref{thm:lb-pimin,thm:lb-gap}).
     Essentially, \emph{every} state may need to be visited about
     $\log(d)/\gap$ times, on average, before an accurate estimate of
-    the mixing time can be provided, regardless of what
-    estimation procedure one uses.
+    the mixing time can be provided, regardless of the estimation
+    procedure used.
 
   \item
-    In \cref{thm:empirical}, we develop a procedure for computing
+    In \cref{sec:empirical}, we develop a procedure for computing
     \emph{valid empirical confidence intervals} for $\pimin$ and
@@
     The procedure uses just a single sample path and does not require
-    any prior knowledge of the chain's parameters, such as $\pimin$ or $\gap$.
-    We also quantify the rate at which the width of the confidence
-    intervals shrink as a function of the length of the sample path.
+    any prior knowledge of the chain's parameters, such as $\pimin$ or
+    $\gap$.
+    We quantify the rate at which the width of the confidence
+    intervals shrink as a function of the length of the sample path
+    (\cref{thm:empirical}).
+    We also explain how to combine the empirical confidence intervals
+    from \cref{alg:empest} with the non-empirical bounds from
+    \cref{thm:err} to produce valid empirical confidence intervals.
+    We prove in \cref{thm:combined} that the width of these new
+    intervals converge to zero asymptotically at least as fast as
+    those from either \cref{thm:err} and \cref{thm:empirical}.
 
-  \item
-    Finally, in \cref{thm:combined}, we show how to combine the
-    empirical confidence intervals from \cref{alg:empest} with the
-    non-empirical bounds from \cref{thm:err} to produce valid
-    empirical confidence intervals;
-    the width of these intervals converge to zero asymptotically as
-    fast as those from \cref{thm:err}.
     \todoc{Is it true from the construction of \cref{thm:combined}

### 2015-06-02 intro-nips.tex
main results road map

@@
   \item
-    In \cref{sec:rates-lower}, we show in
-    \cref{thm:lb-pimin,thm:lb-gap} that $n = \Omega\parens{ (d\log
+    In \cref{sec:rates-lower}, we show that $n = \Omega\parens{ (d\log
     d)/\gap + 1/\pimin}$ observations are necessary for any procedure

### 2015-06-02 setting.tex
combine notation/setting

@@
-%!TEX root =  matrix-est.tex
+\subsection{Notations}
+\label{sec:notation}
+
+We denote the set of positive integers by $\bbN$, and
+the set of the first $d$ positive integers $\{1,2,\dotsc,d\}$ by $\iset{d}$.
+The non-negative part of a real number $x$ is $[x]_+ := \max\{0,x\}$,
+$\log$ denotes the natural logarithm.
+Boldface symbols are used for vectors and matrices (e.g., $\vv$,
+$\vM$), and their entries are referenced by subindexing (e.g., $v_i$,
+$M_{i,j}$).
+%The $i$-th coordinate basis vector is denoted by $\ve_i$.
+For a vector $\vv$, $\norm{\vv}$ denotes its Euclidean norm; for a
+matrix $\vM$, $\norm{\vM}$ denotes its spectral norm.
+We use $\Diag(\vv)$ to denote the diagonal matrix whose $(i,i)$-th
+entry is $v_i$.
+The probability simplex is denoted by $\Delta^{d-1} = \{ \vp
+\in [0,1]^d : \sum_{i=1}^d p_i = 1 \}$, and we regard vectors in
+$\Delta^{d-1}$ as row vectors.
+%Finally, the total variation between two probability distributions
+%$\vp, \vq \in \Delta^{d-1}$ is denoted by $\norm{\vp-\vq}_{\tv}$.
+
+\subsection{Setting}
+\label{sec:setting}
+
 Let $\vP \in (\Delta^{d-1})^d \subset [0,1]^{d \times d}$ be a $d

### 2015-06-02 rates.tex
Cleaning up

@@
 
-\todoc{Turn this paragraph into a remark?}
 Note that because the eigenvalues of $\vL$ are the same as that of the
@@
 
-Unfortunately, \cref{thm:err} falls short of being directly
-usable for obtaining a fully empirical confidence interval for $\gap$
-and $\pimin$, at least without further assumptions.
+Let us now return to the question of obtaining a fully empirical confidence 
+interval for $\gap$ and $\pimin$.
+Unfortunately, \cref{thm:err} falls short of being directly suitable for this,
+at least without further assumptions.
 This is because the deviation terms themselves depend inversely both

### 2015-06-02 rates.tex
minor

--- 2015-06-02 rates.tex
@@
 this empirical estimate $\wh\vP$.
+(In fact, this would allow the results to extend to non-reversible
+chains.)
 The difficulty with this approach is that $\vP$ lacks the structure
@@
 \subsection{Lower bounds}
+\label{sec:rates-lower}
 

### 2015-06-02 intro-nips.tex

@@
 Under this assumption, the chain converges to its unique stationary
-\todoc{Maybe steady state is better?} distribution $\vpi =
+distribution $\vpi =
 (\pi_i)_{i=1}^d$ regardless of the initial state distribution $\vq$: 

### 2015-06-02 setting.tex
minor

@@
+%!TEX root =  matrix-est.tex
+
 \subsection{Notations}
@@
 (We will strongly exploit the symmetry in our results.)
-\todoc{Say more? Maybe not..}
 %will be advantageous due to the existence of stronger perturbation results
@@
 positive on account of ergodicity.
-\todoc{Do our results generalize to the spectral gap? Where do we need that we talk about the absolute spectral gap?
+\todoc{Do our results generalize to the spectral gap? No.
+Where do we need that we talk about the absolute spectral gap?
 What is the difference between the spectral and the absolute spectral gap? Say, in their relationship to $\tmix$?}
@@
 \cref{eq:mixing-time-bound}.
-\hide{,
-\[
-  \tmix
-  :=
-  \min\Braces{
-    t \in \bbN :
-    \sup_{\vq \in \Delta^{d-1}}
-    \Norm{
-      \vq \vP^t - \vpi
-    }_{\tv} \leq 1/4
-  }
-  ,
-\]
-can be bounded in terms of $\pimin$ and $\gap$~\citep[Theorems~12.3
-and~12.4]{LePeWi08}:
-\[
-  \Parens{\frac1{\gap} - 1} \ln2
-  \ \leq \ \tmix
-  \ \leq \ \frac1{\gap} \log \frac4{\pimin}
-  .
-\]}%
 Moreover, convergence rates for empirical processes on Markov chain
@@
 $\gap$ (as we will show in the proof of our first result).
-\todoc{I removed the citations (saving space), referring to the proof.}
 %~\citep[e.g.,][]{MR1921877,MoRo08,MoRo09}. 

### 2015-06-02 empirical.tex
Mostly abstract/intro
Making abstract more precise.
New reference.
\ln -> \log replacement (should be the other way around?)
Changed some section titles

@@
     :=
-    \sqrt{\frac{9\wh{P}_{i,j}(1-\wh{P}_{i,j})\ln(2d/\delta)}{2N_i}}
-    + \frac{4\ln(2d/\delta)+1.5\abs{(1/d)-\wh{P}_{i,j}}}{N_i}
+    \sqrt{\frac{9\wh{P}_{i,j}(1-\wh{P}_{i,j})\log(2d/\delta)}{2N_i}}
+    + \frac{4\log(2d/\delta)+1.5\abs{(1/d)-\wh{P}_{i,j}}}{N_i}
     .
@@
 
-%$\qquad$ $\wh\Delta_{i,j}(\delta) \gets \sqrt{\frac{9\wh{P}_{i,j}(1-\wh{P}_{i,j})\ln(2d/\delta)}{2N_i}}
-%  + \frac{4\ln(2d/\delta)+1.5\abs{\alpha-d\alpha\wh{P}_{i,j}}}{N_i}$%,  $(i,j) \in [d]^2$.
+%$\qquad$ $\wh\Delta_{i,j}(\delta) \gets \sqrt{\frac{9\wh{P}_{i,j}(1-\wh{P}_{i,j})\log(2d/\delta)}{2N_i}}
+%  + \frac{4\log(2d/\delta)+1.5\abs{\alpha-d\alpha\wh{P}_{i,j}}}{N_i}$%,  $(i,j) \in [d]^2$.
 %\STATE   $\qquad$ $\wh\Delta_{i,j}(\delta) \gets \min( \wh\Delta_{i,j}(\delta) , \max(\wh{P}_{i,j}, 1-\wh{P}_{i,j}))$,   $(i,j) \in [d]^2$.
@@
     O\Parens{ 
-     \frac{\kappa}{\pimin} \sqrt{\frac{\ln(d/\delta)}{\pimin n}} + 
-      \sqrt{ \frac{d\ln(d/\delta)}{ \pimin n}}
+      \frac{\kappa}{\pimin} \sqrt{\frac{\log(d/\delta)}{\pimin n}} + 
+      \sqrt{ \frac{d\log(d/\delta)}{ \pimin n}}
     }
@@
 
-An easy way to improve the asymptotic width is to combine
-\cref{thm:err,thm:empirical}.
+It is easy to combine \cref{thm:err,thm:empirical} to yield intervals
+whose widths shrink at at least as fast as the non-empirical intervals
+from \cref{thm:err} and the empirical intervals from
+\cref{thm:empirical}.
 Specifically, determine lower bounds on $\pimin$ and $\gap$ using
@@
 }
+Finally, we can take the intersection of these new intervals with the
+corresponding intervals from \cref{alg:empest}.
+% XXX now take the intersection
 This is summarized in the following \lcnamecref{thm:combined}, which
@@
     O\Parens{
+      \min\Braces{
         \sqrt{\frac{\log\frac{d}{\delta}\cdot\log(n)}{\pimin\gap n}}
+        ,\,
+        \hat{w}
+      }
     }

### 2015-06-02 empirical.tex
minor

@@
 \] 
-The widths of these intervals are somewhat larger than the corresponding
-(non-empirical) intervals from \cref{thm:err}.
+%The widths of these intervals are somewhat larger than the corresponding
+%(non-empirical) intervals from \cref{thm:err}.
 \todoc{This needs to be updated once we finalize the results. By how much are these larger?}
@@
 It is easy to combine \cref{thm:err,thm:empirical} to yield intervals
-whose widths shrink at at least as fast as the non-empirical intervals
-from \cref{thm:err} and the empirical intervals from
+whose widths shrink at at least as fast as both the non-empirical
+intervals from \cref{thm:err} and the empirical intervals from
 \cref{thm:empirical}.
@@
     }
-    .
   \]
+  where $\hat{w}$ is the width from \cref{alg:empest}.
 \end{theorem}

### 2015-06-02 intro-nips.tex

@@
 \paragraph{Main results.}
-We develop the first \todoc{Maybe leave out efficient? 
-What is efficient is a question of interpretation. We 
-should maybe rather add a note on the computational complexity.}
-procedure for constructing non-trivial
+We develop the first procedure for constructing non-trivial
 and fully empirical confidence intervals for Markov mixing time.
 Consider a reversible ergodic Markov chain on $d$ states with absolute
-spectral gap $\gap$ and stationary distribution minorized by $\pimin$. \todoc{Comment on reversibility somewhere?}
+spectral gap $\gap$ and stationary distribution minorized by $\pimin$. 
 As is well-known \citep[Theorems~12.3 and~12.4]{LePeWi08},
@@
 and hence it suffices to estimate $\gap$ and $\pimin$. 
-\todoc{I would prefer $\ln$ over $\log$ throughout the paper.}
 %these quantities $\gap$ and $\pimin$ are well-known to govern
@@
 barriers for many MCMC diagnostic problems~\citet{BhaBoMo11}.
-\todoc{Explain the organization of the paper?}
 %%Performance guarantees play a key role in the analysis and design of machine learning algorithms.

### 2015-06-03 conclusions.tex
clean up
adding and resolving todos, as per our discussion.
bib file improvements
generalized inverse clean up, both in appendix and in main text
ideas for future work in conclusion
empirical results: comment on O(.) notation
minor tweak to intro
Some todos for myself
New discussion + minor edits

@@
 %!TEX root =  matrix-est.tex
+Generalizing the construction of  \cref{thm:combined}, 
+given any confidence interval $I_n=I_n(\gap,\pimin)$ for the spectral gap $\gap$ that itself depends 
+on $\gap$ and $\pimin$, under general conditions,
+plugging in the empirical bounds of \cref{alg:empest} into the expression defining $I_n$
+gives a fully empirical confidence interval whose
+width matches that of $I_n$ up to lower order terms.
+In particular, this suggests that future work should focus on 
+closing the gap between the lower and upper bounds on the accuracy
+of point-estimation. Another interesting direction is to 
+reduce the computation cost: The current cubic cost in the number of states
+can be too high even when the number of states is moderate.
 
+Perhaps more important, however, is to extend 
+our results to large state space Markov chains:
+In most practical applications the state space is continuous
+or is exponentially large in some natural parameters.
+As follows from our lower bounds, without further assumptions,
+the problem of fully data dependent estimation of the mixing time
+is intractable for information theoretical reasons.
+Interesting directions for future work thus must consider Markov
+chains with specific structure. Parametric classes of Markov chains,
+including but not limited to Markov chains with factored transition kernels
+with a few factors, are a promising candidate for such future investigations.
+While we realize the limitations of our work, 
+we think that the area of fully data dependent estimation of Markov chain 
+parameters based on a single sample path is full of exciting, important
+and challenging research questions and we hope that our work will
+serve as a source of inspiration for others to consider these questions.
+\if0
 The algorithm from Section~\ref{sec:empirical} demonstrates the
@@
 Theorems~\ref{thm:lb-pimin} and \ref{thm:lb-gap}. 
-\todoc{We need to move the references needed in the appendix to after the appendix. 
-This should be possible with some latex package for splitting references..}
+\todoc{Reduce space complexity? Assume more structure of $\vP$?
+Sometimes $\gap$ is known, but $\pimin$ is not. Does this help?
+How about the other way around?
+}
+\fi
+%\todoc{We need to move the references needed in the appendix to after the appendix. 
+%This should be possible with some latex package for splitting references..}
 %and investigate

### 2015-06-03 empirical.tex
New discussion + minor edits

@@
 %(non-empirical) intervals from \cref{thm:err}.
-\todoc{This needs to be updated once we finalize the results. By how much are these larger?}
+%\todoc{This needs to be updated once we finalize the results. By how much are these larger?}
 
@@
   Furthermore, the widths of these intervals almost surely satisfy
-  (as $n \to \infty$) \todoc{Do we also care about $\pimin\to 0$, $\gap\to 0$? Do we need then other terms for both? Replicating the form of bounds from \cref{thm:err}?}
-  \todod{I don't understand the first question. For the second, we
-  could do this, but it would take some non-trivial work\ldots}
+  (as $n \to \infty$) 
   \[

### 2015-06-03 rates.tex
New discussion + minor edits

@@
 
-Let us now return to the question of obtaining a fully empirical confidence 
-interval for $\gap$ and $\pimin$.
-Unfortunately, \cref{thm:err} falls short of being directly suitable for this,
+Returning to the question of obtaining a fully empirical confidence 
+interval for $\gap$ and $\pimin$, we notice that,
+unfortunately, \cref{thm:err} falls short of being directly suitable for this,
 at least without further assumptions.

### 2015-06-03 setting.tex
New discussion + minor edits
Typo fix + log/ln difference explanation

@@
 the set of the first $d$ positive integers $\{1,2,\dotsc,d\}$ by $\iset{d}$.
-The non-negative part of a real number $x$ is $[x]_+ := \max\{0,x\}$,
-$\log$ denotes the natural logarithm.
+The non-negative part of a real number $x$ is $[x]_+ := \max\{0,x\}$.
+We use $\ln(\cdot)$ in expressions with exact constants, while we use $\log(\cdot)$ elsewhere.
+%In expression with exact numerical constants, we will use 
+%the natural logarithm function, $\ln(x)$, while elsewhere we use $\log(x)$.
 Boldface symbols are used for vectors and matrices (e.g., $\vv$,
@@
 \todoc{Do our results generalize to the spectral gap? No.
-Where do we need that we talk about the absolute spectral gap?
-What is the difference between the spectral and the absolute spectral gap? Say, in their relationship to $\tmix$?}
+Discuss this somewhere.}
 %Throughout the paper, both $\pimin$ and $\gap$ are assumed

### 2015-06-03 rates.tex
.

@@
 this empirical estimate $\wh\vP$.
-(In fact, this would allow the results to extend to non-reversible
-chains.)
+In fact, this approach is a way to extend to non-reversible chains, as
+we would no longer rely on the symmetry of $\vM$ or $\vL$.
 The difficulty with this approach is that $\vP$ lacks the structure

### 2015-06-03 conclusions.tex
Aryeh's changes
mostly grammatical/spelling; nontrivial ones indicated by \todoa

@@
 %!TEX root =  matrix-est.tex
-Generalizing the construction of  \cref{thm:combined}, 
-given any confidence interval $I_n=I_n(\gap,\pimin)$ for the spectral gap $\gap$ that itself depends 
-on $\gap$ and $\pimin$, under general conditions,
-plugging in the empirical bounds of \cref{alg:empest} into the expression defining $I_n$
-gives a fully empirical confidence interval whose
-width matches that of $I_n$ up to lower order terms.
+The construction of \cref{thm:combined}
+suggests a general procedure for converting
+any confidence interval $I_n=I_n(\gap,\pimin)$ for the spectral gap $\gap$,
+which
+%itself depends 
+%on
+is expressed in terms of
+$\gap$ and $\pimin$,
+to a fully empirical confidence interval whose
+width matches that of $I_n$ up to lower order terms:
+Plug in the empirical bounds of \cref{alg:empest} into the expression defining $I_n$.
+\todoa{reworded}
 In particular, this suggests that future work should focus on 
@@
 with a few factors, are a promising candidate for such future investigations.
-While we realize the limitations of our work, 
-we think that the area of fully data dependent estimation of Markov chain 
-parameters based on a single sample path is full of exciting, important
-and challenging research questions and we hope that our work will
-serve as a source of inspiration for others to consider these questions.
+The results presented here are a first step in the ambitious research agenda
+outlined above, and we hope that they will
+%pave the way
+serve as a point of departure
+for
+further insights
+in the area of fully empirical estimation of Markov chain 
+parameters based on a single sample path.
+%is full of exciting, important
+%and challenging research questions and we hope that our work will
+%serve as a source of inspiration for others to consider these questions.
+\todoa{rewrote ending}
 \if0

### 2015-06-03 intro-nips.tex
New discussion + minor edits
log -> ln
Aryeh's changes
mostly grammatical/spelling; nontrivial ones indicated by \todoa

@@
 
-This work considers the construction of fully empirical bounds on the
+This work
+tackles the challenge of constructing
+\todoa{was:
+considers the construction of}
+fully empirical bounds on the
 mixing time of Markov chains based on a single sample path.
@@
 \]
-The \emph{mixing time} $\tmix$ of the Markov chain is the time that it
-takes for the chain to (approximately) reach its stationary
-distribution,
+The \emph{mixing time} $\tmix$ of the Markov chain is the
+number of time steps required
+%time that it
+%takes
+for the chain to
+be within a fixed threshold of
+%(approximately) reach
+its stationary
+distribution:
+\todoa{slightly changed wording}
 \begin{align}
@@
 [0,\infty]$, based only on the observed sample path
-$(X_1,X_2,\dotsc,X_n)$ and $\delta \in (0,1)$, that captures the
-mixing time $\tmix$ with the pre-specified probability $1-\delta$.
+$(X_1,X_2,\dotsc,X_n)$ and $\delta \in (0,1)$, that
+succeeds with probability $1-\delta$
+in trapping
+the value of the
+mixing time $\tmix$.
+\todoa{changed wording}
 
 This problem is motivated by the numerous scientific applications and
-machine learning tasks in which one is interested in estimating the
-mean $\vpi(f) = \sum_i \pi_i f(i)$ of a function $f$ of the states of
+machine learning tasks in which
+the quantity of interest is
+%one is interested in estimating
+the
+mean $\vpi(f) = \sum_i \pi_i f(i)$
+for some function $f$ of the states of
 a Markov chain.
-This is the realms of the celebrated Markov Chain Monte Carlo (MCMC)
+\todoa{changed wording}
+This is the setting of the celebrated Markov Chain Monte Carlo (MCMC)
 paradigm~\cite{liu2001monte}, but the problem also arises in
@@
 bounds on $\Pr\parens{ |\vpi_n(f)-\vpi(f)| > \eps }$, thus providing
-invaluable tools for non-asymptotic probabilistic
-analysis~\cite{gillman1998chernoff,leon2004optimal,DBLP:conf/valuetools/KontoyiannisLM06,paulin15}.
+valuable tools for non-asymptotic probabilistic
+analysis~\cite{gillman1998chernoff,leon2004optimal,DBLP:conf/valuetools/KontoyiannisLM06,
+%meroi2012-jap,
+paulin15}.
 These probability bounds are larger than corresponding bounds for
@@
 It is possible to eliminate many of the difficulties presented above
-with additional access to the Markov chain.
-For example, with access to a sampling oracle that generates
+%with additional access 
+when allowed more flexible
+to the Markov chain.
+\todoa{slightly changed wording}
+For example, given a sampling oracle that generates
 independent transitions from any given state (akin to a ``reset''

### 2015-06-03 abstract.tex
New discussion + minor edits
Aryeh's changes
mostly grammatical/spelling; nontrivial ones indicated by \todoa
Merge with Aryeh's version + some further edits

@@
 The width of the interval converges to zero at a
-$\sqrt{n}$ rate, where $n$ is the length of the sample path.
+$\sqrt{n}$ rate, where $n$ is the length of the sample path. \todoc{Aryeh prefers: .. at a rate of $\sqrt{n}$. }
 Upper and lower bounds are given on the number of samples required to
@@
 seeing each state at least $\Omega(\tmix)$ times on the average.
-Future directions of research are identified.
-\todoc{Such as: closing the gap. Extension to large state spaces, 
-based on additional assumptions. Even if $\gap$ is large, we need many samples.
-Maybe we need even stronger assumptions, like the decay rate of eigenvalues.
-}
+Finally, future directions of research are identified.
 %If $\gap$ denotes the absolute spectral gap of the unknown Markov chain

### 2015-06-03 intro-nips.tex
Merge with Aryeh's version + some further edits

@@
 tackles the challenge of constructing
-\todoa{was:
-considers the construction of}
 fully empirical bounds on the
@@
 a Markov chain.
-\todoa{changed wording}
+\todoa{changed wording} \todoc{I am not sure that the new wording is better. It
+fails to express that there is an estimation problem. "Quantity of interest"
+is way to vague for me.} 
 This is the setting of the celebrated Markov Chain Monte Carlo (MCMC)
@@
 \end{equation}
-%$(\frac{1}\gap  - 1)\log(2)  \le \tmix \le \frac{1}\gap \log \frac4\pimin$,
+%$(\frac{1}\gap  - 1)\ln(2)  \le \tmix \le \frac{1}\gap \ln \frac4\pimin$,
 and hence it suffices to estimate $\gap$ and $\pimin$. 
@@
     those from either \cref{thm:err} and \cref{thm:empirical}.
-
-    \todoc{Is it true from the construction of \cref{thm:combined}
-    that if we had a better point estimator
-    then we would automatically get an interval estimator
-    with the same accuracy?
-    This seems to be true.
-    This would mean that to close the gap
-    one should really work on coming up with better point estimators.
-    }
 
@@
 It is possible to eliminate many of the difficulties presented above
-%with additional access 
-when allowed more flexible
+%with additional access to the Markov chain.
+when allowed more flexible access
 to the Markov chain.
-\todoa{slightly changed wording}
 For example, given a sampling oracle that generates

### 2015-06-04 rates.tex
Lower/upper bound switch, more explanation of the difficulty in the empirical construction

@@
 
-In this section, we present upper and lower bounds on achievable rates
+In this section, we present lower and upper bounds on achievable rates
 for estimating the spectral gap as a function of the length of the
@@
 
+\subsection{Lower bounds}
+\label{sec:rates-lower}
+The purpose of this section is to show lower bounds on the number of observations
+necessary to achieve a fixed multiplicative (or even just additive) accuracy in estimating the spectral gap $\gap$.
+By \cref{eq:mixing-time-bound}, the multiplicative accuracy lower bound for $\gap$
+gives the same lower bound for estimating the mixing time.
+%It is important to emphasize that these are algorithm independent lower bounds, i.e.,
+%any algorithm is bound by them.
+Our first result holds even for two state Markov chains and shows that a sequence length of $\Omega(1/\pimin)$
+is necessary to achieve even a constant \emph{additive} accuracy in estimating $\gap$.
+%We begin with a result showing that any algorithm for estimating
+%$\gap$ (even for just constant additive accuracy) requires an
+%observation sequence of length $\Omega(1/\pimin)$
+%even when the number of states is only two.
+%%%
+\begin{theorem}
+  \label{thm:lb-pimin}
+  There is an absolute constant $c>0$ such that the following holds.
+  Pick any $\bar\pi \in (0,1/4)$.
+  Consider any estimator $\hatgap$ that takes as input a random sample
+  path of length $n < c/\bar\pi$ from a Markov chain starting from any
+  desired initial state distribution.
+  There exists a two-state ergodic and reversible Markov chain
+  distribution with spectral gap $\gap \geq 1/2$ and minimum
+  stationary probability $\pimin \geq \bar\pi$ such that
+  \[
+    \Pr\Brackets{ |\hatgap - \gap| \geq 1/6 } \geq 1/8 .
+  \]
+\end{theorem}
+%%%
+Next, considering $d$ state chains, we show that 
+a sequence of length $\Omega(d\log(d)/\gap)$ is required
+to estimate $\gap$ up to a constant multiplicative accuracy.
+Essentially, the sequence may have to visit all $d$ states $\log(d)/\gap$ times each on average.
+This holds \emph{even} if $\pimin$ is within a factor of two of the
+\emph{largest} possible value of $1/d$ that it can take, i.e., when
+$\vpi$ is nearly uniform.
+\begin{theorem}
+  \label{thm:lb-gap}
+  There is an absolute constant $c>0$ such that the following holds.
+  Pick any positive integer $d \geq 3$ and any $\bar\gamma \in
+  (0,1/2)$.
+  Consider any estimator $\hatgap$ that takes as input a random sample
+  path of length $n < c d\log(d) / \bar\gamma$ from a $d$-state
+  reversible Markov chain starting from any desired initial state
+  distribution.
+  There is an ergodic and reversible Markov chain distribution
+  with spectral gap $\gap \in [\bar\gamma,2\bar\gamma]$ and minimum
+  stationary probability $\pimin \geq 1/(2d)$ such that
+  \[
+    \Pr\Brackets{ |\hatgap - \gap| \geq \bar\gamma} \geq 1/4 .
+  \]
+\end{theorem}
+
+The proofs of \cref{thm:lb-pimin,thm:lb-gap} are given in
+\cref{app:lower}. 
+%In summary, the lower bounds show that no algorithm can hope to achieve a fixed multiplicative
+%accuracy before seeing $\Omega( d \log(d)/\gamma^* + 1/\pimin)$ states.
+
 \subsection{A plug-in based point estimator and its accuracy}
 \label{sec:rates-upper}
-
-Our upper bound is demonstrated using estimators based on natural
-plug-in estimates.
-We also provide an estimator for the minimum stationary probability
-in the process, so that one can use the bounds from
-\cref{eq:mixing-time-bound} to trap
+Let us now consider the problem of estimating $\gap$.
+For this, we construct a natural plug-in estimator.
+Along the way, we also provide an estimator for the minimum stationary probability,
+allowing one to use the bounds from \cref{eq:mixing-time-bound} to trap
 the mixing time.
@@
 \]
+By \cref{eq:mixing-time-bound},
+the second of these is also a bound on the required sequence length to estimate $\tmix$. % the mixing time.
 
@@
 not need to provide empirical confidence bounds.
-
-\subsection{Lower bounds}
-\label{sec:rates-lower}
-
-We complement \cref{thm:err} with lower bounds showing the
-necessity of some of the quantities appearing in the upper bound.
-We begin with a result showing that any algorithm for estimating
-$\gap$ (even for just constant additive accuracy) requires an
-observation sequence of length $\Omega(1/\pimin)$
-even when the number of states is only two.
-
-\begin{theorem}
-  \label{thm:lb-pimin}
-  There is an absolute constant $c>0$ such that the following holds.
-  Pick any $\bar\pi \in (0,1/4)$.
-  Consider any estimator $\hatgap$ that takes as input a random sample
-  path of length $n < c/\bar\pi$ from a Markov chain starting from any
-  desired initial state distribution.
-  There exists a two-state ergodic and reversible Markov chain
-  distribution with spectral gap $\gap \geq 1/2$ and minimum
-  stationary probability $\pimin \geq \bar\pi$ such that
-  \[
-    \Pr\Brackets{ |\hatgap - \gap| \geq 1/6 } \geq 1/8 .
-  \]
-\end{theorem}
-
-Next, we show that in order to estimate
-$\gap$ to within a constant multiplicative accuracy in a $d$ state chain,
-a sequence of length $\Omega(d\log(d)/\gap)$ is required.
-Essentially, the sequence may have to visit all $d$ states $\log(d)/\gap$ times each on average.
-This hold \emph{even} if $\pimin$ is within a factor of two of the
-\emph{largest} possible value of $1/d$ that it can take, i.e., when
-$\vpi$ is nearly uniform.
-\begin{theorem}
-  \label{thm:lb-gap}
-  There is an absolute constant $c>0$ such that the following holds.
-  Pick any positive integer $d \geq 3$ and any $\bar\gamma \in
-  (0,1/2)$.
-  Consider any estimator $\hatgap$ that takes as input a random sample
-  path of length $n < c d\log(d) / \bar\gamma$ from a $d$-state
-  reversible Markov chain starting from any desired initial state
-  distribution.
-  There is an ergodic and reversible Markov chain distribution
-  with spectral gap $\gap \in [\bar\gamma,2\bar\gamma]$ and minimum
-  stationary probability $\pimin \geq 1/(2d)$ such that
-  \[
-    \Pr\Brackets{ |\hatgap - \gap| \geq \bar\gamma} \geq 1/4 .
-  \]
-\end{theorem}
-
-The proofs of \cref{thm:lb-pimin,thm:lb-gap} are given in
-\cref{app:lower}.

### 2015-06-09 nips2015/intro.tex
reorganize files

%!TEX root =  matrix-est.tex

\begin{sloppypar}
Predicting diverse characteristics of various physical and simulated phenomena, 
	performance monitoring and diagnostics of controlled systems, 
        as well as
	various speech recognition, video or natural language processing tasks 
	are all examples that involve temporally dependent data.
Classical results of statistics and statistical learning theory extend 
	to the dependent data in a more or less seamless fashion,
	at least when the data is sufficiently ``mixing''.
Such results include the central limit theorem, 
	exponential tail bounds for empirical processes, 
	performance guarantees for complexity regularization
	using sieves or roughness penalties, as well as 
	results based on stability arguments or Rademacher-complexity 
	\citep{Woo86,Dou94,Yu94,ZheChu96,dlPenaGine99,Mei00,
MR1921877,vid2003,gamarnik03,LoKuScha05,MoRo08,MoRo09,
DBLP:conf/nips/SteinwartC09,Steinwart2009175,
FaSze11,
london:nips12asalsn,london:icml13,
DBLP:conf/nips/ShaliziK13}.
\end{sloppypar}

While under sufficiently fast mixing, the form of the results is preserved,
	the performance guarantees are typically dependent on the \emph{unknown}
	mixing time of the underlying stochastic process.	
A simple example where this problem arises is when a data-dependent bound
	on the performance of a small number of predictors is needed for the selection
	of the best estimator.
\citet{Mei00} 	considers the more challenging problem of complexity regularization and notes that
	this technique remains ``a theoretician's dream'' in the absence of an estimate of the mixing time.
%As noted by \citet{Mei00}, in order to form better predictions, 
An analogue issue arises in the convergence diagnostics of Monte-Carlo Markov Chain (MCMC)
	samplers, or when a single sample path of a stochastic controlled system is used to
	estimate the performance of a control policy.
Owning to the wide applicability of MCMC methods to various computational problems,
	much effort has been devoted to developing effective convergence diagnostic methods
	with the aim of deriving stopping rules or stationarity diagnoses.
While a large number of heuristic techniques have been developed 
	\citep[e.g.,][]{MCMCDiscussion93,CoCa96},
	it is widely recognized that all these methods may fail to meet the guarantees
	that they are designed for.

In the theory community, 
	\citet{BaFoRuSmiWhi00,BaFoRuSmiWhi13} develop algorithms for testing whether a given Markov
	chain over a finite state space of size $d$ is rapidly mixing.
In particular, based on testing closeness of distributions when a sampling device is available,
	their algorithm can distinguish between chains that in $t$ steps for a fixed $t$ reach a distribution
	close to the stationary distribution by simulating $\tilde{O}(t d^{5/3})$ transitions.
It is remarkable that when $t$ is small enough, the procedure requires sublinearly many samples
	in the total number of transitions in the Markov chain.
The algorithm is assumes that transitions can be simulated from any given state.

\citet{BhaBoMo11} study a similar question for Markov chains with very large state spaces.
	In particular, they assume an circuit based description of the transition probability structure of the matrix
	and derive several hardness results for Markov chains whose state space 
	consists of all binary strings of a given length.
For example, they prove that given a Markov chain that mixes rapidly, it is $S\!Z\!K$-hard
	to distinguish whether the chain is close to stationarity by time $t$ or far from stationarity at time $ct$ for a constant $c$
	when started from a given initial state ($S\!Z\!K \subset A\!M \cap \mathrm{co}A\!M$, 
	which is conjectured to extend beyond $B\!P\!P$ \citep{GoVa11}).%
\footnote{$SZK$ stands for statistical zero-knowledge, $BPP$ stands for probabilistic polynomial-time, 
$AM$ stands for ``Arthur-Merlin protocol''.}

Other works include that of \citet{McDoShaSche11} who propose to estimate the $\beta$-mixing coefficients 
	of a stationary process possessing a density based on histogram estimates of the underlying joint density.
Their main result is a consistency result for a fixed ``lag''. For fast mixing processes (e.g., for $k$-step Markov processes)
their result yields a convergence rate, which is given in terms of the unknown mixing coefficients.
%Catch-22 problem remains: The rate of convergence of their estimates depends on the unknown mixing coefficients
%(Theorem 2.4)
% $\exp(W(\ln(n)))$ is the window length, where $W$ is Lambert's $W$-function (i.e., this is between $\ln \ln(n)$ and $\ln n$).
An earlier paper of the same spirit is that of \citet{Nobel06} who considers the generalization of 
existing discernibility results for i.i.d. processes to weakly uniformly ergodic families and shows how
this can be used to study, and in some cases estimate, polynomial mixing times for dependent processes.

 \todoc[inline]{\url{http://arxiv.org/abs/1212.2016}, Non-asymptotic confidence intervals for MCMC in practice
Benjamin M. Gyori, Daniel Paulin, could be interesting.
At the end of the abstract they say:
``We propose estimators of the parameters appearing in the bounds, and illustrate our results with simulations on lattice models in statistical physics as well as an example of Bayesian model averaging.''
See also \url{http://arxiv.org/abs/1409.7986v1}.
}
%\todoc[inline]{Add:
%Computable exponential bounds for screened estimation and simulation
%I Kontoyiannis, S Meyn 
%AREYH: is that one relevant? They deal with iid sequences
%and
%Exponential bounds and stopping rules for MCMC and general Markov chains
%I Kontoyiannis, L Lastras-Montano, S Meyn
%AREYH: I don't think this one is too relevant either:
%their bounds depend on unknown quantities
%}
%\citet{DBLP:conf/valuetools/KontoyiannisLM06}
%give large deviation bounds

\paragraph{Main results.}
Consider a reversible ergodic Markov chain on $d$ states with spectral
gap $\gap$ (which is well-known to govern mixing time)
and stationary distribution minorized by $\pimin$. \todoc{Should we add definitions? Reversible, ergodic, spectral gap? Where?}
Our main results are summarized as follows.
\begin{enumerate}
  \item
    In Theorem~\ref{thm:err}, we give a procedure that estimates
    $\gap$ to within a constant multiplicative factor from \emph{a single
    sample path} of length
    $\tilde{O}\parens{1/(\pimin\gap^3)}$.\footnote{The
    $\tilde{O}(\cdot)$ notation suppresses logarithmic factors.}

  \item
    In Theorems~\ref{thm:lb-pimin} and \ref{thm:lb-gap}, we show that
    $\Omega\parens{ (d\ln d)/\gap + 1/\pimin }$ observations are
    necessary for constant multiplicative accuracy.

  \item
    Finally, in Theorem~\ref{thm:empirical} we develop a procedure for
    providing \emph{fully data-dependent confidence bounds} on the
    spectral gap $\gap$.
    The procedure uses just a single sample path and does not require
    knowing anything about $\gap$ or $\pimin$.
    We also quantify the rate at which the confidence interval
    shrinks as a function of the length of the sample path,
    which is
    $     O\Parens{
      \frac1{\pimin}\sqrt{\frac{d\ln(d/\delta)}{ n}}
      +
      \frac{d}{\pimin\gap} \sqrt{\frac{\ln(d/\delta)}{\pimin n}}
    }.$

%    \todoc{please add rate}

\end{enumerate}

\hide{
Our approach:
Interval can be given in terms of the mixing time of the chain 
and the smallest visit probability. \todoc{Is this a necessary dependence?}
We estimate both quantities. % and use a plug-in estimate.

Difficulty: %Whatever empirical data is used, the speed of convergence will be dependent on the same unknown quantities.
For both quantities,  the speed of convergence will be dependent on the same unknown quantities.
We show how to overcome this Catch-22 problem and come up with the 
first fully data-dependent interval estimate.

(Almost) matching lower bounds.
}

\if0
%%%
ALGORITHMS FOR LARGE GRAPHS
A Sarma
Uses Batu's method.
Estimating page rank on graph streams
JACM paper

%%
Intersection and mixing times for reversible chains
Y Peres, T Sauerwald, P Sousi, A Stauffer
Sandwitches the intersection time in between the mixing time and the maximum hitting time.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
http://projecteuclid.org/euclid.aos/1362493041
Convergence rate of Markov chain methods for genomic motif discovery
D Woodard, J Rosenthal
Ours are some of the first meaning- ful bounds on the convergence rate of a Markov chain method for sampling from a multimodal posterior distribution, as a function of statistical quantities like the number of observations.

we show that, due to multimodality of the posterior distribution, the rate of convergence often decreases exponentially as a function of the length of the DNA sequence.

Characterizing the dependence of the convergence rate on statistical quantities like the number of observations or the number of parameters is critical in justifying the use of a Markov chain method. However, there are several difficulties in doing so. First, the posterior distribution of a statistical model has a much more complex form than the stylized distributions for which Markov chain convergence rates are typically obtained [Borgs et al. (1999), Bhatnagar and Randall (2004), Woodard, Schmidler and Huber (2009b)]. Second, the data, and thus the convergence rate of the Markov chain, are stochastic and depend on the data-generating mechanism.

T irreducible, aperiodic, reversible and nonnegative definite,

Since the state space X is finite, Gap(T ) > 0 and the chain is called uniformly ergodic [Roberts and Rosenthal (2004)]. The quanti- ties ?? and Gap(T ) are related via [Sinclair (1992)]

(2.9)

Markov chain Monte Carlo (MCMC) sampling tech- niques are currently the most widely used approach to approximating the high-dimensional integrals arising in Bayesian statistics, as well as in related areas such as statistical mechanics.

Adaptive MCMC: effort to speed up..

Relation between mixing time and spectral gap
http://arxiv.org/abs/1405.0028
Proposition 8(a) of Aldous (1982) provides a lower bound on ?? in terms of the spectral radius of T

-----------
Peter J. Diggle:

Peter Green, in his presentation, emphasized that many practical problems
can be expressed as the estimator of the expectation, Er (f), of some functional f with respect to the
distribution 7r;

Adrian Smith adopted a more ambitious stance in suggesting that the
full panoply of graphical data analytic methods could be used to make inferences about the whole
full panoply of graphical data analytic methods could be used to make inferences about the whole
distribution 7r from the empirical marginal distribution of the ÂÂ£1'.

-------
Persi Diaconis and Jeffrey S. Rosenthal

The asymptotic rate of
convergence to stationarity is determined by the second-largest eigenvalue (Besag and Green,
Section 2). This can be deceptive.
They give an example when based on the second largest eigenvalue, on the symmetry group of $n$ elements
the convergence to stationary comes out as $n^2 \ln n$, whereas a more precise analysis gives $n \ln n$.

Another class of techniques for proving convergence results involves bounds on eigenvalues by using
the underlying geometry of the chain. They were pioneered by computer scientists (see, for example,
Jerrum and Sinclair {1989)) to study randomized algorithms and have been developed (Lawler and Sokal,
1988; Diaconis and Stroock, 1991; Diaconis and Saloff-Coste, 1992) to give essentially sharp rates in
a variety of messy problems.

Jerrum, M. and Sinclair, A. (1989) Approximating the permanent. SIAM J. Comput., l, 1149-1178.

Lawler, G. F. and Sokal, A. D. (1988) Bounds on the L2 spectrum for Markov chains and Markov processes: a
generalization of Cheeger's inequality. Trans. Am. Math. Soc., 309, 557-580.

Diaconis, P. and Stroock, D. (1991) Geometric bounds for eigenvalues of Markov chains. Ann. Appl. Probab.,
1, 36-61

Diaconis, P. and Saloff-Coste, L. (1992) Comparison theorems for reversible Markov chains. Technical Report.
Department of Mathematics, Harvard University, Cambridge.

Permanent:
Thepermanentfunction
arises naturally in a number of fields, including algebra, combinatorial enumeration, and the physical sciences, and has been an object of study by mathematicians since first appearing in 1812 in the work of Cauchy and Binet (see [26] for background). Despite considerable effort, and in contrast with the syntactically very similar deter- minant, no efficient procedure for computing this function is known.

16,Lemma6.1: probability boosting
Uses conductance

Combinatorial enumeration problems.

Let (X,),=o be a time-homogeneous Markov chain with finite state space 3f and transition matrix P=(Pq)i,sx. (All chains in this paper are assumed to be of this form.) If the chain is ergodic we let or= (rri)ix denote its stationary distribution, the unique vector satisfying rP--r and i 7ri 1. In this case, if the chain is allowed to evolvefor stepsfromanyinitialstatethedistributionofitsfinalstateapproaches as t--> oe. Necessary and sufficient conditions for ergodicity are that the chain is (a) irreducible, i.e., any state can be reached from any other in some number of steps; and (b)aperiodic,i.e.,gcd{s: isreachablefromjinssteps} 1foralli,j 2f.

Analysis of time-reversible Markov chains is simplified by the following observation. LEMMA2.1. Suppose(Xt)isanergodicMarkovchainwithfinitestatespaceA;and transition matrix P. Let Tri)ix be any vector satisfying the detailed balance condition (1) and the norrnalisation condition Y 7r 1. Then the Markov chain (Xt) is time-
reversible and r is its (unique) stationary distribution.
%%%%%

The authors replied later, in writing, as follows.
A. F. M. Smith and G.O. Roberts:

Many contributors have commented on issues related to 'convergence diagnostics'. In particular, various
doubts are raised about reliance on diagnostics based on just a few summary statistics. Unfortunately,
the complexity of problems on which MCMC methods are currently being employed means that analytic
bounds are usually unavailable, so that some form of output analysis is necessary. Of course this does
not detract from the value of exploratory techniques, such as those suggested by Dr Jennison, and many
others, although the efficiency of estimates based on importance sampling using heated 1r in high
dimensional problems is likely to be poor.

%%%%%%%
Dr Jennison's example illustrates a basic principle: it is not sensible to attempt to assess convergence
on the basis of output from one long run of the chain. Within the operations research literature many
authors (see for example Kelton and Law (1984) and Whitt (1991), and their references) have considered
the irnplementational question of whether to run multiple replications of a stationary time series. However,
this body of literature concentrates on efficiency of subsequent estimates. The main advantages of running
multiple replications are in the areas of preliminary exploration, the design of 'production run' chains
and assessment of convergence. These aspects of the MCMC method are vital but are less amenable
to quantitative analysis than, say, the standard error of an estimate based on an ergodic average. It
must be remembered that simulations are experiments and, as such, should be approached with many
of the same kinds of caution and protocol that attach to other areas of experimentation. In particular,
pilot experiments and replication will typically be required. Similar pragmatic considerations apply to
using the output from MCMC simulations, as Professor Geyer seems to recognize in part of his discussion,
albeit after attributing to us views that we do not hold.

Output from a single chain can never diagnose mixing of the chain. The only way to try to do this
is to use multiple chains from different starting points, and to perform comparisons between chains.
Merely taking averages across chains can be as misleading as running single chains. In the example given
by Dr Jennison , using multiple starting points would almost certainly detect the problem, unless we
were so unlucky as to choose all our starting points from the same component. However, the example
would almost certainly catch out the Raftery and Lewis method.
The coupling approach of Professor Frigessi represents the most direct approach to diagnosing mixing.
The use of stopping times and the fundamental theorem of coupling has also been used to study
convergence in Diaconis (1988) and is alluded to in the comment by Professor Diaconis and Professor
Rosenthal. The disadvantage of Professor Frigessi's suggestion is the need to have an initial value,
simulated from 1r

.. quantification of Monte Carlo errors..

``Output from a single chain can never diagnose mixing of the chain. ''

NOTES:
We do not exploit that $f$ may be special.
Advantage and disadvantage..

Although practitioners are mainly interested in stopping rules and stationarity diagnoses,

%It has been observed that if the features of the posterior distribution are completely
%unknown, then convergence diagnosis for a single Markov chain is inherently an intractable
%problem (Clifford 1993; Bhatnagar, Bogdanov and Mossel 2010). The possible presence
%of modes with very small Òbasins of attractionÓ (Section 3.1) implies that a convergence
%diagnostic would have to exhaustively enumerate the space in order to guarantee that the
%chain had not missed any such mode. The use of multiple chains, as in the Gelman-Rubin
%approach, does not change the intractability problem.

BaFoRuSmiWhi13

What is the problem:

Single sample path

Time series problems:

This paper studies the scenario where the observations are drawn from a stationary ?-mixing or ?-mixing sequence, a widely adopted assumption in the study of non-i.i.d. processes that im- plies a dependence between observations weakening over time (Yu, 1994; Meir, 2000; Vidyasagar, 2003; Lozano et al., 2006; Mohri and Rostamizadeh, 2007).

------------------------------------------
Meir:
 many of the problems to which Machine Learning techniques are applied are inherently temporal in nature. Some obvious examples are stock market prediction, analysis of financial markets, monitoring and diagnosing complex control systems and speech recognition, to name but a few..

 One approach to incorporating temporal structure in order to form better predictors, by more appropriate complexity regularization, is described in this work. In particular, the optimal memory size that should be used in order to form a predictor is in principle derivable from the procedure (see also (Modha \& Masry, 1998)), given information about the mixing nature of the time series (see Section 4 for a definition of mixing).

  If precise knowledge of the mixing parameters is lacking, the procedure requires estimation of these parameters. Unfortunately, as far as we are aware, there is no efficient practical approach known at this stage for estimation of mixing parameters.

Meir's main concern: complexity regularization.
Needs the knowledge of mixing-rate.

There remain several issues, which need to be addressed in future work. First, the adaptive algorithm, as well as the bounds rely heavily on a knowledge of the mixing nature of the process. As mentioned in Section 4, no tools are currently available for establishing mixing properties, which are therefore more of a theoreticianÕs dream than a practical tool. 
------------------------------------------

Idea: Online to batch reduction with correlated data.

----------  
Vidyasagar, 2003; cares a lot about time-series.
----------

Lozano:
Convergence and consistency of regularized boosting algorithms with stationary beta-mixing observations

We study the statistical convergence and consistency of regularized Boosting methods, where the samples are not independent and identi- cally distributed (i.i.d.) but come from empirical processes of stationary ?-mixing sequences. Utilizing a technique that constructs a sequence of independent blocks close in distribution to the original samples, we prove the consistency of the composite classifiers resulting from a regulariza- tion achieved by restricting the 1-norm of the base classifiersÕ weights. When compared to the i.i.d. case, the nature of sampling manifests in the consistency result only through generalization of the original condition on the growth of the regularization parameter.

A practical motivation for our study of non i.i.d. sam- pling is that in many learning applications observations are intrinsically temporal and hence often weakly dependent. Ignoring this dependency could seriously undermine the perfor- mance of the learning process (for instance, information related to the time-dependent or- dering of samples would be lost). Recognition of this issue has led to several studies of non i.i.d. sampling [6, 7, 8, 9, 10, 11, 12].

Basically these guys also figure that for consistency the regularization has to be adapted to the mixing rate.
----------------------

Estimating $\ beta $-mixing coefficients
Daniel J McDonald, Cosma Rohilla Shalizi, Mark Schervish

Good intro (short para, summarizing many previous works).

We have shown that our estimator of the ?-mixing coefficients is consistent for the true coefficients ?(a) under some conditions on the data generating process. There are numerous results in the statistics and ma- chine learning literatures which assume knowledge of the ?-mixing coefficients, yet as far as we know, this is the first estimator for them. An ability to estimate these coefficients will allow researchers to apply ex- isting results to dependent data without the need to arbitrarily assume their values. Despite the obvious utility of this estimator, as a consequence of its novelty, it comes with a number of potential extensions which warrant careful exploration as well as some drawbacks.
The reader will note that Theorem 2.3 does not pro- vide a convergence rate. The rate in Theorem 2.4 ap- plies only to the difference between ?×Â¦d(a) and ?d(a). In order to provide a rate in Theorem 2.3, we would need a better understanding of the non-stochastic con- vergence of ?d(a) to ?(a). It is not immediately clear that this quantity can converge at any well- defined rate. In particular, it seems likely that the rate of convergence depends on the tail of the sequence { ? ( a ) } ÂÂ°a = 1 .

Indeed, conv. rate depends on itself. This is one thing we resolve.
--------------------
RL examples

--------------------
Computable exponential bounds for screened estimation and simulation
I Kontoyiannis, S Meyn

Exponential bounds and stopping rules for MCMC and general Markov chains
I Kontoyiannis, L Lastras-MontaÂÂ–o, S Meyn

 We develop explicit, general bounds for the probability that the empirical sample averages of a func- tion of a Markov chain on a general alphabet will exceed the steady-state mean of that function by a given amount. Our bounds combine simple information-theoretic ideas together with techniques from optimization and some fairly elementary tools from analysis. In one direction, motivated by central problems in simulation, we de- velop bounds for the general class of Ògeometrically ergodicÓ Markov chains. These bounds take a form that is particularly suited to simulation problems, and they naturally lead to a new class of sampling criteria. These are illustrated by several examples.

 In this section we present a new, general-purpose stopping rule, which can be easily applied to a va- riety of Markov Chain Monte Carlo problems. In Section III-A we treat a simple example illustrating the stopping rule in a simple setting, Section III-B contains the description of the general stopping rule, and in Section III-C we present a non-asymptotic bound that justifies this stopping rule and also offers precise performance guarantees for its use.

--

Markov Processes

Markov Reward Processes
Output sequence of a Markov chain

--------------------
Negative interpretation:
In some Markov chains estimation will be slow (tied to the size of the state space).
Can we adapt to the actual complexity?
Think of aggregated chains.

--------------------

\fi

### 2015-06-09 nips2015/matrix-est.tex
reorganize files

--- 2015-06-01 matrix-est.tex
@@
 \input{__includes}
-\renewcommand\citet\citep % since NIPS is using numbered refs
+\renewcommand\citet\cite % since NIPS is using numbered refs
+\renewcommand\citep\cite % since NIPS is using numbered refs
 
@@
 \title{%
-  Mixing time estimation in reversible Markov chains from a single
-  sample path%
+  Mixing Time Estimation in Reversible Markov Chains from a Single
+  Sample Path%
 }
@@
 \begin{document}
+\begin{bibunit}[myunsrt]
   \maketitle
@@
 
-\section{Notation}\label{sec:notation}
-\input{notation}
+%  \section{Notation}\label{sec:notation}
+%  \input{notation}
 
-\section{Setting}\label{sec:setting}
+  \section{Preliminaries}\label{sec:prelim}
   \input{setting}
 
-%\section{Results}
-%\input{results}
-
-\section{Convergence rates}\label{sec:rates}
+  \section{Point estimation}\label{sec:rates}
   \input{rates}
@@
   \subsubsection*{References}
-{\def\section*#1{}\small \bibliography{all} \bibliographystyle{myunsrt}}
+  {\def\section*#1{}\small\putbib[all]}
+\end{bibunit}
 
@@
 
+\begin{bibunit}[myunsrt]
   \section{Proof of \cref{thm:err}}\label{app:upper}
@@
 
-\section{Proofs of lower bounds}\label{app:lower}
+  \section{Proofs of the lower bounds}\label{app:lower}
   \input{lowerbound}
@@
 
+  \subsubsection*{References}
+  {\def\section*#1{}\small\putbib[all]}
+\end{bibunit}
+
 \end{document}

### 2015-06-09 nips2015/notation.tex
reorganize files

--- 2015-06-03 setting.tex
@@
 %!TEX root =  matrix-est.tex
-
-\subsection{Notations}
-\label{sec:notation}
-
 We denote the set of positive integers by $\bbN$, and
@@
 The non-negative part of a real number $x$ is $[x]_+ := \max\{0,x\}$.
-We use $\ln(\cdot)$ in expressions with exact constants, while we use $\log(\cdot)$ elsewhere.
-%In expression with exact numerical constants, we will use 
-%the natural logarithm function, $\ln(x)$, while elsewhere we use $\log(x)$.
+We will use $\ln(x)$, to denote the natural logarithm of $x$
+in expressions with exact numerical constants,
+while we use $\log(x)$ elsewhere.
 Boldface symbols are used for vectors and matrices (e.g., $\vv$,
@@
 %$\vp, \vq \in \Delta^{d-1}$ is denoted by $\norm{\vp-\vq}_{\tv}$.
-
-\subsection{Setting}
-\label{sec:setting}
-
-Let $\vP \in (\Delta^{d-1})^d \subset [0,1]^{d \times d}$ be a $d
-\times d$ row-stochastic matrix for an ergodic (i.e., irreducible and
-aperiodic) Markov chain.
-This implies there is a unique stationary distribution $\vpi \in
-\Delta^{d-1}$ with $\pi_i > 0$ for all $i \in [d]$~\citep[Corollary
-1.17]{LePeWi08}.
-We also assume that $\vP$ is \emph{reversible} (with respect to
-$\vpi$):
-\begin{align}
-\label{eq:reversibility}
-  \pi_i P_{i,j} = \pi_j P_{j,i} ,
-  \quad i,j \in [d] .
-\end{align}
-The minimum stationary probability is denoted by $\pimin := \min_{i
-\in [d]} \pi_i$.
-
-Define the matrices
-\begin{align*}
-\vM := \Diag(\vpi) \vP \quad \text{and} \quad
-\vL := \Diag(\vpi)^{-1/2} \vM \Diag(\vpi)^{-1/2}\,.
-\end{align*}
-The $(i,j)$th entry of the matrix $M_{i,j}$ contains the \emph{doublet
-probabilities} associated with $\vP$: $M_{i,j} = \pi_i P_{i,j}$ is the
-probability of seeing state $i$ followed by state $j$ when the chain
-is started from its stationary distribution.
-The matrix $\vM$ is symmetric on account of the reversibility of
-$\vP$, and hence it follows that $\vL$ is also symmetric.
-(We will strongly exploit the symmetry in our results.)
-%will be advantageous due to the existence of stronger perturbation results
-%for symmetric matrices.
-Further, $\vL = \Diag(\vpi)^{1/2} \vP \Diag(\vpi)^{-1/2}$, hence $\vL$
-and $\vP$ are similar and thus their eigenvalue systems are identical.
-Ergodicity and reversibility imply that the eigenvalues of $\vL$ are
-contained in the interval $\opencloseint{-1,1}$, and that $1$ is an
-eigenvalue of $\vL$ with multiplicity $1$~\citep[Lemmas 12.1 and
-12.2]{LePeWi08}.
-Denote and order the eigenvalues of $\vL$ as
-\[
-  1 = \lambda_1 > \lambda_2 \geq \dotsb \geq \lambda_d > -1 .
-\]
-Let $\slem := \max\{ \lambda_2,\, |\lambda_d| \}$, and define the
-(absolute) spectral gap to be $\gap := 1-\slem$, which is strictly
-positive on account of ergodicity.
-\todoc{Do our results generalize to the spectral gap? No.
-Discuss this somewhere.}
-%Throughout the paper, both $\pimin$ and $\gap$ are assumed
-%to be positive.
-
-Let $(X_t)_{t\in\bbN}$ be a Markov chain whose transition
-probabilities are governed by $\vP$.
-For each $t \in \bbN$, let $\vpi^{(t)} \in \Delta^{d-1}$ denote the
-marginal distribution of $X_t$, so
-\[
-  \vpi^{(t+1)} = \vpi^{(t)} \vP ,
-  \quad t \in \bbN .
-\]
-Note that the initial distribution $\vpi^{(1)}$ is arbitrary,
-and need not be the stationary distribution $\vpi$.
-
-The goal is to estimate $\pimin$ and $\gap$ from the length $n$ sample
-path $(X_t)_{t \in [n]}$, and also to construct fully empirical
-confidence intervals that $\pimin$ and $\gap$ with high probability;
-in particular, the construction of the intervals should not depend on
-any unobservable quantities, including $\pimin$ and $\gap$ themselves.
-As mentioned in the introduction,
-it is well-known that the \emph{mixing time} of the Markov chain
-$\tmix$ (defined in Eq.~\ref{eq:mixtimedef})
-is bounded in terms of $\pimin$ and $\gap$, as shown in
-\cref{eq:mixing-time-bound}.
-Moreover, convergence rates for empirical processes on Markov chain
-sequences are also often given in terms of mixing coefficients that
-can ultimately be bounded in terms of $\pimin$ and
-$\gap$ (as we will show in the proof of our first result).
-%~\citep[e.g.,][]{MR1921877,MoRo08,MoRo09}. 
-Therefore, valid confidence intervals for $\pimin$ and $\gap$ can be
-used to make these rates fully observable.
-
-%Therefore, confidence intervals for $\pimin$ and $\gap$ can be used to
-%make these rates data-dependent and hence usable in procedures such
-%as Structural Risk Minimization.
-%\todoc{citation for SRM, and other applications?}
-%
-%Additionally, we aim to provide
-%asymptotic convergence rates
-%as well as data-dependent confidence intervals;
-%the latter should not depend on any unknown quantities
-%(such as $\pimin$ and $\gap$).
-%
-%\todoc[inline]{What is the data? What is the problem?}

### 2015-06-09 nips2015/setting.tex
reorganize files

--- 2015-06-03 setting.tex
@@
 used to make these rates fully observable.
+%All probabilities in this paper are over the space of state sequences with the given Markov measure.
+\todoa{Add: ``All probabilities in this paper are over the space of state sequences with the given Markov measure.'' 
+Cs: I think the probability space could be arbitrary
+and the probabilities are over this (implicit) probability space as usual in probability papers.
+I would suggest to remove this sentence. Also, we are short of space.}
 

### 2015-06-09 arxiv-v1/appendix-combined.tex
arxiv version?

\newcommand\LB{\ensuremath{\operatorname{lb}}}
\newcommand\piminlb{\ensuremath{\hat\pi_{\star,\LB}}}
\newcommand\gaplb{\ensuremath{\hat\gamma_{\star,\LB}}}
Let $\piminlb$ and $\gaplb$ be the lower bounds on $\pimin$ and
$\gap$, respectively, computed from \cref{alg:empest}.
Let $\hatpimin$ and $\hatgap$ be the estimates of $\pimin$ and $\gap$
computed using the estimators from \cref{thm:err}.
By a union bound, we have by \cref{thm:err} and \cref{thm:empirical}
that with probability at least $1-2\delta$,
\begin{equation}
  \Abs{\hatpimin-\pimin}
  \le
  C \,
  \Parens{
    \sqrt{\frac{\pimin\log\frac{d}{\piminlb\delta}}{\gaplb n}}
    +
    \frac{\log\frac{d}{\piminlb\delta}}{\gaplb n}
  }
  \label{eq:pimin-plugin}
\end{equation}
and
\begin{equation}
  \Abs{\hatgap-\gap}
  \leq
  C \,
  \Parens{
    \sqrt{\frac{\log\frac{d}{\delta}\cdot\log\frac{n}{\piminlb\delta}}{\piminlb\gaplb n}}
    +
    \frac{\log\frac{d}{\delta}\cdot\log\frac{n}{\piminlb\delta}}{\piminlb\gaplb n}  
    + \frac{\log\frac{1}{\gaplb}}{\gaplb n}  
  }
  \,.
  \label{eq:gap-plugin}
\end{equation}
The bound on $\abs{\hatgap-\gap}$ in \cref{eq:gap-plugin}---call it
$\hat{w}'$---is fully observable and hence yields a confidence
interval for $\gap$.
The bound on $\abs{\hatpimin-\pimin}$ in \cref{eq:pimin-plugin}
depends on $\pimin$, but from it one can derive
\[
  \Abs{\hatpimin-\pimin}
  \le
  C' \,
  \Parens{
    \sqrt{\frac{\hatpimin\log\frac{d}{\piminlb\delta}}{\gaplb n}}
    +
    \frac{\log\frac{d}{\piminlb\delta}}{\gaplb n}
  }
\]
using the approach from the proof of \cref{lem:P-obs-bound}.
Here, $C'>0$ is an absolute constant that depends only on $C$.
This bound---call it $\hat{b}'$---is now also fully observable.
We have established that in the $1-2\delta$ probability event from
above,
\[
  \pimin \in \wh{U} := [\hatpimin-\hat{b}', \hatpimin+\hat{b}']
  , \quad
  \gap \in \wh{V} := [\hatgap-\hat{w}', \hatgap+\hat{w}']
  .
\]
It is easy to see that almost surely (as $n \to \infty$),
\[
  \sqrt{n / \log(n)} \hat{w}'
  = O\Parens{
    \sqrt{\frac{\log(d/\delta)}{\pimin\gap}}
  }
\]
and
\[
  \sqrt{n} \hat{b}'
  = O\Parens{
    \sqrt{\frac{\pimin \log\frac{d}{\pimin\delta}}{\gap}}
  }
  .
\]
This completes the proof of \cref{thm:combined}.
\hfill\qed

### 2015-06-09 arxiv-v1/conclusions.tex
arxiv version?

--- 2015-06-03 conclusions.tex
@@
 %!TEX root =  matrix-est.tex
-The construction of \cref{thm:combined}
-suggests a general procedure for converting
-any confidence interval $I_n=I_n(\gap,\pimin)$ for the spectral gap $\gap$,
-which
-%itself depends 
-%on
-is expressed in terms of
-$\gap$ and $\pimin$,
-to a fully empirical confidence interval whose
-width matches that of $I_n$ up to lower order terms:
-Plug in the empirical bounds of \cref{alg:empest} into the expression defining $I_n$.
-\todoa{reworded}
+The construction used in \cref{thm:combined} applies more generally:
+%to convert confidence intervals trapping the spectral gap under some conditions:
+Given a confidence interval of the form $I_n = I_n(\gap,\pimin,\delta)$ 
+for some confidence level $\delta$
+and a fully empirical confidence set $E_n(\delta)$ for $(\gap,\pimin)$  for the same level,
+$I_n' = E_n(\delta) \cap \cup_{(\gamma,\pi)\in E_n(\delta)} I_n(\gamma,\pi,\delta)$ is a valid
+fully empirical $2\delta$-level confidence interval whose asymptotic width
+matches that of $I_n$ up to lower order terms under reasonable assumptions on $E_n$ and $I_n$.
+\todoa{reworded} \todoc{more rewording}
 In particular, this suggests that future work should focus on 
@@
 %serve as a source of inspiration for others to consider these questions.
-\todoa{rewrote ending}
 \if0

### 2015-06-09 arxiv-v1/paper.tex
arxiv version?

--- 2015-06-09 nips2015/matrix-est.tex
@@
-%\documentclass[11pt]{article}
-%\documentclass[anon,12pt]{colt2015} 
 \documentclass{article}
-\usepackage{nips15submit_e,times}
 \input{__includes}
@@
 
-%\title{Single sample path based estimation of the mixing time in finite, reversible Markov chains}
 \title{%
@@
 
-\author{
-David S.~Hippocampus\thanks{ Use footnote for providing further information
-about author (webpage, alternative address)---\emph{not} for acknowledging
-funding agencies.} \\
-Department of Computer Science\\
-Cranberry-Lemon University\\
-Pittsburgh, PA 15213 \\
-\texttt{hippo@cs.cranberry-lemon.edu} \\
-\And
-Coauthor \\
-Affiliation \\
-Address \\
-\texttt{email} \\
-\AND
-Coauthor \\
-Affiliation \\
-Address \\
-\texttt{email} \\
-\And
-Coauthor \\
-Affiliation \\
-Address \\
-\texttt{email} \\
-\And
-Coauthor \\
-Affiliation \\
-Address \\
-\texttt{email} \\
-(if needed)\\
-}
+\author[1]{\mbox{Daniel Hsu}}
+\author[2]{\mbox{Aryeh Kontorovich}}
+\author[3]{\mbox{Csaba Szepesv\'ari}}
+\affil[1]{\mbox{Columbia University}}
+\affil[2]{\mbox{Ben-Gurion University}}
+\affil[3]{\mbox{University of Alberta}}
 
 \begin{document}
-\begin{bibunit}[myunsrt]
 \maketitle
@@
 
-  \if0
-  \begin{keywords}
-    Markov chain; mixing time; spectral gap;
-    fully data-dependent confidence interval
-  \end{keywords}
-  \fi
-
 \section{Introduction}\label{sec:intro}
-  \input{intro-nips}
-
-%  \section{Notation}\label{sec:notation}
-%  \input{notation}
+\input{intro}
 
@@
 
-  \subsubsection*{References}
-  {\def\section*#1{}\small\putbib[all]}
-\end{bibunit}
-
-\newpage
+\bibliography{all}
+\bibliographystyle{plain}
 
@@
 
-\begin{bibunit}[myunsrt]
 \section{Proof of \cref{thm:err}}\label{app:upper}
-  \input{upperbound-new}
+\input{upperbound}
 
@@
 
-  \subsubsection*{References}
-  {\def\section*#1{}\small\putbib[all]}
-\end{bibunit}
-
 \end{document}

### 2015-06-09 arxiv-v1/appendix-empirical.tex
remove comments

In this section, we derive \cref{alg:empest} and prove
\cref{thm:empirical}.

\subsection{Estimators for $\vpi$ and $\gap$}

The algorithm forms the estimator $\wh\vP$ of $\vP$ using Laplace smoothing:
\[
  \wh{P}_{i,j}
  := \frac{N_{i,j} + \alpha}{N_i + d\alpha}
\]
where
\[
  N_{i,j} := \Abs{ \Braces{ t \in [n-1] : (X_t,X_{t+1}) = (i,j) } } ,
  \quad
  N_i := \Abs{ \Braces{ t \in [n-1] : X_t = i } }
\]
and $\alpha>0$ is a positive constant, which we set beforehand as $\alpha := 1/d$
for simplicity.

As a result of the smoothing, all entries of $\wh\vP$ are positive,
and hence $\wh\vP$ is a transition probability
matrix for an ergodic Markov chain.
We let $\hat\vpi$ be the unique stationary distribution for $\wh\vP$.
Using $\hat\vpi$, we form an estimator $\Sym(\wh\vL)$ of $\vL$ using:
\[
  \Sym(\wh\vL) := \frac12 \parens{ \wh\vL + \wh\vL^\t }
  ,
  \qquad
  \wh\vL := \Diag(\hat\vpi)^{1/2} \wh\vP \Diag(\hat\vpi)^{-1/2}
  .
\]
Let $\hat\lambda_1 \geq \hat\lambda_2 \geq \dotsb \geq \hat\lambda_d$
be the eigenvalues of $\Sym(\wh\vL)$ (and in fact, we have $1 =
\hat\lambda_1 > \hat\lambda_2$ and $\hat\lambda_d > -1$).
The algorithm estimates the spectral gap $\gap$ using
\[
  \hatgap := 1 - \max\{ \hat\lambda_2, |\hat\lambda_d| \} .
\]

\subsection{Empirical bounds for $\vP$}
\label{sec:P-obs-bound}

\begin{lemma}
  \label{lem:P-unobs-bound}
  The following holds with probability at least $1-\delta$: for all
  $(i,j) \in \iset{d}^2$,
  \begin{equation}
    \abs{ \wh{P}_{i,j} - P_{i,j} }
    \leq
    \sqrt{\Parens{\frac{N_i}{N_i+d\alpha}}\frac{2P_{i,j}(1-P_{i,j})\ln(2d/\delta)}{N_i+d\alpha}}
    + \frac{\frac23\ln(2d/\delta)}{N_i+d\alpha}
    + \frac{\abs{\alpha-d\alpha P_{i,j}}}{N_i+d\alpha}
    .
    \label{eq:P-unobs-bound}
  \end{equation}
\end{lemma}
\begin{proof}
  For each $t \in \bbN$, define $\cF_t$ to be the $\sigma$-field
  generated by $X_1,X_2,\dotsc,X_t$.
  Fix a pair $(i,j) \in [d]^2$, and define
  \[
    Y_{t+1} := \Ind{X_t = i} (\Ind{X_{t+1}=j} - P_{i,j})
    , \quad t \in [n-1]
  \]
  so that
  \[
    \sum_{t=1}^{n-1} Y_{t+1}
    = N_{i,j} - N_i P_{i,j}
    .
  \]
  By the Markov property, the stochastic process $\{ Y_{t+1} \}_{t \in
  [n-1]}$ is an $\cF_t$-adapted martingale difference sequence:
  $Y_{t+1}$ is $\cF_{t+1}$-measurable and $\EE{Y_{t+1}| \cF_t } = 0$,
  for each $t \in [n-1]$.
  Moreover, for each $t \in [n-1]$,
  \[
    \EE{Y_{t+1}^2| \cF_t } = \Ind{X_t=i} P_{i,j} (1-P_{i,j})
  \]
  and
  \[
    Y_{t+1} \in [-P_{i,j},1-P_{i,j}] .
  \]
  Therefore, by Freedman's inequality~\citep[Theorem 1.6]{Fre75} and
  union bounds, we have
  \[
    \abs{ N_{i,j} - N_i P_{i,j} }
    \leq \sqrt{2 N_i P_{i,j} (1-P_{i,j}) \ln(2d/\delta)} + \frac23
    \ln\left(\frac{2d}{\delta}\right)
  \]
  for all $(i,j) \in [d]^2$.
\end{proof}

\Cref{eq:P-unobs-bound} can be viewed as constraints on the possible
value that $P_{i,j}$ may have (with high probability).
Since $P_{i,j}$ is the only unobserved quantity in the bound from
\cref{eq:P-unobs-bound}, we can numerically maximize $|\wh{P}_{i,j} -
P_{i,j}|$ subject to the constraint in \cref{eq:P-unobs-bound}
(viewing $P_{i,j}$ as the optimization variable).
Let $B_{i,j}^*$ be this maximum value, so we have
\[
  P_{i,j} \in
  \Brackets{
    \wh{P}_{i,j} - B_{i,j}^*,\,
    \wh{P}_{i,j} + B_{i,j}^*
  }
\]
in the same event where \cref{eq:P-unobs-bound} holds.

In the algorithm, we give a simple alternative to computing
$B_{i,j}^*$ that avoids numerical optimization, derived in the spirit
of empirical Bernstein bounds~\citep{audibert2009}.
Specifically, we compute
\begin{equation}
  B_{i,j} :=
  \sqrt{\frac{9\wh{P}_{i,j}(1-\wh{P}_{i,j})\ln(2d/\delta)}{2N_i}}
  + \frac{4\ln(2d/\delta)+1.5\abs{\alpha-d\alpha\wh{P}_{i,j}}}{N_i}
  \label{eq:P-obs-bound}
\end{equation}
for each $(i,j) \in [d]^2$.
We show in \cref{lem:P-obs-bound} that
\[
  P_{i,j} \in
  \Brackets{
    \wh{P}_{i,j} - B_{i,j},\,
    \wh{P}_{i,j} + B_{i,j}
  }
\]
again, in the same event where \cref{eq:P-unobs-bound} holds.
The observable bound in \cref{eq:P-obs-bound} is not too far from the
unobservable bound in~\cref{eq:P-unobs-bound}.

\begin{lemma}
  \label{lem:P-obs-bound}
  In the same $1-\delta$ event as from \cref{lem:P-unobs-bound},
  we have for all $(i,j) \in [d]^2$,
  \begin{equation*}
    \abs{\wh{P}_{i,j} - P_{i,j}}
    \leq
    \sqrt{\frac{9\wh{P}_{i,j}(1-\wh{P}_{i,j})\ln(2d/\delta)}{2N_i}}
    + \frac{4\ln(2d/\delta)+1.5\abs{\alpha-d\alpha\wh{P}_{i,j}}}{N_i}
    .
  \end{equation*}
\end{lemma}
\begin{proof}
  Recall that in the $1-\delta$ probability event from
  \cref{lem:P-unobs-bound}, we have for all $(i,j) \in [d]^2$,
  \begin{multline*}
    \abs{ \wh{P}_{i,j} - P_{i,j} }
    =
    \Abs{
      \frac{N_{i,j} - N_i P_{i,j}}{N_i + d\alpha}
      + \frac{\alpha - d\alpha P_{i,j}}{N_i + d\alpha}
    } \\
    \leq
    \sqrt{\frac{2N_iP_{i,j}(1-P_{i,j})\ln(2d/\delta)}{(N_i+d\alpha)^2}}
    + \frac{\frac23\ln(2d/\delta)}{N_i+d\alpha}
    + \frac{\abs{\alpha-d\alpha P_{i,j}}}{N_i+d\alpha}
    .
  \end{multline*}
  Applying the triangle inequality to the right-hand side, we obtain
  \begin{align}
    \abs{ \wh{P}_{i,j} - P_{i,j} }
    & \leq
    \sqrt{
      \frac{
        2N_i\parens{
          \wh{P}_{i,j}(1-\wh{P}_{i,j})
          + \abs{\wh{P}_{i,j} - P_{i,j}}
        } \ln(2d/\delta)
      }{(N_i+d\alpha)^2}
    }
    + \frac{\frac23\ln(2d/\delta)}{N_i+d\alpha}
    \notag \\
    & \qquad
    + \frac{
      \abs{\alpha-d\alpha \wh{P}_{i,j}}
      + d\alpha\abs{\wh{P}_{i,j}-P_{i,j}}
    }{N_i+d\alpha}
    .
    \notag
  \end{align}
  Since $\sqrt{A+B} \leq \sqrt{A} + \sqrt{B}$ for non-negative $A,B$, we
  loosen the above inequality and rearrange it to obtain
  \begin{align*}
    \Parens{ 1 - \frac{d\alpha}{N_i+d\alpha} }
    \abs{\wh{P}_{i,j} - P_{i,j}}
    & \leq
    \sqrt{\abs{\wh{P}_{i,j} - P_{i,j}}} \cdot
    \sqrt{\frac{2N_i\ln(2d/\delta)}{(N_i+d\alpha)^2}}
    \\
    & \qquad
    +
    \sqrt{\frac{2N_i\wh{P}_{i,j}(1-\wh{P}_{i,j})\ln(2d/\delta)}{(N_i+d\alpha)^2}}
    + \frac{\frac23\ln(2d/\delta) + \abs{\alpha -
    d\alpha\wh{P}_{i,j}}}{N_i+d\alpha}
    .
  \end{align*}
  Whenever $N_i > 0$, we can solve a quadratic inequality and apply
  AM/GM to conclude the claimed inequalities.
\end{proof}

\subsection{Empirical bounds for $\vpi$}

Recall that $\hat\vpi$ is obtained as the unique stationary
distribution for $\wh\vP$.
Let $\wh\vA := \vI - \wh\vP$, and let $\giAh$ be the \emph{group
inverse} of $\wh\vA$---i.e., the unique square matrix satisfying the
following equalities:
\[
  \wh\vA\giAh\wh\vA = \wh\vA , \quad
  \giAh\wh\vA\giAh = \giAh , \quad
  \giAh\wh\vA = \wh\vA\giAh .
\]
The matrix $\giAh$, which is well defined no matter what transition probability matrix $\wh\vP$ we start with 
\citep{meyer1975role},
 is a central quantity that captures many properties
of the ergodic Markov chain with transition matrix
$\wh\vP$~\citep{meyer1975role}.
We denote the $(i,j)$-th entry of $\giAh$ by $\giAh_{i,j}$.
Define
\begin{equation}
  \hat\kappa :=
  \frac12
  \max
  \Braces{
    \giAh_{j,j}
    - \min\Braces{ \giAh_{i,j} : i \in [d] }
    : j \in [d]
  } 
  .
  \notag
\end{equation}
Analogously define
\begin{align*}
  \vA & := \vI - \vP , \\
  \giA & := \text{group inverse of $\vA$} , \\
  \kappa & :=
  \frac12
  \max
  \Braces{
    \giA_{j,j}
    - \min\Braces{ \giA_{i,j} : i \in [d] }
    : j \in [d]
  } 
  .
\end{align*}
We now use the following perturbation bound from \citep[Section
3.3]{cho2001comparison} (derived
from~\citet{haviv1984perturbation,kirkland1998applications}).
\begin{lemma}[\citep{haviv1984perturbation,kirkland1998applications}]
  \label{lem:pi-perturb}
  If
  $\abs{\wh{P}_{i,j} - P_{i,j}} \leq \wh{B}_{i,j}$ for each $(i,j) \in
  [d]^2$, then
  \begin{align}
    \max\Braces{
      \abs{\hat\pi_i - \pi_i}
      : i \in [d]
    }
    & \leq \min\{\kappa,\hat\kappa\} \max
    \braces{
      \wh{B}_{i,j}
      : (i,j) \in [d]^2
    }
    \notag \\
    & \leq \hat\kappa \max
    \braces{
      \wh{B}_{i,j}
      : (i,j) \in [d]^2
    }
    .
    \notag
  \end{align}
\end{lemma}
This establishes the validity of the confidence intervals for the
$\pi_i$ in the same event from \cref{lem:P-unobs-bound}.

We now establish the validity of the bounds for the ratio quantities
$\sqrt{\hat\pi_i/\pi_i}$ and $\sqrt{\pi_i/\hat\pi_i}$.
\begin{lemma}
  \label{lem:pi-ratio-perturb}
  If $\max\braces{\abs{\hat\pi_i-\pi_i} : i \in [d]} \leq \hat{b}$,
  then
  \begin{equation}
    \max\bigcup_{i\in[d]}
    \braces{
      \abs{\sqrt{\pi_i/\hat\pi_i}-1},\,
      \abs{\sqrt{\hat\pi_i/\pi_i}-1}
    }
    \leq
    \frac12 \max \bigcup_{i\in[d]}
    \Braces{
      \frac{\hat{b}}{\hat\pi_i},\,
      \frac{\hat{b}}{\brackets{\hat\pi_i-\hat{b}}_+}
    }
    .
    \notag
  \end{equation}
\end{lemma}
\begin{proof}
  By \cref{lem:pi-perturb}, we have for each $i \in [d]$,
  \begin{equation}
    \frac{\abs{\hat\pi_i - \pi_i}}{\hat\pi_i}
    \leq \frac{\hat{b}}{\hat\pi_i}
    , \quad
    \frac{\abs{\hat\pi_i - \pi_i}}{\pi_i}
    \leq 
    \frac{\hat{b}}{\pi_i}
    \leq
    \frac{\hat{b}}{\brackets{\hat{\pi}_i-\hat{b}}_+}
    .
    \notag
  \end{equation}
  Therefore, using the fact that for any $x>0$,
  \[
    \max\Braces{
      |\sqrt{x}-1| ,\, |\sqrt{1/x}-1|
    } \leq \frac12 \max\Braces{ |x-1| ,\, |1/x-1| }
  \]
  we have for every $i \in [d]$,
  \begin{align*}
    \max\Braces{
      \abs{\sqrt{\pi_i/\hat\pi_i}-1}
      ,\,
      \abs{\sqrt{\hat\pi_i/\pi_i}-1}
    }
    & \leq
    \frac12 \max\Braces{
      \abs{\pi_i/\hat\pi_i-1}
      ,\,
      \abs{\hat\pi_i/\pi_i-1}
    }
    \\
    & \leq
    \frac12 \max\Braces{
      \frac{\hat{b}}{\hat\pi_i}
      ,\,
      \frac{\hat{b}}{\brackets{\hat{\pi}_i-\hat{b}}_+}
    }
    .
    \qedhere
  \end{align*}
\end{proof}

\subsection{Empirical bounds for $\vL$}

By Weyl's inequality and the triangle inequality,
\[
  \max_{i \in [d]} |\lambda_i - \hat\lambda_i|
  \leq \norm{\vL - \Sym(\wh\vL)}
  \leq \norm{\vL - \wh\vL} .
\]
It is easy to show that $|\hatgap - \gap|$ is bounded by the same
quantity.
Therefore, it remains to establish an empirical bound on $\norm{\vL -
\wh\vL}$.

\begin{lemma}
  \label{lem:L-obs-bound}
  If
  $\abs{\wh{P}_{i,j} - P_{i,j}} \leq \wh{B}_{i,j}$ for each $(i,j) \in
  [d]^2$ and
  $\max\braces{\abs{\hat\pi_i-\pi_i} : i \in [d]} \leq \hat{b}$,
  then
  \[
    \norm{\wh\vL - \vL}
    \leq
    2\hat\rho + \hat\rho^2
    + (1+2\hat\rho+\hat\rho^2)
    \Biggl(
      \sum_{(i,j)\in[d]^2} \frac{\hat\pi_i}{\hat\pi_j} \hat{B}_{i,j}^2
    \Biggr)^{1/2}
    ,
  \]
  where
  \[
    \hat\rho := \frac12 \max \bigcup_{i\in[d]}
    \Braces{
      \frac{\hat{b}}{\hat\pi_i},\,
      \frac{\hat{b}}{\brackets{\hat\pi_i-\hat{b}}_+}
    }
    .
  \]
\end{lemma}
\begin{proof}
  We use the following decomposition of $\vL - \wh\vL$:
  \[
    \vL - \wh\vL
    = \errP
    + \errpil \wh\vL
    + \wh\vL \errpir
    + \errpil \errP
    + \errP \errpir
    + \errpil \wh\vL \errpir
    + \errpil \errP \errpir
  \]
  where
  \begin{align*}
    \errP
    & := \Diag(\hat\vpi)^{1/2} \parens{ \vP - \wh\vP } \Diag(\hat\vpi)^{-1/2} ,
    \\
    \errpil
    & := \Diag(\vpi)^{1/2} \Diag(\hat\vpi)^{-1/2} - \vI ,
    \\
    \errpir
    & := \Diag(\hat\vpi)^{1/2} \Diag(\vpi)^{-1/2} - \vI .
  \end{align*}
  Therefore
  \begin{align*}
    \norm{\vL - \wh\vL}
    & \leq
    \norm{\errpil} + \norm{\errpir} + \norm{\errpil} \norm{\errpir}
    \\
    & \quad
    + \Parens{
      1 + \norm{\errpil} + \norm{\errpir} + \norm{\errpil} \norm{\errpir}
    }
    \norm{\errP}
    .
  \end{align*}
  Observe that for each $(i,j) \in [d]^2$, the $(i,j$)-th entry of
  $\errP$ is bounded in absolute value by
  \[
    |(\errP)_{i,j}|
    = \hat{\pi}_i^{1/2} \hat{\pi}_j^{-1/2} |P_{i,j} - \wh{P}_{i,j}|
    \leq
    \hat\pi_i^{1/2} \hat{\pi}_j^{-1/2} \wh{B}_{i,j}
    .
  \]
  Since the spectral norm of $\errP$ is bounded above by its Frobenius
  norm,
  \[
    \norm{\errP}
    \leq \Biggl(
      \sum_{(i,j) \in [d]^2}
      (\errP)_{i,j}^2
    \Biggr)^{1/2}
    \leq \Biggl(
      \sum_{(i,j) \in [d]^2}
      \frac{\pi_i}{\pi_j} \wh{B}_{i,j}^2
    \Biggr)^{1/2}
    .
  \]
  Finally, the spectral norms of $\errpil$ and $\errpir$ satisfy
  \[
    \max\Braces{ \norm{\errpil} ,\, \norm{\errpir} }
    =
    \max\bigcup_{i\in[d]}
    \braces{
      \abs{\sqrt{\pi_i/\hat\pi_i}-1},\,
      \abs{\sqrt{\hat\pi_i/\pi_i}-1}
    }
    ,
  \]
  which can be bounded using \cref{lem:pi-ratio-perturb}.
\end{proof}

This establishes the validity of the confidence interval for $\gap$ in
the same event from \cref{lem:P-unobs-bound}.

\subsection{Asymptotic widths of intervals}
\label{sec:asymptotic}

Let us now turn to the asymptotic behavior of the interval widths
(regarding $\hat{b}$, $\hat\rho$, and $\hat{w}$ all as
functions of $n$).

A simple calculation gives that, almost surely, as $n\to \infty$,
\begin{align*}
  \sqrt{n} \hat{b}
  & =
  O\Parens{
    \max_{i,j} \kappa \sqrt{\frac{P_{i,j} \log(d/\delta)}{\pi_i}}
  }
  ,
  \\
  \sqrt{n} \hat\rho
  & =
  O\Parens{
    \frac{\kappa}{\pimin} \sqrt{\frac{\log(d/\delta)}{\pimin}}
  }
  .
\end{align*}
Here, we use the fact that $\hat\kappa \to \kappa$ as $n\to\infty$
since $\giAh \to \giA$ as $\wh\vP \to
\vP$~\citep{li2001improvement,benitez2012continuity}.

Further, since
\begin{align*}
  \sqrt{n}
  \Biggl(
    \sum_{i,j} \frac{\hat{\pi}_i}{\hat{\pi}_j} \wh{B}_{i,j}^2
  \Biggr)^{1/2}
  & =
  O\Parens{
    \Biggl(
      \sum_{i,j} \frac{{\pi}_i}{{\pi}_j}
      \cdot \frac{P_{i,j}(1-P_{i,j})\log(d/\delta)}{\pi_i}
    \Biggr)^{1/2}
  }
  & =
  O\Parens{ \sqrt{ \frac{d\log(d/\delta)}{ \pimin}} }
  .
\end{align*}
we thus have
\begin{align*}
  \sqrt{n} \hat{w}
  =
  O\Parens{ 
    \frac{\kappa}{\pimin} \sqrt{\frac{\log(d/\delta)}{\pimin}} + 
    \sqrt{ \frac{d\log(d/\delta)}{ \pimin}}
  }
  .
\end{align*}
This completes the proof of \cref{thm:empirical}.
\hfill $\qed$

The following claim provides a bound on $\kappa$ in terms of the
number of states and spectral gap.
\begin{claim}
  \label{claim:kappa-bound}
  $\kappa \leq d/\gap$.
\end{claim}
\begin{proof}
  It is established in~\citep{cho2001comparison} that
  \[
    \kappa
    \leq \max_{i,j} |\giA_{i,j}|
    \leq
    \sup_{\norm{\vv}_1 = 1 , \dotp{\vv,\v1} = 0} \norm{\vv^\t \giA}_1
  \]
  (our $\kappa$ is the $\kappa_4$ quantity from
  \citep{cho2001comparison}), and \citep{seneta1993sensitivity}
  establishes
  \[
    \sup_{\norm{\vv}_1 = 1 , \dotp{\vv,\v1} = 0} \norm{\vv^\t \giA}_1
    \leq
    \frac{d}{\gap}
    .
    \qedhere
  \]
\end{proof}

### 2015-06-09 arxiv-v1/conclusions.tex
remove comments

--- 2015-06-03 conclusions.tex
@@
-%!TEX root =  matrix-est.tex
-Generalizing the construction of  \cref{thm:combined}, 
-given any confidence interval $I_n=I_n(\gap,\pimin)$ for the spectral gap $\gap$ that itself depends 
-on $\gap$ and $\pimin$, under general conditions,
-plugging in the empirical bounds of \cref{alg:empest} into the expression defining $I_n$
-gives a fully empirical confidence interval whose
-width matches that of $I_n$ up to lower order terms.
+The construction used in \cref{thm:combined} applies more generally:
+Given a confidence interval of the form $I_n = I_n(\gap,\pimin,\delta)$ 
+for some confidence level $\delta$
+and a fully empirical confidence set $E_n(\delta)$ for $(\gap,\pimin)$  for the same level,
+$I_n' = E_n(\delta) \cap \cup_{(\gamma,\pi)\in E_n(\delta)} I_n(\gamma,\pi,\delta)$ is a valid
+fully empirical $2\delta$-level confidence interval whose asymptotic width
+matches that of $I_n$ up to lower order terms under reasonable assumptions on $E_n$ and $I_n$.
 In particular, this suggests that future work should focus on 
@@
 with a few factors, are a promising candidate for such future investigations.
-While we realize the limitations of our work, 
-we think that the area of fully data dependent estimation of Markov chain 
-parameters based on a single sample path is full of exciting, important
-and challenging research questions and we hope that our work will
-serve as a source of inspiration for others to consider these questions.
-\if0
-The algorithm from Section~\ref{sec:empirical} demonstrates the
-feasibility of fully empirical finite-sample confidence bounds on
-Markov chain mixing time.
-This makes it possible to apply learning methods that require
-empirical bounds to many non-iid settings that are commonplace in many
-applications.
-%One limitation of our techniques is that they are limited to Markov
-%chains over finite state spaces.
-We expect that future results will generalize and tighten our
-analysis, possibly extend it to large (or infinite) state spaces by
-adding extra assumptions.
-
-A salient limitation of our approach is the reversibility assumption.
-Without the latter, one could still obtain non-trivial finite-sample
-estimates.
-One possible starting point is to work with a pointwise estimate $\wh
-P_{ij}$ and invoke the Ostrowski-Elsner theorem
-\citep{stewart1990matrix} to bound the deviation between the empirical
-$\hatgap$ and the true $\gap$.
-Unfortunately, this bound has an exponential dependence on $d$.
-Another problem we leave open for now is to close the gap between the
-upper bound in Theorem~\ref{thm:err} and the lower bounds in
-Theorems~\ref{thm:lb-pimin} and \ref{thm:lb-gap}. 
-\todoc{Reduce space complexity? Assume more structure of $\vP$?
-Sometimes $\gap$ is known, but $\pimin$ is not. Does this help?
-How about the other way around?
-}
-\fi
-%\todoc{We need to move the references needed in the appendix to after the appendix. 
-%This should be possible with some latex package for splitting references..}
-%and investigate
-%whether the width of the fully empirical confidence interval can be
-%shortened to match the width of the theoretical confidence interval.
-
-%The upshot is that now the various Markov PAC learning results stated in
-%terms of unknown mixing coefficients
-%\citep{gamarnik03,MoRo09}
-%can be applied to give nontrivial data-dependent bounds,
-%albeit in a limited setting.
+The results presented here are a first step in the ambitious research agenda
+outlined above, and we hope that they will
+serve as a point of departure
+for
+further insights
+in the area of fully empirical estimation of Markov chain 
+parameters based on a single sample path.

### 2015-06-09 arxiv-v1/empirical.tex
arxiv version?
remove comments

--- 2015-06-03 empirical.tex
@@
-%!TEX root =  matrix-est.tex
 In this section, we address the shortcoming of \cref{thm:err} and give
@@
 use of the smoothed estimates), $\giAh$ can be computed at the cost of
-inverting an $(d-1)\times (d-1)$ matrix~\citep[Theorem
+inverting an $(d{-}1){\times}(d{-}1)$ matrix~\citep[Theorem
 5.2]{meyer1975role}.%
-\footnote{
+\footnote{%
 \label{ftnote:group-inverse}
@@
 $\vA^\#\vA = \vA\vA^\#$.
-%It is efficiently computable,
-%and for $\vA=\vI-\vP$, where $\vP$ is ergodic with stationary distribution $\vpi$,
-%has the
-%following simple
-%form: $\vA^\#=(\vI-\vP+\v1\vpi^\t)^{-1}-\v1\vpi^\t$
-%\citep{seneta1993sensitivity,meyer1975role}.
 }
@@
 complexity is $O(d^2)$.
-\todoc{Implement and test?}
 
-\begin{algorithm}[t]
+\begin{algorithm}
 \caption{Empirical confidence intervals}
@@
     \end{aligned}
-    %\label{eq:frequencies}
     \notag
@@
     :=
-    \sqrt{\frac{9\wh{P}_{i,j}(1-\wh{P}_{i,j})\log(2d/\delta)}{2N_i}}
-    + \frac{4\log(2d/\delta)+1.5\abs{(1/d)-\wh{P}_{i,j}}}{N_i}
+    \sqrt{\frac{9\wh{P}_{i,j}(1-\wh{P}_{i,j})\ln(2d/\delta)}{2N_i}}
+    + \frac{4\ln(2d/\delta)+1.5\abs{(1/d)-\wh{P}_{i,j}}}{N_i}
     .
-%    \label{eq:P-obs-bound}
     \notag
@@
   \label{step:P-bound}
-
-%$\qquad$ $\wh\Delta_{i,j}(\delta) \gets \sqrt{\frac{9\wh{P}_{i,j}(1-\wh{P}_{i,j})\log(2d/\delta)}{2N_i}}
-%  + \frac{4\log(2d/\delta)+1.5\abs{\alpha-d\alpha\wh{P}_{i,j}}}{N_i}$%,  $(i,j) \in [d]^2$.
-%\STATE   $\qquad$ $\wh\Delta_{i,j}(\delta) \gets \min( \wh\Delta_{i,j}(\delta) , \max(\wh{P}_{i,j}, 1-\wh{P}_{i,j}))$,   $(i,j) \in [d]^2$.
-
-%  \STATE Empirical bound for $\max_{i,j}|\wh{P}_{i,j}-P_{i,j}|$:
-%  \quad $B(\delta) := \max_{i,j} B_{i,j}(\delta)$.
-  %$\wh\Delta_{\vP}(\delta) \gets \sqrt{\sum_{i,j} \wh\Delta_{i,j}^2(\delta)}$,
-%  $\wh\Delta^*_{\vP}(\delta) \gets \max_{i,j} \wh\Delta_{i,j}(\delta)$.
 
@@
     .
-%    \label{eq:hatkappa}
     \notag
@@
     .
-%    \label{eq:pi-obs-bound}
     \notag
@@
     \Biggr)^{1/2} .
-%    \label{eq:gap-obs-bound}
     \notag
@@
 
-%  \STATE Empirical bounds for $\hat{\pi}_i$:
-%  \begin{align*}
-%    \rho_i(\delta)
-%    & := \frac{\wh\kappa B(\delta)}{\hat\pi_i}
-%    , \quad i \in [d] ;
-%    &
-%    \rho_i'(\delta)
-%    & := \frac{\wh\kappa B(\delta)}{[\hat{\pi}_i-\wh\kappa B(\delta)]_+}
-%    , \quad i \in [d] .
-%  \end{align*}
-
-%\STATE $\rho(\delta) \gets \frac12 \max_{i\in \iset{d}} \rho_i(\delta)$; $\rho'(\delta) \gets \frac12 \max_{i\in \iset{d}} \rho_i'(\delta)$.
-%\STATE $\hat{\rho}(\delta) \gets 2\max(\rho(\delta),\rho'(\delta)) + \max(\rho(\delta),\rho'(\delta))^2$.
-%%$\hat{\rho}(\delta) \gets \rho(\delta) + \rho'(\delta) + \rho'(\delta)\rho(\delta)$. % THIS WAS BUGGY
-%\STATE Half-width of interval: $w \gets 
-%\hat{\rho}(\delta)+ \parens{ 1+ \hat{\rho}(\delta) } \, \sqrt{ \sum_{i,j} \frac{\hat{\pi}_i}{\hat{\pi}_j} \wh{\Delta}^2_{i,j}(\delta)}
-%$.
-%\STATE Confidence interval for $\gap$: $U_n(\delta) \gets [\hatgap-w,\hatgap+w]\cap [0,1]$.
 \end{algorithmic}
@@
   \]
-%  \todoc{
-%    Precisely,
-%    what we mean by the $O(\cdot)$ notation is as follows:
-%  for a random sequence of nonnegative numbers $(X_n)_n$
-%  and another sequence of nonrandom positive numbers $(a_n^\eps)_n$ 
-%  indexed by $\eps$ we
-%  say that $X_n = O(a_n^\eps)$ holds almost surely as $n\to \infty$
-%  if for some universal constant $C>0$, 
-%  $\limsup_{n\to \infty} X_n/a_n^\eps \le C$ holds almost surely, regardless of
-%  the choice of $\eps$. I think this should be made clear.
-%  }
 \end{theorem}
@@
 \]
-%The widths of these intervals are somewhat larger than the corresponding
-%(non-empirical) intervals from \cref{thm:err}.
-%\todoc{This needs to be updated once we finalize the results. By how much are these larger?}
 
@@
 corresponding intervals from \cref{alg:empest}.
-% XXX now take the intersection
 This is summarized in the following \lcnamecref{thm:combined}, which
@@
 \end{theorem}
-
-%The rest of the section is devoted to the proof of \cref{thm:empirical}.
-
-%In what follows, we will (re-)define  $\wh{\Delta}_{i,j}(\delta)$ to be the RHS of \eqref{eq:P-obs-bound} when $N_i>0$,
-% \todoc{Abbrev RHS should be introduced.}
-%while we let $\wh{\Delta}_{i,j}(\delta)=\max(\wh{P}_{i,j},1-\wh{P}_{i,j})$ otherwise.
-%%Clearly, $\wt{\Delta}_{i,j}(\delta) \ge \wh{\Delta}_{i,j}(\delta)$.
-%%In what follows, we calculate with the tighter $\wh{\Delta}_{i,j}(\delta)$. 
-%%In fact, the algorithm was presented for $\wt{\Delta}_{i,j}(\delta)$ for easier comprehension only.
-%

### 2015-06-09 arxiv-v1/intro.tex
remove comments

--- 2015-06-03 intro-nips.tex
@@
-%!TEX root =  matrix-est.tex
-
-%Performance guarantees play a key role in the analysis and design of machine learning algorithms.
-
 This work
@@
 number of time steps required
-%time that it
-%takes
 for the chain to
 be within a fixed threshold of
-%(approximately) reach
 its stationary
 distribution:
-\todoa{slightly changed wording}
 \begin{align}
@@
     }
-%    \Norm{
-%      \vq \vP^t - \vpi
-%    }_{\tv} 
     \leq 1/4
@@
 mixing time $\tmix$.
-\todoa{changed wording}
 
@@
 the quantity of interest is
-%one is interested in estimating
 the
@@
 a Markov chain.
-\todoa{changed wording} \todoc{I am not sure that the new wording is better. It
-fails to express that there is an estimation problem. "Quantity of interest"
-is way to vague for me.} 
 This is the setting of the celebrated Markov Chain Monte Carlo (MCMC)
@@
   \label{eq:mixing-time-bound}
-  \Parens{\frac1{\gap} - 1} \ln2
+  \Parens{\mfrac1{\gap} - 1} \ln2
   \ \leq \ \tmix
-  \ \leq \ \frac1{\gap} \ln \frac4{\pimin}
+  \ \leq \ \mfrac1{\gap} \ln \mfrac4{\pimin}
   ,
 \end{equation}
-%$(\frac{1}\gap  - 1)\ln(2)  \le \tmix \le \frac{1}\gap \ln \frac4\pimin$,
 and hence it suffices to estimate $\gap$ and $\pimin$. 
-%these quantities $\gap$ and $\pimin$ are well-known to govern
-%mixing time.
 Our main results are summarized as follows.
 \begin{enumerate}
+  \item
+    In \cref{sec:rates-lower}, we show that in some problems 
+    $n = \Omega\parens{ (d\log d)/\gap + 1/\pimin}$ observations are necessary for any procedure
+    to guarantee constant multiplicative accuracy in estimating $\gap$
+    (\cref{thm:lb-pimin,thm:lb-gap}).
+    Essentially, in some problems \emph{every} state may need to be visited about
+    $\log(d)/\gap$ times, on average, before an accurate estimate of
+    the mixing time can be provided, regardless of the estimation
+    procedure used.
+
   \item
@@
   \item
-    In \cref{sec:rates-lower}, we show that $n = \Omega\parens{ (d\log
-    d)/\gap + 1/\pimin}$ observations are necessary for any procedure
-    to guarantee constant multiplicative accuracy in estimating $\gap$
-    (\cref{thm:lb-pimin,thm:lb-gap}).
-    Essentially, \emph{every} state may need to be visited about
-    $\log(d)/\gap$ times, on average, before an accurate estimate of
-    the mixing time can be provided, regardless of the estimation
-    procedure used.
-
-  \item
-    In \cref{sec:empirical}, we develop a procedure for computing
-    \emph{valid empirical confidence intervals} for $\pimin$ and
-    $\gap$ (\cref{alg:empest}).
-    The procedure uses just a single sample path and does not require
-    any prior knowledge of the chain's parameters, such as $\pimin$ or
-    $\gap$.
-    We quantify the rate at which the width of the confidence
-    intervals shrink as a function of the length of the sample path
+    In \cref{sec:empirical}, the construction of
+    valid \emph{fully empirical confidence intervals} for $\pimin$ and
+    $\gap$ are considered. First, the difficulty of the task is explained,
+    i.e., why the standard approach of turning the finite time confidence intervals of
+    \cref{thm:err} into a fully empirical one fails.
+	Combining several results from perturbation theory in a novel fashion
+	we propose a new procedure and prove that it avoids slow convergence
     (\cref{thm:empirical}).
@@
     those from either \cref{thm:err} and \cref{thm:empirical}.
-
-%    which is
-%    $     O\Parens{
-%      \frac1{\pimin}\sqrt{\frac{d\log(d/\delta)}{ n}}
-%      +
-%      \frac{d}{\pimin\gap} \sqrt{\frac{\log(d/\delta)}{\pimin n}}
-%    }.$
-
-%    \todoc{please add rate}
 
@@
 ,SwaJoa15:LoggedBandit%
-}. %BaHaVa10:activelearning,ShaVo08:conformalpred
+}.
 To address this need, numerous works have developed Chernoff-type
@@
 analysis~\cite{gillman1998chernoff,leon2004optimal,DBLP:conf/valuetools/KontoyiannisLM06,
-%meroi2012-jap,
 paulin15}.
@@
 It is possible to eliminate many of the difficulties presented above
-%with additional access to the Markov chain.
 when allowed more flexible access
@@
 barriers for many MCMC diagnostic problems~\citet{BhaBoMo11}.
-%%Performance guarantees play a key role in the analysis and design of machine learning algorithms.
-%
-%In this paper we study the construction of fully empirical bounds on the mixing time of Markov chains based on a single sample path.
-%Let $(X_t)_{t=1,2,\dotsc}$ be an irreducible, aperiodic time-homogeneous Markov chain on a finite state space $[d]$ of $d$ states numbered by positive integers with transition matrix $\vP$.
-Under this assumption, the chain converges to its unique stationary \todoc{Maybe steady state is better?}
-%distribution $\vpi = (\pi_i)_{i=1}^d$ regardless of the distribution $\vq$ of $X_1$: 
-%$\lim_{t\to\infty} \Pr_{\vq}\Parens{X_t = i} = \lim_{t\to\infty} (\vq \vP^t)_i = \pi_i$, $1\le i \le d$.
-%The mixing time $\tmix$ of the Markov chain is the time that it takes for the chain to reach its stationary distribution, say,
-%\[
-%  \tmix
-%  :=
-%  \min\Braces{
-%    t \in \bbN :
-%    \sup_{\vq}
-%    \max_{A\subset [d]} |\textstyle\Pr_{\vq}\Parens{ X_t \in A } - \vpi(A)|
-%%    \Norm{
-%%      \vq \vP^t - \vpi
-%%    }_{\tv} 
-%    \leq 1/4
-%  }\,.
-%\]
-%Here $\vpi(A) = \sum_{i\in A} \pi_i$ is the total probability assigned
-%to set $A$ by $\vpi$, and the supremum is over
-%all possible initial distributions $\vq$.
-%The problem studied in the paper is the construction of a confidence
-%set $C_n = C_n(X_1,X_2,\dotsc,X_n,\delta) \subset [0,\infty]$ based on
-%the knowledge of $X_1,X_2,\dotsc,X_n$ only that captures the mixing
-%time $\tmix$ with a pre-specified probability $1-\delta \in (0,1)$. 
-%In the remainder of the introduction, we describe our motivation of studying this problem, followed by a 
-%brief description of our main results.
-%\todod{``might be interesting'' sounds weak.} \todoc{better now?}
-%
-%In numerous scientific applications and some machine learning tasks
-%one is interested in estimating the mean $\vpi(f) = \sum_i \pi_i f(i)$ of some function $f$ of the states of a Markov chain. 
-%For example, this is the realms of the celebrated Markov Chain Monte Carlo (MCMC) paradigm, but 
-%the problem also arises in performance prediction when working with time correlated data, as is common in,
-%for example, in reinforcement learning.
-%%as 
-%% (cf. \cref{sec:mot}) 
-%Under our assumptions,
-%the law of large numbers guarantees that the sample mean $\vpi_n(f) \defeq \frac1n \sum_{t=1}^n f(X_t)$ 
-%converges almost surely to $\vpi(f)$, while the central limit theorem tells us that as $n\to \infty$, 
-%the distribution of the deviation $\sqrt{n}( \vpi_n(f)-\vpi(f))$ will be normal with zero mean and asymptotic
-%variance $\sigma^2_{\mathrm{as}}(f) \defeq \lim_{n\to\infty} n \Var\Parens{ \vpi_n(f) }$. \todoc{shall we add a citation here?}
-%While these results help us to understand how the sample mean over a Markov chain behaves in the limit,
-%they tell us little about the finite time, non-asymptotic behavior of the sample mean,
-%which are needed for prudent evaluation of one's method or even for algorithm design 
-%\cite{
-%MCMCDiscussion93%
-%,DBLP:conf/valuetools/KontoyiannisLM06%
-%,BBL06%
-%,MniSzeAu08%
-%,MauPo09%
-%,LiLiWaSt11:KWIK%
-%,flegal2011implementing%
-%,Gyori-paulin15%
-%,SwaJoa15:LoggedBandit%
-%}. %BaHaVa10:activelearning,ShaVo08:conformalpred
-%To address this need, numerous works have been devoted to deriving
-%Chernoff-type, exponential bounds for upper bounding the tail probabilities $p(r) = \Pr\Parens{ |\vpi_n(f)-\vpi(f)|> r }$,
-%also known as ``concentration bounds'', e.g., 
-%\cite{gillman1998chernoff,leon2004optimal,DBLP:conf/valuetools/KontoyiannisLM06,paulin15}.
-%Compared with the case when $(X_t)_t$ is an independent, identically distributed (iid) sequence of
-%random variables, these tail probabilities must be larger due to the temporal dependence of the sequence $(X_t)_t$.
-%Intuitively, for the Markov chain to forget the state where it is at time $t$, or in other words to get a 
-%``fresh'' sample $X_{t'}$ that is approximately ``independent'' from $X_t$, one must wait $\Theta(\tmix)$ time steps.
-%Thus, in the bounds on $p(r)$, one expects to see that the sample size $n$ 
-%gets appropriately reduced as $\tmix$ grows and this is indeed the case in bounds mentioned above.
-%is replaced by $n/\tmix$, which one
-%might think of as an approximate sample size. 
-%Indeed, all the above mentioned works get bounds on $p(r)$ that depend on various functions of $\vP$.
-%Namely, Kontoyiannis et al. obtain bounds that depend on certain information-theoretic quantities defined implicitly
-%as a function of $\vP$ and $f$, which are after simplification is bounded based on quantities related to 
-%a regeneration-type quantity derived from $\vP$
-%\cite{DBLP:conf/valuetools/KontoyiannisLM06}, while the other works mentioned above, in the case of reversible chains,
-%derive bounds that depend on the spectral gap of $\vP$.
-%Note that the spectral gap for such chains is essentially inversely related to the mixing time of the chain (see \cref{sec:setting}).
-%%\cite{gillman1998chernoff,leon2004optimal,,paulin15}.
-%Much research have been devoted to finding appropriate \emph{a priori} bounds on the various quantities that appear 
-%in these bounds, mostly in the theory community (e.g., \cite{saloff2004random}). 
-%To address the lack or looseness of such a priori bounds, a significant effort also went into developing methods
-%that estimate these quantities based on data, mostly in the MCMC community in an effort to develop
-%effective methods that are able to diagnose whether a chain has reached its stationary state 
-%(e.g., \cite{MCMCDiscussion93,flegal2011implementing,Gyori-paulin15}).
-%While the typical results state the asymptotic consistency of these estimators, or a central limit theorem (which 
-%are of little help to achieve a fully empirical, non-asymptotic bound on $p(r)$),
-%Gyori and Paulin give a finite time error bound for these estimates. 
-%Unsurprisingly, these error bounds themselves
-%depend on other unknown quantities of the Markov chain (which can be bounded in terms of the mixing time, again)
-%\cite{Gyori-paulin15}.
-%
-%This situation is similar to what appears when one attempts to use a Bernstein-type deviation bound to 
-%control the deviation between the sample mean of a sequence of iid random variables from their common mean
-%with a known range, but unknown variance: For a tight bound, one wishes to use the unknown variance, but in the deviation
-%bound for the error of a variance estimate, the variance appears again.
-%While in the case of Bernstein-bounds this circular reference leads to a nontrivial bound
-%\cite{audibert2009}, it appears that the situation is different in the Markov setting.
-%Focusing on the spectral gap, we will see that if one attempts to use a similar technique,
-%the set of feasible values based on the two error bounds always will never rule out that the spectral gap is zero,
-%or that the mixing time is infinite. Can one even hope to obtain 
-%after finitely many samples a nontrivial upper bound on the mixing time?
-%Obviously, for any finite sample size, if the mixing time of the chain is larger than the sample size than a sound method
-%must return infinity as the upper bound on the mixing time. Hence, the question is whether one can design a method
-%that never fails and if the sample size is appropriately large, returns a nontrivial bound on the mixing time, adapting the time it needs to trap the mixing time in a fixed size confidence interval to the unknown mixing time.
-%
-%\hide{
-%\subsection{Motivation}
-%\label{sec:mot}
-%One key competence in statistical learning is the ability to predict one's performance on future data.
-%To make performance prediction possible, one starts by posing some assumptions on how the data arises.
-%Simpler assumptions lead to simpler algorithms and analysis, while they may also limit applicability of the
-%obtained results and algorithms.
-%The most common assumption employed postulates that all data points, including past and future data points,
-%are sampled independently of each other from a common underlying distribution (e.g., \cite{SSBD14:UnderstandingML}).
-%This is helpful on two accounts: 
-%The lack of dependence between the data points means that sample averages of bounded functions of the data 
-%will converge at a uniform rate (irrespective of the distribution), while the assumption that all data points share the same
-%distribution means that one can learn about the distribution of future data points based on past data.
-%}
-%
-%\hide{
-%It is well-known that contrary to the idealized assumptions of the standard
-%learning model, real data is almost never iid.
-%Fortunately, many classical results of statistics and statistical
-%learning theory extend to the dependent data in a more or less
-%seamless fashion, at least when the data is sufficiently ``mixing''.
-%Such results include the central limit theorem, 
-%exponential tail bounds for empirical processes, 
-%performance guarantees for complexity regularization
-%using sieves or roughness penalties, as well as 
-%results based on stability arguments or Rademacher complexity 
-%\citep{Woo86,Dou94,Yu94,ZheChu96,dlPenaGine99,Mei00,
-%  MR1921877,vid2003,gamarnik03,LoKuScha05,MoRo08,MoRo09,
-%  DBLP:conf/nips/SteinwartC09,Steinwart2009175,
-%  FaSze11,
-%  london:nips12asalsn,london:icml13,
-%  DBLP:conf/nips/ShaliziK13}.
-%
-%Unfortunately, a distribution-free theory of non-iid learning is simply
-%impossible, when arbitrary dependence among the sample points is allowed.
-%Hence, one must quantify this dependence via the various mixing coefficients,
-%which appear in the non-iid generalization bounds mentioned above.
-%A frequently raised concern is that now the bounds depend on {\em
-%unknown} properties of the distribution, which precludes their
-%actual use in algorithmic applications (e.g., active
-%learning~\citep{BBL06,DHM07}).
-%%, and any algorithm aiming at a fully empirical error bound must
-%%also estimate the mixing coefficients.
-%In addition to learning-theoretic generalization bounds, the need for
-%fully empirical confidence bounds arises in Markov Chain Monte Carlo
-%(MCMC) simulations---so as to halt the run when sufficiently close to
-%the stationary distribution \citep{galin06}, and in Markov Reward
-%Processes \citep{sutton98}, as this will affect the learner's policy.
-%As \citet{Mei00} notes, in the absence of accurate estimates of the
-%mixing time, such fully empirical bounds remain ``a theoretician's
-%dream''.
-%
-%Our goal here is to develop an algorithm that, given a single sample
-%path of a Markov chain, computes non-trivial bounds on the mixing time
-%of the Markov chain.
-%We remark on the inherent chicken-and-egg nature of the challenge:
-%confidence intervals derived from a Markov chain---say, on the the
-%chain's mixing time---are typically given in terms of the chain's
-%mixing time.
-%Our way out of this conundrum is to give {\em empirical} (i.e.,
-%\emph{data-dependent}) as opposed to {\em a priori} guarantees.
-%To illustrate the intuition, consider a Markov chain on two states,
-%$\braces{0,1}$.
-%A learner who is stuck in state $0$ for a long time learns very little
-%about the chain and hence cannot be expected to give meaningful upper
-%bounds on its mixing time.
-%Consider, however, a learner who is lucky enough
-%to be in a chain that switches frequently between the two states.
-%Surely at some finite time, the learner will possess meaningful and nontrivial
-%information about the chain's mixing time --- crucially, without any
-%{\em prior} knowledge regarding this quantity. This simple example captures
-%the nature of our results: in some cases, we will not be able to say
-%anything meaningful (but the learner will recognize that this is the case),
-%while in ``luckier'' situations, we will be able to provide nontrivial
-%upper estimates on the mixing time.
-%}
-%
-%\paragraph{Main results.}
-%We propose what is apparently the first efficient and fully empirical
-%estimator of Markov mixing time.
-%\todod{``apparently'' sounds weak.}
-%Consider a reversible ergodic Markov chain on $d$ states with spectral
-%gap $\gap$ (which is well-known to govern mixing time)
-%and stationary distribution minorized by $\pimin$.
-%Our main results are summarized as follows.
-%\begin{enumerate}
-%  \item
-%    In \cref{thm:err}, we give an estimator for $\gap$ that
-%    achieves multiplicative accuracy from \emph{a single sample path}
-%    of length $\tilde{O}\parens{1/(\pimin\gap^3)}$.\footnote{The
-%    $\tilde{O}(\cdot)$ notation suppresses logarithmic factors.}
-%    We also provide an estimator for $\pimin$ that requires a sample
-%    path of length $\tilde{O}\parens{1/(\pimin\gap)}$.
-%    However, the valid confidence intervals suggested by
-%    \cref{thm:err} depend on the unknown quantities $\pimin$ and
-%    $\gap$.
-%
-%  \item
-%    In \cref{thm:lb-pimin,thm:lb-gap}, we show that $\Omega\parens{
-%    (d\log d)/\gap + 1/\pimin }$ observations are necessary for
-%    constant multiplicative accuracy in estimating $\gap$.
-%
-%  \item
-%    In \cref{thm:empirical}, we develop a procedure for computing
-%    \emph{valid empirical confidence intervals} for $\pimin$ and
-%    $\gap$.
-%    The procedure uses just a single sample path and does not require
-%    any prior knowledge about $\pimin$ or $\gap$.
-%    We also quantify the rate at which the width of the confidence
-%    intervals shrink as a function of the length of the sample path.
-%
-%    We show in \cref{thm:combined} how to combine these empirical
-%    confidence intervals with the non-empirical bounds from
-%    \cref{thm:err} to produce valid empirical confidence intervals;
-%    the width of these intervals converge to zero asymptotically as
-%    fast as those from \cref{thm:err}.
-%
-%%    which is
-%%    $     O\Parens{
-%%      \frac1{\pimin}\sqrt{\frac{d\log(d/\delta)}{ n}}
-%%      +
-%%      \frac{d}{\pimin\gap} \sqrt{\frac{\log(d/\delta)}{\pimin n}}
-%%    }.$
-%
-%
-%    
-%%    \todoc{please add rate}
-%
-%\end{enumerate}
-%
-%
-%
-%\paragraph{Related work.} \todoc{Move this later? It also needs to be updated, to 
-%adjust it to the current intro.}
-%Owing to the wide applicability of MCMC methods to various computational
-%problems, much effort has been devoted to developing effective convergence
-%diagnostic methods the aim of deriving stopping rules or stationarity
-%diagnoses. While a large number of heuristic techniques have been developed
-%\citep[e.g.,][]{MCMCDiscussion93,CoCa96}, it is widely recognized that all
-%these methods may fail to meet the guarantees that they are designed for.
-%Some recent advance in this area \citep{1209.0703,Gyori-paulin15} have
-%yielded practical algorithms for estimating the rates of convergence
-%of various MCMC algorithms, but they fall short of providing intervals
-%that are valid with any finite length sequence.
-%
-%In the theory community, \citet{BaFoRuSmiWhi00,BaFoRuSmiWhi13}
-%developed algorithms for testing whether a given Markov chain over a
-%finite state space of size $d$ is rapidly mixing.
-%In particular, assuming access to a sampling oracle that can generate
-%independent transitions from any given state, their algorithm can
-%distinguish between chains that in $t$ steps for a fixed $t$ reach a
-%distribution close to the stationary distribution using $\tilde{O}(t
-%d^{5/3})$ oracle calls.
-%It is remarkable that when $t$ is small enough, the procedure requires
-%sublinearly many samples in the total number of transitions in the
-%Markov chain.
-%For Markov chains with exponentially-large state spaces (with a
-%circuit-based description of the transition probability structure),
-%\citet{BhaBoMo11} establishes complexity-theoretic barriers for
-%similar MCMC diagnostic problems.
-%%A similar question was studied by \citet{BhaBoMo11} in the context of
-%%Markov chains with very large state spaces.
-%%They assume an circuit based description of the
-%%transition probability structure of the matrix and derive several hardness
-%%results for Markov chains whose state space consists of all binary strings of
-%%a given length. For example,
-%%they prove that given a Markov chain that mixes
-%%rapidly, it is $\mathsf{SZK}$-hard to distinguish whether the chain is close to
-%%stationarity by time $t$ or far from stationarity at time $ct$ for a
-%%constant $c$ when started from a given initial state ($\mathsf{SZK}
-%%\subset \mathsf{AM} \cap \mathsf{coAM}$, which is conjectured to
-%%extend beyond $\mathsf{BPP}$ \citep{GoVa11}).\footnote{%
-%%  $\mathsf{SZK}$ stands for statistical zero-knowledge, $\mathsf{BPP}$
-%%  stands for probabilistic polynomial-time, $\mathsf{AM}$ stands for
-%%  ``Arthur-Merlin protocol''.%
-%%}
-%
-%Other works include that of \citet{McDoShaSche11} who propose to estimate
-%the $\beta$-mixing coefficients of a stationary process possessing a density
-%based on histogram estimates of the underlying joint density. Their main
-%result is a consistency result for a fixed ``lag''. For fast mixing processes
-%(e.g., for $k$-step Markov processes) their result yields a convergence rate,
-%which unfortunately is given in terms of the unknown mixing coefficients.
-%%Catch-22 problem remains: The rate of convergence of their estimates depends on the unknown mixing coefficients
-%%(Theorem 2.4)
-%% $\exp(W(\log(n)))$ is the window length, where $W$ is Lambert's $W$-function (i.e., this is between $\log \log(n)$ and $\log n$).
-%The earlier study of \citet{Nobel06} considers the generalization of
-%existing discernibility results for iid processes to weakly uniformly
-%ergodic families, and shows how this can be used to study---and in
-%some cases estimate---polynomial mixing rates for dependent processes.
-%
-%

### 2015-06-09 arxiv-v1/lowerbound.tex
remove comments

\begin{proof}[Proof of \cref{thm:lb-pimin}]
  Fix $\bar\pi \in (0,1/3)$, and set $\veps := 2\bar\pi/(1-\bar\pi)$.
  Consider two Markov chains given by the following stochastic
  matrices:
  \[
    \vP^{(1)} :=
    \begin{bmatrix}
      1-\veps & \veps \\
      1-\veps & \veps
    \end{bmatrix}
    , \quad
    \vP^{(2)} :=
    \begin{bmatrix}
      1-\veps & \veps \\
      1/2 & 1/2
    \end{bmatrix}
    .
  \]
  Each Markov chain is ergodic and reversible; their stationary
  distributions are, respectively, $\vpi^{(1)} = (1-\veps,\veps)$
  and $\vpi^{(2)} = (\veps/(2+\veps),2/(2+\veps))$.
  We have $\pimin \geq \bar\pi$ in both cases.
  For the first Markov chain, $\slem = 0$, and hence the spectral gap
  is $1$; for the second Markov chain, $\slem = 1/2-\veps$, so the
  spectral gap is $1/2+\veps$.

  In order to guarantee $|\hatgap - \gap| < 1/6 < 1/2-\veps$, it must
  be possible to distinguish the two Markov chains.
  Assume that the initial state distribution has mass at least $1/2$
  on state $1$.
  (If this is not the case, we swap the roles of states $1$ and $2$ in
  the constructions above.)
  With probability at least half, the initial state is $1$; and both
  chains have the same transition probabilities from state $1$.
  The chains are indistinguishable less the sample path eventually
  reaches state $2$.
  But with probability at least $1/4$, a sample path of length $n <
  c/\bar\pi = c(2+\veps)/\veps$ starting from state $1$ always
  remains in the same state.
\end{proof}

\begin{proof}[Proof of \cref{thm:lb-gap}]
  We consider $d$-state Markov chains of the following form:
  \[
    P_{i,j} =
    \begin{cases}
      1-\veps_i & \text{if $i = j$} ; \\
      \displaystyle\frac{\veps_i}{d-1} & \text{if $i \neq j$}
    \end{cases}
  \]
  for some $\veps_1, \veps_2, \dotsc, \veps_d \in (0,1)$.
  Such a chain is ergodic and reversible, and its unique stationary
  distribution $\vpi$ satisfies
  \[
    \pi_i = \frac{1/\veps_i}{\sum_{j=1}^d 1/\veps_j}
    .
  \]
  We fix $\veps := \frac{d-1}{d/2}\bar\gamma$ and set $\veps' :=
  \frac{d/2-1}{d-1} \veps < \veps$.
  Consider the following $d+1$ different Markov chains of the type
  described above:
  \begin{itemize}
    \item
      $\vP^{(0)}$: $\veps_1 = \dotsb = \veps_d = \veps$.
      For this Markov chain, $\lambda_2 = \lambda_d = \slem =
      1-\frac{d}{d-1}\veps$.

    \item
      $\vP^{(i)}$ for $i \in [d]$: $\veps_j = \veps$ for $j \neq i$,
      and $\veps_i = \veps'$.
      For these Markov chains, $\lambda_2 = 1 - \veps' -
      \frac{1}{d-1}\veps = 1 - \frac{d/2}{d-1} \veps$,
      and $\lambda_d = 1 - \frac{d}{d-1} \veps$.
      So $\slem = 1-\frac{d/2}{d-1} \veps$.

  \end{itemize}
  The spectral gap in each chain satisfies $\gap \in
  [\bar\gamma,2\bar\gamma]$; in $\vP^{(i)}$ for $i \in [d]$, it is half
  of what it is in $\vP^{(0)}$.
  Also $\pi_i \geq 1/(2d)$ for each $i \in [d]$.

  In order to guarantee $|\hatgap - \gap| \leq \bar\gamma$, it must
  be possible to distinguish $\vP^{(0)}$ from each $\vP^{(i)}$,
  $i\in[d]$.
  But $\vP^{(0)}$ is identical to $\vP^{(i)}$ except for the transition
  probabilities from state $i$.
  Therefore, regardless of the initial state, the sample path must
  visit all states in order to distinguish $\vP^{(0)}$ from each
  $\vP^{(i)}$, $i \in [d]$.
  For any of the $d+1$ Markov chains above, the earliest time in which
  a sample path visits all $d$ states
  stochastically dominates a generalized coupon collection time $T = 1 +
  \sum_{i=1}^{d-1} T_i$, where $T_i$ is the number of steps required to
  see the $(i+1)$-th distinct state in the sequence beyond the first
  $i$.
  The random variables $T_1,T_2,\dotsc,T_{d-1}$ are independent, and are
  geometrically distributed, $T_i \sim \Geom(\veps -
  (i-1)\veps/(d-1))$.
  We have that
  \[
    \bbE[T_i] = \frac{d-1}{\veps(d-i)} , \quad
    \var(T_i) = \frac{1 - \veps \frac{d-i}{d-1}}{%
      \Parens{ \veps \frac{d-i}{d-1} }^2
    }
    .
  \]
  Therefore
  \[
    \bbE[T] = 1 + \frac{d-1}{\veps} H_{d-1} , \quad
    \var(T) \leq \Parens{ \frac{d-1}{\veps} }^2 \frac{\pi^2}{6} 
  \]
  where $H_{d-1} = 1 + 1/2 + 1/3 + \dotsb + 1/(d-1)$.
  Since $n < c d\log(d) / \bar\gamma$, with probability at least $1/4$,
  the sample path does not visit all $d$ states.
\end{proof}

### 2015-06-09 arxiv-v1/rates.tex
remove comments

--- 2015-06-04 rates.tex
@@
-%!TEX root =  matrix-est.tex
-
 In this section, we present lower and upper bounds on achievable rates
@@
 gives the same lower bound for estimating the mixing time.
-%It is important to emphasize that these are algorithm independent lower bounds, i.e.,
-%any algorithm is bound by them.
 Our first result holds even for two state Markov chains and shows that a sequence length of $\Omega(1/\pimin)$
 is necessary to achieve even a constant \emph{additive} accuracy in estimating $\gap$.
-%We begin with a result showing that any algorithm for estimating
-%$\gap$ (even for just constant additive accuracy) requires an
-%observation sequence of length $\Omega(1/\pimin)$
-%even when the number of states is only two.
-%%%
 \begin{theorem}
@@
 \end{theorem}
-%%%
 Next, considering $d$ state chains, we show that 
@@
 \cref{app:lower}.
-%In summary, the lower bounds show that no algorithm can hope to achieve a fixed multiplicative
-%accuracy before seeing $\Omega( d \log(d)/\gamma^* + 1/\pimin)$ states.
 
@@
       \sqrt{\frac{\log\frac{d}{\delta}\cdot\log\frac{n}{\pimin\delta}}{\pimin\gap n}}
-%      + \frac{\log\frac{d}{\delta}\cdot\log\frac{n}{\pimin\delta}}{\pimin\gap n}  
       + \frac{\log\frac{1}{\gap}}{\gap n}  
@@
 By \cref{eq:mixing-time-bound},
-the second of these is also a bound on the required sequence length to estimate $\tmix$. % the mixing time.
+the second of these is also a bound on the required sequence length to estimate $\tmix$.
 
@@
 $d$ larger.
-%\todoc{In terms of the dependence on the number of states, right? Can we be more specific? Factor of $d$ more, right?}
 Instead, we adapt probability tail bounds for sums of independent
@@
 matrix $\vA$ and its perturbation $\vA+\vE$ by $O(\norm{\vE}^{1/d})$.
-%\footnote{Actually, this result is not even readily applicable: 
-%the Haussdorff distance is actually an optimistic estimate of the deviation between
-%$\gap$ and its estimate.}
 Since $\norm{\wh\vP-\vP}$ is expected to be of size $O(n^{-1/2})$,
@@
 of perturbation results that exploit more structure.
-%(In \cref{sec:empirical}, we develop another estimator of
-%$\vL$---in order to exploit symmetry---that is based on estimating
-%$\vP$ for reasons discussed below.)
 

### 2015-06-09 arxiv-v1/setting.tex
remove comments

--- 2015-06-03 setting.tex
@@
-%!TEX root =  matrix-est.tex
-
 \subsection{Notations}
@@
 We use $\ln(\cdot)$ in expressions with exact constants, while we use $\log(\cdot)$ elsewhere.
-%In expression with exact numerical constants, we will use 
-%the natural logarithm function, $\ln(x)$, while elsewhere we use $\log(x)$.
 Boldface symbols are used for vectors and matrices (e.g., $\vv$,
@@
 $M_{i,j}$).
-%The $i$-th coordinate basis vector is denoted by $\ve_i$.
 For a vector $\vv$, $\norm{\vv}$ denotes its Euclidean norm; for a
@@
 $\Delta^{d-1}$ as row vectors.
-%Finally, the total variation between two probability distributions
-%$\vp, \vq \in \Delta^{d-1}$ is denoted by $\norm{\vp-\vq}_{\tv}$.
 
@@
 (We will strongly exploit the symmetry in our results.)
-%will be advantageous due to the existence of stronger perturbation results
-%for symmetric matrices.
 Further, $\vL = \Diag(\vpi)^{1/2} \vP \Diag(\vpi)^{-1/2}$, hence $\vL$
@@
 positive on account of ergodicity.
-\todoc{Do our results generalize to the spectral gap? No.
-Discuss this somewhere.}
-%Throughout the paper, both $\pimin$ and $\gap$ are assumed
-%to be positive.
 
@@
 $\gap$ (as we will show in the proof of our first result).
-%~\citep[e.g.,][]{MR1921877,MoRo08,MoRo09}. 
 Therefore, valid confidence intervals for $\pimin$ and $\gap$ can be
 used to make these rates fully observable.
-
-%Therefore, confidence intervals for $\pimin$ and $\gap$ can be used to
-%make these rates data-dependent and hence usable in procedures such
-%as Structural Risk Minimization.
-%\todoc{citation for SRM, and other applications?}
-%
-%Additionally, we aim to provide
-%asymptotic convergence rates
-%as well as data-dependent confidence intervals;
-%the latter should not depend on any unknown quantities
-%(such as $\pimin$ and $\gap$).
-%
-%\todoc[inline]{What is the data? What is the problem?}

### 2015-06-09 arxiv-v1/upperbound.tex
remove comments

In this section, we prove \cref{thm:err}.

\subsection{Accuracy of $\hatpimin$}

We start by proving the deviation bound on $\pimin-\hatpimin$, from
which we may easily deduce \cref{eq:piminbound} in \cref{thm:err}.
\begin{lemma}
  \label{lem:hatpimin}
  Pick any $\delta \in (0,1)$, and let
  \begin{equation}
    \veps_n :=
    \frac{ \ln\Parens{\frac{d}\delta\sqrt{\frac{2}{\pimin}}} }{\gap n}
    .
    \label{eq:singleton-veps}
  \end{equation}
  With probability at least $1-\delta$, the following inequalities
  hold simultaneously:
  \begin{align}
    \Abs{\hat{\pi}_i-\pi_i}
    & \le
    \sqrt{8\pi_i(1-\pi_i) \veps_n}
    + 20 \veps_n
    \quad \text{for all $i \in [d]$}
    ;
    \label{eq:singletonboundsimple}
    \\
    \Abs{\hatpimin - \pimin}
    & \leq
    4\sqrt{\pimin\veps_n}
    + 47\veps_n
    .
    \label{eq:hatpimin}
  \end{align}
\end{lemma}
\begin{proof}
  We use the following Bernstein-type inequality for Markov chains
  from \citet[][Theorem~3.8]{paulin15}: letting $\bbP^{\vpi}$ denote
  the probability with respect to the stationary chain (where the
  marginal distribution of each $X_t$ is $\vpi$), we have for every
  $\eps>0$,
  \[
    \bbP^{\vpi}
    \Parens{
      \abs{ \hat\pi_i-\pi_i} > \eps
    } \le
    2\exp\Parens{
      -\frac{
        n\gap\eps^2
      }{
        4\pi_i(1-\pi_i)+10\eps
      }
    },
    \qquad i\in[d]
    .
  \]
  To handle possibly non-stationary chains, as is our case, we combine
  the above inequality with \citet[][Proposition 3.14]{paulin15}, to
  obtain for any $\eps>0$,
  \begin{align*}
    \bbP
    \Parens{
      \abs{ \hat\pi_i-\pi_i} > \eps
    }
    &\le
    \sqrt{
      \frac1\pimin
      \bbP^{\vpi}
      \Parens{
        \abs{ \hat\pi_i-\pi_i} > \eps
      }
    } 
    \le
    \sqrt{\frac2\pimin}
    \exp\Parens{
      -\frac{
        n\gap\eps^2
      }{
        8\pi_i(1-\pi_i)+20\eps
      }
    }.
  \end{align*}
  Using this tail inequality with $\eps :=
  \sqrt{8\pi_i(1-\pi_i)\veps_n} + 20\veps_n$ and a union bound over
  all $i \in [d]$ implies that the inequalities
  in~\cref{eq:singletonboundsimple} hold with probability at least
  $1-\delta$.

  Now assume this $1-\delta$ probability event holds; it remains to
  prove that \cref{eq:hatpimin} also holds in this event.
  Without loss of generality, we assume that $\pimin = \pi_1 \leq
  \pi_2 \leq \dotsb \leq \pi_d$.
  Let $j \in [d]$ be such that $\hatpimin = \hat\pi_j$.
  By \cref{eq:singletonboundsimple}, we have $|\pi_i - \hat\pi_i| \leq
  \sqrt{8\pi_i\veps_n} + 20\veps_n$ for each $i \in \{1,j\}$.
  Since $\hatpimin \leq \hat\pi_1$,
  \[
    \hatpimin - \pimin
    \leq \hat\pi_1 - \pi_1
    \leq \sqrt{8\pimin\veps_n} + 20\veps_n
    \leq \pimin + 22\veps_n
  \]
  where the last inequality follows by the AM/GM inequality.
  Furthermore, using the fact that $a \leq b\sqrt{a} + c \Rightarrow a
  \leq b^2 + b\sqrt{c} + c$ for nonnegative numbers $a, b, c \geq
  0$~\citep[see, e.g.,][]{BBL04} with the inequality $\pi_j \leq
  \sqrt{8\veps_n} \sqrt{\pi_j} + \parens{\hat\pi_j + 20\veps_n}$ gives
  \[
    \pi_j
    \leq 
    \hat\pi_j
    + \sqrt{8(\hat\pi_j + 20\veps_n)\veps_n}
    + 28\veps_n
    .
  \]
  Therefore
  \[
    \pimin - \hatpimin
    \leq \pi_j - \hat\pi_j
    \leq \sqrt{8(\hatpimin + 20\veps_n)\veps_n} + 28\veps_n
    \leq \sqrt{8(2\pimin + 42\veps_n)\veps_n} + 28\veps_n
    \leq 4\sqrt{\pimin\veps_n} + 47\veps_n
  \]
  where the second-to-last inequality follows from the above bound on
  $\hatpimin - \pimin$, and the last inequality uses $\sqrt{a+b} \leq
  \sqrt{a} + \sqrt{b}$ for nonnegative $a,b \geq 0$.
\end{proof}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Accuracy of $\hatgap$}

Let us now turn to proving \cref{eq:gapbound}, i.e., 
the bound on the error of the spectral gap estimate $\hatgap$.
The accuracy of $\hatgap$ is based on the accuracy of $\Sym(\wh\vL)$
in approximating $\vL$ via Weyl's inequality:
\begin{align*}
  |\hat\lambda_i - \lambda_i|
  \leq \norm{\Sym(\wh\vL) - \vL}
  \quad \text{for all }i \in [d] .
\end{align*}
Moreover, the triangle inequality implies that symmetrizing $\wh\vL$
can only help:
\begin{align*}
  \norm{\Sym(\wh\vL) - \vL} \leq \norm{\wh\vL - \vL} .
\end{align*}
Therefore, we can deduce \cref{eq:gapbound} in \cref{thm:err} from the
following lemma.
\begin{lemma}
  \label{lem:gap}
  There exists an absolute constant $C>0$ such that the following
  holds.
  For any $\delta \in (0,1)$, with probability at least $1-\delta$,
  the bounds from \cref{lem:hatpimin} hold, and
  \begin{equation*}
    \norm{\wh\vL - \vL}
    \leq
    C\,
      \sqrt{
        \frac{
          \log\Parens{\frac{d}{\delta}}
          \log\Parens{\frac{n}{\pimin\delta}}
        }{\pimin\gap n}
      }
      +
      C\,
      \frac{\log\Parens{\frac{1}{\gap}}}{\gap n}
  \end{equation*}
\end{lemma}
The remainder of this section is devoted to proving this
\lcnamecref{lem:gap}.

The error $\wh\vL - \vL$ may be written as
\[
  \wh\vL - \vL
  = \errm + \errp \vL + \vL \errp + \errp \vL \errp
  + \errp \errm + \errm \errp + \errp \errm \errp\,,
\]
where
\begin{align*}
  \errp & := \Diag(\hat\vpi)^{-1/2} \Diag(\vpi)^{1/2} - \vI , \\
  \errm & := \Diag(\vpi)^{-1/2} \Parens{
    \wh\vM - \vM
  } \Diag(\vpi)^{-1/2} .
\end{align*}
Therefore
\begin{align*}
  \norm{\wh\vL - \vL}
  \leq \norm{\errm} +
  \Parens{
    \norm{\errm} + \norm{\vL}
  }
  \Parens{
    2\norm{\errp} + \norm{\errp}^2
  }
  .
\end{align*}
Since $\norm{\vL} \leq 1$ and $\norm{\wh\vL} \leq 1$~\citep[Lemma
12.1]{LePeWi08}, we also have $\norm{\wh\vL - \vL} \leq 2$.
Therefore,
\begin{align}
  \norm{\wh\vL - \vL}
  \leq
  \min\Braces{
    \norm{\errm} +
    \Parens{
      \norm{\errm} + \norm{\vL}
    }
    \Parens{
      2\norm{\errp} + \norm{\errp}^2
    }
    ,\,
    2
  }
  \leq 3(\norm{\errm}+\norm{\errp})
  .
  \label{eq:vlsimple}
\end{align}

\subsection{A bound on $\norm{\errp}$}

Since $\errp$ is diagonal,
\begin{align*}
  \norm{\errp}
  = \max_{i\in[d]}
  \Abs{
    \sqrt{\frac{\pi_i}{\hat\pi_i}} - 1
  }
  .
\end{align*}
Assume that
\begin{equation}
  n \geq \frac{108 \ln\Parens{\frac{d}\delta\sqrt{\frac{2}{\pimin}}}
  }{\pimin\gap}
  ,
  \label{eq:min-n1}
\end{equation}
in which case
\[
  \sqrt{8\pi_i(1-\pi_i) \veps_n} + 20\veps_n
  \leq \frac{\pi_i}{2}
\]
where $\veps_n$ is as defined in \cref{eq:singleton-veps}.
Therefore, in the $1-\delta$ probability event from
\cref{lem:hatpimin}, we have $|\pi_i - \hat\pi_i| \leq \pi_i/2$ for
each $i \in [d]$, and moreover, $2/3 \leq \pi_i/\hat\pi_i \leq 2$ for
each $i \in [d]$.
For this range of $\pi_i/\hat\pi_i$, we have
\[
  \Abs{ \sqrt{\frac{\pi_i}{\hat\pi_i}} - 1 }
  \leq 
  \Abs{ \frac{\hat\pi_i}{\pi_i} - 1 }
  .
\]
We conclude that if $n$ satisfies \cref{eq:min-n1}, then in this
$1-\delta$ probability event from \cref{lem:hatpimin},
\begin{align}
  \norm{\errp}
  & \leq
  \max_{i \in [d]}
  \Abs{
    \frac{\hat\pi_i}{\pi_i} - 1
  }
  \leq
  \max_{i \in [d]}
  \frac{
    \sqrt{8\pi_i(1-\pi_i) \veps_n} + 20 \veps_n
  }{\pi_i}
  \notag \\
  & \leq
  \sqrt{\frac{8\veps_n}{\pimin}} + \frac{20\veps_n}{\pimin}
  =
  \sqrt{
    \frac{
      8\ln\Parens{\frac{d}\delta\sqrt{\frac{2}{\pimin}}}
    }{
      \pimin\gap n
    }
  }
  +
  \frac{
    20\ln\Parens{\frac{d}\delta\sqrt{\frac{2}{\pimin}}}
  }{
    \pimin\gap n
  }
  .
  \label{eq:singletonbound}
\end{align}

\subsection{Accuracy of doublet frequency estimates (bounding $\norm{\errm}$)}
\label{sec:pairwise}
In this section we prove a bound on $\norm{\errm}$.
For this, we decompose
$\errm = \Diag(\vpi)^{-1/2}(\wh\vM-\vM)\Diag(\vpi)^{-1/2} $
into $\EE{\errm}$ and 
$\errm-\EE{\errm}$, the first measuring the effect of a non-stationary start of the chain,
while the second measuring the variation due to randomness.
\if0
into the sum of
\[
  \Diag(\vpi)^{-1/2}
  \left(\bbE( \wh\vM )- \vM\right)
  \Diag(\vpi)^{-1/2}
\]
and
\begin{multline*}
  \Diag(\vpi)^{-1/2} \left(\wh\vM- \bbE(\wh\vM)\right) \Diag(\vpi)^{-1/2}
  \\
  =
  \frac{n_H}{n-1}
  \Diag(\vpi)^{-1/2}\Parens{
    \wh{\vM}_H - \bbE(\wh{\vM}_H)
  }\Diag(\vpi)^{-1/2}
  \\
  + \frac{n_T}{n-1}
  \Diag(\vpi)^{-1/2}\Parens{
    \wh{\vM}_T - \bbE(\wh{\vM}_T)
  }\Diag(\vpi)^{-1/2}
  .
\end{multline*}
\fi

\subsubsection{Bounding $\norm{\EE{\errm}}$: The price of a non-stationary start.}
Let $\vpi^{(t)}$ be the distribution of states at time step $t$.
We will make use of the following proposition, which can be derived by
following \citet[Proposition 1.12]{MoTe06}:
\begin{proposition}
For $t\ge 1$, let $\vUpsilon^{(t)}$ be the vector with $\Upsilon^{(t)}_i= \frac{\pi_i^{(t)}}{\pi_i}$ and let
$\norm{\cdot}_{2,\vpi}$ denote the $\vpi$-weighted $2$-norm
\begin{align}
\label{eq:vv}
  \norm{\vv}_{2,\vpi} := \Parens{ \sum_{i=1}^d \pi_i v_i^2 }^{1/2}
  .
\end{align}
 Then,
\begin{equation}
  \label{eq:L2-contraction}
  \norm{ \vUpsilon^{(t)} - \v1 }_{2,\vpi} \leq
  \frac{(1-\gap)^{t-1}}{\sqrt{\pimin}} \,.
\end{equation}
\end{proposition}
An immediate corollary of this result is that 
\begin{align}
\label{eq:dvpi-ratio-bound}
\Norm{  \Dvpit \Dvpi^{-1} - \vI  } \le \frac{(1-\gap)^{t-1}}{\sqrt{\pimin}}\,.
\end{align}
Now note that
\begin{align*}
\bbE (\wh\vM) = \frac{1}{n-1} \sum_{t=1}^{n-1} \Dvpit \vP\,
\end{align*}
and thus
\begin{align*}
\EE{\errm} &=
  \Dvpi^{-1/2}
  \left(\bbE( \wh\vM )- \vM\right)
  \Dvpi^{-1/2}\\
& =  
  \frac{1}{n-1} \sum_{t=1}^{n-1}
  \Dvpi^{-1/2} (\Dvpit  - \Dvpi) \vP \Dvpi^{-1/2} \\
& =  
  \frac{1}{n-1} \sum_{t=1}^{n-1}
  \Dvpi^{-1/2} (\Dvpit \Dvpi^{-1} - \vI) \vM \Dvpi^{-1/2} \\
& =  
  \frac{1}{n-1} \sum_{t=1}^{n-1}
   (\Dvpit \Dvpi^{-1} - \vI) \vL \,.
\end{align*}
Combining this, $\norm{\vL}\le 1$ and~\cref{eq:dvpi-ratio-bound}, we get
\begin{align}
  \norm{\bbE(\errm)}
  \le \frac{1}{(n-1)\sqrt{\pimin}} \sum_{t=1}^{n-1} (1-\gap)^{t-1}
  \le \frac{1}{(n-1)\gap\sqrt{\pimin}}
  .
  \label{eq:biasbound}
\end{align}

\subsubsection{
Bounding $\norm{\errm-\EE{\errm}}$: 
Application of a matrix tail inequality}
\label{sec:pairwise-tailbound}

In this section we analyze the deviations of $\errm-\EE{\errm}$. By the definition of $\errm$,
\begin{align}
\label{eq:errmdev1}
\norm{\errm-\EE{\errm}}
 =  \norm{\Diag(\vpi)^{-1/2}\Parens{\wh\vM - \bbE \wh\vM } \Diag(\vpi)^{-1/2}}
 \,.
\end{align}
The matrix $\wh\vM - \EE{\wh\vM}$  is defined 
as a sum of dependent centered random matrices.
We will use the blocking technique of \citet{Bernstein27} 
to relate the likely deviations of this matrix to that of
a sum of independent centered random matrices.
The deviations of these will then bounded
with the help of a Bernstein-type matrix tail inequality due to \citet{tropp2015intro}.

We divide $[n-1]$ into contiguous blocks of time steps; each has size
$a \leq n/3$ except possibly the first block, which has size between
$a$ and $2a-1$.
Formally, let $a' := a + ((n-1) \bmod a) \leq 2a-1$, and define
\begin{align*}
  F & := [a'] , \\
  H_s & := \{ t \in [n-1] : a' + 2(s-1)a + 1 \leq t \leq a' + (2s-1)a \} , \\
  T_s & := \{ t \in [n-1] : a' + (2s-1)a + 1 \leq t \leq a' + 2sa \} ,
\end{align*}
for $s=1,2,\dotsc$.
Let $\mu_H$ (resp., $\mu_T$) be the number of non-empty $H_s$ (resp., $T_s$)
blocks.
Let $n_H := a\mu_H$ (resp., $n_T := a\mu_T$) be the number of time
steps in $\cup_s H_s$ (resp., $\cup_s T_s$).
We have
\begin{align}
  \wh\vM
  & = \frac1{n-1} \sum_{t=1}^{n-1} \ve_{X_t} \ve_{X_{t+1}}^\t
  \notag \\
  & = \frac{a'}{n-1} \cdot
    \underbrace{
      \frac{1}{a'} \sum_{t\in F} \ve_{X_t} \ve_{X_{t+1}}^\t
     }_{\wh\vM_F} +
  \frac{n_H}{n-1} \cdot
  \underbrace{
    \frac1{\mu_H}
    \sum_{s=1}^{\mu_H}
    \Parens{
      \frac1a \sum_{t \in H_s} \ve_{X_t} \ve_{X_{t+1}}^\t
    }
  }_{\wh{\vM}_H}
  \notag \\
  & \qquad
  + \frac{n_T}{n-1} \cdot
  \underbrace{
    \frac1{\mu_T}
    \sum_{s=1}^{\mu_T}
    \Parens{
      \frac1a \sum_{t \in T_s} \ve_{X_t} \ve_{X_{t+1}}^\t
    }
  }_{\wh{\vM}_T}
  .
  \label{eq:blocking}
\end{align}
Here, $\ve_i$ is the $i$-th coordinate basis vector, so $\ve_i
\ve_j^\t \in \{0,1\}^{d \times d}$ is a $d \times d$ matrix of all
zeros except for a $1$ in the $(i,j)$-th position.

The contribution of the first block is easily bounded using the
triangle inequality:
\begin{multline}
  \frac{a'}{n-1}
  \Norm{
    \Diag(\vpi)^{-1/2}
    \Parens{
      \wh\vM_F - \bbE(\wh\vM_F)
    }
    \Diag(\vpi)^{-1/2}
  }
  \\
  \leq
  \frac1{n-1} \sum_{t \in F}
  \Braces{
    \Norm{
      \frac{
        \ve_{X_t} \ve_{X_{t+1}}^\t
      }{\sqrt{\pi_{X_t}\pi_{X_{t+1}}}}
    }
    +
    \Norm{
      \bbE\Parens{
        \frac{
          \ve_{X_t} \ve_{X_{t+1}}^\t
        }{\sqrt{\pi_{X_t}\pi_{X_{t+1}}}}
      }
    }
  }
  \leq
  \frac{2a'}{\pimin(n-1)}
  .
  \label{eq:first-block}
\end{multline}

It remains to bound the contributions of the $H_s$ blocks and the
$T_s$ blocks.
We just focus on the the $H_s$ blocks, since the analysis is identical
for the $T_s$ blocks.

Let
\[
  \vY_s := \frac1a \sum_{t \in H_s} \ve_{X_t} \ve_{X_{t+1}}^\t ,
  \quad s \in [\mu_H] ,
\]
so
\[
  \wh{\vM}_H = \frac1{\mu_H} \sum_{s=1}^{\mu_H} \vY_s ,
\]
an average of the random matrices $\vY_s$.
For each $s \in [\mu_H]$, the random matrix $\vY_s$ is a function of
\[ (X_t : a' + 2(s-1)a + 1 \leq t \leq a' + (2s-1)a + 1) \]
(note the $+1$ in the upper limit of $t$),
so $\vY_{s+1}$ is $a$ time steps ahead of $\vY_s$.
When $a$ is sufficiently large, we will be able to effectively treat
the random matrices $\vY_s$ as if they were independent.
In the sequel, we shall always assume that the block length $a$
satisfies
\begin{equation}
  \label{eq:block-length}
  a \geq
  a_\delta
  :=
  \frac{1}{\gap} \ln \frac{2(n-2)}{\delta\pimin}
\end{equation}
for $\delta \in (0,1)$.

Define
\begin{align*}
  \vpi^{(H_s)}  := \frac1a \sum_{t \in H_s} \vpi^{(t)} , \qquad \qquad
  \vpi^{(H)}  := \frac1{\mu_H} \sum_{s=1}^{\mu_H} \vpi^{(H_s)} .
\end{align*}
Observe that
\[
  \bbE(\vY_s)
  = \Diag(\vpi^{(H_s)}) \vP
\]
so
\[
  \bbE\Parens{
    \frac1{\mu_H} \sum_{s=1}^{\mu_H} \vY_s
  } = \Diag(\vpi^{(H)}) \vP
  .
\]

Define
\[
  \vZ_s
  := \Diag(\vpi)^{-1/2} \left( \vY_s 
  - \bbE(\vY_s) \right)\Diag(\vpi)^{-1/2}
  .
\]
We apply a matrix tail inequality to the average of \emph{independent}
copies of the $\vZ_s$'s.
More precisely, we will apply the tail inequality to independent
copies $\wt{\vZ}_s$, $s\in \iset{\mu_H}$ of the random variables
$\vZ_s$ and then relate the average of $\wt{\vZ}_s$ to that of
$\vZ_s$.
The following probability inequality is from \citet[Theorem
6.1.1.]{tropp2015intro}.
\begin{theorem}[Matrix Bernstein inequality]
\label{thm:mxbernstein}
Let $\vQ_1,\vQ_2,\dotsc,\vQ_m$ be a sequence of independent, random
$d_1 \times d_2$ matrices.
Assume that $\EE{\vQ_i} = \v0$ and $\Norm{\vQ_i}\le R$ for each $1\le
i \le m$.
Let $\vS = \sum_{i=1}^m \vQ_i$ and let 
\begin{align*}
v = \max\left\{ \norm{\bbE \textstyle\sum_i \vQ_i \vQ_i^\top  }, 
						      \norm{\bbE \textstyle\sum_i \vQ_i^\top \vQ_i  }
			\right\}\,.
\end{align*}
Then, for all $t\ge 0$, 
\[
\bbP\left( \Norm{\vS} \ge t \right) \le 2(d_1 + d_2) \exp\left( -\frac{t^2/2}{v+R t/3}\right)\,.
\]
In other words, for any $\delta \in (0,1)$,
\[
  \bbP\Parens{
    \norm{\vS} > \sqrt{2 v \ln \frac{2(d_1+d_2)}{\delta}} + \frac{2R}{3}
    \ln\frac{2(d_1+d_2)}{\delta}
  } \leq \delta
  \,.
\]
\end{theorem}

To apply \cref{thm:mxbernstein}, it suffices to bound the spectral
norms of $\vZ_s$ (almost surely), $\bbE(\vZ_s \vZ_s^\t)$, and
$\bbE(\vZ_s^\t \vZ_s)$.

\paragraph{Range bound.}
By the triangle inequality,
\[
  \norm{\vZ_s}
  \leq \norm{\Diag(\vpi)^{-1/2} \vY_s \Diag(\vpi)^{-1/2}}
  + \norm{\Diag(\vpi)^{-1/2} \bbE(\vY_s) \Diag(\vpi)^{-1/2}}
  \,.
\]
For the first term, we have
\begin{align}
\label{eq:vysbound}
  \norm{\Diag(\vpi)^{-1/2} \vY_s \Diag(\vpi)^{-1/2}} \leq \frac1{\pimin} .
\end{align}
For the second term, we use the fact $\norm{\vL}\le 1$ to bound
\begin{align*}
  \norm{\Diag(\vpi)^{-1/2} (\bbE(\vY_s) -\vM)\Diag(\vpi)^{-1/2}}
  & = \norm{ \bigl(\Diag(\vpi^{(H_s)}) \Diag(\vpi)^{-1} - \vI  \bigr) \vL }
  \\
  & \le \norm{ \Diag(\vpi^{(H_s)}) \Diag(\vpi)^{-1} - \vI }\,.
\end{align*}
Then, using~\cref{eq:dvpi-ratio-bound},
\begin{equation}
  \norm{ \Diag(\vpi^{(H_s)}) \Diag(\vpi)^{-1} - \vI }
  \le \frac{(1-\gap)^{a'+2(s-1)a}}{\sqrt{\pimin}}
  \le \frac{(1-\gap)^{a}}{\sqrt{\pimin}} \le 1\,,
  \label{eq:ysbias}
\end{equation}
where the last inequality follows from the assumption that the
block length $a$ satisfies~\cref{eq:block-length}.
Combining this with
$\norm{\Diag(\vpi)^{-1/2} \vM\Diag(\vpi)^{-1/2}} = \norm{\vL}\le 1$,
it follows that
\begin{align}
\label{eq:evysbound}
\norm{\Diag(\vpi)^{-1/2} \bbE(\vY_s) \Diag(\vpi)^{-1/2}} \le 2
\end{align}
by the triangle inequality.
Therefore, together with~\cref{eq:vysbound}, we obtain the range bound
\[
  \norm{\vZ_s} \leq \frac1{\pimin} + 2
  .
\]

\paragraph{Variance bound.}
We now determine bounds on the spectral norms of $\bbE(\vZ_s
\vZ_s^\t)$ and $\bbE(\vZ_s^\t \vZ_s)$.
Observe that
\begin{align}
  \lefteqn{
    \bbE(\vZ_s\vZ_s^\t)
  } \notag \\
  & =
  \frac1{a^2} \sum_{t \in H_s}
  \bbE\Parens{
    \Diag(\vpi)^{-1/2}
    \ve_{X_t} \ve_{X_{t+1}}^\t 
    \Diag(\vpi)^{-1}
    \ve_{X_{t+1}} \ve_{X_t}^\t
    \Diag(\vpi)^{-1/2}
  }
  \label{eq:var1}
  \\
  & \quad
  + \frac1{a^2} \sum_{ \substack{ t \neq t' \\ t,t'\in H_s} } 
  \bbE\Parens{
    \Diag(\vpi)^{-1/2}
    \ve_{X_t} \ve_{X_{t+1}}^\t 
    \Diag(\vpi)^{-1}
    \ve_{X_{t'+1}} \ve_{X_{t'}}^\t
    \Diag(\vpi)^{-1/2}
  }
  \label{eq:var2}
  \\
  & \quad
  - \Diag(\vpi)^{-1/2}
  \bbE(\vY_s)
  \Diag(\vpi)^{-1} 
  \bbE(\vY_s^\t)
  \Diag(\vpi)^{-1/2}
  .
  \label{eq:var3}
\end{align}
The first sum, \cref{eq:var1}, easily simplifies to the diagonal matrix
\begin{align*}
  \lefteqn{
    \frac1{a^2}
    \sum_{t \in H_s}
    \sum_{i=1}^d \sum_{j=1}^d \Pr(X_t = i, X_{t+1} = j) \cdot
    \frac{1}{\pi_i\pi_j} \ve_i\ve_j^\t\ve_j\ve_i^\t
  } \\
  & =
  \frac1{a^2}
  \sum_{t \in H_s}
  \sum_{i=1}^d \sum_{j=1}^d \pi_i^{(t)} P_{i,j} \cdot
  \frac{1}{\pi_i\pi_j} \ve_i\ve_i^\t
  =
  \frac1{a}
  \sum_{i=1}^d \frac{\pi_i^{(H_s)}}{\pi_i}
  \Parens{ \sum_{j=1}^d \frac{P_{i,j}}{\pi_j} }
   \ve_i \ve_i^\t
  .
\end{align*}
For the second sum, \cref{eq:var2},  a symmetric matrix, consider
\[
  \vu^\t
  \Parens{
    \frac1{a^2} \sum_{ \substack{ t \neq t' \\ t,t'\in H_s} } 
    \bbE\Parens{
      \Diag(\vpi)^{-1/2}
      \ve_{X_t} \ve_{X_{t+1}}^\t 
      \Diag(\vpi)^{-1}
      \ve_{X_{t'+1}} \ve_{X_{t'}}^\t
      \Diag(\vpi)^{-1/2}
    }
  } \vu
\]
for an arbitrary unit vector $\vu$.
By Cauchy-Schwarz and AM/GM, this is bounded from above by
\begin{multline*}
  \frac1{2a^2}
  \sum_{ \substack{ t \neq t' \\ t,t'\in H_s} }
  \biggl[
  \bbE\Parens{
    \vu^\t
    \Diag(\vpi)^{-1/2} \ve_{X_t} \ve_{X_{t+1}}^\t \Diag(\vpi)^{-1}
    \ve_{X_{t+1}} \ve_{X_t}^\t \Diag(\vpi)^{-1/2}
    \vu
  }
  \\
  +
  \bbE\Parens{
    \vu^\t
    \Diag(\vpi)^{-1/2} \ve_{X_{t'}} \ve_{X_{t'+1}}^\t \Diag(\vpi)^{-1}
    \ve_{X_{t'+1}} \ve_{X_{t'}}^\t \Diag(\vpi)^{-1/2}
    \vu
  }
  \biggr]
  ,
\end{multline*}
which simplifies to
\[
  \frac{a-1}{a^2}
  \vu^\t \bbE\Parens{
    \sum_{t \in H_s}
    \Diag(\vpi)^{-1/2} \ve_{X_t} \ve_{X_{t+1}}^\t \Diag(\vpi)^{-1}
    \ve_{X_{t+1}} \ve_{X_t}^\t \Diag(\vpi)^{-1/2}
  } \vu\,.
\]
The expectation is the same as that for the first term,
\cref{eq:var1}.

Finally, the spectral norm of the third term, \cref{eq:var3}, is
bounded using \cref{eq:evysbound}:
\[
  \norm{\Diag(\vpi)^{-1/2} \bbE(\vY_s) \Diag(\vpi)^{-1/2}}^2
  \leq 4
  .
\]

Therefore, by the triangle inequality, the bound $\pi_i^{(H)}/\pi_i
\leq 2$ from \cref{eq:ysbias}, and simplifications, 
\begin{align*}
  \Norm{
    \bbE(\vZ_s \vZ_s^\t)
  }
  &\leq
  \max_{i\in[d]}
  \Parens{ \sum_{j=1}^d \frac{P_{i,j}}{\pi_j} }
  \frac{\pi_i^{(H)}}{\pi_i}
  + 4
  \leq
  2\max_{i\in[d]}
  \Parens{ \sum_{j=1}^d \frac{P_{i,j}}{\pi_j} }
  + 4
  .
\end{align*}

We can bound $\bbE(\vZ_s^\t \vZ_s)$
in a similar way; the only difference is that the reversibility needs
to be used at one place to simplify an expectation:
\begin{align*}
\MoveEqLeft
    \frac1{a^2} \sum_{t \in H_s}
    \bbE\Parens{
      \Diag(\vpi)^{-1/2}
      \ve_{X_{t+1}} \ve_{X_t}^\t
      \Diag(\vpi)^{-1}
      \ve_{X_t}\ve_{X_{t+1}}^\t
      \Diag(\vpi)^{-1/2}
    }
	\\
  & =
  \frac1{a^2} \sum_{t \in H_s}
  \sum_{i=1}^d \sum_{j=1}^d \Pr(X_t = i, X_{t+1} = j) \cdot
  \frac{1}{\pi_i\pi_j} \ve_j\ve_j^\t
  \\
  & =
  \frac1{a^2} \sum_{t \in H_s}
  \sum_{i=1}^d \sum_{j=1}^d \pi_i^{(t)} P_{i,j} \cdot
  \frac{1}{\pi_i\pi_j} \ve_j\ve_j^\t
  \\
  & =
  \frac1{a^2} \sum_{t \in H_s}
  \sum_{j=1}^d \Parens{
    \sum_{i=1}^d \frac{\pi_i^{(t)}}{\pi_i} \cdot \frac{P_{j,i}}{\pi_i}
  } \ve_j\ve_j^\t
\end{align*}
where the last step uses \cref{eq:reversibility}.
As before, we get
\begin{align*}
  \Norm{
    \bbE(\vZ_s^\t\vZ_s)
  }
  &\leq
  \max_{i\in[d]}
  \Parens{
    \sum_{j=1}^d \frac{P_{i,j}}{\pi_j}
    \cdot \frac{\pi_j^{(H)}}{\pi_j}
  }
  + 4
  \leq
  2 \max_{i\in[d]}
  \Parens{
    \sum_{j=1}^d \frac{P_{i,j}}{\pi_j}
  }
  + 4
\end{align*}
again using the bound $\pi_i^{(H)}/\pi_i \leq 2$ from
\cref{eq:ysbias}.

\paragraph{Independent copies bound.}
Let $\wt\vZ_s$ for $s \in [\mu_H]$ be independent copies of
$\vZ_s$ for $s \in [\mu_H]$.
Applying \cref{thm:mxbernstein} to the average of these random
matrices, we have
\begin{equation}
  \bbP\Parens{
    \Norm{ \frac1{\mu_H} \sum_{s=1}^{\mu_H} \wt\vZ_s}
    >
    \sqrt{
      \frac{4\Parens{ d_{\vP} + 2 } \ln\frac{4d}{\delta}}
      {\mu_H}
    }
    + \frac{2\Parens{ \frac1{\pimin} + 2 } \ln\frac{4d}{\delta}}
    {3\mu_H}
  } \leq \delta
  \label{eq:indpt-bound}
\end{equation}
where
\begin{align*}
  d_{\vP}
  & := \max_{i \in [d]} \sum_{j=1}^d \frac{P_{i,j}}{\pi_j}
  \leq
  \frac{1}{\pimin}
  \,.
\end{align*}

\paragraph{The actual bound.}
To bound the probability that $\norm{\sum_{s=1}^{\mu_H} \vZ_s/\mu_H}$
is large, we appeal to the following result, which is a consequence of
\citep[Corollary 2.7]{Yu94}.
For each $s \in [\mu_H]$, let $X^{(H_s)} := (X_t : a' + 2(s-1)a + 1
\leq t \leq a' + (2s-1)a + 1)$, which are the random variables
determining $\vZ_s$.
Let $\bbP$ denote the joint distribution of $(X^{(H_s)} : s \in
[\mu_H])$; let $\bbP_s$ be its marginal over $X^{(H_s)}$, and let
$\bbP_{1:s+1}$ be its marginal over
$(X^{(H_1)},X^{(H_2)},\dotsc,X^{(H_{s+1})})$.
Let $\wt\bbP$ be the product distribution formed from the marginals
$\bbP_1,\bbP_2,\dotsc,\bbP_{\mu_H}$, so $\wt\bbP$ governs the joint
distribution of $(\wt\vZ_s : s \in [\mu_H])$.
The result from \citep[Corollary 2.7]{Yu94} implies for any event $E$,
\[
  |\bbP(E) - \wt\bbP(E)| \leq (\mu_H-1) \beta(\bbP)
\]
where
\[
  \beta(\bbP)
  :=
  \max_{1 \leq s \leq \mu_H-1}
  \bbE\Parens{
    \Norm{
      \bbP_{1:s+1}(\cdot\,|X^{(H_1)},X^{(H_2)},\dotsc,X^{(H_s)}) - \bbP_{s+1}
    }_{\tv}
  }
  \,.
\]
Here, $\norm{\cdot}_{\tv}$ denotes the total variation norm.
The number $\beta(\bbP)$ can be recognized to be the
\emph{$\beta$-mixing coefficient} of the stochastic process
$\{X^{(H_s)}\}_{s \in [\mu_H]}$.
This result implies that the bound from \cref{eq:indpt-bound} for
$\norm{\sum_{s=1}^{\mu_H} \wt\vZ_s/\mu_H}$ also holds for
$\norm{\sum_{s=1}^{\mu_H} \vZ_s/\mu_H}$, except the probability
bound increases from $\delta$ to $\delta + (\mu_H-1)\beta(\bbP)$:
\begin{equation}
  \bbP\Parens{
    \Norm{ \frac1{\mu_H} \sum_{s=1}^{\mu_H} \vZ_s}
    >
    \sqrt{
      \frac{4\Parens{ d_{\vP} + 2 } \ln\frac{4d}{\delta}}
      {\mu_H}
    }
    + \frac{2\Parens{ \frac1{\pimin} + 2 } \ln\frac{4d}{\delta}}
    {3\mu_H}
  } \leq \delta + (\mu_H-1)\beta(\bbP)
  .
  \label{eq:dpt-bound}
\end{equation}
By the triangle inequality,
\[
  \beta(\bbP)
  \leq
  \max_{1 \leq s \leq \mu_H-1}
  \bbE\Parens{
    \vphantom{\bigg\vert}
    \Norm{
      \bbP_{1:s+1}(\cdot\,|X^{(H_1)},X^{(H_2)},\dotsc,X^{(H_s)})
      - \bbP^{\vpi}
    }_{\tv}
    +
    \Norm{
      \bbP_{s+1}
      - \bbP^{\vpi}
    }_{\tv}
  }
\]
where $\bbP^{\vpi}$ is the marginal distribution of $X^{(H_1)}$ under
the stationary chain.
Using the Markov property and integrating out $X_t$ for $t >
\min H_{s+1} = a'+2sa+1$,
\[
  \Norm{
    \bbP_{1:s+1}(\cdot\,|X^{(H_1)},X^{(H_2)},\dotsc,X^{(H_s)})
    - \bbP^{\vpi}
  }_{\tv}
  =
  \Norm{
    \cL(X_{a'+2sa+1}\,|X_{a'+(2s-1)a+1})
    - \vpi
  }_{\tv}
\]
where $\cL(Y|Z)$ denotes the conditional distribution of $Y$ given
$Z$.
We bound this distance using standard arguments for bounding the
mixing time in terms of the \emph{relaxation time}
$1/\gap$~\citep[see, e.g., the proof of Theorem 12.3 of][]{LePeWi08}:
for any $i \in [d]$,
\[
  \Norm{
    \cL(X_{a'+2sa+1}\,|X_{a'+(2s-1)a+1}=i)
    - \vpi
  }_{\tv}
  =
  \Norm{
    \cL(X_{a+1}\,|X_1=i)
    - \vpi
  }_{\tv}
  \leq
  \frac{
    \exp\Parens{ -a\gap }
  }{\pimin}
  .
\]
The distance $\norm{ \bbP_{s+1} - \bbP^{\vpi} }_{\tv}$ can be bounded
similarly:
\begin{align*}
  \Norm{
    \bbP_{s+1}
    - \bbP^{\vpi}
  }_{\tv}
  & =
  \Norm{
    \cL(X_{a'+2sa+1})
    - \vpi
  }_{\tv}
  \\
  & =
  \Norm{
    \sum_{i=1}^d \bbP(X_1 = i)
    \cL(X_{a'+2sa+1}\,|X_1 = i)
    - \vpi
  }_{\tv}
  \\
  & \leq
  \sum_{i=1}^d \bbP(X_1 = i)
  \Norm{
    \cL(X_{a'+2sa+1}\,|X_1 = i)
    - \vpi
  }_{\tv}
  \\
  & \leq
  \frac{
    \exp\Parens{-(a'+2sa)\gap}
  }{\pimin}
  \leq
  \frac{
    \exp\Parens{-a\gap}
  }{\pimin}
  .
\end{align*}
We conclude
\[
  (\mu_H-1)\beta(\bbP)
  \leq (\mu_H-1)\frac{2\exp(-a\gap)}{\pimin}
  \leq \frac{2(n-2)\exp(-a\gap)}{\pimin}
  \leq \delta
\]
where the last step follows from the block length assumption
\cref{eq:block-length}.

We return to the decomposition from \cref{eq:blocking}.
We apply \cref{eq:dpt-bound} to both the $H_s$ blocks and the $T_s$
blocks, and combine with \cref{eq:first-block} to obtain the following
probabilistic bound.
Pick any $\delta \in (0,1)$, let the block length be
\[
  a
  := \ceil{ a_\delta }
  =
  \Ceil{
    \frac{1}{\gap}\ln\frac{2(n-2)}{\pimin\delta}
  }
  ,
\]
so
\[
  \min\{\mu_H,\mu_T\}
  =
  \Floor{
    \frac{n-1-a'}{2a}
  }
  \geq
  \frac{n-1}
  {
    2\Parens{
      1 + \frac1{\gap}\ln\frac{2(n-2)}{\pimin\delta}
    }
  }
  -2
  =: \mu
  .
\]
If
\begin{equation}
  n \geq 7 + \frac6{\gap} \ln \frac{2(n-2)}{\pimin\delta}
  \geq 3a
  ,
  \label{eq:min-n2}
\end{equation}
then with probability at least $1-4\delta$,
\begin{multline*}
  \Norm{\Diag(\vpi)^{-1/2}\Parens{\wh\vM - \bbE[\wh\vM]} \Diag(\vpi)^{-1/2} }
  \\
  \leq
  \frac{4
    \Ceil{
      \frac{1}{\gap}\ln\frac{2(n-2)}{\pimin\delta}
    }
  }{\pimin(n-1)}
  +
  \sqrt{
    \frac{4\Parens{ d_{\vP} + 2 } \ln\frac{4d}{\delta}}
    {\mu}
  }
  + \frac{2\Parens{ \frac1{\pimin} + 2 } \ln\frac{4d}{\delta}}
  {3\mu}
  .
\end{multline*}

\subsubsection{The bound on $\norm{\errm}$}
Combining the probabilistic bound from above with the bound on the
bias from \cref{eq:biasbound}, we obtain the following.
Assuming the condition on $n$ from \cref{eq:min-n2}, with probability
at least $1-4\delta$,
\begin{equation}
  \norm{\errm}
  \leq
  \frac{1}{(n-1)\gap\sqrt{\pimin}}
  +
  \frac{4
    \Ceil{
      \frac{1}{\gap}\ln\frac{2(n-2)}{\pimin\delta}
    }
  }{\pimin(n-1)}
  +
  \sqrt{
    \frac{4\Parens{ d_{\vP} + 2 } \ln\frac{4d}{\delta}}
    {\mu}
  }
  + \frac{2\Parens{ \frac1{\pimin} + 2 } \ln\frac{4d}{\delta}}
  {3\mu}
  .
  \label{eq:doubletbound}
\end{equation}

\subsection{Overall error bound}
Assume the sequence length $n$ satisfies \cref{eq:min-n1} and
\cref{eq:min-n2}.
Consider the $1-5\delta$ probability event in which
\cref{eq:singletonboundsimple,eq:hatpimin,eq:doubletbound} hold.
In this event, \cref{eq:singletonbound} also holds, and hence by
\cref{eq:vlsimple},
\begin{multline*}
  \norm{\wh\vL - \vL}
  \leq
  3
  \Parens{
    \sqrt{
      \frac{
        8\ln\Parens{\frac{d}\delta\sqrt{\frac{2}{\pimin}}}
      }{
        \pimin\gap n
      }
    }
    +
    \frac{
      20\ln\Parens{\frac{d}\delta\sqrt{\frac{2}{\pimin}}}
    }{
      \pimin\gap n
    }
  }
  \\
  +
  3\Parens{
    \frac{1}{(n-1)\gap\sqrt{\pimin}}
    +
    \frac{4
      \Ceil{
        \frac{1}{\gap}\ln\frac{2(n-2)}{\pimin\delta}
      }
    }{\pimin(n-1)}
    +
    \sqrt{
      \frac{4\Parens{ d_{\vP} + 2 } \ln\frac{4d}{\delta}}
      {\mu}
    }
    + \frac{2\Parens{ \frac1{\pimin} + 2 } \ln\frac{4d}{\delta}}
    {3\mu}
  }
  \\
  =
  O
  \Parens{
    \sqrt{
      \frac{
        \log\Parens{\frac{d}{\delta}}
        \log\Parens{\frac{n}{\pimin\delta}}
      }{\pimin\gap n}
    }
  }
  .
\end{multline*}

To finish the proof of \cref{lem:gap}, we replace $\delta$ with
$\delta/5$, and now observe that the bound on $\norm{\wh\vL-\vL}$ is
trivial if \cref{eq:min-n1} is violated (recalling that
$\norm{\wh\vL-\vL} \leq 2$ always holds).
We tack on an additional term $\log(1/(\pimin\gap\delta)) / (\gap n)$
to also ensure a trivial bound if \cref{eq:min-n2} is violated.
\hfill $\qed$

### 2015-06-09 arxiv-v1/intro.tex
orphans

@@
     $\log(d)/\gap$ times, on average, before an accurate estimate of
-    the mixing time can be provided, regardless of the estimation
+    the mixing time can be provided, regardless of the actual estimation
     procedure used.

### 2015-08-03 nips2015/rebuttal.txt
prelim rebuttal

We thank the referees for the helpful comments. A detailed response follows.

Tightness of bounds: We don't claim tightness of the bounds -- they
are indeed rather loose -- but our results give the first non-trivial
(upper and lower) bounds in this challenging setting. Our work are
motivated by practical applications where fully data-dependent
intervals were sought but not actually achieved. Our results are a
first step towards a practical solution: we demonstrate feasibility
and also prove limitations on what is possible.

Relaxation vs mixing time: In the revision, we will be more explicit
about estimating the relaxation time rather than mixing time proper
(throughout the abstract and paper). We agree that this is an
important distinction that should be emphasized.

Reviewer_1

3. Qualitatively, Theorems 1 and 2 show that if n is less than some
function of (d,pi,gamma), then for any estimator based on a sample
path of length n, there is some Markov chain such that the error is
large.

4. If n > max{16C^2,4C}/(pi_* gamma_*), then the RHS in (4) (without
the logarithmic factor) is less than pi_*/2.

If n > max{16C^2,4C}/(pi_* gamma_*^3), then the RHS in (5) (without
the logarithmic factor) is less than gamma_*/2.

5. Combining Theorems 1 and 2 with a standard argument yields the
lower bound in the introduction. We will include the details in the
revision.

6.

7. To conclude Theorem 2, it follows from Paley-Zygmund and the moment
bounds that, with probability > 1/4, the earliest time in which a
sample path visits all d states is at least c d log(d) / \bar\gamma.
This implies that an estimator using a shorter sample path cannot
distinguish P^0 from P^i for some i (with probability at least 1/4),
and hence must have large estimation error.

Reviewer_2

\tilde{O}: We'll use this more standard use in the revision.

Theorem 2: Indeed, including an extra constant factor in the claim (or
minding inequality strictness) resolves this issue.

Lemma 2: We'll include more details in the revision.

Theorem 2 (line 958): The estimator cannot rule out P^i without having
vistited state i. We'll clarify this point in the revision.

Reviewer_7

Perfect sampling: Indeed, this line of work is related, although these
methods generally do not apply in our setting of estimation from a
single sample path. We'll discuss this in the revision.

### 2015-08-03 nips2015/rebuttal.txt
rebuttal

@@
-We thank the referees for the helpful comments. A detailed response follows.
+We thank the referees for the helpful comments. We'll incorporate the
+suggestions in a revision of the paper, especially where more detailed
+explanations are needed.
+
+The following two concerns were common to many of the reviews.
 
@@
 important distinction that should be emphasized.
+
+Additional responses to individual reviews follow.
 
@@
 
-6.
+6. ???
 
@@
 This implies that an estimator using a shorter sample path cannot
-distinguish P^0 from P^i for some i (with probability at least 1/4),
-and hence must have large estimation error.
+distinguish P^0 from P^i for some i (with probability > 1/4), and
+hence must have large estimation error. We'll include these details in
+the revision.
 
@@
 minding inequality strictness) resolves this issue.
-
-Lemma 2: We'll include more details in the revision.
 

### 2015-08-05 nips2015/rebuttal.txt
rebuttal

@@
 about estimating the relaxation time rather than mixing time proper
-(throughout the abstract and paper). We agree that this is an
-important distinction that should be emphasized.
+(throughout the abstract and paper). We fully agree that this is a
+very important distinction that should be emphasized, and we'll amend
+the overstated claims throughout.
 
@@
 
+We will add these details in the revision.
+
 5. Combining Theorems 1 and 2 with a standard argument yields the
@@
 
-7. To conclude Theorem 2, it follows from Paley-Zygmund and the moment
-bounds that, with probability > 1/4, the earliest time in which a
-sample path visits all d states is at least c d log(d) / \bar\gamma.
-This implies that an estimator using a shorter sample path cannot
-distinguish P^0 from P^i for some i (with probability > 1/4), and
-hence must have large estimation error. We'll include these details in
-the revision.
+7. To conclude Theorem 2, it follows the second-moment method (e.g.,
+Paley-Zygmund) and the moment bounds that, with probability > 1/4, the
+earliest time in which a sample path visits all d states is at least c
+d log(d) / \bar\gamma.  This implies that an estimator using a shorter
+sample path cannot distinguish P^0 from P^i for some i (with
+probability > 1/4), and hence must have large estimation error. We'll
+include these details in the revision.
 
@@
 
-\tilde{O}: We'll use this more standard use in the revision.
+\tilde{O}: We'll adopt this more standard use in the revision.
 

### 2015-08-07 nips2015/rebuttal.txt
typos
.

@@
 
-The following two concerns were common to many of the reviews.
+The following two concerns were common to multiple reviews.
 
@@
 are indeed rather loose -- but our results give the first non-trivial
-(upper and lower) bounds in this challenging setting. Our work are
+(upper and lower) bounds in this challenging setting. Our work is
 motivated by practical applications where fully data-dependent
@@
 
+2. We'll make this edit in the revision.
+
 3. Qualitatively, Theorems 1 and 2 show that if n is less than some
@@
 path of length n, there is some Markov chain such that the error is
-large.
+large. We'll add additional clarifying interpretations in the
+revision.
 
@@
 
-6. ???
+6. Unfortunately, we were not able to understand the counterexample,
+likely due to typos in the review. We hope to resolve this via the
+program chairs/area chair.
 
@@
 
+Theorem 5: We get an improvement when the min in the bound for
+|\hat{V}| is achieved by the first term. A sheepish caveat: this is
+not always the case for all sufficiently large n, due to the
+sqrt{log(n)} term in the numerator, but will be true for a very large
+range of n. There is more discussion in Section 5.
+
 \tilde{O}: We'll adopt this more standard use in the revision.
+
+kappa: It is defined just above Theorem 4.
 
@@
 
-Theorem 2 (line 958): The estimator cannot rule out P^i without having
+Theorem 2 (line 958): An estimator cannot rule out P^i without having
 vistited state i. We'll clarify this point in the revision.

### 2015-08-07 nips2015/rebuttal.txt
.

@@
-We thank the referees for the helpful comments. We'll incorporate the
-suggestions in a revision of the paper, especially where more detailed
-explanations are needed.
+We thank the referees for the helpful comments. We'll incorporate the suggestions in a revision of the paper, especially where more detailed explanations are needed.
 
@@
 
-Tightness of bounds: We don't claim tightness of the bounds -- they
-are indeed rather loose -- but our results give the first non-trivial
-(upper and lower) bounds in this challenging setting. Our work is
-motivated by practical applications where fully data-dependent
-intervals were sought but not actually achieved. Our results are a
-first step towards a practical solution: we demonstrate feasibility
-and also prove limitations on what is possible.
+Tightness of bounds: We don't claim tightness of the bounds -- they are indeed rather loose -- but our results give the first non-trivial (upper and lower) bounds in this challenging setting.
+Our work is motivated by practical applications where fully data-dependent intervals were sought but not actually achieved.
+Our results are a first step towards a practical solution: we demonstrate feasibility and also prove limitations on what is possible.
 
-Relaxation vs mixing time: In the revision, we will be more explicit
-about estimating the relaxation time rather than mixing time proper
-(throughout the abstract and paper). We fully agree that this is a
-very important distinction that should be emphasized, and we'll amend
-the overstated claims throughout.
+Relaxation vs mixing time: In the revision, we will be more explicit about estimating the relaxation time rather than mixing time proper (throughout the abstract and paper).
+We fully agree that this is a very important distinction that should be emphasized, and we'll amend the overstated claims throughout.
 
@@
 
-3. Qualitatively, Theorems 1 and 2 show that if n is less than some
-function of (d,pi,gamma), then for any estimator based on a sample
-path of length n, there is some Markov chain such that the error is
-large. We'll add additional clarifying interpretations in the
-revision.
+3.
+Qualitatively, Theorems 1 and 2 show that if n is less than some function of (d,pi,gamma), then for any estimator based on a sample path of length n, there is some Markov chain such that the error is large.
+We'll add additional clarifying interpretations in the revision.
 
-4. If n > max{16C^2,4C}/(pi_* gamma_*), then the RHS in (4) (without
-the logarithmic factor) is less than pi_*/2.
+4. If n > max{16C^2,4C}/(pi_* gamma_*), then the RHS in (4) (without the logarithmic factor) is less than pi_*/2.
 
-If n > max{16C^2,4C}/(pi_* gamma_*^3), then the RHS in (5) (without
-the logarithmic factor) is less than gamma_*/2.
+If n > max{16C^2,4C}/(pi_* gamma_*^3), then the RHS in (5) (without the logarithmic factor) is less than gamma_*/2.
 
@@
 
-5. Combining Theorems 1 and 2 with a standard argument yields the
-lower bound in the introduction. We will include the details in the
-revision.
+5. Combining Theorems 1 and 2 with a standard argument yields the lower bound in the introduction. We will include the details in the revision.
 
-6. Unfortunately, we were not able to understand the counterexample,
-likely due to typos in the review. We hope to resolve this via the
-program chairs/area chair.
+6. Regarding Theorem 1: Unfortunately, we were not able to understand the counterexample, partly due to typos in the review. We hope to resolve this via the program chairs/area chair.
 
-7. To conclude Theorem 2, it follows the second-moment method (e.g.,
-Paley-Zygmund) and the moment bounds that, with probability > 1/4, the
-earliest time in which a sample path visits all d states is at least c
-d log(d) / \bar\gamma.  This implies that an estimator using a shorter
-sample path cannot distinguish P^0 from P^i for some i (with
-probability > 1/4), and hence must have large estimation error. We'll
-include these details in the revision.
+However, there were some minor numerical calculation errors in the proof of Theorem 1, which we can clarify here.
+We set \varepsilon := \bar\pi, and observe that \pi^{(2)} = (1/(1+2\varepsilon),2\varepsilon/(1+2\varepsilon)).
+The rest of the proof is correct as is.
+
+7.
+To conclude Theorem 2, it follows the second-moment method (e.g., Paley-Zygmund) and the moment bounds that, with probability > 1/4, the earliest time in which a sample path visits all d states is at least c d log(d) / \bar\gamma.
+This implies that an estimator using a shorter sample path cannot distinguish P^0 from P^i for some i (with probability > 1/4), and hence must have large estimation error.
+We'll include these details in the revision.
 
@@
 
-Theorem 5: We get an improvement when the min in the bound for
-|\hat{V}| is achieved by the first term. A sheepish caveat: this is
-not always the case for all sufficiently large n, due to the
-sqrt{log(n)} term in the numerator, but will be true for a very large
-range of n. There is more discussion in Section 5.
+Theorem 5: We get an improvement when the min in the bound for |\hat{V}| is achieved by the first term.
+A sheepish caveat: this is not always the case for all sufficiently large n, due to the sqrt{log(n)} term in the numerator, but will be true for a very large range of n.
+There is more discussion in Section 5.
 
@@
 
-Theorem 2: Indeed, including an extra constant factor in the claim (or
-minding inequality strictness) resolves this issue.
+Theorem 2 (line 188): Indeed, including an extra constant factor in the claim (or minding inequality strictness) resolves this issue.
 
-Theorem 2 (line 958): An estimator cannot rule out P^i without having
-vistited state i. We'll clarify this point in the revision.
+Theorem 2 (line 958): An estimator cannot rule out P^i without having visited state i. We'll clarify this point in the revision.
 
@@
 
-Perfect sampling: Indeed, this line of work is related, although these
-methods generally do not apply in our setting of estimation from a
-single sample path. We'll discuss this in the revision.
+Perfect sampling: Indeed, this line of work is related, although these methods generally do not apply in our setting of estimation from a single sample path. We'll discuss this in the revision.

### 2015-08-07 nips2015/rebuttal.txt
.

@@
 However, there were some minor numerical calculation errors in the proof of Theorem 1, which we can clarify here.
-We set \varepsilon := \bar\pi, and observe that \pi^{(2)} = (1/(1+2\varepsilon),2\varepsilon/(1+2\varepsilon)).
+We need to change the "1/6" in the theorem statement and proof to "1/8".
+Furthermore, in the proof, \bar\pi should be taken in (0,1/4), and we should set \varepsilon := \bar\pi.
+Observe that \pi^{(2)} = (1/(1+2\varepsilon),2\varepsilon/(1+2\varepsilon)).
+Thus, \pi_\star \geq \bar\pi in both P^{(1)} and P^{(2)}.
+To guarantee |\hat\gamma_\star - \gamma_\star| < 1/8, it must be possible to distinguish the two Markov chains.
 The rest of the proof is correct as is.
@@
 
-Theorem 2 (line 188): Indeed, including an extra constant factor in the claim (or minding inequality strictness) resolves this issue.
+Theorem 2 (line 188): Indeed, including an extra constant factor in the claim resolves this issue.
 

### 2015-08-07 arxiv-v2/abstract.tex
arxiv-v2 (not yet submitted)

--- 2015-06-03 abstract.tex
@@
-%!TEX root =  matrix-est.tex
-This article provides the first procedure for computing a fully
-data-dependent interval that traps the mixing time $\tmix$ of a finite
-reversible ergodic Markov chain at a prescribed confidence level.
-The interval is computed from a single finite-length sample
-path from the Markov chain, and does not require the 
-knowledge of any parameters of the chain.
-%bounds on the mixing time itself.
-This stands in contrast to previous approaches, which either 
-only provide point estimates, or require a
-reset mechanism, or additional prior knowledge.
-%Most notably, 
-%the procedure produces a fully empirical confidence
-%interval (i.e., it depends only on the sample path and a given
-%confidence parameter $\delta$) that is always valid with probability
-%$1-\delta$; and 
-The width of the interval converges to zero at a
-$\sqrt{n}$ rate, where $n$ is the length of the sample path. \todoc{Aryeh prefers: .. at a rate of $\sqrt{n}$. }
-Upper and lower bounds are given on the number of samples required to
-achieve constant-factor multiplicative accuracy.
-\todoc{We may need to add some explanation to connect $\gap$ estimates
-to $\tmix$ estimates.}
-The lower bounds indicate that, unless further restrictions are placed
-on the chain, no procedure can achieve this accuracy level before
-seeing each state at least $\Omega(\tmix)$ times on the average.
+This article provides the first procedure for computing a fully data-dependent interval that traps the mixing time $\tmix$ of a finite reversible ergodic Markov chain at a prescribed confidence level.
+The interval is computed from a single finite-length sample path from the Markov chain, and does not require the knowledge of any parameters of the chain.
+This stands in contrast to previous approaches, which either only provide point estimates, or require a reset mechanism, or additional prior knowledge.
+The width of the interval converges to zero at a $\sqrt{n}$ rate, where $n$ is the length of the sample path.
+Upper and lower bounds are given on the number of samples required to achieve constant-factor multiplicative accuracy.
+The lower bounds indicate that, unless further restrictions are placed on the chain, no procedure can achieve this accuracy level before seeing each state at least $\Omega(\tmix)$ times on the average.
 Finally, future directions of research are identified.
-%If $\gap$ denotes the absolute spectral gap of the unknown Markov chain
-%and the chain's stationary distribution is minorized by $\pimin$,
-%our procedure requires $\tilde{O}( \frac{1}{\pimin\gap^3})$ observations
-% to estimate the mixing time up to a constant multiplicative accuracy, 
-%while we also establish that any algorithm to achieve the same
-%goal would need $\Omega( d\log(d)/\gap + 1/\pimin)$ samples.
-
-%and the dependence on the main  chain's
-
-%We provide confidence guarantees that are fully data-dependent in the sense that they yield non-trivial
-%finite-sample deviation bounds without any knowledge of the mixing coefficients.
-%
-%
-%The novelty of our approach lies in several key aspects.
-%Crucially,
-%we do not assume any a priori bounds
-%on the mixing time,
-%which
-%sets our approach apart from
-%most others that do
-%require some advance knowledge of this quantity.
-%Another
-%advantage of our result over
-%common practice is the sufficiency of a single long sampling sequence,
-%without access to a restart mechanism.
-%Perhaps most importantly, we overcome an obstacle inherent in the chicken-and-egg nature of
-%estimating mixing coefficients: to give confidence bounds one typically needs to know the mixing time.
-%We provide confidence guarantees that are fully data-dependent in the sense that they yield non-trivial
-%finite-sample deviation bounds without any knowledge of the mixing coefficients.
-%%Upper and lower sample complexity bounds are provided and challenging open problems are posed.
-%%\todod{I would remove the last sentence, at least the part about
-%%``challenging'' open problems.}
-%%%test

### 2015-08-07 arxiv-v2/conclusions.tex
arxiv-v2 (not yet submitted)

--- 2015-06-09 arxiv-v1/conclusions.tex
@@
 reduce the computation cost: The current cubic cost in the number of states
-can be too high even when the number of states is moderate.
+can be too high even when the number of states is only moderately
+large.
 

### 2015-08-07 arxiv-v2/lowerbound.tex
arxiv-v2 (not yet submitted)

--- 2015-06-09 arxiv-v1/lowerbound.tex
@@
 \begin{proof}[Proof of \cref{thm:lb-pimin}]
-  Fix $\bar\pi \in (0,1/3)$, and set $\veps := 2\bar\pi/(1-\bar\pi)$.
+  Fix $\bar\pi \in (0,1/4)$.
   Consider two Markov chains given by the following stochastic
@@
     \begin{bmatrix}
-      1-\veps & \veps \\
-      1-\veps & \veps
+      1-\bar\pi & \bar\pi \\
+      1-\bar\pi & \bar\pi
     \end{bmatrix}
@@
     \begin{bmatrix}
-      1-\veps & \veps \\
+      1-\bar\pi & \bar\pi \\
       1/2 & 1/2
@@
   Each Markov chain is ergodic and reversible; their stationary
-  distributions are, respectively, $\vpi^{(1)} = (1-\veps,\veps)$
-  and $\vpi^{(2)} = (\veps/(2+\veps),2/(2+\veps))$.
+  distributions are, respectively, $\vpi^{(1)} = (1-\bar\pi,\bar\pi)$
+  and $\vpi^{(2)} = (1/(1+2\bar\pi),2\bar\pi/(1+2\bar\pi))$.
   We have $\pimin \geq \bar\pi$ in both cases.
   For the first Markov chain, $\slem = 0$, and hence the spectral gap
-  is $1$; for the second Markov chain, $\slem = 1/2-\veps$, so the
-  spectral gap is $1/2+\veps$.
+  is $1$; for the second Markov chain, $\slem = 1/2-\bar\pi$, so the
+  spectral gap is $1/2+\bar\pi$.
 
-  In order to guarantee $|\hatgap - \gap| < 1/6 < 1/2-\veps$, it must
-  be possible to distinguish the two Markov chains.
+  In order to guarantee $|\hatgap - \gap| < 1/8 < |1 -
+  (1/2+\bar\pi)|/2$, it must be possible to distinguish the two Markov
+  chains.
   Assume that the initial state distribution has mass at least $1/2$
@@
   But with probability at least $1/4$, a sample path of length $n <
-  c/\bar\pi = c(2+\veps)/\veps$ starting from state $1$ always
-  remains in the same state.
+  c/\bar\pi$ starting from state $1$ always remains in the same state.
 \end{proof}

### 2015-08-07 arxiv-v2/lowerbound.tex
.

@@
   reaches state $2$.
-  But with probability at least $1/4$, a sample path of length $n <
-  c/\bar\pi$ starting from state $1$ always remains in the same state.
+  But with probability at least $1/3$, a sample path of length $n <
+  1/(4\bar\pi)$ starting from state $1$ always remains in the same
+  state (this follows from properties of the geometric distribution).
 \end{proof}
@@
   where $H_{d-1} = 1 + 1/2 + 1/3 + \dotsb + 1/(d-1)$.
-  Since $n < c d\log(d) / \bar\gamma$, with probability at least $1/4$,
-  the sample path does not visit all $d$ states.
+  By the Paley-Zygmund inequality,
+  \[
+    \Pr\Parens{ T > \frac13\bbE[T] }
+    \geq \frac{1}{1 + \frac{\var(T)}{(4/9)\bbE[T]^2}}
+    \geq \frac{1}{1 + \frac{\Parens{ \frac{d-1}{\veps} }^2
+    \frac{\pi^2}{6}}{(4/9)\Parens{ \frac{d-1}{\veps} H_2}^2}}
+    \geq \frac14
+    .
+  \]
+  Since $n < c d\log(d) / \bar\gamma \leq (1/3) (1 + d H_{d-1} /
+  (2\bar\gamma)) = \bbE[T]/3$ (for an appropriate absolute constant
+  $c$), with probability at least $1/4$, the sample path does not
+  visit all $d$ states.
 \end{proof}

### 2015-08-07 arxiv-v2/rates.tex
.

--- 2015-06-09 arxiv-v1/rates.tex
@@
   \label{thm:lb-pimin}
-  There is an absolute constant $c>0$ such that the following holds.
   Pick any $\bar\pi \in (0,1/4)$.
   Consider any estimator $\hatgap$ that takes as input a random sample
-  path of length $n < c/\bar\pi$ from a Markov chain starting from any
-  desired initial state distribution.
+  path of length $n \leq 1/(4\bar\pi)$ from a Markov chain starting
+  from any desired initial state distribution.
   There exists a two-state ergodic and reversible Markov chain
@@
   \[
-    \Pr\Brackets{ |\hatgap - \gap| \geq 1/6 } \geq 1/8 .
+    \Pr\Brackets{ |\hatgap - \gap| \geq 1/8 } \geq 1/6 .
   \]

### 2015-08-07 nips2015/rebuttal.txt
.

@@
 
-6. Regarding Theorem 1: Unfortunately, we were not able to understand the counterexample, partly due to typos in the review. We hope to resolve this via the program chairs/area chair.
+6. Regarding Theorem 1: Note that "c" is an absolute constant in the theorem, not something that can be chosen arbitrarily. In particular, one value that works in the theorem is c=1/4.
 
-However, there were some minor numerical calculation errors in the proof of Theorem 1, which we can clarify here.
+There were some minor numerical calculation errors in the proof of Theorem 1, which we can clarify here.
 We need to change the "1/6" in the theorem statement and proof to "1/8".

### 2015-08-08 arxiv-v2/lowerbound.tex
.

@@
 
-  In order to guarantee $|\hatgap - \gap| \leq \bar\gamma$, it must
+  In order to guarantee $|\hatgap - \gap| < \bar\gamma/2$, it must
   be possible to distinguish $\vP^{(0)}$ from each $\vP^{(i)}$,
@@
     \Pr\Parens{ T > \frac13\bbE[T] }
-    \geq \frac{1}{1 + \frac{\var(T)}{(4/9)\bbE[T]^2}}
+    \geq \frac{1}{1 + \frac{\var(T)}{(1-1/3)^2\bbE[T]^2}}
     \geq \frac{1}{1 + \frac{\Parens{ \frac{d-1}{\veps} }^2
@@
   \]
-  Since $n < c d\log(d) / \bar\gamma \leq (1/3) (1 + d H_{d-1} /
+  Since $n < c d\log(d) / \bar\gamma \leq (1/3) (1 + (d-1) H_{d-1} /
   (2\bar\gamma)) = \bbE[T]/3$ (for an appropriate absolute constant

### 2015-08-08 arxiv-v2/lowerbound.tex
.

@@
   reaches state $2$.
-  But with probability at least $1/3$, a sample path of length $n <
+  But with probability at least $3/4$, a sample path of length $n <
   1/(4\bar\pi)$ starting from state $1$ always remains in the same
-  state (this follows from properties of the geometric distribution).
+  state (this follows from properties of the geometric distribution
+  and the assumption $\bar\pi < 1/4$).
 \end{proof}
@@
   \sum_{i=1}^{d-1} T_i$, where $T_i$ is the number of steps required to
-  see the $(i+1)$-th distinct state in the sequence beyond the first
+  see the $(i+1)$-th distinct state in the sample path beyond the first
   $i$.

### 2015-08-09 nips2015/rebuttal.txt
Minor changes to rebuttal

@@
 
-Tightness of bounds: We don't claim tightness of the bounds -- they are indeed rather loose -- but our results give the first non-trivial (upper and lower) bounds in this challenging setting.
+Tightness of bounds: We don't claim tightness of the bounds (we did not attempt to optimize the constants) but our results give the first non-trivial (upper and lower) bounds in this challenging problem.
+In particular, the estimator is based on a less than straighforward, indirect approach while avoiding slow rates (as explained at the beginning of Section 4) that builds on advanced perturbation results from the theory of stochastic matrices.
 Our work is motivated by practical applications where fully data-dependent intervals were sought but not actually achieved.
-Our results are a first step towards a practical solution: we demonstrate feasibility and also prove limitations on what is possible.
+Our results are a first step towards a practical solution: we demonstrate feasibility and also prove limitations on what is possible, both add significantly to our knowledge.
 
@@
 7.
-To conclude Theorem 2, it follows the second-moment method (e.g., Paley-Zygmund) and the moment bounds that, with probability > 1/4, the earliest time in which a sample path visits all d states is at least c d log(d) / \bar\gamma.
+The conclusion of Theorem 2 follows using the second-moment method (e.g., Paley-Zygmund) and the moment bounds that, with probability > 1/4, the earliest time in which a sample path visits all d states is at least c d log(d) / \bar\gamma.
 This implies that an estimator using a shorter sample path cannot distinguish P^0 from P^i for some i (with probability > 1/4), and hence must have large estimation error.
 We'll include these details in the revision.
+
+Minor comment 3: We respectfully disagree.
+The definition of a confidence interval does not distinguish intervals based on what is known, which itself may depend on the application (in specific applications additional knowledge if often available).
 
@@
 
+Theorem 2 (line 188): Indeed, including an extra constant factor in the claim resolves this issue.
+
 \tilde{O}: We'll adopt this more standard use in the revision.
@@
 
-Theorem 2 (line 188): Indeed, including an extra constant factor in the claim resolves this issue.
+line 926: Thanks. See response to Issue 6 by Reviewer 1.
 

### 2015-09-12 arxiv-v2/abstract.tex
issues about relaxation time

@@
-This article provides the first procedure for computing a fully data-dependent interval that traps the mixing time $\tmix$ of a finite reversible ergodic Markov chain at a prescribed confidence level.
-The interval is computed from a single finite-length sample path from the Markov chain, and does not require the knowledge of any parameters of the chain.
-This stands in contrast to previous approaches, which either only provide point estimates, or require a reset mechanism, or additional prior knowledge.
-The width of the interval converges to zero at a $\sqrt{n}$ rate, where $n$ is the length of the sample path.
-Upper and lower bounds are given on the number of samples required to achieve constant-factor multiplicative accuracy.
-The lower bounds indicate that, unless further restrictions are placed on the chain, no procedure can achieve this accuracy level before seeing each state at least $\Omega(\tmix)$ times on the average.
-Finally, future directions of research are identified.
+This article provides the first procedure for computing a fully
+data-dependent interval that traps the mixing time
+$t_{\text{mix}}$ of a finite
+reversible ergodic Markov chain at a prescribed confidence level.  The
+interval is computed from a single finite-length sample path from the
+Markov chain, and does not require the knowledge of any parameters of
+the chain.  This stands in contrast to previous approaches, which
+either only provide point estimates, or require a reset mechanism, or
+additional prior knowledge.
+The interval is constructed around the relaxation time
+$t_{\text{relax}}$, which is strongly related to the mixing time, and
+the width of the interval converges to zero
+at a $\sqrt{n}$ rate, where $n$ is the length of the sample path.
+Upper and lower bounds are given on the number of samples required to
+achieve constant-factor multiplicative accuracy.  The lower bounds
+indicate that, unless further restrictions are placed on the chain, no
+procedure can achieve this accuracy level before seeing each state at
+least $\Omega(t_{\text{relax}})$ times on the average.  Finally, future
+directions of research are identified.

### 2015-09-12 arxiv-v2/lowerbound.tex
more details on lowerbound claim in intro

@@
-\begin{proof}[Proof of \cref{thm:lb-pimin}]
+\begin{theorem}[Theorem~\ref{thm:lb-pimin} restated]
+  Pick any $\bar\pi \in (0,1/4)$.
+  Consider any estimator $\hatgap$ that takes as input a random sample
+  path of length $n \leq 1/(4\bar\pi)$ from a Markov chain starting
+  from any desired initial state distribution.
+  There exists a two-state ergodic and reversible Markov chain
+  distribution with spectral gap $\gap \geq 1/2$ and minimum
+  stationary probability $\pimin \geq \bar\pi$ such that
+  \[
+    \Pr\Brackets{ |\hatgap - \gap| \geq 1/8 } \geq 3/8 .
+  \]
+\end{theorem}
+\begin{proof}
   Fix $\bar\pi \in (0,1/4)$.
@@
 
-\begin{proof}[Proof of \cref{thm:lb-gap}]
+\begin{theorem}[Theorem~\ref{thm:lb-gap} restated]
+  There is an absolute constant $c>0$ such that the following holds.
+  Pick any positive integer $d \geq 3$ and any $\bar\gamma \in
+  (0,1/2)$.
+  Consider any estimator $\hatgap$ that takes as input a random sample
+  path of length $n < c d\log(d) / \bar\gamma$ from a $d$-state
+  reversible Markov chain starting from any desired initial state
+  distribution.
+  There is an ergodic and reversible Markov chain distribution
+  with spectral gap $\gap \in [\bar\gamma,2\bar\gamma]$ and minimum
+  stationary probability $\pimin \geq 1/(2d)$ such that
+  \[
+    \Pr\Brackets{ |\hatgap - \gap| \geq \bar\gamma/2} \geq 1/4 .
+  \]
+\end{theorem}
+\begin{proof}
   We consider $d$-state Markov chains of the following form:
@@
 \end{proof}
+
+We claim in Section~\ref{sec:intro} that a sample path of length
+$\Omega\parens{ (d \log d)/\gap + 1/\pimin}$ is required to guarantee
+constant multiplicative accuracy in estimating $\gap$.
+This follows by combining Theorems~\ref{thm:lb-pimin}
+and~\ref{thm:lb-gap} in a standard, straightforward way.
+Specifically, if the length of the sample path $n$ is smaller than
+$(n_1 + n_2) / 2$---where $n_1$ is the lower bound from
+Theorem~\ref{thm:lb-pimin}, and $n_2$ is the lower bound from
+Theorem~\ref{thm:lb-gap}---then $n$ is smaller than
+$\max\braces{n_1,n_2}$.
+So at least one of Theorem~\ref{thm:lb-pimin} and
+Theorem~\ref{thm:lb-gap} implies the existence of an ergodic and
+reversible Markov chain distribution with (say) $1/4 \leq \gap \leq
+1/2$ and $\pimin \geq 1/(2d)$ such that
+\[
+  \Pr\Brackets{ |\hatgap - \gap| \geq \gap/4 } \geq 1/4
+  .
+\]

### 2015-10-15 nips2015ws/abstract.tex
workshop version

--- 2015-09-12 arxiv-v2/abstract.tex
@@
 the width of the interval converges to zero
-at a $\sqrt{n}$ rate, where $n$ is the length of the sample path.
+at a $\tilde{O}(\sqrt{n})$ rate, where $n$ is the length of the sample path.
 Upper and lower bounds are given on the number of samples required to

### 2015-10-15 nips2015ws/empirical.tex
workshop version

--- 2015-06-09 arxiv-v1/empirical.tex
@@
-In this section, we address the shortcoming of \cref{thm:err} and give
-fully empirical confidence intervals for the stationary probabilities
-and the spectral gap $\gap$.
-The main idea is to use the Markov property to eliminate the
-dependence of the confidence intervals on the unknown quantities
-(including $\pimin$ and $\gap$).
-Specifically, we estimate the transition probabilities from the sample
-path using simple frequency estimates: as a consequence of the Markov
-property, for each state, the frequency estimates converge at a rate
-that depends only on the number of visits to the state, and in
-particular the rate (given the visit count of the state) is
-independent of the mixing time of the chain.
-
-As discussed in \cref{sec:rates}, it is possible to form a confidence
-interval for $\gap$ based on the eigenvalues of an estimated
-transition probability matrix by appealing to the
-Ostrowski-Elsner theorem.
-However, as explained earlier, this would lead to a slow
-$O(n^{-1/(2d)})$ rate.
-We avoid this slow rate by using an estimate of the symmetric matrix
-$\vL$, so that we can use a stronger perturbation result (namely Weyl's
-inequality, as in the proof of \cref{thm:err}) available for symmetric matrices.
-
-To form an estimate of $\vL$ based on an estimate of the transition
-probabilities, one possibility is to estimate $\vpi$ using a
-frequency-based estimate for $\vpi$ as was done in \cref{sec:rates},
-and appeal to the relation $\vL = \Diag(\vpi)^{1/2} \vP
-\Diag(\vpi)^{-1/2}$ to form a plug-in estimate.
-However, as noted in \cref{sec:rates-upper}, confidence intervals for
-the entries of $\vpi$ formed this way may depend on the mixing time.
-Indeed, such an estimate of $\vpi$ does not exploit the Markov
-property.
-
-We adopt a different strategy for estimating $\vpi$, which leads to
-our construction of empirical confidence intervals, detailed in
-\cref{alg:empest}.
-We form the matrix $\wh\vP$ using smoothed frequency estimates of
-$\vP$ (Step~\ref{step:P}), then compute the so-called group inverse
-$\giAh$ of $\wh\vA = I - \wh\vP$ (Step~\ref{step:gi}), followed by
-finding the unique stationary distribution $\hat\vpi$ of $\wh\vP$
-(Step~\ref{step:pi}), this way decoupling the bound on the accuracy of $\hat\vpi$
-from the mixing time.
-The group inverse $\giAh$ of $\wh\vA$ is uniquely defined;
- and if
-$\wh\vP$ defines an ergodic chain (which is the case here due to the
-use of the smoothed estimates), $\giAh$ can be computed at the cost of
-inverting an $(d{-}1){\times}(d{-}1)$ matrix~\citep[Theorem
-5.2]{meyer1975role}.%
-\footnote{%
-\label{ftnote:group-inverse}
-The group inverse of a square matrix $\vA$, a special case of the {\em Drazin inverse},
-is the unique matrix $\vA^\#$ satisfying
-$ \vA\vA^\#\vA = \vA$,
-$\vA^\#\vA\vA^\# = \vA^\#$ and
-$\vA^\#\vA = \vA\vA^\#$.
-}
-Further, once given $\giAh$, the unique stationary distribution
-$\hat\vpi$ of $\wh\vP$ can be read out from the last row of
-$\giAh$~\citep[Theorem 5.3]{meyer1975role}.
-The group inverse is also be used to compute the sensitivity of
-$\vpi$.
-Based on $\hat\vpi$ and $\wh\vP$, we construct the plug-in estimate
-$\wh\vL$ of $\vL$, and use the eigenvalues of its symmetrization to
-form the estimate $\hatgap$ of the spectral gap (Steps~\ref{step:eig}
-and~\ref{step:gap}).
-In the remaining steps, 
-we use perturbation analyses to relate $\hat\vpi$ and $\vpi$, viewing
-$\vP$ as the perturbation of $\wh\vP$; and also to relate $\hatgap$
-and $\gap$, viewing $\vL$ as a perturbation of $\Sym(\wh\vL)$.
-Both analyses give error bounds entirely in terms of observable
-quantities (e.g., $\hat\kappa$), tracing back to empirical error
-bounds for the smoothed frequency estimates of $\vP$.
-
-The most computationally expensive step in \cref{alg:empest} is the
-computation of the group inverse $\giAh$, which, as noted reduces to
-matrix inversion.
-Thus, with a standard implementation of matrix inversion, the
-algorithm's time complexity is $O(n + d^3)$, while its space
-complexity is $O(d^2)$.
-
+\iffalse
 \begin{algorithm}
@@
 \end{algorithm}
+\fi
 
-To state our main theorem concerning \cref{alg:empest}, we first define
-$\kappa$ to be analogous to $\hat{\kappa}$ from Step~\ref{step:sens},
-with $\giAh$ replaced by the group inverse $\giA$ of $\vA := \vI -
-\vP$.
-The result is as follows.
+In this section, we address the shortcoming of \cref{thm:err} and give
+fully empirical confidence intervals for the stationary probabilities
+and the spectral gap $\gap$.
+The main idea is to use the Markov property to eliminate the
+dependence of the confidence intervals on the unknown quantities
+(including $\pimin$ and $\gap$).
+Specifically, we estimate the transition probabilities from the sample
+path using simple frequency estimates: as a consequence of the Markov
+property, for each state, the frequency estimates converge at a rate
+that depends only on the number of visits to the state, and in
+particular the rate (given the visit count of the state) is
+independent of the mixing time of the chain.
+
+As discussed in \cref{sec:rates}, it is possible to form a confidence
+interval for $\gap$ based on the eigenvalues of an estimated
+transition probability matrix by appealing to the
+Ostrowski-Elsner theorem.
+However, as explained earlier, this would lead to a slow
+$O(n^{-1/(2d)})$ rate.
+We avoid this slow rate by using an estimate of the symmetric matrix
+$\vL$, so that we can use a stronger perturbation result (namely
+Weyl's inequality).
+%, as in the proof of \cref{thm:err}) available for
+%symmetric matrices.
+%
+To form an estimate of $\vL$ based on an estimate of the transition
+probabilities, one possibility is to estimate $\vpi$ using a
+frequency-based estimate for $\vpi$ as was done in \cref{sec:rates},
+and appeal to the relation $\vL = \Diag(\vpi)^{1/2} \vP
+\Diag(\vpi)^{-1/2}$ to form a plug-in estimate.
+However, as noted in \cref{sec:rates-upper}, confidence intervals for
+the entries of $\vpi$ formed this way may depend on the mixing time.
+Indeed, such an estimate of $\vpi$ does not exploit the Markov
+property.
+
+We adopt a different strategy for estimating $\vpi$, which leads to
+our construction of empirical confidence intervals.
+%, detailed in
+%\cref{alg:empest}.
+We form the matrix $\wh\vP$ using smoothed frequency estimates of
+$\vP$, then compute the so-called group inverse
+$\giAh$ of $\wh\vA = I - \wh\vP$, followed by
+finding the unique stationary distribution $\hat\vpi$ of $\wh\vP$, this way decoupling the bound on the accuracy of $\hat\vpi$
+from the mixing time.
+The group inverse $\giAh$ of $\wh\vA$ is uniquely defined;
+ and if
+$\wh\vP$ defines an ergodic chain (which is the case here due to the
+use of the smoothed estimates), $\giAh$ can be computed at the cost of
+inverting an $(d{-}1){\times}(d{-}1)$ matrix~\citep[Theorem
+5.2]{meyer1975role}.%
+\footnote{%
+\label{ftnote:group-inverse}
+The group inverse of a square matrix $\vA$, a special case of the {\em Drazin inverse},
+is the unique matrix $\vA^\#$ satisfying
+$ \vA\vA^\#\vA = \vA$,
+$\vA^\#\vA\vA^\# = \vA^\#$ and
+$\vA^\#\vA = \vA\vA^\#$.
+}
+Further, once given $\giAh$, the unique stationary distribution
+$\hat\vpi$ of $\wh\vP$ can be read out from the last row of
+$\giAh$~\citep[Theorem 5.3]{meyer1975role}.
+The group inverse is also be used to compute the sensitivity of
+$\vpi$.
+Based on $\hat\vpi$ and $\wh\vP$, we construct the plug-in estimate
+$\wh\vL$ of $\vL$, and use the eigenvalues of its symmetrization to
+form the estimate $\hatgap$ of the spectral gap.
+In the remaining steps, 
+we use perturbation analyses to relate $\hat\vpi$ and $\vpi$, viewing
+$\vP$ as the perturbation of $\wh\vP$; and also to relate $\hatgap$
+and $\gap$, viewing $\vL$ as a perturbation of $\Sym(\wh\vL)$.
+Both analyses give error bounds entirely in terms of observable
+quantities, tracing back to empirical error
+bounds for the smoothed frequency estimates of $\vP$.
+
+%The most computationally expensive step in \cref{alg:empest} is the
+%computation of the group inverse $\giAh$, which, as noted reduces to
+%matrix inversion.
+%Thus, with a standard implementation of matrix inversion, the
+%algorithm's time complexity is $O(n + d^3)$, while its space
+%complexity is $O(d^2)$.
+
 \begin{theorem}
   \label{thm:empirical}
-  Suppose \cref{alg:empest} is given as input a sample path of length
+  There is an algorithm that, if
+  given as input a sample path of length
   $n$ from an ergodic and reversible Markov chain and confidence
-  parameter $\delta \in (0,1)$.
+  parameter $\delta \in (0,1)$,
+  returns intervals $[\hat\pi_i-\hat{b},\hat\pi_i+\hat{b}]$
+  and $[\hatgap-\hat{w},\hatgap+\hat{w}]$ satisfying the following.
   Let $\gap>0$ denote the spectral gap, $\vpi$ the unique stationary
   distribution, and $\pimin>0$ the minimum stationary probability.
-  Then, on an event of probability at least $1-\delta$,
+  On an event of probability at least $1-\delta$,
   \[
@@
       \max_{(i,j) \in [d]^2}
-      \kappa
+      \frac{d}{\gap}
       \sqrt{\frac{P_{i,j}\log(d/\delta)}{\pi_i n}}
@@
     O\Parens{ 
-      \frac{\kappa}{\pimin} \sqrt{\frac{\log(d/\delta)}{\pimin n}} + 
+      \frac{d}{\pimin\gap} \sqrt{\frac{\log(d/\delta)}{\pimin n}} + 
       \sqrt{ \frac{d\log(d/\delta)}{ \pimin n}}
@@
     .\footnote{%
-    In \cref{thm:empirical,thm:combined}, our use of big-$O$ notation
+    In \cref{thm:empirical}, our use of big-$O$ notation
     is as follows.
@@
 \end{theorem}
+\iffalse
 The proof of \cref{thm:empirical} is given in \cref{app:empirical}.
@@
 intervals decrease as the sequence length increases.
-It is known that $\kappa \le d/\gap$ (see \cref{sec:asymptotic}),
-which gives
+We show in \cref{sec:asymptotic} that
+$\kappa \le d/\gap$, and hence
 \[
@@
 \end{theorem}
+\fi

### 2015-10-15 nips2015ws/intro.tex
workshop version

--- 2015-06-09 arxiv-v1/intro.tex
@@
   \label{eq:mixing-time-bound}
-  \Parens{\mfrac1{\gap} - 1} \ln2
+  \Parens{\trelax - 1} \ln2
   \ \leq \ \tmix
-  \ \leq \ \mfrac1{\gap} \ln \mfrac4{\pimin}
-  ,
+  \ \leq \ \trelax \ln \mfrac4{\pimin}
 \end{equation}
-and hence it suffices to estimate $\gap$ and $\pimin$. 
-Our main results are summarized as follows.
-\begin{enumerate}
+where $\trelax := 1/\gap$ is the \emph{relaxation time}.
+Hence, it suffices to estimate $\gap$ and $\pimin$. 
+Our main results are:
+\begin{enumerate}[leftmargin=1em]
   \item
@@
     (\cref{thm:empirical}).
-    We also explain how to combine the empirical confidence intervals
-    from \cref{alg:empest} with the non-empirical bounds from
+    In the full version of the paper,
+    we also explain how to combine the empirical confidence intervals
+    from \cref{thm:empirical} with the non-empirical bounds from
     \cref{thm:err} to produce valid empirical confidence intervals.
-    We prove in \cref{thm:combined} that the width of these new
+    We prove that the width of these new
     intervals converge to zero asymptotically at least as fast as
@@
 
+\iffalse
 \paragraph{Related work.}
@@
 barriers for many MCMC diagnostic problems~\citet{BhaBoMo11}.
+\fi

### 2015-10-15 nips2015ws/rates.tex
workshop version

--- 2015-08-07 arxiv-v2/rates.tex
@@
-In this section, we present lower and upper bounds on achievable rates
-for estimating the spectral gap as a function of the length of the
-sample path $n$.
+%We present lower/upper bounds on rates
+%for estimating spectral gap as a function of the length of the
+%sample path $n$.
 
@@
 \label{sec:rates-lower}
-The purpose of this section is to show lower bounds on the number of observations
-necessary to achieve a fixed multiplicative (or even just additive) accuracy in estimating the spectral gap $\gap$.
-By \cref{eq:mixing-time-bound}, the multiplicative accuracy lower bound for $\gap$
-gives the same lower bound for estimating the mixing time.
+We first show lower bounds on the number of observations necessary to
+achieve a fixed multiplicative (or even just additive) accuracy in
+estimating the spectral gap $\gap$.
+By \cref{eq:mixing-time-bound}, the multiplicative accuracy lower
+bound for $\gap$ gives the same lower bound for estimating the mixing
+time.
 Our first result holds even for two state Markov chains and shows that a sequence length of $\Omega(1/\pimin)$
@@
   stationary probability $\pimin \geq \bar\pi$ such that
-  \[
-    \Pr\Brackets{ |\hatgap - \gap| \geq 1/8 } \geq 1/6 .
-  \]
+  $\Pr\Brackets{ |\hatgap - \gap| \geq 1/8 } \geq 3/8$.
+%  \[
+%    \Pr\Brackets{ |\hatgap - \gap| \geq 1/8 } \geq 3/8 .
+%  \]
 \end{theorem}
@@
 to estimate $\gap$ up to a constant multiplicative accuracy.
-Essentially, the sequence may have to visit all $d$ states $\log(d)/\gap$ times each on average.
+Essentially, the sequence may have to visit all $d$ states at least
+$\log(d)/\gap$ times each, on average.
 This holds \emph{even} if $\pimin$ is within a factor of two of the
@@
   stationary probability $\pimin \geq 1/(2d)$ such that
-  \[
-    \Pr\Brackets{ |\hatgap - \gap| \geq \bar\gamma} \geq 1/4 .
-  \]
+  $\Pr\Brackets{ |\hatgap - \gap| \geq \bar\gamma/2} \geq 1/4$.
+%  \[
+%    \Pr\Brackets{ |\hatgap - \gap| \geq \bar\gamma/2} \geq 1/4 .
+%  \]
 \end{theorem}
 
-The proofs of \cref{thm:lb-pimin,thm:lb-gap} are given in
-\cref{app:lower}.
+%The proofs of \cref{thm:lb-pimin,thm:lb-gap} are given in
+%\cref{app:lower}.
 
@@
 \label{sec:rates-upper}
-Let us now consider the problem of estimating $\gap$.
-For this, we construct a natural plug-in estimator.
-Along the way, we also provide an estimator for the minimum stationary probability,
-allowing one to use the bounds from \cref{eq:mixing-time-bound} to trap
-the mixing time.
+We now consider the problem of estimating $\gap$ using a natural
+plug-in estimator.
+%For this, we construct a natural plug-in estimator.
+%Along the way, we also provide an estimator for the minimum stationary
+%probability, allowing one to use the bounds from
+%\cref{eq:mixing-time-bound} to trap the mixing time.
 
+\iffalse
 Define the random matrix $\wh\vM \in [0,1]^{d \times d}$ and random
@@
 \]
-
 These estimators have the following accuracy guarantees:
+\fi
 \begin{theorem}
@@
   For any $\delta \in (0,1)$, with probability at least $1-\delta$,
-  \begin{equation}\label{eq:piminbound}
-    \Abs{\hatpimin-\pimin}
+  the plug-in estimates $\hatpimin$ and $\hatgap$ of $\pimin$ and
+  $\gap$ satisfy
+   $\Abs{\hatpimin-\pimin}
     \le
     C \,
-    \Parens{
-      \sqrt{\frac{\pimin\log\frac{d}{\pimin\delta}}{\gap n}}
+    \parens{
+      \sqrt{\pimin\log\frac{d}{\pimin\delta}/(\gap n)}
       +
-      \frac{\log\frac{d}{\pimin\delta}}{\gap n}
-    }
-  \end{equation}
+      \log\frac{d}{\pimin\delta}/(\gap n)
+    }$
+%  \begin{equation}\label{eq:piminbound}
+%    \Abs{\hatpimin-\pimin}
+%    \le
+%    C \,
+%    \Parens{
+%      \sqrt{\frac{\pimin\log\frac{d}{\pimin\delta}}{\gap n}}
+%      +
+%      \frac{\log\frac{d}{\pimin\delta}}{\gap n}
+%    }
+%  \end{equation}
   and
-  \begin{equation}\label{eq:gapbound}
-    \Abs{\hatgap-\gap}
+   $\Abs{\hatgap-\gap}
     \leq
     C \,
-    \Parens{
-      \sqrt{\frac{\log\frac{d}{\delta}\cdot\log\frac{n}{\pimin\delta}}{\pimin\gap n}}
-      + \frac{\log\frac{1}{\gap}}{\gap n}  
-    }
-    \,.
-  \end{equation}
+    \parens{
+      \sqrt{\log\frac{d}{\delta}\cdot\log\frac{n}{\pimin\delta}/(\pimin\gap n)}
+      + \log\frac{1}{\gap}/(\gap n)
+    }$.
+%  \begin{equation}\label{eq:gapbound}
+%    \Abs{\hatgap-\gap}
+%    \leq
+%    C \,
+%    \Parens{
+%      \sqrt{\frac{\log\frac{d}{\delta}\cdot\log\frac{n}{\pimin\delta}}{\pimin\gap n}}
+%      + \frac{\log\frac{1}{\gap}}{\gap n}  
+%    }
+%    \,.
+%  \end{equation}
 \end{theorem}
 
-\Cref{thm:err} implies that the sequence lengths required to estimate
-$\pimin$ and
-$\gap$ to within constant multiplicative factors are, respectively,
-\[
-  \tilde{O}\Parens{\frac1{\pimin\gap}}
-  \quad\text{and}\quad
-  \tilde{O}\Parens{\frac1{\pimin\gap^3}}
-  .
-\]
-By \cref{eq:mixing-time-bound},
-the second of these is also a bound on the required sequence length to estimate $\tmix$.
+%\Cref{thm:err} implies that the sequence lengths required to estimate
+%$\pimin$ and
+%$\gap$ to within constant multiplicative factors are, respectively,
+%\[
+%  \tilde{O}\Parens{\frac1{\pimin\gap}}
+%  \quad\text{and}\quad
+%  \tilde{O}\Parens{\frac1{\pimin\gap^3}}
+%  .
+%\]
+%By \cref{eq:mixing-time-bound},
+%the second of these is also a bound on the required sequence length to estimate $\tmix$.
 
-The proof of \cref{thm:err} is based on analyzing the
-convergence of the sample averages $\wh{\vM}$ and
-$\hat\vpi$ to their expectation, and then using perturbation bounds
-for eigenvalues to derive a bound on the error of $\hatgap$.
-However, since these averages are formed using a \emph{single sample
-path} from a (possibly) non-stationary Markov chain, we cannot use
-standard large deviation bounds; moreover applying Chernoff-type
-bounds for Markov chains to each entry of $\wh{\vM}$ would result in a
-significantly worse sequence length requirement, roughly a factor of
-$d$ larger.
-Instead, we adapt probability tail bounds for sums of independent
-random matrices~\citep{tropp2015intro} to our non-iid setting by
-directly applying a blocking technique of~\citet{Bernstein27} as
-described in the article of~\citet{Yu94}.
-Due to ergodicity, the convergence rate can be bounded without any
-dependence on the initial state distribution $\vpi^{(1)}$.
-The proof of \cref{thm:err} is given in \cref{app:upper}.
+%The proof of \cref{thm:err} is based on analyzing the
+%convergence of the sample averages $\wh{\vM}$ and
+%$\hat\vpi$ to their expectation, and then using perturbation bounds
+%for eigenvalues to derive a bound on the error of $\hatgap$.
+%However, since these averages are formed using a \emph{single sample
+%path} from a (possibly) non-stationary Markov chain, we cannot use
+%standard large deviation bounds; moreover applying Chernoff-type
+%bounds for Markov chains to each entry of $\wh{\vM}$ would result in a
+%significantly worse sequence length requirement, roughly a factor of
+%$d$ larger.
+%Instead, we adapt probability tail bounds for sums of independent
+%random matrices~\citep{tropp2015intro} to our non-iid setting by
+%directly applying a blocking technique of~\citet{Bernstein27} as
+%described in the article of~\citet{Yu94}.
+%Due to ergodicity, the convergence rate can be bounded without any
+%dependence on the initial state distribution $\vpi^{(1)}$.
+%The proof of \cref{thm:err} is given in \cref{app:upper}.
 

### 2015-10-15 nips2015ws/setting.tex
workshop version

--- 2015-06-03 setting.tex
@@
-%!TEX root =  matrix-est.tex
-
-\subsection{Notations}
-\label{sec:notation}
-
+\paragraph{Notations.}
+%\label{sec:notation}
+%
 We denote the set of positive integers by $\bbN$, and
@@
 $\Delta^{d-1}$ as row vectors.
-%Finally, the total variation between two probability distributions
-%$\vp, \vq \in \Delta^{d-1}$ is denoted by $\norm{\vp-\vq}_{\tv}$.
 
-\subsection{Setting}
-\label{sec:setting}
-
+\paragraph{Setting.}
+%\label{sec:setting}
+%
 Let $\vP \in (\Delta^{d-1})^d \subset [0,1]^{d \times d}$ be a $d
@@
 $\vpi$):
-\begin{align}
-\label{eq:reversibility}
-  \pi_i P_{i,j} = \pi_j P_{j,i} ,
-  \quad i,j \in [d] .
-\end{align}
+$\pi_i P_{i,j} = \pi_j P_{j,i}$ for all $i,j \in [d]$.
+%\begin{align}
+%\label{eq:reversibility}
+%  \pi_i P_{i,j} = \pi_j P_{j,i} ,
+%  \quad i,j \in [d] .
+%\end{align}
 The minimum stationary probability is denoted by $\pimin := \min_{i
@@
 Define the matrices
-\begin{align*}
-\vM := \Diag(\vpi) \vP \quad \text{and} \quad
-\vL := \Diag(\vpi)^{-1/2} \vM \Diag(\vpi)^{-1/2}\,.
-\end{align*}
+$\vM := \Diag(\vpi) \vP$ and
+$\vL := \Diag(\vpi)^{-1/2} \vM \Diag(\vpi)^{-1/2}$.
+%\begin{align*}
+%\vM := \Diag(\vpi) \vP \quad \text{and} \quad
+%\vL := \Diag(\vpi)^{-1/2} \vM \Diag(\vpi)^{-1/2}\,.
+%\end{align*}
 The $(i,j)$th entry of the matrix $M_{i,j}$ contains the \emph{doublet
@@
 $\vP$, and hence it follows that $\vL$ is also symmetric.
-(We will strongly exploit the symmetry in our results.)
-%will be advantageous due to the existence of stronger perturbation results
-%for symmetric matrices.
+%(We will strongly exploit the symmetry in our results.)
 Further, $\vL = \Diag(\vpi)^{1/2} \vP \Diag(\vpi)^{-1/2}$, hence $\vL$
@@
 Denote and order the eigenvalues of $\vL$ as
-\[
-  1 = \lambda_1 > \lambda_2 \geq \dotsb \geq \lambda_d > -1 .
-\]
+$1 = \lambda_1 > \lambda_2 \geq \dotsb \geq \lambda_d > -1$.
+%\[
+%  1 = \lambda_1 > \lambda_2 \geq \dotsb \geq \lambda_d > -1 .
+%\]
 Let $\slem := \max\{ \lambda_2,\, |\lambda_d| \}$, and define the
@@
 marginal distribution of $X_t$, so
-\[
-  \vpi^{(t+1)} = \vpi^{(t)} \vP ,
-  \quad t \in \bbN .
-\]
+$\vpi^{(t+1)} = \vpi^{(t)} \vP$ for all $t \in \bbN$.
+%\[
+%  \vpi^{(t+1)} = \vpi^{(t)} \vP ,
+%  \quad t \in \bbN .
+%\]
 Note that the initial distribution $\vpi^{(1)}$ is arbitrary,

### 2015-10-15 arxiv-v2/empirical.tex
.

--- 2015-06-09 arxiv-v1/empirical.tex
@@
-In this section, we address the shortcoming of \cref{thm:err} and give
-fully empirical confidence intervals for the stationary probabilities
-and the spectral gap $\gap$.
-The main idea is to use the Markov property to eliminate the
-dependence of the confidence intervals on the unknown quantities
-(including $\pimin$ and $\gap$).
-Specifically, we estimate the transition probabilities from the sample
-path using simple frequency estimates: as a consequence of the Markov
-property, for each state, the frequency estimates converge at a rate
-that depends only on the number of visits to the state, and in
-particular the rate (given the visit count of the state) is
-independent of the mixing time of the chain.
-
-As discussed in \cref{sec:rates}, it is possible to form a confidence
-interval for $\gap$ based on the eigenvalues of an estimated
-transition probability matrix by appealing to the
-Ostrowski-Elsner theorem.
-However, as explained earlier, this would lead to a slow
-$O(n^{-1/(2d)})$ rate.
-We avoid this slow rate by using an estimate of the symmetric matrix
-$\vL$, so that we can use a stronger perturbation result (namely Weyl's
-inequality, as in the proof of \cref{thm:err}) available for symmetric matrices.
-
-To form an estimate of $\vL$ based on an estimate of the transition
-probabilities, one possibility is to estimate $\vpi$ using a
-frequency-based estimate for $\vpi$ as was done in \cref{sec:rates},
-and appeal to the relation $\vL = \Diag(\vpi)^{1/2} \vP
-\Diag(\vpi)^{-1/2}$ to form a plug-in estimate.
-However, as noted in \cref{sec:rates-upper}, confidence intervals for
-the entries of $\vpi$ formed this way may depend on the mixing time.
-Indeed, such an estimate of $\vpi$ does not exploit the Markov
-property.
-
-We adopt a different strategy for estimating $\vpi$, which leads to
-our construction of empirical confidence intervals, detailed in
-\cref{alg:empest}.
-We form the matrix $\wh\vP$ using smoothed frequency estimates of
-$\vP$ (Step~\ref{step:P}), then compute the so-called group inverse
-$\giAh$ of $\wh\vA = I - \wh\vP$ (Step~\ref{step:gi}), followed by
-finding the unique stationary distribution $\hat\vpi$ of $\wh\vP$
-(Step~\ref{step:pi}), this way decoupling the bound on the accuracy of $\hat\vpi$
-from the mixing time.
-The group inverse $\giAh$ of $\wh\vA$ is uniquely defined;
- and if
-$\wh\vP$ defines an ergodic chain (which is the case here due to the
-use of the smoothed estimates), $\giAh$ can be computed at the cost of
-inverting an $(d{-}1){\times}(d{-}1)$ matrix~\citep[Theorem
-5.2]{meyer1975role}.%
-\footnote{%
-\label{ftnote:group-inverse}
-The group inverse of a square matrix $\vA$, a special case of the {\em Drazin inverse},
-is the unique matrix $\vA^\#$ satisfying
-$ \vA\vA^\#\vA = \vA$,
-$\vA^\#\vA\vA^\# = \vA^\#$ and
-$\vA^\#\vA = \vA\vA^\#$.
-}
-Further, once given $\giAh$, the unique stationary distribution
-$\hat\vpi$ of $\wh\vP$ can be read out from the last row of
-$\giAh$~\citep[Theorem 5.3]{meyer1975role}.
-The group inverse is also be used to compute the sensitivity of
-$\vpi$.
-Based on $\hat\vpi$ and $\wh\vP$, we construct the plug-in estimate
-$\wh\vL$ of $\vL$, and use the eigenvalues of its symmetrization to
-form the estimate $\hatgap$ of the spectral gap (Steps~\ref{step:eig}
-and~\ref{step:gap}).
-In the remaining steps, 
-we use perturbation analyses to relate $\hat\vpi$ and $\vpi$, viewing
-$\vP$ as the perturbation of $\wh\vP$; and also to relate $\hatgap$
-and $\gap$, viewing $\vL$ as a perturbation of $\Sym(\wh\vL)$.
-Both analyses give error bounds entirely in terms of observable
-quantities (e.g., $\hat\kappa$), tracing back to empirical error
-bounds for the smoothed frequency estimates of $\vP$.
-
-The most computationally expensive step in \cref{alg:empest} is the
-computation of the group inverse $\giAh$, which, as noted reduces to
-matrix inversion.
-Thus, with a standard implementation of matrix inversion, the
-algorithm's time complexity is $O(n + d^3)$, while its space
-complexity is $O(d^2)$.
-
 \begin{algorithm}
@@
 
+In this section, we address the shortcoming of \cref{thm:err} and give
+fully empirical confidence intervals for the stationary probabilities
+and the spectral gap $\gap$.
+The main idea is to use the Markov property to eliminate the
+dependence of the confidence intervals on the unknown quantities
+(including $\pimin$ and $\gap$).
+Specifically, we estimate the transition probabilities from the sample
+path using simple frequency estimates: as a consequence of the Markov
+property, for each state, the frequency estimates converge at a rate
+that depends only on the number of visits to the state, and in
+particular the rate (given the visit count of the state) is
+independent of the mixing time of the chain.
+
+As discussed in \cref{sec:rates}, it is possible to form a confidence
+interval for $\gap$ based on the eigenvalues of an estimated
+transition probability matrix by appealing to the
+Ostrowski-Elsner theorem.
+However, as explained earlier, this would lead to a slow
+$O(n^{-1/(2d)})$ rate.
+We avoid this slow rate by using an estimate of the symmetric matrix
+$\vL$, so that we can use a stronger perturbation result (namely Weyl's
+inequality, as in the proof of \cref{thm:err}) available for symmetric matrices.
+
+To form an estimate of $\vL$ based on an estimate of the transition
+probabilities, one possibility is to estimate $\vpi$ using a
+frequency-based estimate for $\vpi$ as was done in \cref{sec:rates},
+and appeal to the relation $\vL = \Diag(\vpi)^{1/2} \vP
+\Diag(\vpi)^{-1/2}$ to form a plug-in estimate.
+However, as noted in \cref{sec:rates-upper}, confidence intervals for
+the entries of $\vpi$ formed this way may depend on the mixing time.
+Indeed, such an estimate of $\vpi$ does not exploit the Markov
+property.
+
+We adopt a different strategy for estimating $\vpi$, which leads to
+our construction of empirical confidence intervals, detailed in
+\cref{alg:empest}.
+We form the matrix $\wh\vP$ using smoothed frequency estimates of
+$\vP$ (Step~\ref{step:P}), then compute the so-called group inverse
+$\giAh$ of $\wh\vA = \vI - \wh\vP$ (Step~\ref{step:gi}), followed by
+finding the unique stationary distribution $\hat\vpi$ of $\wh\vP$
+(Step~\ref{step:pi}), this way decoupling the bound on the accuracy of $\hat\vpi$
+from the mixing time.
+The group inverse $\giAh$ of $\wh\vA$ is uniquely defined;
+ and if
+$\wh\vP$ defines an ergodic chain (which is the case here due to the
+use of the smoothed estimates), $\giAh$ can be computed at the cost of
+inverting an $(d{-}1){\times}(d{-}1)$ matrix~\citep[Theorem
+5.2]{meyer1975role}.%
+\footnote{%
+\label{ftnote:group-inverse}
+The group inverse of a square matrix $\vA$, a special case of the {\em Drazin inverse},
+is the unique matrix $\vA^\#$ satisfying
+$ \vA\vA^\#\vA = \vA$,
+$\vA^\#\vA\vA^\# = \vA^\#$ and
+$\vA^\#\vA = \vA\vA^\#$.
+}
+Further, once given $\giAh$, the unique stationary distribution
+$\hat\vpi$ of $\wh\vP$ can be read out from the last row of
+$\giAh$~\citep[Theorem 5.3]{meyer1975role}.
+The group inverse is also be used to compute the sensitivity of
+$\vpi$.
+Based on $\hat\vpi$ and $\wh\vP$, we construct the plug-in estimate
+$\wh\vL$ of $\vL$, and use the eigenvalues of its symmetrization to
+form the estimate $\hatgap$ of the spectral gap (Steps~\ref{step:eig}
+and~\ref{step:gap}).
+In the remaining steps, 
+we use perturbation analyses to relate $\hat\vpi$ and $\vpi$, viewing
+$\vP$ as the perturbation of $\wh\vP$; and also to relate $\hatgap$
+and $\gap$, viewing $\vL$ as a perturbation of $\Sym(\wh\vL)$.
+Both analyses give error bounds entirely in terms of observable
+quantities (e.g., $\hat\kappa$), tracing back to empirical error
+bounds for the smoothed frequency estimates of $\vP$.
+
+The most computationally expensive step in \cref{alg:empest} is the
+computation of the group inverse $\giAh$, which, as noted reduces to
+matrix inversion.
+Thus, with a standard implementation of matrix inversion, the
+algorithm's time complexity is $O(n + d^3)$, while its space
+complexity is $O(d^2)$.
+
 To state our main theorem concerning \cref{alg:empest}, we first define

### 2015-10-15 nips2015ws/empirical.tex
put in log log n
simplify big O bound
.

@@
 $\vP$, then compute the so-called group inverse
-$\giAh$ of $\wh\vA = I - \wh\vP$, followed by
+$\giAh$ of $\wh\vA = \vI - \wh\vP$, followed by
 finding the unique stationary distribution $\hat\vpi$ of $\wh\vP$, this way decoupling the bound on the accuracy of $\hat\vpi$
@@
       \frac{d}{\gap}
-      \sqrt{\frac{P_{i,j}\log(d/\delta)}{\pi_i n}}
+      \sqrt{\frac{P_{i,j}\log\log n}{\pi_i n}}
     }
@@
     O\Parens{
-      \frac{d}{\pimin\gap} \sqrt{\frac{\log(d/\delta)}{\pimin n}} + 
-      \sqrt{ \frac{d\log(d/\delta)}{ \pimin n}}
+      \frac{d}{\pimin^{3/2}\gap}
+      \sqrt{\frac{\log\log n}{n}}
+%      \frac{d}{\pimin\gap} \sqrt{\frac{\log(d/\delta)}{\pimin n}} + 
+%      \sqrt{ \frac{d\log(d/\delta)}{ \pimin n}}
     }

### 2015-10-27 arxiv-v2/appendix-empirical.tex
added freedman inequality

--- 2015-06-09 arxiv-v1/appendix-empirical.tex
@@
 \label{sec:P-obs-bound}
+
+We make use of Freedman's inequality for martingales~\citep[Theorem
+1.6]{Fre75}, which we state below in a simplified form.
+\begin{theorem}
+  \label{thm:freedman}
+  Let $Y_1, Y_2, \dotsc, Y_n$ be a bounded martingale difference
+  sequence with respect to the filtration $\cF = (\cF_t)_{1 \leq t
+  \leq n}$ and with $|Y_t| \leq b$ for some $b>0$.
+  Let $S := \sum_{t=1}^n Y_t$ and $V := \EE{Y_t^2|\cF_{t-1}}$.
+  For all $t, v > 0$,
+  \[
+    \Pr\Brackets{
+      S > \sqrt{2vt} + 2bt/3
+      \,\wedge\,
+      V \leq v
+    }
+    \leq e^{-t}
+    .
+  \]
+\end{theorem}
+\begin{corollary}
+  \label{cor:freedman}
+  Under the same setting as \Cref{thm:freedman}, for any $t > 0$ and
+  $c > 1$,
+  \[
+    \Pr\Brackets{
+      S > \sqrt{2cVt} + 5bt/3
+    }
+    \leq
+    \Parens{ 1 + \ceil{\log_c(n)} }
+    e^{-t}
+    .
+  \]
+\end{corollary}
+\begin{proof}
+  Define $v_i := c^i b^2t/2$ for $i = 0, 1, \dotsc, \ceil{ \log_c(n)
+  }$, and let $v_{-1} := -\infty$.
+  Then, since $V \leq b^2n$,
+  \begin{align*}
+    \Pr\Brackets{
+      S > \sqrt{2\max\braces{v_0,cV}t} + 2bt/3
+    }
+    & =
+    \sum_{i=0}^{\ceil{\log_c(n)}}
+    \Pr\Brackets{
+      S > \sqrt{2\max\braces{v_0,cV}t} + 2bt/3
+      \,\wedge\, v_{j-1} < V \leq v_j
+    }
+    \\
+    & \leq
+    \sum_{i=0}^{\ceil{\log_c(n)}}
+    \Pr\Brackets{
+      S > \sqrt{2\max\braces{v_0,cv_{j-1}}t} + 2bt/3
+      \,\wedge\, v_{j-1} < V \leq v_j
+    }
+    \\
+    & \leq
+    \sum_{i=0}^{\ceil{\log_c(n)}}
+    \Pr\Brackets{
+      S > \sqrt{2v_jt} + 2bt/3
+      \,\wedge\, V \leq v_j
+    }
+    \\
+    & \leq
+    \sum_{i=0}^{\ceil{\log_c(n)}}
+    e^{-t}
+  \end{align*}
+  where the final inequality uses \Cref{thm:freedman}.
+  The conclusion now follows because
+  \[
+    \sqrt{2cVt} + 5bt/3
+    \geq \sqrt{2\max\braces{v_0,cV}t} + 2bt/3
+    .
+    \qedhere
+  \]
+\end{proof}
 

### 2015-10-27 arxiv-v2/appendix-empirical.tex
fixed freedman inequality

@@
 
-We make use of Freedman's inequality for martingales~\citep[Theorem
-1.6]{Fre75}, which we state below in a simplified form.
-\begin{theorem}
+We make use of a simple corollary of Freedman's inequality for
+martingales~\citep[Theorem 1.6]{Fre75}.
+\begin{theorem}[Freedman's inequality]
   \label{thm:freedman}
   Let $Y_1, Y_2, \dotsc, Y_n$ be a bounded martingale difference
-  sequence with respect to the filtration $\cF = (\cF_t)_{1 \leq t
-  \leq n}$ and with $|Y_t| \leq b$ for some $b>0$.
-  Let $S := \sum_{t=1}^n Y_t$ and $V := \EE{Y_t^2|\cF_{t-1}}$.
+  sequence with respect to the filtration $\cF = (\cF_i)_{1 \leq i
+  \leq n}$ and with $|Y_i| \leq b$ for some $b>0$.
+  Let $V := \sum_{i=1}^n \EE{Y_i^2|\cF_{i-1}}$ and $S := \max_{1 \leq
+  m \leq n} \sum_{i=1}^m Y_i$.
   For all $t, v > 0$,
@@
     \leq
-    \Parens{ 1 + \ceil{\log_c(n)} }
+    \Parens{ 1 + \ceil{\log_c(2n/t)} }
     e^{-t}
@@
 \begin{proof}
-  Define $v_i := c^i b^2t/2$ for $i = 0, 1, \dotsc, \ceil{ \log_c(n)
-  }$, and let $v_{-1} := -\infty$.
-  Then, since $V \leq b^2n$,
+  Define $v_i := c^i b^2t/2$ for $i = 0, 1, 2, \dotsc, \ceil{
+  \log_c(2n/t) }$, and let $v_{-1} := -\infty$.
+  Then, since $V \in [0,b^2n]$,
   \begin{align*}
@@
     & =
-    \sum_{i=0}^{\ceil{\log_c(n)}}
+    \sum_{i=0}^{\ceil{\log_c(2n/t)}}
     \Pr\Brackets{
@@
     & \leq
-    \sum_{i=0}^{\ceil{\log_c(n)}}
+    \sum_{i=0}^{\ceil{\log_c(2n/t)}}
     \Pr\Brackets{
@@
     & \leq
-    \sum_{i=0}^{\ceil{\log_c(n)}}
+    \Parens{ 1 + \ceil{\log_c(2n/t)} }
     e^{-t}

### 2015-10-27 arxiv-v2/appendix-empirical.tex
fixed constants in lemma 4

@@
   \label{lem:P-unobs-bound}
-  The following holds with probability at least $1-\delta$: for all
-  $(i,j) \in \iset{d}^2$,
+  The following holds for any constant $c>1$ with probability at least
+  $1-\delta$: for all $(i,j) \in \iset{d}^2$,
   \begin{equation}
@@
     \leq
-    \sqrt{\Parens{\frac{N_i}{N_i+d\alpha}}\frac{2P_{i,j}(1-P_{i,j})\ln(2d/\delta)}{N_i+d\alpha}}
-    + \frac{\frac23\ln(2d/\delta)}{N_i+d\alpha}
+    \sqrt{\Parens{\frac{N_i}{N_i+d\alpha}}\frac{2cP_{i,j}(1-P_{i,j})\emptail}{N_i+d\alpha}}
+    + \frac{(5/3)\emptail}{N_i+d\alpha}
     + \frac{\abs{\alpha-d\alpha P_{i,j}}}{N_i+d\alpha}
-    .
     \label{eq:P-unobs-bound}
+  \end{equation}
+  where
+  \begin{equation}
+    \emptail
+    := \inf
+    \Braces{
+      t\geq0 :
+      2d^2 \Parens{ 1 + \ceil{\log_c(2n/t)} } e^{-t} \leq \delta
+    }
+    = O\Parens{ \log\Parens{ \frac{d\log(n)}{\delta} } }
+    \,.
+    \label{eq:Tdelta}
   \end{equation}
@@
   \]
-  Therefore, by Freedman's inequality~\citep[Theorem 1.6]{Fre75} and
-  union bounds, we have
+  Therefore, by \Cref{cor:freedman} and union bounds, we have
   \[
     \abs{ N_{i,j} - N_i P_{i,j} }
-    \leq \sqrt{2 N_i P_{i,j} (1-P_{i,j}) \ln(2d/\delta)} + \frac23
-    \ln\left(\frac{2d}{\delta}\right)
+    \leq \sqrt{2c N_i P_{i,j} (1-P_{i,j}) \emptail} + \frac53
+    \emptail
   \]
@@
     \leq
-    \sqrt{\frac{9\wh{P}_{i,j}(1-\wh{P}_{i,j})\ln(2d/\delta)}{2N_i}}
-    + \frac{4\ln(2d/\delta)+1.5\abs{\alpha-d\alpha\wh{P}_{i,j}}}{N_i}
+    \sqrt{\frac{9c\wh{P}_{i,j}(1-\wh{P}_{i,j})\emptail}{2N_i}}
+    + \frac{(3c+2.5)\ln(2d/\delta)+1.5\abs{\alpha-d\alpha\wh{P}_{i,j}}}{N_i}
     .
@@
       + \frac{\alpha - d\alpha P_{i,j}}{N_i + d\alpha}
-    } \\
+    } %\\
     \leq
-    \sqrt{\frac{2N_iP_{i,j}(1-P_{i,j})\ln(2d/\delta)}{(N_i+d\alpha)^2}}
-    + \frac{\frac23\ln(2d/\delta)}{N_i+d\alpha}
+    \sqrt{\frac{2cN_iP_{i,j}(1-P_{i,j})\emptail}{(N_i+d\alpha)^2}}
+    + \frac{(5/3)\emptail}{N_i+d\alpha}
     + \frac{\abs{\alpha-d\alpha P_{i,j}}}{N_i+d\alpha}
@@
       \frac{
-        2N_i\parens{
+        2cN_i\parens{
           \wh{P}_{i,j}(1-\wh{P}_{i,j})
           + \abs{\wh{P}_{i,j} - P_{i,j}}
-        } \ln(2d/\delta)
+        } \emptail
       }{(N_i+d\alpha)^2}
     }
-    + \frac{\frac23\ln(2d/\delta)}{N_i+d\alpha}
+    + \frac{(5/3)\emptail}{N_i+d\alpha}
     \notag \\
@@
     \sqrt{\abs{\wh{P}_{i,j} - P_{i,j}}} \cdot
-    \sqrt{\frac{2N_i\ln(2d/\delta)}{(N_i+d\alpha)^2}}
+    \sqrt{\frac{2cN_i\emptail}{(N_i+d\alpha)^2}}
     \\
@@
     +
-    \sqrt{\frac{2N_i\wh{P}_{i,j}(1-\wh{P}_{i,j})\ln(2d/\delta)}{(N_i+d\alpha)^2}}
-    + \frac{\frac23\ln(2d/\delta) + \abs{\alpha -
+    \sqrt{\frac{2cN_i\wh{P}_{i,j}(1-\wh{P}_{i,j})\emptail}{(N_i+d\alpha)^2}}
+    + \frac{(5/3)\emptail + \abs{\alpha -
     d\alpha\wh{P}_{i,j}}}{N_i+d\alpha}

### 2015-10-28 arxiv-v2/empirical.tex
.

@@
   \in [d]^2$:
+  let $\emptail
+    := \inf
+    \braces{
+      t\geq0 :
+      2d^2 \parens{ 1 + \ceil{\ln(2n/t)} } e^{-t} \leq \delta
+    }$ and
   \begin{equation}
@@
     :=
-    \sqrt{\frac{9\wh{P}_{i,j}(1-\wh{P}_{i,j})\ln(2d/\delta)}{2N_i}}
-    + \frac{4\ln(2d/\delta)+1.5\abs{(1/d)-\wh{P}_{i,j}}}{N_i}
+    \sqrt{\frac{9e\wh{P}_{i,j}(1-\wh{P}_{i,j})\emptail}{2N_i}}
+    + \frac{(3e+2.5)\emptail+1.5\abs{(1/d)-\wh{P}_{i,j}}}{N_i}
     .
@@
       \kappa
-      \sqrt{\frac{P_{i,j}\log(d/\delta)}{\pi_i n}}
+      \sqrt{\frac{P_{i,j}\log\log n}{\pi_i n}}
     }
@@
     O\Parens{ 
-      \frac{\kappa}{\pimin} \sqrt{\frac{\log(d/\delta)}{\pimin n}} + 
-      \sqrt{ \frac{d\log(d/\delta)}{ \pimin n}}
+      \frac{\kappa}{\pimin} \sqrt{\frac{\log\log n}{\pimin n}} + 
+      \sqrt{ \frac{d\log\log n}{ \pimin n}}
     }
@@
 We establish fully observable upper and lower bounds on the entries of
-$\vP$ that converge at a $\sqrt{n}$ rate using standard martingale
-tail inequalities; this justifies the validity of the bounds from
-Step~\ref{step:P-bound}.
+$\vP$ that converge at a $\sqrt{n/\log\log n}$ rate using standard
+martingale tail inequalities; this justifies the validity of the
+bounds from Step~\ref{step:P-bound}.
 Properties of the group inverse~\citep{meyer1975role,cho2001comparison} and eigenvalue
@@
     \frac{d}{\gap}
-    \sqrt{\frac{P_{i,j}\log(d/\delta)}{\pi_i n}}
+    \sqrt{\frac{P_{i,j}\log\log n}{\pi_i n}}
   }
@@
   O\Parens{ 
-    \frac{d}{\pimin\gap} \sqrt{\frac{\log(d/\delta)}{\pimin n}}
+    \frac{d}{\pimin\gap} \sqrt{\frac{\log\log n}{\pimin n}}
   }

### 2015-10-28 arxiv-v2/appendix-empirical.tex
fixed constants in lemma 4
.

@@
     \,.
-    \label{eq:Tdelta}
+    \label{eq:emptail}
   \end{equation}
@@
 of empirical Bernstein bounds~\citep{audibert2009}.
-Specifically, we compute
+Specifically, with $c := 1.01$ (an arbitrary choice), we compute
 \begin{equation}
-  B_{i,j} :=
-  \sqrt{\frac{9\wh{P}_{i,j}(1-\wh{P}_{i,j})\ln(2d/\delta)}{2N_i}}
-  + \frac{4\ln(2d/\delta)+1.5\abs{\alpha-d\alpha\wh{P}_{i,j}}}{N_i}
+  \wh{B}_{i,j}
+  :=
+  \min_{\veps>0}
+  (1+\veps)
+  \sqrt{\frac{2c\wh{P}_{i,j}(1-\wh{P}_{i,j})\emptail}{N_i}}
+  +
+  \Parens{
+    \Parens{ 1 + \frac1{2\veps} + \frac{\veps}{2} }c
+    + \frac{5(1+\veps)}{3}
+  } \frac{\emptail}{N_i}
+  + (1+\veps) \frac{\abs{\alpha-d\alpha\wh{P}_{i,j}}}{N_i}
   \label{eq:P-obs-bound}
 \end{equation}
-for each $(i,j) \in [d]^2$.
+for each $(i,j) \in [d]^2$, where $\emptail$ is defined in
+\cref{eq:emptail}.
 We show in \cref{lem:P-obs-bound} that
@@
   \Brackets{
-    \wh{P}_{i,j} - B_{i,j},\,
-    \wh{P}_{i,j} + B_{i,j}
+    \wh{P}_{i,j} - \wh{B}_{i,j},\,
+    \wh{P}_{i,j} + \wh{B}_{i,j}
   }
@@
   In the same $1-\delta$ event as from \cref{lem:P-unobs-bound},
-  we have for all $(i,j) \in [d]^2$,
-  \begin{equation*}
-    \abs{\wh{P}_{i,j} - P_{i,j}}
-    \leq
-    \sqrt{\frac{9c\wh{P}_{i,j}(1-\wh{P}_{i,j})\emptail}{2N_i}}
-    + \frac{(3c+2.5)\ln(2d/\delta)+1.5\abs{\alpha-d\alpha\wh{P}_{i,j}}}{N_i}
-    .
-  \end{equation*}
+  we have
+  $P_{i,j} \in \brackets{ \wh{P}_{i,j} - \wh{B}_{i,j},\, \wh{P}_{i,j}
+  + \wh{B}_{i,j} }$
+  for all $(i,j) \in [d]^2$,
+%  \begin{equation*}
+%    \abs{\wh{P}_{i,j} - P_{i,j}}
+%    \leq
+%    (1+\veps)
+%    \sqrt{\frac{2c\wh{P}_{i,j}(1-\wh{P}_{i,j})\emptail}{N_i}}
+%    +
+%    \Parens{
+%      \Parens{ 1 + \frac1{2\veps} + \frac{\veps}{2} }c
+%      + \frac{5(1+\veps)}{3}
+%    } \frac{\emptail}{N_i}
+%    + (1+\veps) \frac{\abs{\alpha-d\alpha\wh{P}_{i,j}}}{N_i}
+%  \end{equation*}
+  where $\wh{B}_{i,j}$ is defined in \cref{eq:P-obs-bound}.
 \end{lemma}
@@
 \begin{align*}
-  \sqrt{n} \hat{b}
+  \sqrt{\frac{n}{\log\log n}}
+  \hat{b}
   & =
   O\Parens{
-    \max_{i,j} \kappa \sqrt{\frac{P_{i,j} \log(d/\delta)}{\pi_i}}
+    \max_{i,j} \kappa \sqrt{\frac{P_{i,j}}{\pi_i}}
   }
@@
   \\
-  \sqrt{n} \hat\rho
+  \sqrt{\frac{n}{\log\log n}}
+  \hat\rho
   & =
   O\Parens{
-    \frac{\kappa}{\pimin} \sqrt{\frac{\log(d/\delta)}{\pimin}}
+    \frac{\kappa}{\pimin^{3/2}}
   }
@@
 \begin{align*}
-  \sqrt{n}
+  \sqrt{\frac{n}{\log\log n}}
   \Biggl(
@@
       \sum_{i,j} \frac{{\pi}_i}{{\pi}_j}
-      \cdot \frac{P_{i,j}(1-P_{i,j})\log(d/\delta)}{\pi_i}
+      \cdot \frac{P_{i,j}(1-P_{i,j})}{\pi_i}
     \Biggr)^{1/2}
   }
-  & =
-  O\Parens{ \sqrt{ \frac{d\log(d/\delta)}{ \pimin}} }
+  =
+  O\Parens{ \sqrt{ \frac{d}{\pimin}} }
   .
@@
 \begin{align*}
-  \sqrt{n} \hat{w}
+  \sqrt{\frac{n}{\log\log n}}
+  \hat{w}
   =
   O\Parens{ 
-    \frac{\kappa}{\pimin} \sqrt{\frac{\log(d/\delta)}{\pimin}} + 
-    \sqrt{ \frac{d\log(d/\delta)}{ \pimin}}
+    \frac{\kappa}{\pimin^{3/2}} + 
+    \sqrt{\frac{d}{\pimin}}
   }

### 2015-10-28 arxiv-v2/appendix-empirical.tex
.

@@
   :=
-  \min_{\veps>0}
-  (1+\veps)
+  \Parens{
+    \sqrt{\frac{c\emptail}{2N_i}}
+    + \sqrt{
+      \frac{c\emptail}{2N_i}
+      + 
       \sqrt{\frac{2c\wh{P}_{i,j}(1-\wh{P}_{i,j})\emptail}{N_i}}
-  +
-  \Parens{
-    \Parens{ 1 + \frac1{2\veps} + \frac{\veps}{2} }c
-    + \frac{5(1+\veps)}{3}
-  } \frac{\emptail}{N_i}
-  + (1+\veps) \frac{\abs{\alpha-d\alpha\wh{P}_{i,j}}}{N_i}
+      + \frac{(5/3)\emptail + \abs{\alpha-d\alpha\wh{P}_{i,j}}}{N_i}
+    }
+  }^2
   \label{eq:P-obs-bound}
@@
   for all $(i,j) \in [d]^2$,
-%  \begin{equation*}
-%    \abs{\wh{P}_{i,j} - P_{i,j}}
-%    \leq
-%    (1+\veps)
-%    \sqrt{\frac{2c\wh{P}_{i,j}(1-\wh{P}_{i,j})\emptail}{N_i}}
-%    +
-%    \Parens{
-%      \Parens{ 1 + \frac1{2\veps} + \frac{\veps}{2} }c
-%      + \frac{5(1+\veps)}{3}
-%    } \frac{\emptail}{N_i}
-%    + (1+\veps) \frac{\abs{\alpha-d\alpha\wh{P}_{i,j}}}{N_i}
-%  \end{equation*}
   where $\wh{B}_{i,j}$ is defined in \cref{eq:P-obs-bound}.
@@
   \end{align*}
-  Whenever $N_i > 0$, we can solve a quadratic inequality and apply
-  AM/GM to conclude the claimed inequalities.
+  Whenever $N_i > 0$, we can solve a quadratic inequality to conclude
+  $\abs{\wh{P}_{i,j}-P_{i,j}} \leq \wh{B}_{i,j}$.
 \end{proof}

### 2015-10-28 arxiv-v2/abstract.tex
abstract

@@
 $t_{\text{relax}}$, which is strongly related to the mixing time, and
-the width of the interval converges to zero
+the width of the interval converges to zero roughly
 at a $\sqrt{n}$ rate, where $n$ is the length of the sample path.

### 2015-10-29 nips2015final/intro.tex
nips final

--- 2015-06-09 arxiv-v1/intro.tex
@@
   \label{eq:mixing-time-bound}
-  \Parens{\mfrac1{\gap} - 1} \ln2
+  \Parens{\trelax - 1} \ln2
   \ \leq \ \tmix
-  \ \leq \ \mfrac1{\gap} \ln \mfrac4{\pimin}
-  ,
+  \ \leq \ \trelax \ln \mfrac4{\pimin}
 \end{equation}
-and hence it suffices to estimate $\gap$ and $\pimin$. 
+where $\trelax := 1/\gap$ is the \emph{relaxation time}.
+Hence, it suffices to estimate $\gap$ and $\pimin$. 
 Our main results are summarized as follows.
@@
 prudent evaluation of a method or even its algorithmic design
-\cite{
-MCMCDiscussion93%
+\cite{%
+%MCMCDiscussion93%
 ,DBLP:conf/valuetools/KontoyiannisLM06%
@@
 order to provide data-dependent assessments of estimation
-accuracy~\cite[e.g.,][]{MCMCDiscussion93,GaSmi00:eigval,jones2001,flegal2011implementing,1209.0703,Gyori-paulin15}.
+accuracy~\cite[e.g.,][]{%
+%  MCMCDiscussion93,%
+  GaSmi00:eigval,%
+  jones2001,%
+  flegal2011implementing,%
+  1209.0703,%
+  Gyori-paulin15%
+}.
 However, these approaches generally only provide asymptotic

### 2015-10-29 nips2015final/setting.tex
nips final

--- 2015-06-09 arxiv-v1/setting.tex
@@
 the set of the first $d$ positive integers $\{1,2,\dotsc,d\}$ by $\iset{d}$.
-The non-negative part of a real number $x$ is $[x]_+ := \max\{0,x\}$.
-We use $\ln(\cdot)$ in expressions with exact constants, while we use $\log(\cdot)$ elsewhere.
+The non-negative part of a real number $x$ is $[x]_+ := \max\{0,x\}$,
+and $\ceil{x}_+ := \max\{0,\ceil{x}\}$.
+We use $\ln(\cdot)$ for natural logarithm, and $\log(\cdot)$ for
+logarithm with an arbitrary constant base.
 Boldface symbols are used for vectors and matrices (e.g., $\vv$,

### 2015-10-30 nips2015final/intro.tex
cite arxiv version

@@
 \cite{%
-%MCMCDiscussion93%
+MCMCDiscussion93%
 ,DBLP:conf/valuetools/KontoyiannisLM06%
@@
 accuracy~\cite[e.g.,][]{%
-%  MCMCDiscussion93,%
+  MCMCDiscussion93,%
   GaSmi00:eigval,%

### 2015-10-30 nips2015final/rates.tex
cite arxiv version

--- 2015-08-07 arxiv-v2/rates.tex
@@
   \[
-    \Pr\Brackets{ |\hatgap - \gap| \geq 1/8 } \geq 1/6 .
+    \Pr\Brackets{ |\hatgap - \gap| \geq 1/8 } \geq 3/8 .
   \]
@@
   \[
-    \Pr\Brackets{ |\hatgap - \gap| \geq \bar\gamma} \geq 1/4 .
+    \Pr\Brackets{ |\hatgap - \gap| \geq \bar\gamma/2} \geq 1/4 .
   \]
@@
 The proofs of \cref{thm:lb-pimin,thm:lb-gap} are given in
-\cref{app:lower}.
+\cref{app:lower}.\footnote{%
+  A full version of this paper, with appendices, is available on
+  arXiv~\citep{DBLP:journals/corr/HsuKS15}.%
+}
 

### 2015-10-30 nips2015final/empirical.tex
minor formatting; makefile
stupid fullpage

--- 2015-10-28 arxiv-v2/empirical.tex
@@
-\begin{algorithm}
+\begin{algorithm}[t]
 \caption{Empirical confidence intervals}
@@
         }
-      }
-      , \quad (i,j) \in [d]^2 ; \\
+      },
+%      , \quad (i,j) \in [d]^2 ; \\
+      &
       \wh{P}_{i,j}
@@
 
-  \STATE Empirical bounds for $|\wh{P}_{i,j}-P_{i,j}|$ for each $(i,j)
-  \in [d]^2$:
-  let $\emptail
+  \STATE Empirical bounds for $|\wh{P}_{i,j}{-}P_{i,j}|$ for $(i,j) \in [d]^2$:
+  $c:=1.01$,
+  $\emptail
     := \inf
@@
       t\geq0 :
-      2d^2 \parens{ 1 + \ceil{\ln(2n/t)} } e^{-t} \leq \delta
-    }$ and
+      2d^2 \parens{ 1 + \ceil{\log_c\frac{2n}{t}}_+ } e^{-t} \leq
+      \delta
+    }$,
   \begin{equation}
+    \text{and} \quad
     \wh{B}_{i,j}
     :=
-    \sqrt{\frac{9e\wh{P}_{i,j}(1-\wh{P}_{i,j})\emptail}{2N_i}}
-    + \frac{(3e+2.5)\emptail+1.5\abs{(1/d)-\wh{P}_{i,j}}}{N_i}
+    \Parens{
+      \sqrt{\frac{c\emptail}{2N_i}}
+      + \sqrt{
+        \frac{c\emptail}{2N_i}
+        + 
+        \sqrt{\frac{2c\wh{P}_{i,j}(1-\wh{P}_{i,j})\emptail}{N_i}}
+        + \frac{(5/3)\emptail + \abs{\wh{P}_{i,j}-1/d}}{N_i}
+      }
+    }^2
     .
@@
 $\kappa \le d/\gap$, and hence
-\[
-  \hat{b}
+%\[
+ $\hat{b}
   =
@@
     \sqrt{\frac{P_{i,j}\log\log n}{\pi_i n}}
-  }
-  , \quad
-  \hat{w}
+  }$%
+  , %\quad
+ $\hat{w}
   =
@@
     \frac{d}{\pimin\gap} \sqrt{\frac{\log\log n}{\pimin n}}
-  }
+  }$%
   .
-\]
+%\]
 
@@
 \cref{alg:empest},
-\[
-  \pimin \geq \min_{i \in [d]} \brackets{ \hat\pi_i - \hat{b} }_+
-  , \quad
-  \gap \geq \brackets{ \hatgap - \hat{w} }_+
+%\[
+ $\pimin \geq \min_{i \in [d]} \brackets{ \hat\pi_i - \hat{b} }_+$
+  , %\quad
+ $\gap \geq \brackets{ \hatgap - \hat{w} }_+$%
   ;
-\]
+%\]
 then plug-in these lower bounds for $\pimin$ and $\gap$ in the
@@
   (as $n \to \infty$) 
-  \[
-    |\wh{U}|
+%  \[
+   $|\wh{U}|
     =
@@
       \sqrt{\frac{\pimin\log\frac{d}{\pimin\delta}}{\gap n}}
-    }
+    }$%
     ,
-    \quad
-    |\wh{V}|
+%    \quad
+   $|\wh{V}|
     =
@@
       }
-    }
-  \]
+    }$,
+%  \]
   where $\hat{w}$ is the width from \cref{alg:empest}.

### 2015-10-30 nips2015final/rates.tex
stupid fullpage

@@
 Our estimator of the minimum stationary probability $\pimin$ is
-\[
-  \hatpimin := \min_{i \in [d]} \hat\pi_i ,
-\]
+%\[
+ $\hatpimin := \min_{i \in [d]} \hat\pi_i$,
+%\]
 and our estimator of the spectral gap $\gap$ is
-\[
-  \hatgap := 1 - \max\{ \hat\lambda_2, |\hat\lambda_d| \} .
-\]
+%\[
+ $\hatgap := 1 - \max\{ \hat\lambda_2, |\hat\lambda_d| \}$.
+%\]
 
@@
 $\gap$ to within constant multiplicative factors are, respectively,
-\[
-  \tilde{O}\Parens{\frac1{\pimin\gap}}
-  \quad\text{and}\quad
-  \tilde{O}\Parens{\frac1{\pimin\gap^3}}
+%\[
+ $\tilde{O}\Parens{\frac1{\pimin\gap}}$
+ and
+%  \quad\text{and}\quad
+ $\tilde{O}\Parens{\frac1{\pimin\gap^3}}$
   .
-\]
+%\]
 By \cref{eq:mixing-time-bound},

### 2015-10-30 nips2015final/empirical.tex
typo

@@
 It is easy to combine \cref{thm:err,thm:empirical} to yield intervals
-whose widths shrink at at least as fast as both the non-empirical
+whose widths shrink at least as fast as both the non-empirical
 intervals from \cref{thm:err} and the empirical intervals from

### 2015-10-31 nips2015final/appendix-empirical.tex
nips final
minor typographical edits

--- 2015-10-28 arxiv-v2/appendix-empirical.tex
@@
     \leq
-    \Parens{ 1 + \ceil{\log_c(2n/t)} }
+    \Parens{ 1 + \ceil{\log_c(2n/t)}_+ }
     e^{-t}
@@
   Define $v_i := c^i b^2t/2$ for $i = 0, 1, 2, \dotsc, \ceil{
-  \log_c(2n/t) }$, and let $v_{-1} := -\infty$.
+  \log_c(2n/t) }_+$, and let $v_{-1} := -\infty$.
   Then, since $V \in [0,b^2n]$,
@@
     & =
-    \sum_{i=0}^{\ceil{\log_c(2n/t)}}
+    \sum_{i=0}^{\ceil{\log_c(2n/t)}_+}
     \Pr\Brackets{
       S > \sqrt{2\max\braces{v_0,cV}t} + 2bt/3
-      \,\wedge\, v_{j-1} < V \leq v_j
+      \,\wedge\, v_{i-1} < V \leq v_i
     }
@@
     & \leq
-    \sum_{i=0}^{\ceil{\log_c(2n/t)}}
+    \sum_{i=0}^{\ceil{\log_c(2n/t)}_+}
     \Pr\Brackets{
-      S > \sqrt{2\max\braces{v_0,cv_{j-1}}t} + 2bt/3
-      \,\wedge\, v_{j-1} < V \leq v_j
+      S > \sqrt{2\max\braces{v_0,cv_{i-1}}t} + 2bt/3
+      \,\wedge\, v_{i-1} < V \leq v_i
     }
@@
     & \leq
-    \sum_{i=0}^{\ceil{\log_c(2n/t)}}
+    \sum_{i=0}^{\ceil{\log_c(2n/t)}_+}
     \Pr\Brackets{
-      S > \sqrt{2v_jt} + 2bt/3
-      \,\wedge\, V \leq v_j
+      S > \sqrt{2v_it} + 2bt/3
+      \,\wedge\, V \leq v_i
     }
@@
     & \leq
-    \Parens{ 1 + \ceil{\log_c(2n/t)} }
-    e^{-t}
+    \Parens{ 1 + \ceil{\log_c(2n/t)}_+ }
+    e^{-t}\,,
   \end{align*}
@@
     + \frac{(5/3)\emptail}{N_i+d\alpha}
-    + \frac{\abs{\alpha-d\alpha P_{i,j}}}{N_i+d\alpha}
+    + \frac{\abs{\alpha-d\alpha P_{i,j}}}{N_i+d\alpha}\,,
     \label{eq:P-unobs-bound}
@@
       t\geq0 :
-      2d^2 \Parens{ 1 + \ceil{\log_c(2n/t)} } e^{-t} \leq \delta
+      2d^2 \Parens{ 1 + \ceil{\log_c(2n/t)}_+ } e^{-t} \leq \delta
     }
@@
   O\Parens{ \sqrt{ \frac{d}{\pimin}} }
-  .
+  \,,
 \end{align*}
@@
 The following claim provides a bound on $\kappa$ in terms of the
-number of states and spectral gap.
+number of states and the spectral gap.
 \begin{claim}
@@
 \begin{proof}
-  It is established in~\citep{cho2001comparison} that
+  It is established by Cho and Meyer~\citep{cho2001comparison} that
   \[
@@
   (our $\kappa$ is the $\kappa_4$ quantity from
-  \citep{cho2001comparison}), and \citep{seneta1993sensitivity}
+  \citep{cho2001comparison}), and Seneta~\citep{seneta1993sensitivity}
   establishes

### 2015-10-31 arxiv-v2/empirical.tex
.
corner cases

@@
 
-  \STATE Empirical bounds for $|\wh{P}_{i,j}-P_{i,j}|$ for each $(i,j)
-  \in [d]^2$:
-  let $\emptail
+  \STATE Empirical bounds for $|\wh{P}_{i,j}{-}P_{i,j}|$ for $(i,j){\in}[d]^2$:
+  $c:=1.01$,
+  $\emptail
     := \inf

### 2015-10-31 arxiv-v2/appendix-empirical.tex
corner cases
freedman

@@
   \label{thm:freedman}
-  Let $Y_1, Y_2, \dotsc, Y_n$ be a bounded martingale difference
-  sequence with respect to the filtration $\cF = (\cF_i)_{1 \leq i
-  \leq n}$ and with $|Y_i| \leq b$ for some $b>0$.
-  Let $V := \sum_{i=1}^n \EE{Y_i^2|\cF_{i-1}}$ and $S := \max_{1 \leq
-  m \leq n} \sum_{i=1}^m Y_i$.
+  Let $(Y_i)_{i \in \bbN}$ be a bounded martingale difference sequence
+  with respect to the filtration $(\cF_i)_{i \in \bbN}$; assume for
+  some $b>0$, $|Y_i| \leq b$ almost surely for all $i \in \bbN$.
+  Let $V_k := \sum_{i=1}^k \EE{Y_i^2|\cF_{i-1}}$ and $S_k :=
+  \sum_{i=1}^k Y_i$ for $k \in \bbN$.
   For all $t, v > 0$,
@@
     \Pr\Brackets{
-      S > \sqrt{2vt} + 2bt/3
+      \exists k \in \bbN \,\centerdot\,
+      S_k > \sqrt{2vt} + 2bt/3
       \,\wedge\,
-      V \leq v
+      V_k \leq v
     }
@@
   \label{cor:freedman}
-  Under the same setting as \Cref{thm:freedman}, for any $t > 0$ and
-  $c > 1$,
+  Under the same setting as \Cref{thm:freedman}, for any $n \geq 1$,
+  $t > 0$, and $c > 1$,
   \[
     \Pr\Brackets{
-      S > \sqrt{2cVt} + 5bt/3
+      \exists k \in [n] \,\centerdot\,
+      S_k > \sqrt{2cV_kt} + 5bt/3
     }
     \leq
-    \Parens{ 1 + \ceil{\log_c(2n/t)} }
+    \Parens{ 1 + \ceil{\log_c(2n/t)}_+ }
     e^{-t}
@@
   Define $v_i := c^i b^2t/2$ for $i = 0, 1, 2, \dotsc, \ceil{
-  \log_c(2n/t) }$, and let $v_{-1} := -\infty$.
-  Then, since $V \in [0,b^2n]$,
+  \log_c(2n/t) }_+$, and let $v_{-1} := -\infty$.
+  Then, since $V_k \in [0,b^2n]$ for all $k\in[n]$,
   \begin{align*}
+    \lefteqn{
       \Pr\Brackets{
-      S > \sqrt{2\max\braces{v_0,cV}t} + 2bt/3
+        \exists k \in [n] \,\centerdot\,
+        S_k > \sqrt{2\max\braces{v_0,cV_k}t} + 2bt/3
       }
+    }
+    \\
     & =
-    \sum_{i=0}^{\ceil{\log_c(2n/t)}}
+    \sum_{i=0}^{\ceil{\log_c(2n/t)}_+}
     \Pr\Brackets{
-      S > \sqrt{2\max\braces{v_0,cV}t} + 2bt/3
-      \,\wedge\, v_{j-1} < V \leq v_j
+      \exists k \in [n] \,\centerdot\,
+      S_k > \sqrt{2\max\braces{v_0,cV_k}t} + 2bt/3
+      \,\wedge\, v_{i-1} < V_k \leq v_i
     }
@@
     & \leq
-    \sum_{i=0}^{\ceil{\log_c(2n/t)}}
+    \sum_{i=0}^{\ceil{\log_c(2n/t)}_+}
     \Pr\Brackets{
-      S > \sqrt{2\max\braces{v_0,cv_{j-1}}t} + 2bt/3
-      \,\wedge\, v_{j-1} < V \leq v_j
+      \exists k \in [n] \,\centerdot\,
+      S_k > \sqrt{2\max\braces{v_0,cv_{i-1}}t} + 2bt/3
+      \,\wedge\, v_{i-1} < V_k \leq v_i
     }
@@
     & \leq
-    \sum_{i=0}^{\ceil{\log_c(2n/t)}}
+    \sum_{i=0}^{\ceil{\log_c(2n/t)}_+}
     \Pr\Brackets{
-      S > \sqrt{2v_jt} + 2bt/3
-      \,\wedge\, V \leq v_j
+      \exists k \in [n] \,\centerdot\,
+      S_k > \sqrt{2v_it} + 2bt/3
+      \,\wedge\, V_k \leq v_i
     }
@@
   \[
-    \sqrt{2cVt} + 5bt/3
-    \geq \sqrt{2\max\braces{v_0,cV}t} + 2bt/3
-    .
-    \qedhere
+    \sqrt{2cV_kt} + 5bt/3
+    \geq \sqrt{2\max\braces{v_0,cV_k}t} + 2bt/3
   \]
+  for all $k \in [n]$.
 \end{proof}

### 2015-10-31 arxiv-v2/appendix-empirical.tex
"such that" macro
freedman
5/3 -> 4/3
.

@@
   \label{thm:freedman}
-  Let $(Y_i)_{i \in \bbN}$ be a bounded martingale difference sequence
-  with respect to the filtration $(\cF_i)_{i \in \bbN}$; assume for
-  some $b>0$, $|Y_i| \leq b$ almost surely for all $i \in \bbN$.
-  Let $V_k := \sum_{i=1}^k \EE{Y_i^2|\cF_{i-1}}$ and $S_k :=
-  \sum_{i=1}^k Y_i$ for $k \in \bbN$.
-  For all $t, v > 0$,
+  Let $(Y_t)_{t \in \bbN}$ be a bounded martingale difference sequence
+  with respect to the filtration $\cF_0 \subset \cF_1 \subset \cF_2
+  \subset \dotsb$; assume for some $b>0$, $|Y_t| \leq b$ almost surely
+  for all $t \in \bbN$.
+  Let $V_k := \sum_{t=1}^k \EE{Y_t^2|\cF_{t-1}}$ and $S_k :=
+  \sum_{t=1}^k Y_t$ for $k \in \bbN$.
+  For all $s, v > 0$,
   \[
     \Pr\Brackets{
-      \exists k \in \bbN \,\centerdot\,
-      S_k > \sqrt{2vt} + 2bt/3
+      \exists k \in \bbN \;\st\,
+      S_k > s
       \,\wedge\,
@@
     }
-    \leq e^{-t}
-    .
+    \leq \Parens{
+      \frac{v}{s+v}
+    }^{s+v}
+    e^s
+    = \exp\Parens{
+      -\frac{v}{b^2} \cdot h\Parens{\frac{bs}{v}}
+    }
+    \,,
   \]
+  where $h(u) := (1+u)\ln(1+u) - u$.
 \end{theorem}
+Observe that in \Cref{thm:freedman}, for any $x>0$, if $s :=
+\sqrt{2vx} + bx/3$ and $z := b^2x/v$, then probability bound becomes
+\[
+  \exp\Parens{
+    -x \cdot \frac{h\Parens{\sqrt{2z}+z/3}}{z}
+  }
+  \leq
+  e^{-x}
+\]
+since $h(\sqrt{2z}+z/3)/z \geq 1$ for all $z > 0$ (see,
+e.g.,~\citep[proof of Lemma 5]{audibert2009}).
+
 \begin{corollary}
@@
   Under the same setting as \Cref{thm:freedman}, for any $n \geq 1$,
-  $t > 0$, and $c > 1$,
+  $x > 0$, and $c > 1$,
   \[
     \Pr\Brackets{
-      \exists k \in [n] \,\centerdot\,
-      S_k > \sqrt{2cV_kt} + 5bt/3
+      \exists k \in [n] \;\st\,
+      S_k > \sqrt{2cV_kx} + 4bx/3
     }
     \leq
-    \Parens{ 1 + \ceil{\log_c(2n/t)}_+ }
-    e^{-t}
+    \Parens{ 1 + \ceil{\log_c(2n/x)}_+ }
+    e^{-x}
     .
@@
 \begin{proof}
-  Define $v_i := c^i b^2t/2$ for $i = 0, 1, 2, \dotsc, \ceil{
-  \log_c(2n/t) }_+$, and let $v_{-1} := -\infty$.
+  Define $v_i := c^i b^2x/2$ for $i = 0, 1, 2, \dotsc, \ceil{
+  \log_c(2n/x) }_+$, and let $v_{-1} := -\infty$.
   Then, since $V_k \in [0,b^2n]$ for all $k\in[n]$,
   \begin{align*}
-    \lefteqn{
+    \MoveEqLeft{
       \Pr\Brackets{
-        \exists k \in [n] \,\centerdot\,
-        S_k > \sqrt{2\max\braces{v_0,cV_k}t} + 2bt/3
+        \exists k \in [n] \;\st\,
+        S_k > \sqrt{2\max\braces{v_0,cV_k}x} + bx/3
       }
@@
     & =
-    \sum_{i=0}^{\ceil{\log_c(2n/t)}_+}
+    \sum_{i=0}^{\ceil{\log_c(2n/x)}_+}
     \Pr\Brackets{
-      \exists k \in [n] \,\centerdot\,
-      S_k > \sqrt{2\max\braces{v_0,cV_k}t} + 2bt/3
+      \exists k \in [n] \;\st\,
+      S_k > \sqrt{2\max\braces{v_0,cV_k}x} + bx/3
       \,\wedge\, v_{i-1} < V_k \leq v_i
@@
     & \leq
-    \sum_{i=0}^{\ceil{\log_c(2n/t)}_+}
+    \sum_{i=0}^{\ceil{\log_c(2n/x)}_+}
     \Pr\Brackets{
-      \exists k \in [n] \,\centerdot\,
-      S_k > \sqrt{2\max\braces{v_0,cv_{i-1}}t} + 2bt/3
+      \exists k \in [n] \;\st\,
+      S_k > \sqrt{2\max\braces{v_0,cv_{i-1}}x} + bx/3
       \,\wedge\, v_{i-1} < V_k \leq v_i
@@
     & \leq
-    \sum_{i=0}^{\ceil{\log_c(2n/t)}_+}
+    \sum_{i=0}^{\ceil{\log_c(2n/x)}_+}
     \Pr\Brackets{
-      \exists k \in [n] \,\centerdot\,
-      S_k > \sqrt{2v_it} + 2bt/3
+      \exists k \in [n] \;\st\,
+      S_k > \sqrt{2v_ix} + bx/3
       \,\wedge\, V_k \leq v_i
@@
     & \leq
-    \Parens{ 1 + \ceil{\log_c(2n/t)}_+ }
-    e^{-t}\,,
+    \Parens{ 1 + \ceil{\log_c(2n/x)}_+ }
+    e^{-x}\,,
   \end{align*}
@@
   \[
-    \sqrt{2cV_kt} + 5bt/3
-    \geq \sqrt{2\max\braces{v_0,cV_k}t} + 2bt/3
+    \sqrt{2cV_kx} + 4bx/3
+    \geq \sqrt{2\max\braces{v_0,cV_k}x} + 2bx/3
   \]
@@
     \sqrt{\Parens{\frac{N_i}{N_i+d\alpha}}\frac{2cP_{i,j}(1-P_{i,j})\emptail}{N_i+d\alpha}}
-    + \frac{(5/3)\emptail}{N_i+d\alpha}
+    + \frac{(4/3)\emptail}{N_i+d\alpha}
     + \frac{\abs{\alpha-d\alpha P_{i,j}}}{N_i+d\alpha}\,,
@@
 \begin{proof}
-  For each $t \in \bbN$, define $\cF_t$ to be the $\sigma$-field
-  generated by $X_1,X_2,\dotsc,X_t$.
-  Fix a pair $(i,j) \in [d]^2$, and define
+  Let $\cF_t$ be the $\sigma$-field generated by $X_1,X_2,\dotsc,X_t$.
+  Fix a pair $(i,j) \in [d]^2$.
+  Let $Y_1 := 0$, and for $t \geq 2$,
   \[
-    Y_{t+1} := \Ind{X_t = i} (\Ind{X_{t+1}=j} - P_{i,j})
-    , \quad t \in [n-1]
+    Y_t := \Ind{X_{t-1} = i} (\Ind{X_t=j} - P_{i,j})
+    ,
   \]
@@
   \[
-    \sum_{t=1}^{n-1} Y_{t+1}
+    \sum_{t=1}^n Y_t
     = N_{i,j} - N_i P_{i,j}
@@
   \]
-  By the Markov property, the stochastic process $\{ Y_{t+1} \}_{t \in
-  [n-1]}$ is an $\cF_t$-adapted martingale difference sequence:
-  $Y_{t+1}$ is $\cF_{t+1}$-measurable and $\EE{Y_{t+1}| \cF_t } = 0$,
-  for each $t \in [n-1]$.
-  Moreover, for each $t \in [n-1]$,
+  The Markov property implies that the stochastic process
+  $(Y_t)_{t\in[n]}$ is
+  an $(\cF_t)_{t\in[n]}$-adapted martingale difference sequence: $Y_t$ is
+  $\cF_t$-measurable and $\EE{Y_t| \cF_{t-1} } = 0$, for each $t$.
+  Moreover, for all $t \in [n]$,
   \[
-    \EE{Y_{t+1}^2| \cF_t } = \Ind{X_t=i} P_{i,j} (1-P_{i,j})
+    Y_t \in [-P_{i,j},1-P_{i,j}]
+    \,,
   \]
-  and
+  and for $t \geq 2$,
   \[
-    Y_{t+1} \in [-P_{i,j},1-P_{i,j}] .
+    \EE{Y_t^2| \cF_{t-1} } = \Ind{X_{t-1}=i} P_{i,j} (1-P_{i,j})
+    \,.
   \]
@@
     \abs{ N_{i,j} - N_i P_{i,j} }
-    \leq \sqrt{2c N_i P_{i,j} (1-P_{i,j}) \emptail} + \frac53
-    \emptail
+    \leq \sqrt{2c N_i P_{i,j} (1-P_{i,j}) \emptail} +
+    \frac{4\emptail}3
   \]
@@
       \sqrt{\frac{2c\wh{P}_{i,j}(1-\wh{P}_{i,j})\emptail}{N_i}}
-      + \frac{(5/3)\emptail + \abs{\alpha-d\alpha\wh{P}_{i,j}}}{N_i}
+      + \frac{(4/3)\emptail + \abs{\alpha-d\alpha\wh{P}_{i,j}}}{N_i}
     }
@@
     \sqrt{\frac{2cN_iP_{i,j}(1-P_{i,j})\emptail}{(N_i+d\alpha)^2}}
-    + \frac{(5/3)\emptail}{N_i+d\alpha}
+    + \frac{(4/3)\emptail}{N_i+d\alpha}
     + \frac{\abs{\alpha-d\alpha P_{i,j}}}{N_i+d\alpha}
@@
     }
-    + \frac{(5/3)\emptail}{N_i+d\alpha}
+    + \frac{(4/3)\emptail}{N_i+d\alpha}
     \notag \\
@@
     \sqrt{\frac{2cN_i\wh{P}_{i,j}(1-\wh{P}_{i,j})\emptail}{(N_i+d\alpha)^2}}
-    + \frac{(5/3)\emptail + \abs{\alpha -
+    + \frac{(4/3)\emptail + \abs{\alpha -
     d\alpha\wh{P}_{i,j}}}{N_i+d\alpha}

### 2015-10-31 arxiv-v2/appendix-empirical.tex
.

@@
   The Markov property implies that the stochastic process
-  $(Y_t)_{t\in[n]}$ is
-  an $(\cF_t)_{t\in[n]}$-adapted martingale difference sequence: $Y_t$ is
-  $\cF_t$-measurable and $\EE{Y_t| \cF_{t-1} } = 0$, for each $t$.
+  $(Y_t)_{t\in[n]}$ is an $(\cF_t)$-adapted martingale difference
+  sequence: $Y_t$ is $\cF_t$-measurable and $\EE{Y_t| \cF_{t-1} } =
+  0$, for each $t$.
   Moreover, for all $t \in [n]$,

### 2015-10-31 arxiv-v2/appendix-empirical.tex
minor typos corrected in Freedman's ineq
scaling with b..

@@
     \leq \Parens{
-      \frac{v}{s+v}
-    }^{s+v}
-    e^s
+      \frac{v/b^2}{s/b+v/b^2}
+    }^{s/b+v/b^2}
+    e^{s/b}
     = \exp\Parens{
@@
 \end{theorem}
-Observe that in \Cref{thm:freedman}, for any $x>0$, if $s :=
-\sqrt{2vx} + bx/3$ and $z := b^2x/v$, then probability bound becomes
+Observe that in \Cref{thm:freedman}, for any $t>0$, if $s :=
+\sqrt{2vt} + bt/3$ and $z := b^2t/v$, then the probability bound
+on the right-hand side becomes
 \[
@@
     \sqrt{2cV_kx} + 4bx/3
-    \geq \sqrt{2\max\braces{v_0,cV_k}x} + 2bx/3
+    \geq \sqrt{2\max\braces{v_0,cV_k}x} + bx/3
   \]

### 2015-10-31 arxiv-v2/appendix-empirical.tex
.
t<->x

@@
 \end{theorem}
-Observe that in \Cref{thm:freedman}, for any $t>0$, if $s :=
-\sqrt{2vt} + bt/3$ and $z := b^2t/v$, then the probability bound
+Observe that in \Cref{thm:freedman}, for any $x>0$, if $s :=
+\sqrt{2vx} + bx/3$ and $z := b^2x/v$, then the probability bound
 on the right-hand side becomes

### 2015-11-03 arxiv-v2/lowerbound.tex
typo
.

@@
   chains have the same transition probabilities from state $1$.
-  The chains are indistinguishable less the sample path eventually
+  The chains are indistinguishable unless the sample path eventually
   reaches state $2$.
@@
 Theorem~\ref{thm:lb-gap} implies the existence of an ergodic and
-reversible Markov chain distribution with (say) $1/4 \leq \gap \leq
-1/2$ and $\pimin \geq 1/(2d)$ such that
+reversible Markov chain distribution (with $\pimin \geq 1/(2d)$) such that
 \[

### 2015-11-03 arxiv-v2/lowerbound.tex
.

@@
 Theorem~\ref{thm:lb-gap} implies the existence of an ergodic and
-reversible Markov chain distribution (with $\pimin \geq 1/(2d)$) such that
+reversible Markov chain distribution with spectral gap $\gap$ and
+stationary distribution minorized by $\pimin$ such that
 \[

### 2015-11-26 code/group_inverse.m
some matlab codes

@@ -1,0 +1,24 @@
+% Function:
+% [Asharp,stationary] = group_inverse(P)
+%
+% Input:
+% 'P'                row stochastic probability transition matrix
+%                    of an ergodic Markov chain
+%
+% Outputs:
+% 'Asharp'           group inverse of I-P
+% 'stationary'       unique stationary distribution
+%
+function [Asharp,stationary] = group_inverse(P)
+n = size(P,1);
+A = eye(n) - P;
+U = A(1:n-1,1:n-1);
+d = A(n,1:n-1);
+invU = inv(U);
+h = d/U;
+delta = -sum(h/U);
+beta = 1-sum(h);
+F = invU - (delta/beta)*eye(n-1);
+Asharp = [invU + (1/delta)*(sum(invU,2)*h/U - sum(F,2)*h*F), -(1/beta)*sum(F,2); ...
+  (1/beta)*h*F, delta/beta^2];
+stationary = [-h,1]/beta;

### 2015-11-26 code/make_lb_P.m

@@ -1,0 +1,19 @@
+% constructs one of the markov chains from the lower bound in the paper
+%
+function [P,gap] = make_lb_P(d,gamma_bar,i)
+
+big_epsilon = 2*(d-1)*gamma_bar/d;
+small_epsilon = (d/2-1)*big_epsilon/(d-1);
+
+P = zeros(d);
+for j=1:d
+  if j == i
+    P(j,:) = small_epsilon/(d-1);
+    P(j,j) = 1-small_epsilon;
+  else
+    P(j,:) = big_epsilon/(d-1);
+    P(j,j) = 1-big_epsilon;
+  end
+end
+l = sort(abs(eig(P)),'descend');
+gap = l(1) - l(2);

### 2015-11-26 code/mcintervals.m

@@ -1,0 +1,33 @@
+function [pimin_hat, gap_hat, b_hat, w_hat, P_hat, pi_hat, Asharp, lambda_hats, P_lb, P_ub, B_hat, kappa_hat, rho_hat] = mcintervals(N, d, delta)
+
+n = sum(N(:));
+
+Nvisit = sum(N,2);
+P_hat = bsxfun(@rdivide, N + 1/d, Nvisit + 1);
+
+tau = fzero(@(t) 2*d^2*(1 + ceil(max(log(2*n/t)/log(1.1),0)))*exp(-t) - delta, [1,100]);
+
+P_ub = zeros(d);
+P_lb = zeros(d);
+
+for i=1:d
+  for j=1:d
+    P_lb(i,j) = fzero(@(p) P_hat(i,j) - p - (sqrt(2.2*tau*Nvisit(i)/(Nvisit(i)+1)^2*p*(1-p)) + (4/3)*tau/(Nvisit(i)+1) + abs(1/d-p)/(Nvisit(i)+1)), [0,1]);
+    P_ub(i,j) = fzero(@(p) p - P_hat(i,j) - (sqrt(2.2*tau*Nvisit(i)/(Nvisit(i)+1)^2*p*(1-p)) + (4/3)*tau/(Nvisit(i)+1) + abs(1/d-p)/(Nvisit(i)+1)), [0,1]);
+  end
+end
+B_hat = max(P_ub - P_hat, P_hat - P_lb);
+
+[Asharp,pi_hat] = group_inverse(P_hat);
+pimin_hat = min(pi_hat);
+
+kappa_hat = 0.5 * max(diag(Asharp) - min(Asharp)');
+b_hat = kappa_hat * max(B_hat(:));
+rho_hat = 0.5 * max([b_hat ./ pi_hat, b_hat ./ max(pi_hat - b_hat,0)]);
+% w_hat = 2*rho_hat + rho_hat^2 + (1+2*rho_hat+rho_hat^2)*norm(diag(sqrt(pi_hat))*B_hat*diag(1./sqrt(pi_hat)),'fro');
+w_hat = 2*rho_hat + rho_hat^2 + (1+2*rho_hat+rho_hat^2)*norm(diag(sqrt(pi_hat))*B_hat*diag(1./sqrt(pi_hat)));
+
+L_hat = diag(sqrt(pi_hat))*P_hat*diag(sqrt(1./pi_hat));
[…]

### 2015-11-26 code/run_expt.m

@@ -1,0 +1,15 @@
+rng(0);
+
+d = 5;
+gamma_bar = 0.25;
+delta = 0.1;
+[P,gap] = make_lb_P(d,gamma_bar,d);
+
+n=2^18;
+N = zeros(d);
+x = 1;
+for epoch=1:1
+  [N,x] = sample_mc_pairs(P,n,x,N);
+end
+
+[pimin_hat, gap_hat, b_hat, w_hat, P_hat, pi_hat, Asharp, lambda_hats, P_lb, P_ub, B_hat, kappa_hat, rho_hat] = mcintervals(N, d, delta)

### 2015-11-26 code/sample_mc_pairs.m

@@ -1,0 +1,17 @@
+% count number of transitions of each type in a random sample path of
+% length n from a markov chain with transition matrix P starting at state
+% x0
+%
+% returns these counts in N and the final state in x0
+%
+function [N,x0] = sample_mc_pairs(P,n,x0,N)
+
+d = size(P,1);
+if nargin < 4
+  N = zeros(d);
+end
+for t=1:n
+  x = randsample(d,1,true,P(x0,:));
+  N(x0,x) = N(x0,x) + 1;
+  x0 = x;
+end

### 2015-11-26 code/mcintervals.m
.

@@ -1,4 +1,4 @@
-function [pimin_hat, gap_hat, b_hat, w_hat, P_hat, pi_hat, Asharp, lambda_hats, P_lb, P_ub, B_hat, kappa_hat, rho_hat] = mcintervals(N, d, delta)
+function [pimin_lb, pimin_ub, gap_lb, gap_ub, pimin_hat, gap_hat, P_hat, pi_hat, Asharp, lambda_hats, P_lb, P_ub, B_hat, b_hat, kappa_hat, rho_hat, w_hat] = mcintervals(N, d, delta)
 
 n = sum(N(:));
 
@@ -31,3 +31,8 @@ L_hat = diag(sqrt(pi_hat))*P_hat*diag(sqrt(1./pi_hat));
 lambda_hats = eig(0.5*(L_hat+L_hat'));
 abs_lambda_hats = sort(abs(lambda_hats), 'descend');
 gap_hat = 1 - abs_lambda_hats(2);
+
+pimin_lb = max(pimin_hat - b_hat,0);
+pimin_ub = min(pimin_hat + b_hat,1);
+gap_lb = max(gap_hat - w_hat,0);
+gap_ub = min(gap_hat + w_hat,1);

### 2015-11-26 code/run_expt.m

@@ -1,15 +1,36 @@
 rng(0);
 
-d = 5;
-gamma_bar = 0.25;
-delta = 0.1;
+d = 4;
+gamma_bar = 0.4;
+delta = 0.05;
 [P,gap] = make_lb_P(d,gamma_bar,d);
 
-n=2^18;
+n_vals = 2.^(15:20);
+T = length(n_vals);
+n = n_vals(1);
+
+pimin_lb = zeros(T,1);
+pimin_ub = zeros(T,1);
+gap_lb = zeros(T,1);
+gap_ub = zeros(T,1);
+
 N = zeros(d);
 x = 1;
-for epoch=1:1
+for i=1:T
+  fprintf('n=%d\n', n_vals(i));
   [N,x] = sample_mc_pairs(P,n,x,N);
+  [pimin_lb(i), pimin_ub(i), gap_lb(i), gap_ub(i)] = mcintervals(N, d, delta);
+  if i < T
+    n = n_vals(i+1) - n_vals(i);
+  end
 end
 
-[pimin_hat, gap_hat, b_hat, w_hat, P_hat, pi_hat, Asharp, lambda_hats, P_lb, P_ub, B_hat, kappa_hat, rho_hat] = mcintervals(N, d, delta)
+subplot(1,2,1);
+plot(n_vals, gap_ub, n_vals, gap_lb);
+xlabel('n');
+legend('\gamma_* ub', '\gamma_* lb');
+
+subplot(1,2,2);
+plot(n_vals, pimin_ub, n_vals, pimin_lb);
+xlabel('n');
+legend('\pi_* ub', '\pi_* lb');
